### 2014-03-19 designdoc.tex
Added design document for r package

\documentclass{article}

\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}
\DeclareMathOperator{\logit}{logit}

\usepackage{geometry}
%\geometry{hmargin=1.025in,vmargin={1.25in,2.5in},nohead,footskip=0.5in} 
%\geometry{hmargin=1.025in,vmargin={1.25in,0.75in},nohead,footskip=0.5in} 
%\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}

\renewcommand{\baselinestretch}{1.25}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} \DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\title{Design Document: Monte Carlo Maximum Likelihood for Two-Stage Hierarchical Models}

\author{Christina Knudson}

\begin{document}
\maketitle{}
\section{Purpose}
The purpose of this package is to perform MCML. It will read in the user's data and the model specifications, simulate random effects, approximate the log likelihood, and use trust to find the parameter estimates that maximize the log likelihood approximation. At the minimum, it will return the point estimates for the parameters along with their standard errors. It will provide confidence intervals. It will be equipped to do a likelihood ratio test in order to do model comparison between nested models.

\section{Notation}
Let $\beta$ be the vector of $p$ fixed effect parameters. Let $X$ be the design matrix for the fixed effects.  Let $U$ be a normal random vector with mean 0, variance matrix $D$, and design matrix $Z$.There are $T$ distinct variance component  $\nu_t,t=1,...,T$.  Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},...\nu_T)$. Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple.

Let $g$ be the link function with inverse $g^{-1}$. Then $\mu$ is the vector of means such that

\begin{align}
g(\mu) = X \beta + Z U
\end{align}

The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$.

For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
\end{align}

\subsection{D }
For now,  $D$ is assumed  diagonal so that $D= \sum_{t=1}^T \nu_t E_t$, where $\sum E_t = I$.\\

 Later versions of this package will allow other covariance structures. With other structures, $D$ will still be $D= \sum_{t=1}^T \nu_t E_t$, but $E_t$ will no longer be diagonal. We will have a D for distance (such as in the car theft data) and for autoregressive 1.

\section{model fitting function} 
This will be the main command the user will implement. The user will need to specify the response and the predictors using the formula coding similar to the lm command.  They'll need to specify the  family (either binomial or Poisson, though  really any exponential family would work and I could add more later).The user will specify the random effects in the same way as for the aster package. Thus, a sample command with fixed predictors $x_1$ and $x_2$ and with random effects $school$ and $classroom$ (in data set as categorical variables ) would look like
\begin{verbatim}
mcml(y ~ x1+ x2, list(~0+school,~0+classroom),  family.mcml="binomial.mcml", data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2) )
    \end{verbatim}
The family.mcml part will be discussed in section \ref{sec:fam}.

$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+? Most models require this, so please check if you meant to include it.''

It's possible that the user could want some variance components to be set equal. For example, in the Coull and Agresti flu problem, there are 4 years and random effects for each year. The authors want the within-year variance components to be equal.  There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since year 4's virus was a repeat). Suppose year is a categorical variable with four levels, and year1 through year4 are dummy variables. Thus, the call could contain these arguments
\begin{verbatim}
mcml(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeet,~0+year3,~0+year4),
 varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeet"),
data=flu,family.mcml=bernoulli.mcml)
\end{verbatim}

Thus model.matrix will make Z into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. What we do is we take the design matrices that share a variance component and then cbind them together. Thus we'll have 3 model matrices in our Z list, one for each random effect. We will want to put them in order (1,2,3 according to varcomps.equal) so that the names for each matrix line up ("subject","year","repeet").  Cbinding the design matrices that share a variance component will not effect the way that we compute $\eta$.  Also, the variance estimates should come out in this same order as well.

    The next steps: do pql to get a mean for the random effects;  sample the random effects from a standard normal; scale those generated random effects to PQL's mean and variance; implement trust on the objective function; and return estimates, standard errors, and the log likelihood evaluated at those estimates.

The value will contain the point estimates, the Monte Carlo approximation to to the log likelihood, its first and second derivatives. Other things may be added later. We need to have the entire Fisher information matrix; just the standard errors are not good enough. This is because the prediction intervals will need the whole Fisher information matrix.

%%Families%%
\section{Families} \label{sec:fam}
This function will be hidden from the user.  These functions (along with the distribution of random effects) are necessary to approximate the log likelihood.

I'm going to have a class called ``mcml.family'' with a few options (for now, binomial and poisson). Each family  function will output a list including the family name (a character string such as ``binomial''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.

The user will provide the family in the model-fitting function. They can either enter the character string (``binomial.mcml''), the function (binomial.mcml()), or the result of invoking it.  The following code for using the input to determine the family is adapted from glm.

\begin{verbatim}
logDensity<-function(family.mcml)
{
	if(is.character(family.mcml))
		family.mcml<-get(family.mcml,mode="function",envir=parent.frame())
	if(is.function(family.mcml))
		family.mcml<-family.mcml()
	if(!inherits(family.mcml,"mcml.family")) 
		stop(" 'family' not recognized") 
	return(family.mcml)
}
\end{verbatim}
If the user has entered the family as a string, go get the R object with that family name, either from the immediate environment or the parent environment.  If this has happened, ``family`` is now a function.  If ``family'' is a function (either because the user entered it as a function or because of the preceding step), invoke that function.  At this point, ``family.mcml'' should have  class ``mcml.family.'' If this is not the case (maybe because of a typo or maybe because they entered ``poisson'' rather than ``poisson.mcml''), then stop and return an error.

For example, one of the family functions will look like this
\begin{verbatim}
poisson.mcml<-function()
{
	family.mcml<- "poisson.mcml"
	c <- function(eta) exp(eta)
	cp <- function(eta) exp(eta)
	cpp<-function(eta) exp(eta)
	out<-list(family.mcml=family.mcml, c=c,cp=cp,cpp=cpp)
	class(out)<-"mcml.family"
	return(out)
}
\end{verbatim}

Then, to use these functions in order to calculate $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$, I can just call 
\begin{verbatim}
family$c(args)
family$cp(args)
family$cpp(args)
\end{verbatim}

For  binomial, these values (c, cp, and cpp) are:
\begin{align}
c(\eta_i)&=\log(1+e^{\eta_i})\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  
\end{align}
and for poisson they are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

Then we use these pieces to create the scalar $c(\eta)$, the vector $c'(\eta)$ and the matrix $c''(\eta)$. We calculate

\begin{align}
c(\eta)= \sum_i c(\eta_i).
\end{align}
 The vector $c'(\eta)$ has components $c'(\eta_i)$. The matrix $c''(\eta)$ is diagonal with diagonal elements $c''(\eta_i)$.

Note: for glm's binomial, the user can choose the link. They can't do that here: we need the link to be logit so that we have an exponential family. I'm going to include the link as one of the things in the family.mcml value just in case they don't already know what the link is.

Also, I'll have a check in here to make sure that the data are valid given the family type. So I'll make sure that if family.mcml is binomial.mcml, the data should be 0's or 1's. If family.mcml is poisson.mcml, then the data should be nonnegative. If that's not true, give an error message.

\section{Log density of the data (el)}

Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= Y' \eta +c(\eta) \\
&= \sum_{i} Y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+ZU.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = Y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l_1}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l_1}} = X_{j{l_1}}
\end{align}

We'd like the derivative of the log of the data density with respect to $\beta$. This can be written using the chain rule as follows:
\begin{align}
\dfrac{\partial}{\partial \beta_{l_1}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l_1}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
&= \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}}
\end{align}

The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
\begin{align}
\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
 \end{align}

Letting $\nabla c(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
\begin{align}
\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ Y- \nabla c(\eta)  \right].
\end{align}

Letting $\nabla^2 c(\eta)$ be a vector with components $c'(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
\begin{align}
   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ - \nabla^2 c(\eta) ] X
\end{align}

%%getEk
\section{getEk}
Recall $D=\sum \nu_t E_t$. The way we  get different forms of D is by changing $E_t$. We'll have a function that gives the $E_t$, depending on what the form of $D$ is.

\subsection{D diagonal}
When D is diagonal, then $\sum E_t = I$, the identity matrix. The $E_t$ are diagonal with either 1 or 0 on the diagonal (1 if that random effect has $\nu_t$ as a variance component, 0 otherwise).

At this point,  mod.mcml\$z is a list with $T$ design matrices (one for each distinct variance component).  We need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list (= the number of random effects based on that variance compnent $\nu_t$).
 Then $\sum q_t$ is the total number of random effects in the model, and $D$ will have that many rows (and columns). Thus, each $E_t$ will have that number of rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

%%distribution of the random effects%%
\section{distribution of the random effects (distRand)}
This will be hidden from the user.  To  approximate the log likelihood and to maximize that approximation, I'm going to need $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$.  We will calculate these quantities with these things: $D$ (the variance matrix for the u's), $\nu$ (the length-t vector of variance components), $U$ (the vector of random effects, which I will have generated from the importance sampling), $\mu$ (the mean of the random effects, which is either 0 or $u^*$ from PQL) and the list $z$ (of length T where T is the number of variance components, and each matrix in the list has $q_t$ columns).

%The model.matrix command has turned each the random effects into a model matrix and has outputted the model matrices as a list. 
%%For example, if a random effect is teacher, and the teacher factor has three levels (``Naslund,'' ``Johnson,'' ``Jergens''), then the model matrix would have ``Naslund,'' ``Johnson,'' ``Jergens'' as the three entries in the $3\times1$ matrix. If another random effect is school, where school is a factor with two levels (``South'', ``Central''), then the corresponding model matrix will be $2\times1$. I will need to remember to use ``drop=F'' to keep some matrices from turning into vectors.
%The number of matrices in the list is equal to the number of random effects $T$. 
%%In the example above, we'd get two model matrices, one for each random effect.

Since the random effects are drawn from $N(\mu, D)$, then the distribution of the random effects is:
\begin{align}
\log f_\theta(u) = (-1/2) \log |D| - (1/2) (U-\mu)' D^{-1} (U-\mu)
\end{align}

We will need first and second derivatives with respect to $\nu$. We'll have to take advantage of the form of D.

\subsection{D Diagonal}
In the simplest case, D is diagonal with variance components $\nu_t$s on the diagonal and 0s off diagonal. Remember that, for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s everywhere else.

We can partition the random effects according to their variance component: $U=(U_1',...,U_T')'$.  Let $D_t$ be the variance matrix for $U_t$ . $D_t$ has $q_t$ rows (and also columns). So $D$ is this:
\begin{align}
D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
\end{align}

Since D is diagonal, then $D^{-1}$ is also diagonal. The diagonal entries are now $\dfrac{1}{\nu_1}$,...,$\dfrac{1}{\nu_T}$.

Calculating det(D) is also easy because D is diagonal:
\begin{align}
det(D)= \nu_1^{q_1}...\nu_T^{q_T}
\end{align}

 The log density for $U$ is
\begin{align}
\log f_\theta(u) &= (-1/2) \log |D| - (1/2) (U-\mu)' D^{-1} (U-\mu)\\
&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} (U_t-\mu_t)'(U_t-\mu_t)   \right]
\end{align}

%Letting $I_{q_t}$ be the identity matrix with $q_t$ rows (and columns), we can write this as a sum as follows:
%\begin{align}
%\log f_\theta(u)&= \sum_{t=1}^T \log f_\theta(u_t)\\
%&=\sum_{t=1}^T - \dfrac{1}{2} \log |D_t| - \dfrac{1}{2} (U_t-\mu_t)' D_t^{-1} (U_t-\mu_t) \\ 
%&=\sum_{t=1}^T - \dfrac{1}{2} \log |\nu_t^{q_t}| - \dfrac{1}{2 \nu_t} (U_t-\mu_t)' I_{q_t} (U_t-\mu_t) \\ 
%&=\sum_{t=1}^T - \dfrac{q_t}{2} \log \nu_t - \dfrac{1}{2 \nu_t}(U_t-\mu_t) '(U_t-\mu_t)
%\end{align}
Taking the first and second derivative of each summand with respect to its associated $\nu_t$ results in
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}(U_t-\mu_t)'(U_t-\mu_t)
\end{align}
and 
\begin{align}
\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} (U_t-\mu_t)'(U_t-\mu_t).
\end{align}
Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
\begin{align}
\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
\end{align}
Also,
\begin{align}
\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
\end{align}
Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is
\begin{align}
\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} (U_1-\mu_1) ' (U_1-\mu_1) & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} (U_T-\mu_T) '(U_T-\mu_T)   \end{bmatrix} 
\end{align}
To calculate this vector, I will create a function to take $q_t$, $U_t$, and $\nu_t$ and spit out $\dfrac{-q_t}{2\nu_t} +\dfrac{1}{2 \nu_t^2} U_t'U_t$. I can then do a loop for t in 1 through T to calculate each entry in the vector.

The Hessian matrix is the following diagonal matrix:
\begin{align}
\dfrac{\partial^2}{\partial \nu^2} \log f_\theta(u) = \begin{bmatrix} \dfrac{q_1}{2 \nu_1^2}- \dfrac{1}{\nu_1^3} U_1'U_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & \dfrac{q_T}{2 \nu_T^2}- \dfrac{1}{\nu_T^3 U_T'U_T} \end{bmatrix}
\end{align}
I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$. Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.

Finally, in the end, I'll return the value (scalar), the gradient vector, and the hessian matrix.

What will the inputs of this function be? Either the vector $\sigma$ of standard deviations or the vector $\nu$ of variance components, U (the vector of random effects), and the list z (from mod.mcml dollar z).

Something we need is $q_t,t=1,...,T$. z is a list of length T, and each item in the list is a matrix with the number of columns equal to $q_t$.

\subsection{Splitting U into $U_t$ and $\mu$ into $\mu_t$}
The last thing we need to do is split $U$ into $U_1,...,U_T$ and $\mu$ into $\mu_1,...,\mu_T$. The same process should work for both, so let's just look at how we'd do it for $U$.  We know that the first $q_1$ items of U are $U_1$, the next $q_2$ items are $U_2$, etc.  In other words, entries 1 through $q_1$ are $U_1$. Items $q_1+1$ through $q_1+q_2$ are $U_2$, etc. In R pseudo code, this means 
\begin{verbatim}
nrand<-lapply(z,ncol) #get the number of columns from each matrix in list z
unlist(nrand) #chance the list into a vector, since each entry in the list is just a scalar

U1<- U[1 through nrand[1]] #first rule is use 1:nrand[1]
Ut<-U[sum(nrand[1:t-1]+1 through sum(nrand[1:t])]
\end{verbatim}

Note that we don't need to actually calculate either A or D. We work only through $\nu$.

\section{PQL}
 Before we get started on describing PQL, we're going to do a bit of changing notation because we don't want to have to deal with a constrained optimization.  Recall that $D=Var(u)$. Let $A^2=D$ so that A can have diagonal components that are positive or negative. We use A rather than D so we have unconstrained optimization.  If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
\begin{align*}
A= \sum_t E_t \sigma_t
\end{align*}
Recall that $E_t$ has either 0s or 1s on the diagonal to show which random effects go together, and $\sum_t E_t$ is the identity matrix. PQL will tell give us the components of A, and we'll take the absolute values of those components to get the standard deviations.

The other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the thing we're trying to optimize.

I will need to do PQL in order to form the importance sampling distribution. There are two ways to do PQL, both of which are described in the  vignette ``re'' from aster. In either case, there is an inner optimization and an outer optimization. The inner optimization is well behaved while the other optimization is a little tougher. I will be using the thing that is not quite PQL but is pretty close and is better behaved.  In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$. Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.

The default of optim is to minimize, but we'd like to  do maximization. If we just reverse the sign of the function and the gradient, it should do what we want.

The {\bf inner} optimization will be done with the trust function in R. We'd like the inner maximization to be a little more precise because any sloppiness in the inner optimization gets carried into the outer optimization.
The inner optimization maximizes the penalized log likelihood, which is calculated as follows:
\begin{align}
\eta=X\beta +ZAs\\
l(\eta)= Y^T \eta - c(\eta) \\
value= l(\eta)- \dfrac{1}{2} s^Ts
\end{align}

The value of the function is a scalar.  We need to give trust derivatives with respect to $s$ and $\beta$ as well. We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.

Since
\begin{align}
l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
\end{align}
then 
\begin{align}
\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i
\end{align}
and the off-diagonal is 0. Create  vector $\mu$ from the components $\mu_i$.  Now we can write the following expression:

\begin{align}
%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s^Ts  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s^Ts  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
&=\sum_i (Y_i-\mu_i) X_{ik}
\end{align}

We find the function's derivative with respect to $s$ as follows:
\begin{align}
\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
&=(Y-\mu)'ZA -s
\end{align}

This gives us the following derivatives:
\begin{align}
\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s^Ts \right]&= X^T (Y-\mu)\\
\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s^Ts \right]&= AZ' (Y-\mu)  -s
\end{align}

Lastly, we need the hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
\begin{enumerate}
\item $ \dfrac{\partial^2}{\partial s^2}$
\item $ \dfrac{\partial^2}{\partial \beta^2}$
\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
\end{enumerate}

We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$.
\begin{align}
  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
&= -AZ' \left[ c''(\eta) \right] ZA - I_q
\end{align}

Note that $c''(eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.  $I_q$ is the identity matrix of dimension $q$.  This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.

Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
\begin{align}
  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
&=-X'  \left[ c''(\eta)\right]  X   \\
\end{align}

Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
\begin{align}
\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
&=-   \left[ c''(\eta) X  \right]' ZA   \\
&=-X'   \left[ c''(\eta)   \right] ZA  
\end{align}

And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood. Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$. Trust will maximize the penalized likelihood.

The {\bf outer} optimization will also be done using optim with the default method of Nelder-Mead. This requires just the function value.

The outer optimization's function is evaluated by
\begin{align}
\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
l(\tilde{\eta}) &= Y^T \tilde{\eta} - c(\tilde{\eta})\\
A&= \sum_k E_k \sigma_k\\
W&=c''(\eta)=diag(c''(\eta_i))\\
maxvalue&= l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}^T \tilde{s} - \dfrac{1}{2} \log det \left( AZ' W ZA +I  \right)\\
minvalue&= - maxvalue
\end{align}
Again, we have to minimize the negative value of the penalized quasi likelihood in order to maximize the value of it.

We do need to be careful about the determinant.  I did a cholesky decomposition.(Maybe it would be better to do a qr or svd decomposition, but I couldn't figure out whether the q matrix in the qr decomposition would have a determinant of positive or negative one. QR is supposed to be more numerically stable, but slower. Charlie used a cholesky decomposition in aster's PQL so I thought it should be good enough).  Let $AZ' W ZA +I=M=LL'$, where $L$ is the lower triangular matrix resulting from the decomposition.
\begin{align}
.5 \log det(M)) &=.5 \log det(LL')\\
&= .5 \log (det(L))^2 \\
%= 2*.5 \log det(L) \\
&=  \log det(L) \\
&= \log \left( \prod diag(L) \right) \\
&= \sum \log diag(L)
\end{align}

If I need to be more careful and more stable, I can use log1p, which is a function that is good for calculating $\log (1+ a)$ where $a$ is a tiny number. Let $M=AZ' W ZA +I$. Let $O$ be orthogonal and $D$ be diagonal with $\lambda_i$ on the diagona, where $\lambda_i$ are the square roots of the eigenvalues of $M'M$ and $MM'$l. Then that  $AZ' W ZA +I$ can be decomposed in the following way,

\begin{align*}
 &AZ' W ZA = ODO' \\
&\Rightarrow AZ' W ZA +I  = O (I+D) O' \\
&\Rightarrow  det \left( AZ' W ZA +I \right) = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
&\Rightarrow \log  det \left( AZ' W ZA +I \right) = \sum \log(1+\lambda_i) = \sum log1p(\lambda_i)
\end{align*}

The outer optimization will use $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them. I'll want to keep track of the very last  $\tilde{s}$ for generating the random effects, so I'm going to store them in the parent environment with the 
\begin{verbatim}<<- \end{verbatim}

The point of doing PQL is to construct a decent importance sampling distribution. Thus, the estimates don't have to be perfect.  It is possible that one of the $\sigma_k$ will be 0 according to PQL. If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.

\section{generating random effects (genRand)}

This will be hidden from the user. It will be called by the model fitting function.

To approximate the log likelihood, we need simulated random effects.  PQL will give us estimates for the $\sigma_t$s and  the s's. Since we have the $E_t$, we can construct A, which will be useful because it is is the cholesky decomposition for D =AA.

Let $s^*$ and $\sigma^*$ be the vectors of PQL estimates. Then we can ``unstandardize'' our random effects this way:
\begin{align}
u^*&=A^*s^*\\
&= \left(\sum E_t \sigma^*_t \right) s^*
\end{align}

Suppose we want to generate random effects $u_k,k=1,...,m$ where $u_k$ is a vector with length q (the number of random effects in the model). Then for each k, draw a sample of size q from the standard normal distribution. Call this sample $\breve{u}_k$. Then we scale it
\begin{align}
u_k = \breve{u}_k A^* + u^*
\end{align}
so that we are effectively drawing our sample from $N(u^*,A^*A^*)$. We are using this distribution because it is our best guess of the distribution of our random effects.

%Then sample from $N(u^*, D^*)$ $m$ times to generate random effects $u_k,k=1,...,m$.  To create a sample from $N(u^*, D^*)$, I'm actually going to get $\sum_{t=1}^T t$ samples from $N(0,1)$ (call these $\breve{u}_k$) then scale this vector to be from $N(u^*, D^*)$. Using the Cholesky decomposition,  $D^* = L L^*$. Then the scaling works as follows:
%\begin{align}
%u_k = \breve{u}_k L + u^*
%\end{align}
%Note: we are using $u^*$ because we want to generate the $u$s so that they are as close as possible to their most likely values. The $u^*$s are the best guess of $u$ that we have.

Lastly, we need an expression for $\log \tilde{f} (u_k)$. We can write it as a function proportional to the normal distribution with mean $u^*$ and variance $D^*=A^*A^*$.  In other words, we will use the distRand function with $u^*$ as the mean vector and $D^*$ as the variance matrix.

\section{objective function: approximated log likelihood}
This will be hidden from the user. It will be called by the model fitting function.

In order to maximize the approximated log likelihood using trust, I will need an objective function that returns the value, the gradient vector, and the Hessian matrix of the log likelihood.  This objective function will get $\log f_\theta (u_k)$, $c(\eta_i)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.    Denote\\
\begin{align}
b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \\
%&= \log f_\theta (u_k) +\sum_i Y_i \eta_i -\sum_i c(\eta_i) - \log \tilde{f} (u_k)
\end{align}

For computational stability, we'll set $a=max(b_k)$. Then the value is expressed as:\\
\begin{align}
\log f_{\theta,m} = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

Define the weights as:\\
\begin{align}
v_\theta(u_k,y) = \dfrac{e^{b_k-a}}{ \sum_{k=1}^m e^{b_k-a}} 
\end{align}

The gradient vector is: \\
\begin{align}
G &= \sum_{k=1}^m \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\
&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] v_\theta(u_k,y)+ \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)
\end{align}

Then the Hessian matrix is: \\
\begin{align}
H = \sum_{k=1}^m \nabla^2 \log f_\theta (u_k,y) v_\theta(u_k,y) +  \sum_{k=1}^m \left[ \nabla  \log f_\theta (u_k,y)  \right] \left[ \nabla  \log f_\theta (u_k,y)  \right]^T  v_\theta(u_k,y) -G G^T
\end{align}
Everything in this has already been defined except:
\begin{align}
\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
\end{align}

\section{checks}
I'll need to perform some checks to make sure that I've coded these right. Finite differences is one.

\section{summary of model}
Typing ``summary(mod)'' will give more detailed info of the model. This will be broken into two pieces (as is usually done for summaries): there will be the summary.mcml and print.summary.mcml. We split this up because we might have a lot of extra stuff in the summary that we don't necessarily want printed each time. So, when a user types ``summary(mod)'', only the  basic information is automatically printed because that's all that most users want. If they want some of the more sophisticated stuff, they can dig into the summary list and find it.

The summary will do all the calculations and its value will be a list of the following:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix with the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column. (note that we are using the standard normal distribution for the reference distribution, not a t distribution like I initially assumed. This is because all inference is asymptotic.)
\item the evaluated Monte Carlo log likelihood along with its first and second derivative
\item maybe other things that I haven't thought of
\end{itemize}

A note on the standard errors: to calculate the standard errors, we take the MCLA hessian matrix, invert it, take the diagonal elements, and square root them. It's possible that the hessian will be noninvertible if it is close enough to singular that the computer thinks it's singular.  Then the standard errors will all be infinite. We also need to warn the user why this is happening.

Then print.summary.mcml will print these things:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix similar to the output of summary.glm, which will be put through printCoefmat in order to get the significance stars that we're used to. We'll have the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column, the significance stars, and the significance legend.
\item maybe other things that I haven't thought of. I'll probably realize more once I get coding.
\end{itemize}

I think for the printCoefmat to work, the fixed effects matrix needs to look like
\begin{verbatim}
            Estimate Std. Error z value     Pr(>|z|)
(Intercept)        2       0.50       4 6.334248e-05
x1                 2       0.25       8 1.244192e-15
\end{verbatim}
(or at least it worked when I had these as the column names and didn't work before that).

Note that the summary and print.summary functions will be S3 generic functions.  What this means is the user will type ``summary(mod)'' and summary is a generic function. R checks the class of the model $mod$ and then automatically uses the summary and print.summary functions for that class of objects. This also might affect the way that I export the function and document it in the manual.  I think that I'll be ok if I copy parts of his summary.aster and print.summary.aster. I need to read about generics more and will check out the hints library.

\section{likelihood ratio test}
This is another S3 generic function that the user can choose to implement if they'd  like  to do likelihood ratio tests for nested models. Eventually I'll want this to handle an arbitrary number of models, but for now we'll stick to comparing two models: $mod2$ nested in $mod1$.  I will assume these models have already been fit.

 There are a few ways that the two models could differ. In other words, we could be testing:
 \begin{enumerate}
 \item whether one or more fixed effects are zero.
 \item whether one variance component is zero.
 \item whether multiple variance components are zero.
 \item whether one or more fixed effects is zero and one or more variance components are zero. 
\end{enumerate}
The hypothesis testing procedure and method for calculating pvalues are different for each of these cases, so I'll go through each one in the following subsubsections.

\subsection{testing whether one or more fixed effects are zero}
Consider the  case of testing whether one or more fixed effects are zero and the random effects are the same between the two models. I think we just do a likelihood ratio test as we're familiar with for linear models.   Let the number of fixed effect parameters in the larger model be $p_1$  and the number in the nested model be $p_2$. I will need the log likelihood values of the models outputted (from the main function at the same time as when I output the MCMLEs and the Fisher information). The log likelihood is simply the ``value'' from the objective function that trust uses. Let's call the log likelihood from the larger model  $l_1$ and the log likelihood from the nested model  $l_2$.

Then the likelihood ratio test statistic is
\begin{align}
t_{LRT}= -2l_1+2l_2
\end{align}
and this follows a $\chi^2$ distribution with $p_1-p_2$ degrees of freedom. The calculation
\begin{align}
P(t_{\chi^2_{p_1-p_2}>LRT} ).
\end{align}
gives us a two-sided pvalue to fit with the two-sided alternative hypothesis.

 \subsection{testing whether one variance component is zero}
  Hypothesis testing for one variance component is easy but not what most people expect.
  In this setup, the two models have the exact same fixed effects. The only difference is the larger model has one more random effect $\nu_{t_1}$. This means we want to test whether $\nu_{t_1}=0$. A variance component must be nonnegative, meaning the alternative hypothesis is that $\nu_{t_1}>0$. In other words, the hypotheses are:
 \begin{align}
 H_0: \nu_{t_1}=0\\
 H_1: \nu_{t_1}>0.
 \end{align}
 Note the alternative hypothesis is one-sided. This means we need to calculate a one-sided pvalue. Let $t_{LRT}=2l_2-2l_1$. If we naively follow the pvalue calculation of
 \begin{align}
P( \chi^2_{1}>t_{LRT})
\end{align}
 we will end up calculating a two-sided pvalue. To fix this, we cut this pvalue in half. In other words, use the standard normal distribution $Z$ and think of calculating the one-sided pvalue this way:
 \begin{align}
 P( Z>\sqrt{|  t_{LRT} |})
 \end{align}

 However, this is not obvious; I never would have thought about having a test statistic of
 \begin{align}
 \sqrt{2 | l_2-l_1   |   }.
 \end{align}

 \subsection{testing whether multiple variance components are zero}
 In this case, the fixed effects are identical between the larger and nested model. The only difference is that the larger model has more than one additional variance components. This gets a lot more complicated (for example, there is no clear way to count parameters so calculating the degrees of freedom is out the window). Luckily, someone's worked out the details already. Charlie suggested a few papers to look at. I'll read about and add this a bit later.  Charlie is hoping that we'll have a mixture of $\chi^2$ distributions, such as
 \begin{align}
 \dfrac{1}{2}P (\chi^2_1> t_{LRT}) +\dfrac{1}{4}P (\chi^2_2> t_{LRT}).
 \end{align}
 The reason this is complicated is because the variance components are restricted to be nonnegative. This means that the likelihood is only defined for nonnegative variance components. Then differentiating the likelihood at $0$ becomes tricky because that's the boundary.

 \subsection{testing whether one or more fixed effects is zero and one or more variance components are zero}\label{sec:combotest}
 This seems very complicated. I don't know if it's any more complicated than the previous case of testing multiple variance components. I'll have to think about it later when I have time.

 \subsection{determining the hypothesis test}
 To follow convention, this command will be called ``anova'' and the two arguments will be the two models to be compared.  The command would look like
\begin{verbatim}
anova(mod1,mod2)
\end{verbatim}

 R will first need to figure out if the fixed effects are different and, if they are, which model is the larger model.
\begin{verbatim}
if(length(coef(mod1))>length(coef(mod2)))
      {bigmod<-mod1;  smallmod <-mod2} 
if(length(coef(mod1))<length(coef(mod2)))
      {bigmod<-mod2; smallmod <-mod1}
\end{verbatim}
Next, if there is a difference in the fixed effects, I'm going to make sure the fixed effects of the small model are nested in those of the big model. Otherwise, produce an error. Then, continue with the testing. Note: this check for nesting isn't perfect, but no other anova checks which model is bigger and whether they're nested.
\begin{verbatim}
if(big mod is defined) {
      pnames1<-names(coef(bigmod))
      pnames2<-names(coef(smallmod))
      if(sum(pnames2 %in% pnames1) != length(pnames2)) {
            stop("The models you provided are not nested.")}

     if(the variance components differ between the two models){
             check big model has more variance components (ow: error)
             check variance components of the small model are nested (ow: error)
             calculate pvalue according to section 3.3.4. return it.
      }
      if(the variance components do not differ between the two models){
            calculate pvalue according to 3.3.1. return it.
      }
}
\end{verbatim}
If and only if we've gotten to the end of this chunk of code without returning anything, bigmod has not been defined, meaning the fixed effects are same. Therefore, we now need to figure out whether the variance components differ by one or by more than one. In order to compare the number of random effects, we need to figure out how to get the number of variance components from each model. I  don't yet have a solid grasp of how formula works its magic, but I have the impression that I can count the number of model matrices returned in order to figure out the number of variance components because there is a model matrix for each one.

\begin{verbatim}
T1<- number of variance components for mod1
T2<- number of variance components for mod2
if(T1>T2){call mod1 the bigmod and mod2 the smallmod}
if(T1<T2){call mod2 the bigmod and mod1 the smallmod}
if(bigmod is defined){
     check  small model is nested in the big model (ow: error)
     (maybe do this by looking at the call?)
     if(T1==T2+1 || T2==T1+1) 
           {calculate and return pvalue according to sec 3.3.2}     
     calculate and return pvalue according to section 3.3.3
     }
if still haven't returned anything, produce error.     
\end{verbatim}

Why the last error? If we have gotten to this point in the code with nothing returned, it ends up that the number of variance components in each model are equal, meaning the user has made some kind of mistake. Either the user provided two identical models or they provided non-nested models (e.g. they have the same number of variance components, but different components in each model). Either way, we can't help them.

This command would return something that first reminds the user what predictors are in each model, then has a table. Each model would have one row of the table. The columns would contain the model name,  the log likelihood for each model, the number of parameters in the model. Then the next columns would have one entry each: the calculated difference between the log likelihoods, the calculated difference between the number of parameters, and the pvalue of the test.

%\section{AIC}
%Typing 
%\begin{verbatim}
%AIC(mod)
%\end{verbatim} 
%will return the AIC of the model. To calculate AIC, I'll need the total number of parameters ($p+T$)  and the Monte Carlo log likelihood evaluated at the MCMLEs $l$. Then the AIC is
%\begin{align}
%2(p+T) - 2 l 
%\end{align}

\section{confidence intervals}
The user can implement this command to create confidence intervals after  fitting a model $mod$ using the main function.  This is an S3 generic.The command would look like
\begin{verbatim}
confint(mod,parm,level=.95)
\end{verbatim}
The only required argument would be the fitted model (the first argument). The third argument (the confidence level) has a default of $.95$.

If the second argument is omitted, confidence intervals would be created for all of the parameters. There are two options to calculate confidence intervals for a subset of the parameters. The user can provide either  the names of the coefficients or a vector with length equal to the number of parameters (entries being 1 if they would like that intervals for that parameter and 0 otherwise).  The code to do this is taken from ``confint.lm'' and is:
\begin{verbatim}
function (object, parm, level = 0.95, ...) 
{
    cf <- coef(object)
    pnames <- names(cf)
    if (missing(parm)) 
        parm <- pnames
    else if (is.numeric(parm)) 
        parm <- pnames[parm]
        stopifnot(parm %in% pnames)

        rest of the code to actually do the confidence intervals goes here
        }
\end{verbatim}
What this says: write down the coefficient names of the object we're given. If parm is missing, we're going to assume they want to create intervals for all the parameters.  If parm is numeric (a vector of 0s and 1s), use that to select the parameter names we want and call that selection parm.  Finally, stop the whole process if parm (the parameter names we want intervals for) don't appear in the model.

The form for a confidence interval is the point estimate plus or minus the standard error of the point estimate times some cutoff.  The point estimate and the standard error are from the summary of the model. The reference distribution used for the cutoff is the standard normal. This will produce an asymptotic confidence interval.

The output would either be a vector of length 2 (if creating confidence interval for only one parameter) or a matrix with 2 columns (one for the lower bound of the confidence interval, one for the upper bond). The column names will be ``$(100-level)/2$'' and ``$50+level/2 $.'' (in the default case, this is``$2.5\%$'' and ``$97.5\%$'').

The only potential hangup is when creating confidence intervals for the variance components $\nu_t, t=1,...,T$.  We know that $\nu_t >0$ for all $t=1,...,T$. We'll find this boundary again makes things complicated. I don't yet know how to deal with this.

To illustrate the problem, consider the scenario that the margin of error for $\nu_t$ is greater than the estimate of $\nu_t$ itself. It wouldn't make sense to produce a confidence interval with a negative lower bound. It could make sense to truncate the interval so that the lower bound starts at 0 and the upper bound remains untouched.

I thought for three or four minutes about a one sided confidence interval, but I don't think that captures what we want. We want the range of all likely values for the variance component, not just an upper bound.

\section{prediction intervals}
This is another S3 generic. The user can implement this command to create prediction intervals after  fitting a model $mod$ using the main function. Still learning about these. I currently know almost nothing. What I do know is that I will need the Monte Carlo log likelihood along with its first and second derivatives. I'll get to practice my first-year-theory delta method skills.

Charlie says that neither predict.glm nor predict.glmmPQL do prediction intervals. The former produces parameter estimates with standard errors. The latter does predictions but does not produce a standard error, so no interval can be formed.

One wrong thing to do: once you find the MCMLEs, you simply plug those into the distribution for $U|Y$ and pretend those parameter estimates have no variability and are  correct. Then the random effect has a normal distribution and it'd be easy to find quantiles. The only problem is this is wrong.

We are hoping to find a way to use Fisher information for the Monte Carlo MLE. Hopefully we can do something more similar to the usual asymptotics.

A last resort is a parametric bootstrap. I haven't spent much time thinking about this, but I think the way this involves plugging parameter estimates  into $\log f_\theta(y|u)$ and $\log f_\theta(u)$. Then you repeat: draw a value of $u$, plug that into $\log f_\theta(y|u)$, draw a value of $y$. I say this is a last resort  because having two Monte Carlo calculations will take a long time.

\end{document}

### 2014-03-20 GoogleProposal.tex
What the fuck.
Just editing the first page and so of my google summer of code proposal.

:i

\documentclass[12pt]{article}

\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}

\usepackage{setspace}
\usepackage[margin=1in]{geometry}

%\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}
\usepackage{times}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\gbar}{\bar g}
\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} \DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\newcommand{\eps}{\epsilon}
\newtheorem{claim}{Claim}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\def\baro{\vskip  .2truecm\hfill \hrule height.5pt \vskip  .2truecm}
\def\barba{\vskip -.1truecm\hfill \hrule height.5pt \vskip .4truecm}

\title{Monte Carlo Maximum Likelihood for Two-Stage Hierarchical Models}

\author{}
\date{}
\doublespacing
\begin{document}
%\maketitle{}
 \centerline{\large \bf Monte Carlo Maximum Likelihood for Two-Stage Hierarchical Models} %% Paper title

 \medskip

 \centerline{Christina Knudson}
 \smallskip
%\subsection*{Background}
\noindent{\textbf{Background}}

Two-stage hierarchical models are popular in many fields from medicine to economics but are held back by limitations in conventional methodology, which I hope to remedy. The utility of a two-stage hierarchical model can be demonstrated by a hypothetical medical study following a group of people at risk for type 2 diabetes. Each patient is randomly assigned one of two preventative treatments. On a monthly basis, researchers record predictors of diabetes--such as weight and age--and whether the patient has developed diabetes. In this two-stage hierarchical model, the patients form one stage and the repeated measurements on a particular patient form the second stage. Such models enable researchers to test hypotheses about parameters (fixed but unknown numbers, such as the quantitative effect of a treatment on the odds of diabetes development) in order to better understand diabetes prevention. Furthermore, these models offer predictive power, which in this case can identify patients likely to develop diabetes on the basis of predictors such as weight or age. Finally, researchers can use these models to estimate the population's variability in random effects, which reflect the unobservable data influencing an individual's susceptibility to diabetes.

The challenge is finding an easy-to-implement and reliable method for fitting and testing two-stage hierarchical models. At present, researchers in many fields commonly turn to the penalized quasi-likelihood (PQL) method for two-stage hierarchical models due to its ease of use. However, PQL relies on many unmet assumptions and ``approximations of undetermined accuracy'' and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}). The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit two-stage hierarchical models.

As an improvement over PQL, I suggest  Monte Carlo Maximum Likelihood (MCML), a computationally intensive method for model-fitting (\ncite{geyer:thom:1992}). MCML is an attractive method for two-stage hierarchical models because of its rigorous theoretical foundation supplied by \citet{geyer:1994} and  \citet{sung:geyer:2007}.   Moreover, despite the fact that maximum likelihood is standard for many simpler models, MCML is the only method that can perform  it for two-stage hierarchical models. Also, an MCML practitioner can control the accuracy and variability of the estimates  by iterating the  procedure (\ncite{geyer:thom:1992}).

%.Moreover, MCML is the only method that performs maximum likelihood--a standard technique for less complicated statistical models--for two-stage hierarchical models.

Despite MCML's solid theoretical underpinnings, it is not yet a widely-used technique due to two main obstacles. First, the extent of programs available to perform MCML is severely limited, which means MCML practitioners  must dedicate weeks to write their own code. 
%The second obstacle is little research has been done to determine the appropriate number of times to iterate MCML. So that practitioners no longer iterate MCML haphazardly, we need  a rule that iterates MCML until some predetermined criterion is satisfied, such as the estimates' variability reaching a certain threshold.
The second obstacle is the unanswered question of ``How many iterations is enough?''  We need  a rule that iterates MCML until some predetermined criterion is satisfied, such as the estimates' variability reaching a certain threshold. Little research has been done in this area, forcing practitioners to iterate MCML haphazardly.

\noindent{\textbf{Goals and objectives}}
%\subsection*{Goals and objectives}

My goals are  (1) to write computer code to implement MCML for a wide variety of two-stage hierarchical models, (2) to develop a stopping criterion to terminate MCML, and (3) to prove new theorems about MCML's theoretical properties. I intend to share these results with other researchers by submitting programs to R-CRAN and publishing theoretical results in statistical journals.  Because R is a free  program that is popular among statisticians and other researchers, submitting to R-CRAN is an attractive way to allow researchers to benefit from my work.

\noindent{\textbf{Progress to date and schedule for completion}}

%\subsection*{Progress to date and schedule for completion}
I consider my  goals separately. I expect to complete all work by May of 2014.

(1) I have completed a program to model the odds  of an event based on a single random effect with a normal distribution. Simulation studies I have performed thus far demonstrate the success of MCML and my code: after generating data from fixed parameter values, the MCML parameter estimates  match the  parameters used to generate the data. I intend to expand my code  to include more than one random effect and to offer a variety of distributions for the random effects (instead of solely the normal distribution). In addition to modeling the odds of an event, I plan to model other measurements as well.

(2) I have performed simulation studies to understand the behavior of parameter estimates resulting from iterations of MCML. I have studied similar behavior for related methods and read  a stopping criterion based on the parameter estimates' variance that I would like to investigate.

(3) My bachelor's degree and graduate coursework in mathematics provide me with a solid foundation to undertake  theoretical problems. Theoretical papers have developed my understanding of properties that are important in methods similar to MCML.  To understand the  theorems that have already been proven for MCML, I have read papers by \citet{geyer:1994} as well as \citet{sung:geyer:2007}. As a start, I have identified theorems that I wish to extend.

\noindent{\textbf{Potential significance of research}}

In conclusion, researchers from all fields have created a high demand for a tool to fit and test two-stage hierarchical models. Ideally, fitting a two-stage hierarchical model should be as user-friendly and theoretically-grounded as fitting a simpler model. Coding MCML for two-stage hierarchical models and determining a stopping criterion will make MCML user-friendly while proving new theorems will solidify MCML's theoretical foundation. Researchers ranging from the earth sciences to psychology will benefit from these efforts.

\noindent{\textbf{References}}
\vspace{-2.3cm}
\renewcommand{\refname}{}
\bibliographystyle{apalike}
\bibliography{brref}

\end{document}

### 2014-03-20 GoogleProposal.tex
Just doing some more work on the Google Proposal!

@@
 
-\title{Monte Carlo Maximum Likelihood for Two-Stage Hierarchical Models}
+\title{Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation}
 
@@
 %\maketitle{}
- \centerline{\large \bf Monte Carlo Maximum Likelihood for Two-Stage Hierarchical Models} %% Paper title
+ \centerline{\large \bf Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation} %% Paper title
 
@@
 
- \centerline{Christina Knudson}
+ \centerline{Christina Knudson\\
+christina@umn.edu\\
+507-384-2220}
  \smallskip
@@
 
-Two-stage hierarchical models are popular in many fields from medicine to economics but are held back by limitations in conventional methodology, which I hope to remedy.
-The utility of a two-stage hierarchical model can be demonstrated by a hypothetical medical study following a group of people at risk for type 2 diabetes.
-Each patient is randomly assigned one of two preventative treatments.
-On a monthly basis, researchers record predictors of diabetes--such as weight and age--and whether the patient has developed diabetes.
-In this two-stage hierarchical model, the patients form one stage and the repeated measurements on a particular patient form the second stage.
-Such models enable researchers to test hypotheses about parameters (fixed but unknown numbers, such as the quantitative effect of a treatment on the odds of diabetes development) in order to better understand diabetes prevention.
-Furthermore, these models offer predictive power, which in this case can identify patients likely to develop diabetes on the basis of predictors such as weight or age.
-Finally, researchers can use these models to estimate the population's variability in random effects, which reflect the unobservable data influencing an individual's susceptibility to diabetes.
+Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics but are limited by conventional methodology.
+The utility of a GLMM can be demonstrated by a hypothetical medical study following a group of people at risk for type 2 diabetes.
+Each patient is randomly assigned one of two preventative treatments: regular exercise or an improved diet.
+Every year, the patients return to their doctors, who record  whether each patient has developed diabetes.
+With this information, we could test whether one preventative treatment is more effective than the other.
+Because the patient's doctor could also have an effect on the patient's motivation to adhere to the preventative treatment, we could have a random effect for each doctor.
+Incorporating this into the model would allow us to test for whether there is significant variability from doctor to doctor.
+If the estimated variance component is large, we may want to investigate further to determine why some doctors are more successful at preventing type 2 diabetes than other doctors.
 
 The challenge is finding an easy-to-implement and reliable method for fitting and testing two-stage hierarchical models.
-At present, researchers in many fields commonly turn to the penalized quasi-likelihood (PQL) method for two-stage hierarchical models due to its ease of use.
+For very simple problems with just a few random effects, the likelihood can be found through numerical integration.
+Since numerical integration is not feasible in most real-world problems, a more commonly used method is penalized quasi-likelihood (PQL), implemented in packages such as lme4 and nlme.
 However, PQL relies on many unmet assumptions and ``approximations of undetermined accuracy'' and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}).
+Also, because PQL maximizes a quasi-likelihood rather than the likelihood, likelihood-based inference (such as testing for whether there is significant variability between doctors in the diabetes example) cannot be performed.
 The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit two-stage hierarchical models.
 
-As an improvement over PQL, I suggest  Monte Carlo Maximum Likelihood (MCML), a computationally intensive method for model-fitting (\ncite{geyer:thom:1992}).
-MCML is an attractive method for two-stage hierarchical models because of its rigorous theoretical foundation supplied by \citet{geyer:1994} and  \citet{sung:geyer:2007}.
-Moreover, despite the fact that maximum likelihood is standard for many simpler models, MCML is the only method that can perform  it for two-stage hierarchical models.
-Also, an MCML practitioner can control the accuracy and variability of the estimates  by iterating the  procedure (\ncite{geyer:thom:1992}).
+As an improvement over PQL, I suggest  Monte Carlo  Likelihood Approximation (MCLA), a computationally intensive method for approximating the likelihood to fit and test GLMMs (\ncite{geyer:thom:1992}).
+Because MCLA approximates the entire likelihood, any type of likelihood-based inference can be performed.
+Inference such as maximum likelihood or likelihood-ratio testing is standard for many simpler models, but MCLA is the only method that can perform these techniques for GLMMs.
+Moreover, MCLA is supported by a rigorous theoretical foundation supplied by \citet{geyer:1994} and  \citet{sung:geyer:2007}.
 
-%.Moreover, MCML is the only method that performs maximum likelihood--a standard technique for less complicated statistical models--for two-stage hierarchical models.
+Despite MCLA's solid theoretical underpinnings, it is not yet a widely-used technique because there are  few packages available that perform MCLA.
+The main problem is that MCLA is  computationally expensive.
+My proposed package would improve MCLA's efficiency by using an importance sampling distribution that is similar to the true distribution of the random effects.
+Random effects are usually assumed to be normally distributed, so all that is left to specify are the mean and variance parameters, which my  package chooses  based on the data at hand.
 
-Despite MCML's solid theoretical underpinnings, it is not yet a widely-used technique due to two main obstacles.
-First, the extent of programs available to perform MCML is severely limited, which means MCML practitioners  must dedicate weeks to write their own code.
-%The second obstacle is little research has been done to determine the appropriate number of times to iterate MCML.
-So that practitioners no longer iterate MCML haphazardly, we need  a rule that iterates MCML until some predetermined criterion is satisfied, such as the estimates' variability reaching a certain threshold.
-The second obstacle is the unanswered question of ``How many iterations is enough?''  We need  a rule that iterates MCML until some predetermined criterion is satisfied, such as the estimates' variability reaching a certain threshold.
-Little research has been done in this area, forcing practitioners to iterate MCML haphazardly.
+\noindent{\textbf{Progress to date}}
+
+%\subsection*{Progress to date and schedule for completion}
+Consider a GLMM with a Poisson or Bernoulli response using the canonical link.
+Assume the random effects are independently drawn from a normal distribution with mean 0 and unknown variances.
+There can be any number of fixed or random effects.
+For this setting, I have been working on an R package to approximate and maximize the likelihood via MCLA.
+The package is in the testing stage and is nearing completion for the setting described earlier in this paragraph.
 
@@
 
-My goals are  (1) to write computer code to implement MCML for a wide variety of two-stage hierarchical models, (2) to develop a stopping criterion to terminate MCML, and (3) to prove new theorems about MCML's theoretical properties.
-I intend to share these results with other researchers by submitting programs to R-CRAN and publishing theoretical results in statistical journals.
-Because R is a free  program that is popular among statisticians and other researchers, submitting to R-CRAN is an attractive way to allow researchers to benefit from my work.
+My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) generalize the model so that models with additional variance structures can be fitted.
 
-\noindent{\textbf{Progress to date and schedule for completion}}
+\noindent{\textbf{Schedule for completion}}
 
-%\subsection*{Progress to date and schedule for completion}
-I consider my  goals separately. I expect to complete all work by May of 2014.
+I consider my  goals separately.
 
@@
 
-In conclusion, researchers from all fields have created a high demand for a tool to fit and test two-stage hierarchical models.
-Ideally, fitting a two-stage hierarchical model should be as user-friendly and theoretically-grounded as fitting a simpler model.
-Coding MCML for two-stage hierarchical models and determining a stopping criterion will make MCML user-friendly while proving new theorems will solidify MCML's theoretical foundation.
-Researchers ranging from the earth sciences to psychology will benefit from these efforts.
+In conclusion, researchers from all fields have created a high demand for a tool to fit and test GLMMs.
+Ideally, fitting a GLMM should be as user-friendly and theoretically-grounded as fitting a simpler model.
+Coding MCLA for GLMMs will make MCLA user-friendly.
 

### 2014-03-20 GoogleProposal.tex
Working on my Google Summer of Code

@@
  \smallskip
-%\subsection*{Background}
 \noindent{\textbf{Background}}
 
-Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics but are limited by conventional methodology.
+Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics.
 The utility of a GLMM can be demonstrated by a hypothetical medical study following a group of people at risk for type 2 diabetes.
-Each patient is randomly assigned one of two preventative treatments: regular exercise or an improved diet.
-Every year, the patients return to their doctors, who record  whether each patient has developed diabetes.
-With this information, we could test whether one preventative treatment is more effective than the other.
+Each patient is randomly assigned to a doctor and to one of two preventative treatments: regular exercise or an improved diet.
+Every year, the doctors record  whether each patient has developed diabetes.
+With this information, we could test whether one treatment is more effective than the other at preventing type 2 diabetes.
 Because the patient's doctor could also have an effect on the patient's motivation to adhere to the preventative treatment, we could have a random effect for each doctor.
@@
 
+\noindent{\textbf{Existing R packages}}
+
 The challenge is finding an easy-to-implement and reliable method for fitting and testing two-stage hierarchical models.
-For very simple problems with just a few random effects, the likelihood can be found through numerical integration.
-Since numerical integration is not feasible in most real-world problems, a more commonly used method is penalized quasi-likelihood (PQL), implemented in packages such as lme4 and nlme.
+For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.
+Most models have crossed random effects or too many random effects for numerical integration to handle.
+Thus, a more commonly used method is penalized quasi-likelihood (PQL), implemented in packages such as lme4 and nlme.
 However, PQL relies on many unmet assumptions and ``approximations of undetermined accuracy'' and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}).
 Also, because PQL maximizes a quasi-likelihood rather than the likelihood,  likelihood-based inference (such as testing for whether there is significant variability between doctors in the diabetes example) cannot be performed.
-The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit two-stage hierarchical models.
+The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit GLMMs.
+
+\noindent{\textbf{Proposed improvement to methodology}}
 
@@
 
-%\subsection*{Progress to date and schedule for completion}
 Consider a GLMM with a Poisson or Bernoulli response using the canonical link.
@@
 The package is in the testing stage and is nearing completion for the setting described earlier in this paragraph.
+This package is part of my doctoral thesis in statistics, which I am earning at the University of Minnesota with Professors Charles Geyer and Galin Jones as my co-advisors.
 
-\noindent{\textbf{Goals and objectives}}
-%\subsection*{Goals and objectives}
+\noindent{\textbf{Goals and objectives for Google Summer of Code}}
 
-My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) generalize the model so that models with additional variance structures can be fitted.
+My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) write additional functions to fit models with correlated random effects.
+These goals are extensions of my current and past work on the topic of using MCLA to fit GLMMs.
 
-\noindent{\textbf{Schedule for completion}}
+\noindent{\textbf{Details and schedule for completion}}
 
-I consider my  goals separately.
+I consider my  goals separately. Completion of one goal does not depend on completion of the other goals, so I could work on the three goals in any order.
 
-(1) I have completed a program to model the odds  of an event based on a single random effect with a normal distribution.
-Simulation studies I have performed thus far demonstrate the success of MCML and my code: after generating data from fixed parameter values, the MCML parameter estimates  match the  parameters used to generate the data.
-I intend to expand my code  to include more than one random effect and to offer a variety of distributions for the random effects (instead of solely the normal distribution).
-In addition to modeling the odds of an event, I plan to model other measurements as well.
+(1) Two steps stand out in my package as time-consuming: the step that decides the parameters for the importance sampling distribution and the step that maximizes the likelihood approximation.
+I learned about using .C to call C from R in a statistical computing course, so I will have those notes to refer to.
+Because I have written functioning R code that performs these, I will be able to compare the R results to the C results to check my work.
+I believe I can accomplish this goal in a month.
 
-(2) I have performed simulation studies to understand the behavior of parameter estimates resulting from iterations of MCML.
-I have studied similar behavior for related methods and read  a stopping criterion based on the parameter estimates' variance that I would like to investigate.
+(2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component (and possibly some fixed effects), the nested models differ by two or more variance components (and possibly by some fixed effects).
+I have worked out the details for calculating the test statistics and p-values for the first two cases and will  be able to write the code for these two in about three weeks.
+Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.
+If I do not finish this last piece during the summer, I will finish it in the fall.
 
-(3) My bachelor's degree and graduate coursework in mathematics provide me with a solid foundation to undertake  theoretical problems.
-Theoretical papers have developed my understanding of properties that are important in methods similar to MCML.
-To understand the  theorems that have already been proven for MCML, I have read papers by \citet{geyer:1994} as well as \citet{sung:geyer:2007}.
-As a start, I have identified theorems that I wish to extend.
+(3) The covariance matrix for the random effects in my currently working code is diagonal, meaning the random effects are independently drawn based on one of possibly many variance components.
+I would like to generalize the covariance matrix in order to fit models with time-dependence and location-dependence.
+This generality would make my R package more usable and practical.
+To fit these types of models, I would like to code two additional variance structures:  first-order autoregressive structure with homogenous variances and exponential decay based on the distance between observations.
+I have written my current package with these future changes in mind.
+Coding and testing with these new covariance structures will take about a month.
+
+I expect to complete this work by August 11.
 
@@
 In conclusion, researchers from all fields have created a high demand for a tool to fit and test GLMMs.
-Ideally, fitting a GLMM should be as user-friendly and theoretically-grounded as fitting a simpler model.
-Coding MCLA for GLMMs will make MCLA user-friendly.
+I would like to make the process of fitting a GLMM as user-friendly and theoretically-grounded as fitting a linear model.
+My package will allow researchers to find maximum likelihood estimates for GLMMs with various covariance structures and perform likelihood ratio tests to compare the fit of nested models.
 

### 2014-03-21 GoogleProposal.tex
Doing the changes that Charlie suggested.

@@
 
-Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics.
-The utility of a GLMM can be demonstrated by a hypothetical medical study following a group of people at risk for type 2 diabetes.
-Each patient is randomly assigned to a doctor and to one of two preventative treatments: regular exercise or an improved diet.
-Every year, the doctors record  whether each patient has developed diabetes.
-With this information, we could test whether one treatment is more effective than the other at preventing type 2 diabetes.
-Because the patient's doctor could also have an effect on the patient's motivation to adhere to the preventative treatment, we could have a random effect for each doctor.
-Incorporating this into the model would allow us to test for whether there is significant variability from doctor to doctor.
-If the estimated variance component is large, we may want to investigate further  to determine why some doctors are more successful at preventing type 2 diabetes than other doctors.
+Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics. The utility of a GLMM is apparent through a Google search: ``GLMM'' return over 242,000 results.
 
@@
 These goals are extensions of my current and past work on the topic of using MCLA to fit GLMMs.
+Because I have been working on this project and am familiar with the limitations and areas of possible improvement, I am the best individual for this project.
+Also, I have access to a wealth of knowledge and experience: one of my advisors is the inventor of MCLA and both of my advisors have submitted and maintained R packages.
 

### 2014-03-21 GoogleProposal.tex
Still working on that Google Proposal. more edits

@@
 
-Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics. The utility of a GLMM is apparent through a Google search: ``GLMM'' return over 242,000 results.
+Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics.
+Researchers clearly want to use GLMMs: a Google search for ``GLMM'' returns about 242,000 results.
+The challenge for researchers is finding an easy-to-implement and reliable method for fitting and testing GLMMs.
+For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.
 
-\noindent{\textbf{Existing R packages}}
-
-The challenge is finding an easy-to-implement and reliable method for fitting and testing two-stage hierarchical models.
-For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.
-Most models have crossed random effects or too many random effects for numerical integration to handle.
-Thus, a more commonly used method is penalized quasi-likelihood (PQL), implemented in packages such as lme4 and nlme.
-However, PQL relies on many unmet assumptions and ``approximations of undetermined accuracy'' and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}).
-Also, because PQL maximizes a quasi-likelihood rather than the likelihood,  likelihood-based inference (such as testing for whether there is significant variability between doctors in the diabetes example) cannot be performed.
+Most models have crossed random effects, which numerical integration cannot handle.
+Thus, a commonly used method is penalized quasi-likelihood (PQL),which is  implemented in packages such as lme4, nlme, and MASS.
+However, PQL relies on approximations of unknown accuracy to approximate the likelihood and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}).
+Since the likelihood is approximated to an unknown accuracy by the quasi-likelihood, any inference performed on the approximated likelihood will also produce results with an unknown level of accuracy.
+Without bootstrapping, the user cannot know how valid their confidence intervals or likelihood ratio test results are when using PQL.
 The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit GLMMs.
 
-\noindent{\textbf{Proposed improvement to methodology}}
-
-As an improvement over PQL, I suggest  Monte Carlo  Likelihood Approximation (MCLA), a computationally intensive method for approximating the likelihood to fit and test GLMMs (\ncite{geyer:thom:1992}).
+Monte Carlo  Likelihood Approximation (MCLA) is another tool for fitting GLMMs.
+This method approximates the likelihood either through Markov Chain Monte Carlo (MCMC) or Ordinary Monte Carlo (OMC), and the resulting likelihood approximation is used to fit and test GLMMs (\ncite{geyer:thom:1992}).
 Because MCLA approximates the entire likelihood, any type of likelihood-based inference can be performed.
@@
 Moreover, MCLA is supported by a rigorous theoretical foundation supplied by \citet{geyer:1994} and  \citet{sung:geyer:2007}.
+Despite MCLA's solid theoretical underpinnings, it is not yet a widely-used technique.
+MCLA via MCMC is  too difficult for most users because they do not know when the Markov chain has run long enough to produce reliable answers.
+\pcite{sung:geyer:2007} version of MCMLA via OMC is more user-friendly but is limited to smaller problems.
 
-Despite MCLA's solid theoretical underpinnings, it is not yet a widely-used technique because there are  few packages available that perform MCLA.
-The main problem is that MCLA is  computationally expensive.
-My proposed package would improve MCLA's efficiency by using an importance sampling distribution that is similar to the true distribution of the random effects.
-Random effects are usually assumed to be normally distributed, so all that is left to specify are the mean and variance parameters, which my  package chooses  based on the data at hand.
-
-\noindent{\textbf{Progress to date}}
-
-Consider a GLMM with a Poisson or Bernoulli response using the canonical link.
-Assume the random effects are independently drawn from a normal distribution with mean 0 and unknown variances.
+My current work performs MCLA via OMC with an improved importance sampling distribution.
+Rather than selecting an importance sampling distribution independently of the data, my  package uses an importance sampling distribution that is similar to the true distribution of the random effects.
+The importance sampling distribution is specified based on the data.
+With this importance sampling distribution, my package performs MCLA for GLMMs with a Poisson or Bernoulli response using the canonical link.
+The package assumes the random effects are independently drawn from a normal distribution with mean 0 and unknown variances.
 There can be any number of fixed or random effects.
-For this setting, I have been working on an R package to approximate and maximize the likelihood via MCLA.
 The package is in the testing stage and is nearing completion for the setting described earlier in this paragraph.
@@
 My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) write additional functions to fit models with correlated random effects.
-These goals are extensions of my current and past work on the topic of using MCLA to fit GLMMs.
+My goals are extensions of my current  work on the topic of using MCLA to fit GLMMs.
 Because I have been working on this project and am familiar with the limitations and areas of possible improvement, I am the best individual for this project.
@@
 (1) Two steps stand out in my package as time-consuming: the step that decides the parameters for the importance sampling distribution and the step that maximizes the likelihood approximation.
-I learned about using .C to call C from R in a statistical computing course, so I will have those notes to refer to.
 Because I have written functioning R code that performs these, I will be able to compare the R results to the C results to check my work.
@@
 Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.
-If I do not finish this last piece during the summer, I will finish it in the fall.
 

### 2014-03-21 GoogleProposal.tex
This fucking git thing is so stupid.
Edited my google proposal more and git is pissing me off.

@@
 Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics.
-Researchers clearly want to use GLMMs: a Google search for ``GLMM'' returns about 242,000 results.
+The 242,000 results from a Google search for ``GLMM'' show the popularity of GLMMs.
 The challenge for researchers is finding an easy-to-implement and reliable method for fitting and testing GLMMs.
 For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.
-
 Most models have crossed random effects, which numerical integration cannot handle.
-Thus, a commonly used method is penalized quasi-likelihood (PQL),which is  implemented in packages such as lme4, nlme, and MASS.
+Thus, a commonly used method is penalized quasi-likelihood (PQL), which is  implemented in packages such as lme4, nlme, and MASS.
 However, PQL relies on approximations of unknown accuracy to approximate the likelihood and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}).
@@
 
-My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) write additional functions to fit models with correlated random effects.
-My goals are extensions of my current  work on the topic of using MCLA to fit GLMMs.
-Because I have been working on this project and am familiar with the limitations and areas of possible improvement, I am the best individual for this project.
-Also, I have access to a wealth of knowledge and experience: one of my advisors is the inventor of MCLA and both of my advisors have submitted and maintained R packages.
+My goals are  (1) to rewrite sections of my package in C to  improve its speed, (2) write functions to perform likelihood ratio tests for comparing nested models, (3) write additional functions to fit models with correlated random effects.
 
@@
 
-(2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component (and possibly some fixed effects), the nested models differ by two or more variance components (and possibly by some fixed effects).
+(2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component and possibly some fixed effects, the nested models differ by two or more variance components and possibly by some fixed effects.
 I have worked out the details for calculating the test statistics and p-values for the first two cases and will  be able to write the code for these two in about three weeks.
@@
 My package will allow researchers to find maximum likelihood estimates for GLMMs with various covariance structures and perform likelihood ratio tests to compare the fit of nested models.
+This proposed work is an extension of my current  research on the topic of using MCLA to fit GLMMs.
+I am the best individual for this project because I am familiar with MCLA's current limitations and areas of possible improvement.
 

### 2014-03-21 GoogleProposal.tex
Following GSOC template
That's all.

@@
 
-\title{Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation}
-
 \author{}
@@
 \begin{document}
-%\maketitle{}
- \centerline{\large \bf Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation} %% Paper title
+ \centerline{\large \bf Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation}
+
+ \centerline{\large  Short Title: Monte Carlo Likelihood Approximation}
+
+ \centerline{ http://users.stat.umn.edu/~christina/SummerofCode.pdf}
+ \centerline{Christina Knudson}
+ \medskip
+
+\noindent{\textbf{Bio}}
+
+I'm a doctoral candidate at the University of Minnesota's School of Statistics.
+I am ABD and about a year away from graduating.
+I earned my BA in Math from Carleton College and graduated magna cum laude.
+I was born and raised in Decorah, Iowa, which is one of the top 20 small towns in America (according to Smithsonian Magazine).
+
+I started coding in the spring of 2007, first in Python and then Java.
+I started using R during the summer of 2007 at the Summer Institute for Training in Biostatistics, then continued programming with R during my summer internship at the National Institutes of Health in 2008.
+Most of my graduate coursework has been in R, and I have taught undergraduate classes in R at the University of Minnesota as well.
+Part of my work for my doctoral thesis is an R package that fits Generalized Linear Mixed Models (GLMMs) using Monte Carlo Likelihood Approximation (MCLA).
+I have written part of my package already and I plan to expand and generalize it this summer.
+ \medskip
+
+\noindent{\textbf{Contact Information}}\\
+Student name: Christina Knudson\\
+Link id: knud0158\\
+Student postal address: 1901 Minnehaha Ave, Apt 317, Minneapolis MN, 55404\\
+Telephone: 1-507-384-2220\\
+Emails: knud0158@umn.edu, christina@umn.edu, cknudson05@gmail.com
 
@@
 
- \centerline{Christina Knudson\\
-christina@umn.edu\\
-507-384-2220}
- \smallskip
+\noindent{\textbf{Student Affiliation}}\\
+Institution: University of Minnesota\\
+Program: Statistics \\
+Stage of completion: Early 2015 \\
+Contact to verify: charlie@stat.umn.edu or galin@stat.umn.edu\\
+Advisors: Charles Geyer and Galin Jones
+ \medskip
+
+\noindent{\textbf{Schedule Conflicts}}\\
+During August 3 through 7, I will  attend the Joint Statistical Meetings to present my research.
+
+ \medskip
+
+\noindent{\textbf{Mentors}}\\
+Mentor names: Charles Geyer and Galin Jones\\
+Mentor emails: charlie@stat.umn.edu and galin@stat.umn.edu\\
+Mentor link ids: cjgeyer\\
+I have been in touch with my mentors.
+I meet with each of them at least weekly, and sometimes I talk to Charlie several times per week.
+They are both expecting me to complete the research I have proposed.
+
+ \medskip
 \noindent{\textbf{Background}}
 
-Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics.
+GLMMs are popular in many fields from ecology to economics.
 The 242,000 results from a Google search for ``GLMM'' show the popularity of GLMMs.

### 2014-03-21 GoogleProposal.tex
Still editing the google proposal for the summer of code.

@@
 
-\noindent{\textbf{Details and schedule for completion}}
+\noindent{\textbf{Details}}
 
-I consider my  goals separately. Completion of one goal does not depend on completion of the other goals, so I could work on the three goals in any order.
+I consider my  goals separately, since completion of one goal does not rely on completion of the other goals.
 
 (1) Two steps stand out in my package as time-consuming: the step that decides the parameters for the importance sampling distribution and the step that maximizes the likelihood approximation.
-Because I have written functioning R code that performs these, I will be able to compare the R results to the C results to check my work.
-I believe I can accomplish this goal in a month.
+Thus, I will need to rewrite these two functions in C.
+The main obstacle here will be coding in C, with which I do not have extensive experience.
+Because I have written functioning R code that performs these steps, I will be able to compare the R results to the C results to verify my functions are correct.
+I have been working with a couple of data sets, including the benchmark \citet{booth:hobert:1999} data set with known maximum likelihood estimates, so I will be able to test my code on these data sets.
+Because I can rewrite one function in C without affecting the other functions, I should be able to write these functions in either order.
+I think it would be better to rewrite the step that maximizes the likelihood approximation first, since  that function is more computationally-intensive and is also more important.
+The function that chooses the parameter values of the importance sampling distribution can be rewritten in C second because it is less time-consuming than the other function as it is.
+The equations for these functions are detailed in my design document, which is on my website (http://users.stat.umn.edu/~knud0158/)
 
 (2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component and possibly some fixed effects, the nested models differ by two or more variance components and possibly by some fixed effects.
-I have worked out the details for calculating the test statistics and p-values for the first two cases and will  be able to write the code for these two in about three weeks.
+I have worked out the details for calculating the test statistics and p-values for the first two cases in my design document on my website (http://users.stat.umn.edu/~knud0158/).
 Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.
+Part of the challenge will be writing a function so that, given two models, the code will know which test statistics and pvalues to calculate and report.
+My advisor Charlie has written a function in his aster package that also does model comparison, so I will look to that for guidance.
+I will be able to test my code  the \citet{coull:agresti:2000} flu data set, by modeling the log odds of catching the flu over four years.
+The model will have a few variance components that I will be able to test: a variance component for a subject-specific random effect, another for a year-to-year random effect, and another for the decreased chance of getting flu if the subject previously contracted that same strain of flew virus.
 
 (3) The covariance matrix for the random effects in my currently working code is diagonal, meaning the random effects are independently drawn based on one of possibly many variance components.
-I would like to generalize the covariance matrix in order to fit models with time-dependence and location-dependence.
+I would like to generalize the covariance matrix in order to fit models with location-dependence.
 This generality would make my R package more usable and practical.
-To fit these types of models, I would like to code two additional variance structures:  first-order autoregressive structure with homogenous variances and exponential decay based on the distance between observations.
-I have written my current package with these future changes in mind.
-Coding and testing with these new covariance structures will take about a month.
+To fit these types of models, I would like to code an additional variance structures with  exponential decay based on the distance between observations.
+I have not written the details of how I will execute these changes into my design document yet, but I have written my current package with these future changes in mind.
+I will test my code on the \citet{caffojj:2005} automobile theft data set by modeling the number of cars stolen in a Baltimore neighborhood based on the distance to sites of other car thefts.
+I will compare my results to the results achieved through Monte Carlo EM.
 
-I expect to complete this work by August 11.
-
-\noindent{\textbf{Potential significance of research}}
-
-In conclusion, researchers from all fields have created a high demand for a tool to fit and test GLMMs.
-I would like to make the process of fitting a GLMM as user-friendly and theoretically-grounded as fitting a linear model.
-My package will allow researchers to find maximum likelihood estimates for GLMMs with various covariance structures and perform likelihood ratio tests to compare the fit of nested models.
-This proposed work is an extension of my current  research on the topic of using MCLA to fit GLMMs.
-I am the best individual for this project because I am familiar with MCLA's current limitations and areas of possible improvement.
+\noindent{\textbf{Proposed Timeline}}
+\begin{itemize}
+\item[] May 19 to May 26: look at Charlie's aster package code for model comparison. Design and write code for my own package to determine how the two models differ.
+\item[] May 26 to June 2: write the hypothesis testing code for the first two cases detailed earlier.
+\item[] June 2 to June 15: test and correct the hypothesis testing code.
+\item[] June 16: Complete documentation for hypothesis testing function.
+\item[] June 16 to June 30: write C function that maximizes the likelihood approximation.
+\item[] June 30 to July 7: test the newly-written C function and compare it to my R results. 
+\item[] July 7 to July 14: write C function that selects the importance sampling distribution.
+\item[]July 14 to July 21: test the newly-written C function and compare it to my R results.
+\item[] July 21 to July 28: write function for distance-based variance structure and incorporate into existing R code.
+\item[] July 28 to August 2: test the new variance structure and compare results to those reported by \citet{caffojj:2005}.
+\item[] August 8 to August 11: document the new variance structure.
+\item[] August 11: Submit R package to CRAN.
+\end{itemize}
+I expect to complete all work by August 11.
 

### 2014-03-21 GoogleProposal.tex
Ok apparently now this stupid thing will work.

@@
 
+\usepackage{hyperref}
+
 \def\baro{\vskip  .2truecm\hfill \hrule height.5pt \vskip  .2truecm}
@@
 
- \centerline{ http://users.stat.umn.edu/~christina/SummerofCode.pdf}
+ \centerline{ \url{http://users.stat.umn.edu/~christina/GoogleProposal.pdf}}
  \centerline{Christina Knudson}
@@
 I am ABD and about a year away from graduating.
-I earned my BA in Math from Carleton College and graduated magna cum laude.
+I graduated cum laude from Carleton college with a BA in Mathematics.
 I was born and raised in Decorah, Iowa, which is one of the top 20 small towns in America (according to Smithsonian Magazine).
@@
 I started coding in the spring of 2007, first in Python and then Java.
-I started using R during the summer of 2007 at the Summer Institute for Training in Biostatistics, then continued programming with R during my summer internship at the National Institutes of Health in 2008.
+I started using R during the summer of 2007 at the Summer Institute for Training in Biostatistics, then I continued programming with R during my summer internship at the National Institutes of Health in 2008.
 Most of my graduate coursework has been in R, and I have taught undergraduate classes in R at the University of Minnesota as well.
@@
 Student name: Christina Knudson\\
-Link id: knud0158\\
+Link id: cknudson05\\
 Student postal address: 1901 Minnehaha Ave, Apt 317, Minneapolis MN, 55404\\
 Telephone: 1-507-384-2220\\
-Emails: knud0158@umn.edu, christina@umn.edu, cknudson05@gmail.com
-
+Emails: knud0158@umn.edu, christina@umn.edu, cknudson05@gmail.com\\
+Website: \url{http://users.stat.umn.edu/~christina/}
  \medskip
@@
 \noindent{\textbf{Schedule Conflicts}}\\
-During August 3 through 7, I will  attend the Joint Statistical Meetings to present my research.
+During August 3 through 7, I will  attend the Joint Statistical Meetings to present my R package.
 
@@
 Mentor link ids: cjgeyer\\
-I have been in touch with my mentors.
-I meet with each of them at least weekly, and sometimes I talk to Charlie several times per week.
-They are both expecting me to complete the research I have proposed.
+I have been in touch with my mentors about my package.  I meet with each of them at least weekly, and sometimes I talk to Charlie several times per week.
 
@@
 GLMMs are popular in many fields from ecology to economics.
-The 242,000 results from a Google search for ``GLMM'' show the popularity of GLMMs.
+The popularity of GLMMs is apparent through a Google search, which yields  242,000 results.
 The challenge for researchers is finding an easy-to-implement and reliable method for fitting and testing GLMMs.
@@
 Since the likelihood is approximated to an unknown accuracy by the quasi-likelihood, any inference performed on the approximated likelihood will also produce results with an unknown level of accuracy.
-Without bootstrapping, the user cannot know how valid their confidence intervals or likelihood ratio test results are when using PQL.
+Without bootstrapping, a PQL user cannot know how valid their confidence intervals or likelihood ratio test results are.
 The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit GLMMs.
@@
 The function that chooses the parameter values of the importance sampling distribution can be rewritten in C second because it is less time-consuming than the other function as it is.
-The equations for these functions are detailed in my design document, which is on my website (http://users.stat.umn.edu/~knud0158/)
+The equations for these functions are detailed in my design document, which is on my website at \url{http://users.stat.umn.edu/~knud0158/designdoc.pdf}.
 
 (2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component and possibly some fixed effects, the nested models differ by two or more variance components and possibly by some fixed effects.
-I have worked out the details for calculating the test statistics and p-values for the first two cases in my design document on my website (http://users.stat.umn.edu/~knud0158/).
+I have worked out the details for calculating the test statistics and p-values for the first two cases in my design document at \url{http://users.stat.umn.edu/~knud0158/designdoc.pdf}.
 Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.
@@
 My advisor Charlie has written a function in his aster package that also does model comparison, so I will look to that for guidance.
-I will be able to test my code  the \citet{coull:agresti:2000} flu data set, by modeling the log odds of catching the flu over four years.
-The model will have a few variance components that I will be able to test: a variance component for a subject-specific random effect, another for a year-to-year random effect, and another for the decreased chance of getting flu if the subject previously contracted that same strain of flew virus.
+I will be able to test my code on the \citet{coull:agresti:2000} flu data set by modeling the log odds of catching the flu over four years.
+The model will have a few variance components that I will be able to test: a variance component for a subject-specific random effect, another for a year-to-year random effect, and another for the decreased chance of getting flu when a  strain of flu virus reappears in a later year.
 
@@
 \item[] June 2 to June 15: test and correct the hypothesis testing code.
-\item[] June 16: Complete documentation for hypothesis testing function.
+\item[] June 16: Complete documentation for hypothesis testing function and submit updated R package to CRAN.
 \item[] June 16 to June 30: write C function that maximizes the likelihood approximation.
 \item[] June 30 to July 7: test the newly-written C function and compare it to my R results. 
+\item[] July 7:  submit updated R package to CRAN.
 \item[] July 7 to July 14: write C function that selects the importance sampling distribution.
 \item[]July 14 to July 21: test the newly-written C function and compare it to my R results.
-\item[] July 21 to July 28: write function for distance-based variance structure and incorporate into existing R code.
+\item[] July 21:  submit updated R package to CRAN.
+\item[] July 21 to July 28: write function for new variance structure and incorporate into package.
 \item[] July 28 to August 2: test the new variance structure and compare results to those reported by \citet{caffojj:2005}.
 \item[] August 8 to August 11: document the new variance structure.
-\item[] August 11: Submit R package to CRAN.
+\item[] August 11: submit final version of fully-updated R package to CRAN.
 \end{itemize}
 I expect to complete all work by August 11.
+If something starts to take longer than predicted, then I may need to postpone the new variance structure to the fall, since the first two goals are more important.
 

### 2014-05-20 glmm/R/getFamily.R
Updated and implemented checkData function
Checks the response to make sure it agrees with the glmm family. (Bernoulli should be 0/1. Poisson should be nonnegative integers)

@@ -21,7 +21,7 @@ function()
 	cpp<-function(eta) exp(eta)
 	checkData<-function(x) {
 		bads<-sum(x<0)
-		if(bads>0) stop("data must be nonnegative.")
+		if(bads>0) stop("data must be nonnegative integers.")
 		return(NULL)	
 		}
 	out<-list(family.glmm=family.glmm, cum=cum,cp=cp,cpp=cpp, checkData=checkData)
@@ -38,8 +38,8 @@ function()
 	cpp<-function(eta) exp(eta)/(1+exp(eta))-exp(2*eta)*(1+exp(eta))^(-2)
 	
 	checkData<-function(x) {
-		bads<-sum(x>1) +sum(x<0)
-		if(bads>0) stop("data must be between 0 and 1.")
+		bads<-sum(x!=1&x!=0)
+		if(bads>0) stop("response must be 0 and 1.")
 		return(NULL)	
 		}
 	out<-list(family.glmm=family.glmm, cum=cum, cp=cp, cpp=cpp, checkData=checkData)

### 2014-05-20 glmm/R/glmm.R

@@ -51,6 +51,8 @@ function(fixed,random,varcomps.names,data,family.glmm,m,varcomps.equal,doPQL=TRU
 		stop("You must name each unique variance component. Check varcomps.names and varcomps.equal.")} 
 	if(min(varcomps.equal)!=1)stop("The vector varcomps.equal must contain numbers starting at 1 to denote which variance components are equal.")	
 	levs<-ordered(unique(varcomps.equal))
+	family.glmm<-getFamily(family.glmm)
+	family.glmm$checkData(y)
 	
 	#this loop is a 2-4-1. We want to check that they're filling in varcomps.equal correctly. 
 	#We also want to group all the design matrices that share a variance components.
@@ -78,6 +80,7 @@ function(fixed,random,varcomps.names,data,family.glmm,m,varcomps.equal,doPQL=TRU
 	      sigma.pql<-pql.out$sigma
 	      nu.pql<-sigma.pql^2
 	      beta.pql<-cache$beta.twid #works through here, cache is ok
+		print("PQL estimates have been identified.")
 	}
 	
 	if(doPQL==FALSE){

### 2014-05-20 CompareExamples/glmmexamples.tex
Added additional examples to compare to other software
Comparing the bacteria dataset fit with glmmPQL and the cbpp (herd) dataset fit with glmer. The glmm package gets similar answers.

\documentclass{article}

\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}

\usepackage{geometry}
%\geometry{hmargin=1.025in,vmargin={1.25in,2.5in},nohead,footskip=0.5in} 
%\geometry{hmargin=1.025in,vmargin={1.25in,0.75in},nohead,footskip=0.5in} 
\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}

\renewcommand{\baselinestretch}{1.25}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\gbar}{\bar g}
\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} \DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\newcommand{\eps}{\epsilon}
\newtheorem{claim}{Claim}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\def\baro{\vskip  .2truecm\hfill \hrule height.5pt \vskip  .2truecm}
\def\barba{\vskip -.1truecm\hfill \hrule height.5pt \vskip .4truecm}

\title{R package glmm}

\author{Christina Knudson}

\begin{document}
\maketitle{}

\section{Bacteria example}
The MASS package contains the command glmmPQL and the bacteria data-set. The manual describes the data set as follows: ``Tests of the presence of the bacteria H. influenzae in children with otitis media in the Northern Territory of Australia.''

The data were fit using glmmPQL, glmm with $m=10^3$, glmm with $m=10^4$, glmm with $m=10^5$. (The data did need to be reformatted, which took only a couple minutes). The parameter estimates are summarized in the following table. More model details can be seen in the output that follows the table.

\begin{tabular}{lccccc}
& Intercept & trtdrug & trtdrug+ & I(week$>2$) TRUE & $\nu$ \\ \hline
glmmPQL & 3.41 & -1.25 & -.75 & -1.61 & 1.99\\
glmm $m=10^3$ & 3.02 & -1.20 & -.89 & -1.44 & .81 \\
glmm $m=10^4$ & 3.65 & -1.36 & -.92 & -1.66 & 2.02 \\
glmm $m=10^5$ & 3.49& -1.65 & -.90 & -1.46 & .90 \\
\end{tabular}

It's safe to say that the bacteria glmm results with a paltry Monte Carlo sample size of $m=10^3$ are not reliable. We conclude this because the estimates change quite a bit when $m=10^4$. The results using $m=10^4$ are very similar to the glmmPQL results.

The glmmPQL results:
\begin{verbatim}
> bac.pql<-glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
+                 family = binomial, data = bacteria)
> summary(bac.pql)
Linear mixed-effects model fit by maximum likelihood
 Data: bacteria 
  AIC BIC logLik
   NA  NA     NA

Random effects:
 Formula: ~1 | ID
        (Intercept)  Residual
StdDev:    1.410637 0.7800511

Variance function:
 Structure: fixed weights
 Formula: ~invwt 
Fixed effects: y ~ trt + I(week > 2) 
                    Value Std.Error  DF   t-value p-value
(Intercept)      3.412014 0.5185033 169  6.580506  0.0000
trtdrug         -1.247355 0.6440635  47 -1.936696  0.0588
trtdrug+        -0.754327 0.6453978  47 -1.168779  0.2484
I(week > 2)TRUE -1.607257 0.3583379 169 -4.485311  0.0000
 Correlation: 
                (Intr) trtdrg trtdr+
trtdrug         -0.598              
trtdrug+        -0.571  0.460       
I(week > 2)TRUE -0.537  0.047 -0.001

Standardized Within-Group Residuals:
       Min         Q1        Med         Q3        Max 
-5.1985361  0.1572336  0.3513075  0.4949482  1.7448845

Number of Observations: 220
Number of Groups: 50 
\end{verbatim}

The bacteria glmm results with a  Monte Carlo sample size of $m=10^3$:
\begin{verbatim}
> set.seed(1234)
> bac.glmm1<-glmm(y2~trt+I(week > 2),list(~0+ID), 
family=bernoulli.glmm, data=bacteria, m=10^3, varcomps.names=c("ID"))

> summary(bac.glmm1)

Call:
glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID),  varcomps.names = c("ID"), 
data = bacteria, family.glmm = bernoulli.glmm,     m = 10^3)

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.0199     0.4667   6.471 9.75e-11 ***
trtdrug          -1.1989     0.4473  -2.680 0.007357 ** 
trtdrug+         -0.8883     0.4742  -1.873 0.061021 .  
I(week > 2)TRUE  -1.4407     0.4258  -3.384 0.000716 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2    
ID   0.8175     0.1635   4.998   2.89e-07 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

\end{verbatim}
The results from fitting the bacteria dataset with glmm and $m=10^4$:

\begin{verbatim}
> set.seed(1234)
> bac.glmm2<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^4, varcomps.names=c("ID"))
> summary(bac.glmm2)

Call:
glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID), 
    varcomps.names = c("ID"), data = bacteria, family.glmm = bernoulli.glmm, 
    m = 10^4)

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.6514     0.6327   5.771  7.9e-09 ***
trtdrug          -1.3645     0.7461  -1.829 0.067427 .  
trtdrug+         -0.9186     0.7619  -1.206 0.227915    
I(week > 2)TRUE  -1.6660     0.4645  -3.587 0.000334 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2   
ID   2.0244     0.7515   2.694    0.00353 **
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

\end{verbatim}

Results  of glmm with $m=10^5$:
\begin{verbatim}
> set.seed(1234)
> bac.glmm3<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^5, varcomps.names=c("ID"))
> summary(bac.glmm3)

Call:
glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID), 
    varcomps.names = c("ID"), data = bacteria, family.glmm = bernoulli.glmm, 
    m = 10^5)

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.4888     0.4861   7.178 7.08e-13 ***
trtdrug          -1.6526     0.5232  -3.159 0.001586 ** 
trtdrug+         -0.8968     0.5160  -1.738 0.082196 .  
I(week > 2)TRUE  -1.4614     0.4334  -3.372 0.000746 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2    
ID   0.8999     0.2471   3.641   0.000136 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

\end{verbatim}

\section{Herd CBPP Example}
The cbpp dataset is located in the lme4 package. The lme4 package describes it thusly: ``Contagious bovine pleuropneumonia (CBPP) is a major disease of cattle in Africa, caused by a
mycoplasma. This dataset describes the serological incidence of CBPP in zebu cattle during a
follow-up survey implemented in 15 commercial herds located in the Boji district of Ethiopia. The
goal of the survey was to study the within-herd spread of CBPP in newly infected herds. Blood
samples were quarterly collected from all animals of these herds to determine their CBPP status.
These data were used to compute the serological incidence of CBPP (new cases occurring during a given time period). Some data are missing (lost to follow-up).''

First, I fit the data using glmer in lme4. Then I fit the data using glmm with $m=10^4$. This took 16.75 minutes on my netbook. The point estimates and the standard errors were very similar between the two  methods of model-fitting. (The data did need to be reformatted, which took only a couple minutes)

First, the results from glmer:
\begin{verbatim}
> summary(gm1)
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial ( logit )
Formula: cbind(incidence, size - incidence) ~ period + (1 | herd)
   Data: cbpp

     AIC      BIC   logLik deviance df.resid 
   194.1    204.2    -92.0    184.1       51

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-2.3816 -0.7889 -0.2026  0.5142  2.8791

Random effects:
 Groups Name        Variance Std.Dev.
 herd   (Intercept) 0.4123   0.6421  
Number of obs: 56, groups: herd, 15

Fixed effects:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  -1.3983     0.2312  -6.048 1.47e-09 ***
period2      -0.9919     0.3032  -3.272 0.001068 ** 
period3      -1.1282     0.3228  -3.495 0.000474 ***
period4      -1.5797     0.4220  -3.743 0.000182 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Correlation of Fixed Effects:
        (Intr) perid2 perid3
period2 -0.363              
period3 -0.340  0.280       
period4 -0.260  0.213  0.198

\end{verbatim}

Next, the results using glmm with $m=10^4$:
\begin{verbatim}
> summary(herd.glmm1)

Call:
glmm(fixed = Y ~ period, random = list(~0 + herd), varcomps.names = c("herd"), 
    data = herddat, family.glmm = bernoulli.glmm, m = 10^4)

Fixed Effects:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  -1.4166     0.2410  -5.879 4.14e-09 ***
period2      -0.9921     0.3078  -3.223 0.001271 ** 
period3      -1.1289     0.3277  -3.445 0.000571 ***
period4      -1.5789     0.4293  -3.678 0.000235 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
     Estimate Std. Error z value Pr(>|z|)/2  
herd   0.4354     0.2488    1.75     0.0401 *
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

\end{verbatim}

\section{Reformatting the Datsets}
\subsection{Bacteria Reformatting}
The issue with the bacteria dataset is the response was y/n rather than 1/0. I created a new response that changed the y to 1 and the n to 0.

\begin{verbatim}
bacteria$y2<-as.integer(bacteria$y)-1
\end{verbatim}
\subsection{CBPP Reformatting}
The cbpp dataset was created for binomial but my package was written for Bernoulli responses. In other words, my package needs a row for each success or failure. I did this in the following way:
\begin{verbatim}
cbpp$nonincidence<-cbpp$size-cbpp$incidence #number of "failures"
herddat<-matrix(data=NA,nrow=842,ncol=3)
colnames(herddat)<-c("Y","period","herd")
rowid<-1
for(i in 1:nrow(cbpp)){
	#make a row for each one of the incidences 
	ntimes<-cbpp[i,2]
	if(ntimes>0){	
		for(j in 1:ntimes){
			herddat[rowid,]<-c(1,cbpp[i,4],cbpp[i,1])
			rowid<-rowid+1
		}
	}

	#make a row for each of the nonincidences
	ntimes<-cbpp[i,5]
	if(ntimes>0){	
		for(j in 1:ntimes){
			herddat[rowid,]<-c(0,cbpp[i,4],cbpp[i,1])
			rowid<-rowid+1
		}
	}
}
herddat<-as.data.frame(herddat)
herddat$herd<-as.factor(herddat$herd)
herddat$period<-as.factor(herddat$period)
\end{verbatim}
\end{document}

### 2014-05-22 CompareExamples/glmmexamples.tex
Added salamander example's results

@@
 
+\section{Salamander}
+
+\begin{verbatim}
+> set.seed(1234)
+> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"),data=salamander,family.glmm=bernoulli.glmm,m=10^4)
+> summary(sal.glmm3)
+
+Call:
+glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
+    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glmm, 
+    m = 10^4)
+
+Fixed Effects:
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)   1.1123     0.2593   4.290 1.79e-05 ***
+CrossRW      -0.5545     0.3558  -1.558    0.119    
+CrossWR      -3.1701     0.4029  -7.867 3.62e-15 ***
+CrossWW      -0.2122     0.3681  -0.577    0.564    
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+  Estimate Std. Error z value Pr(>|z|)/2    
+F   1.2087     0.2534   4.769   9.24e-07 ***
+M   0.9485     0.1809   5.244   7.86e-08 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+\end{verbatim}
+
+Here are the results with $m=10^4$ but I forgot to set the seed.
+\begin{verbatim}
+> summary(sal.glmm2)
+
+Call:
+glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
+    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glmm, 
+    m = 10^4)
+
+Fixed Effects:
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)  0.93879    0.26963   3.482 0.000498 ***
+CrossRW     -0.76157    0.37376  -2.038 0.041593 *  
+CrossWR     -2.72491    0.40736  -6.689 2.24e-11 ***
+CrossWW      0.02256    0.37902   0.060 0.952541    
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+  Estimate Std. Error z value Pr(>|z|)/2    
+F   1.5991     0.2965   5.394   3.45e-08 ***
+M   0.9908     0.1885   5.255   7.41e-08 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+\end{verbatim}
+
 \section{Reformatting the Datsets}

### 2014-05-28 CompareExamples/glmmexamples.tex
Added documentation and updated murder data
Added to glmm examples document

@@
 The point estimates and the standard errors were very similar between the two  methods of model-fitting.
-(The data did need to be reformatted, which took only a couple minutes)
+The documentation says that the default method of model-fitting uses the Laplace approximation, but Charlie wonders if they actually use numerical integration for simple problems like this.
 
@@
 \section{Salamander}
+I fit the salamander data set using  my glmm package and compared those results to the results of Yun Ju Sung and Charlie Geyer.
 
@@
 
-Here are the results with $m=10^4$ but I forgot to set the seed.
+Here are the results with $m=10^4$ but I forgot to set the seed. At least it helps us see a bit of the variability.
 \begin{verbatim}
@@
 \end{verbatim}
+
+\section{Murder}
+Subjects were asked how many victims of homicide they personally knew.
+The data set is from Agresti's \textit{Categorical Data Analysis} book.
+I found similar answers from my glmm package as from glmmPQL.
+
+The results from my package:
+
+\begin{verbatim}
+set.seed(1234)
+murder.glmm<- glmm(y~race ,random=list(~0+black,~0+white), varcomps.names=c("black","white"), data=murder ,family.glmm=poisson.glmm,m=10^4)
+
+> summary(murder.glmm)
+
+Call:
+glmm(fixed = y ~ race, random = list(~0 + black, ~0 + white), 
+    varcomps.names = c("black", "white"), data = murder, family.glmm = poisson.glmm, 
+    m = 10^4)
+
+Fixed Effects:
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)  0.42003    0.06428   6.534 6.40e-11 ***
+raceWhite   -0.33179    0.07021  -4.726 2.29e-06 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+       Estimate Std. Error z value Pr(>|z|)/2
+black 6.468e-10  3.452e-09   0.187      0.426
+white 2.667e-10  1.596e-09   0.167      0.434
+\end{verbatim}
+
+When I make the sample size drastically smaller so that the analysis takes only about a minute, the results are very similar still.
+\begin{verbatim}
+> summary(murder.glmm)
+
+Call:
+glmm(fixed = y ~ race, random = list(~0 + black, ~0 + white), 
+    varcomps.names = c("black", "white"), data = murder, family.glmm = poisson.glmm, 
+    m = 10^2)
+
+Fixed Effects:
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)  0.42003    0.06428   6.534 6.39e-11 ***
+raceWhite   -0.33179    0.07021  -4.726 2.29e-06 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+       Estimate Std. Error z value Pr(>|z|)/2
+black 1.190e-09  1.718e-09   0.693      0.244
+white 1.070e-08  1.440e-08   0.743      0.229
+
+\end{verbatim}
+
+The murder model resulting from glmmPQL:
+\begin{verbatim}
+> murder.pql<-glmmPQL(y~race,random=~1|race,data=murder,family=poisson)iteration 1
+> summary(murder.pql)
+Linear mixed-effects model fit by maximum likelihood
+ Data: murder 
+  AIC BIC logLik
+   NA  NA     NA
+
+Random effects:
+ Formula: ~1 | race
+         (Intercept)  Residual
+StdDev: 9.157794e-06 0.4647694
+
+Variance function:
+ Structure: fixed weights
+ Formula: ~invwt 
+Fixed effects: y ~ race 
+                 Value  Std.Error   DF   t-value p-value
+(Intercept)  0.4200335 0.02989812 1306  14.04883       0
+raceWhite   -0.3317900 0.03265396    0 -10.16079     NaN
+ Correlation: 
+          (Intr)
+raceWhite -0.916
+
+Standardized Within-Group Residuals:
+       Min         Q1        Med         Q3        Max 
+-0.9104427 -0.1899269 -0.1899269 -0.1899269 12.1624898
+
+Number of Observations: 1308
+Number of Groups: 2 
+Warning message:
+In pt(-abs(tTable[, "t-value"]), tTable[, "DF"]) : NaNs produced
+
+\end{verbatim}
+
+%\section{Booth and Hobert}
+%Here are the results from my glmm package fitting a model to the Booth and Hobert data (with $m=10^4$).
+%\begin{verbatim}
+%> set.seed(1234)
+%> mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,family.glmm=bernoulli.glmm,m=10^4,doPQL=TRUE)
+%[1] "PQL estimates have been identified."
+%> summary(mod.mcml1)
+%
+%Call:
+%glmm(fixed = y ~ 0 + x1, random = list(y ~ 0 + z1), varcomps.names = c("z1"), 
+%    data = BoothHobert, family.glmm = bernoulli.glmm, m = 10^4, 
+%    doPQL = TRUE)
+%
+%Fixed Effects:
+%   Estimate Std. Error z value Pr(>|z|)    
+%x1    6.225      1.422   4.377  1.2e-05 ***
+%---
+%Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+%
+%
+%Variance Components for Random Effects (P-values are one-tailed):
+%   Estimate Std. Error z value Pr(>|z|)/2
+%z1    1.907      1.814   1.051      0.147
+%
+%
+%\end{verbatim}
+%
+%Here are the results from glmmPQL:
+%\begin{verbatim}
+%> summary(bh)
+%Linear mixed-effects model fit by maximum likelihood
+% Data: BoothHobert 
+%  AIC BIC logLik
+%   NA  NA     NA
+%
+%Random effects:
+% Formula: ~1 | z1
+%        (Intercept)  Residual
+%StdDev:    1.284666 0.8712694
+%
+%Variance function:
+% Structure: fixed weights
+% Formula: ~invwt 
+%Fixed effects: y ~ 0 + x1 
+%      Value Std.Error  DF  t-value p-value
+%x1 5.857223  1.008231 140 5.809405       0
+%
+%Standardized Within-Group Residuals:
+%        Min          Q1         Med          Q3         Max 
+%-8.55346135  0.07317618  0.17063844  0.35776723  2.15976654 
+%
+%Number of Observations: 150
+%Number of Groups: 10 
+%\end{verbatim}
 

### 2014-06-02 CompareExamples/glmmexamples.tex
Added salamander results, but now thinking I need to fix something.

@@
 First, I fit the data using glmer in lme4.
+The model summary says it was fit with Adaptive Hermite Quadrature with 9 quadrature nodes.
 Then I fit the data using glmm with $m=10^4$.
 This took 16.75 minutes on my netbook.
-The point estimates and the standard errors were very similar between the two  methods of model-fitting.
-The documentation says that the default method of model-fitting uses the Laplace approximation, but Charlie wonders if they actually use numerical integration for simple problems like this.
+The point estimates and some of the standard errors are very similar between the two  methods of model-fitting.
 
-First, the results from glmer:
+\begin{tabular}{lccccc}
+& Intercept & period2 & period3 & period4 & $\nu$ \\ \hline
+glmer (numerical integration) & -1.40 & -.99 & -1.13 & -1.58 & .41\\
+glmm $m=10^4$ & -1.42 & -.99 & -1.13 & -1.58 & .44 \\
+\end{tabular}
+
+Here are the full results from glmer:
 \begin{verbatim}
@@
 \section{Salamander}
-I fit the salamander data set using  my glmm package and compared those results to the results of Yun Ju Sung and Charlie Geyer.
+The salamander dataset represents success or failure in matings between two types of salamanders denoted by R and W.
+Four types of crosses are possible: RR, RW, WR, WW.
+These are the fixed effects.
+There is a random effect for each female salamander and for each male salamander.
+Thus, there are two variance components: one for the females and one for the males.
+The response is binary: whether or not the mating was successful.
+
+I fit the salamander data set using  my glmm package (with $m=10^4$ twice and with $m=10^6$ once)  and compared those results to the MCEM results of Booth and Hobert.
+They do not look so great, and the $m=10^6$ results took 41 hours to get.
+
+\begin{tabular}{lcccccc}
+& Intercept & RW & WR & WW & $\nu_F$ & $\nu_M$ \\ \hline
+MCEM (B+H) & 1.03 & .32 & -1.95 & .99 & 1.4 & 1.25 \\
+glmm $m=10^4$ & 1.11 & -.55 & -3.17& -.21 & 1.21 & .95 \\
+glmm $m=10^4$ & .94 & -.76 & -2.72 & .0023 & 1.60 & .99 \\
+glmm $m=10^6$ &  0.84&  -0.81&  -2.81&  -0.21&1.21& 1.02 \\
+\end{tabular}
 
@@
 > set.seed(1234)
-> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"),data=salamander,family.glmm=bernoulli.glmm,m=10^4)
+> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"),
+data=salamander,family.glmm=bernoulli.glmm,m=10^4)
 > summary(sal.glmm3)
@@
 Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+\end{verbatim}
+
+Here are the results with $m=10^6$. 
+\begin{verbatim}
+> ptm <- proc.time()
+> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F"
+,"M"),data=salamander,family.glmm=bernoulli.glmm,m=10^6)
+>  proc.time()-ptm
+      user     system    elapsed 
+148604.323    595.454 174153.241 
+> summary(sal.glmm3)
+
+Call:
+glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
+    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glm
+m, 
+    m = 10^6)
+
+Fixed Effects:
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)   0.8413     0.2756   3.052  0.00227 ** 
+CrossRW      -0.8074     0.3737  -2.161  0.03072 *  
+CrossWR      -2.8062     0.4097  -6.849 7.43e-12 ***
+CrossWW      -0.2135     0.3887  -0.549  0.58289    
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+  Estimate Std. Error z value Pr(>|z|)/2    
+F   1.2090     0.2443   4.948   3.74e-07 ***
+M   1.0196     0.2013   5.065   2.04e-07 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+> sal.glmm3$beta.pql
+[1]  0.83806013 -0.57917030 -2.44205490 -0.01778234
+> sal.glmm3$nu.pql
+[1] 1.122464 1.036371
+
 \end{verbatim}

### 2014-07-23 CompareExamples/glmmexamples.tex
Added design doc

@@
 MCEM (B+H) & 1.03 & .32 & -1.95 & .99 & 1.4 & 1.25 \\
-glmm $m=10^4$ & 1.11 & -.55 & -3.17& -.21 & 1.21 & .95 \\
-glmm $m=10^4$ & .94 & -.76 & -2.72 & .0023 & 1.60 & .99 \\
-glmm $m=10^6$ &  0.84&  -0.81&  -2.81&  -0.21&1.21& 1.02 \\
+glmm $m=10^5$ &   0.964      -0.645      -2.767      -0.008
 \end{tabular}
-
-\begin{verbatim}
-> set.seed(1234)
-> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"),
-data=salamander,family.glmm=bernoulli.glmm,m=10^4)
-> summary(sal.glmm3)
-
-Call:
-glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
-    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glmm, 
-    m = 10^4)
-
-Fixed Effects:
-            Estimate Std. Error z value Pr(>|z|)    
-(Intercept)   1.1123     0.2593   4.290 1.79e-05 ***
-CrossRW      -0.5545     0.3558  -1.558    0.119    
-CrossWR      -3.1701     0.4029  -7.867 3.62e-15 ***
-CrossWW      -0.2122     0.3681  -0.577    0.564    
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-Variance Components for Random Effects (P-values are one-tailed):
-  Estimate Std. Error z value Pr(>|z|)/2    
-F   1.2087     0.2534   4.769   9.24e-07 ***
-M   0.9485     0.1809   5.244   7.86e-08 ***
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-\end{verbatim}
-
-Here are the results with $m=10^4$ but I forgot to set the seed. At least it helps us see a bit of the variability.
-\begin{verbatim}
-> summary(sal.glmm2)
-
-Call:
-glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
-    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glmm, 
-    m = 10^4)
-
-Fixed Effects:
-            Estimate Std. Error z value Pr(>|z|)    
-(Intercept)  0.93879    0.26963   3.482 0.000498 ***
-CrossRW     -0.76157    0.37376  -2.038 0.041593 *  
-CrossWR     -2.72491    0.40736  -6.689 2.24e-11 ***
-CrossWW      0.02256    0.37902   0.060 0.952541    
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-Variance Components for Random Effects (P-values are one-tailed):
-  Estimate Std. Error z value Pr(>|z|)/2    
-F   1.5991     0.2965   5.394   3.45e-08 ***
-M   0.9908     0.1885   5.255   7.41e-08 ***
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-\end{verbatim}
-
-Here are the results with $m=10^6$. 
-\begin{verbatim}
-> ptm <- proc.time()
-> sal.glmm3<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F"
-,"M"),data=salamander,family.glmm=bernoulli.glmm,m=10^6)
->  proc.time()-ptm
-      user     system    elapsed 
-148604.323    595.454 174153.241 
-> summary(sal.glmm3)
-
-Call:
-glmm(fixed = Mate ~ Cross, random = list(~0 + Female, ~0 + Male), 
-    varcomps.names = c("F", "M"), data = salamander, family.glmm = bernoulli.glm
-m, 
-    m = 10^6)
-
-Fixed Effects:
-            Estimate Std. Error z value Pr(>|z|)    
-(Intercept)   0.8413     0.2756   3.052  0.00227 ** 
-CrossRW      -0.8074     0.3737  -2.161  0.03072 *  
-CrossWR      -2.8062     0.4097  -6.849 7.43e-12 ***
-CrossWW      -0.2135     0.3887  -0.549  0.58289    
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-Variance Components for Random Effects (P-values are one-tailed):
-  Estimate Std. Error z value Pr(>|z|)/2    
-F   1.2090     0.2443   4.948   3.74e-07 ***
-M   1.0196     0.2013   5.065   2.04e-07 ***
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-> sal.glmm3$beta.pql
-[1]  0.83806013 -0.57917030 -2.44205490 -0.01778234
-> sal.glmm3$nu.pql
-[1] 1.122464 1.036371
-
-\end{verbatim}
 

### 2014-07-23 glmm/R/getFamily.R

@@ -6,7 +6,7 @@ getFamily <-function(family.glmm)
 		family.glmm<-family.glmm()
 	#right way to check the class if there might be more than one class (maybe bc of hierarchies)
 	if(!inherits(family.glmm,"glmm.family")) 
-		stop(" 'family' not recognized") 
+		stop(" 'family.glmm' not recognized") 
 	
 	family.glmm	
 }
@@ -14,15 +14,17 @@ getFamily <-function(family.glmm)
 poisson.glmm <-function()
 {
 	family.glmm<- "poisson.glmm"
+	link<-"log"
 	cum <- function(eta) exp(eta)
 	cp <- function(eta) exp(eta)
 	cpp<-function(eta) exp(eta)
 	checkData<-function(x) {
 		bads<-sum(x<0)
 		if(bads>0) stop("data must be nonnegative integers.")
+		if(is.integer(x)==FALSE) stop("data must be nonnegative integers.")
 		return(NULL)	
 		}
-	out<-list(family.glmm=family.glmm, cum=cum,cp=cp,cpp=cpp, checkData=checkData)
+	out<-list(family.glmm=family.glmm, link=link, cum=cum, cp=cp, cpp=cpp, checkData=checkData)
 	class(out)<-"glmm.family"
 	return(out)
 }
@@ -31,6 +33,7 @@ poisson.glmm <-function()
 bernoulli.glmm <-function()
 {
 	family.glmm<- "bernoulli.glmm"
+	link<-"logit (log odds)"
 	cum <- function(eta){ 
 		if(eta>0) eta+log1p(exp(-eta))
 		else log1p(exp(eta))}
@@ -42,7 +45,7 @@ bernoulli.glmm <-function()
 		if(bads>0) stop("response must be 0 and 1.")
 		return(NULL)	
 		}
-	out<-list(family.glmm=family.glmm, cum=cum, cp=cp, cpp=cpp, checkData=checkData)
+	out<-list(family.glmm=family.glmm, link=link, cum=cum, cp=cp, cpp=cpp, checkData=checkData)
 	class(out)<-"glmm.family"
 	return(out)
 }

### 2014-07-23 glmm/R/glmm.R

@@ -132,7 +132,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	genData<-genRand(zeros,D.star,m1)
 	
 	#generate m2 from N(u*,D*)
-	genData2<-genRand(u.star,D.star,m1)
+	genData2<-genRand(u.star,D.star,m2)
 
 	#generate m3 from N(u*,(Z'c''(Xbeta*+zu*)Z+D*^{-1})^-1)
 	Z=do.call(cbind,mod.mcml$z)
@@ -141,7 +141,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	cdouble<-diag(cdouble)
 	Sigmuh.inv<- t(Z)%*%cdouble%*%Z+D.star.inv
 	Sigmuh<-solve(Sigmuh.inv)
-	genData3<-genRand(u.star,Sigmuh,m1)
+	genData3<-genRand(u.star,Sigmuh,m3)
 
 #	#these are from distribution based on data
 #	if(distrib=="tee")genData<-genRand(sigma.gen,s.pql,mod.mcml$z,m1,distrib="tee",gamm)
@@ -167,7 +167,7 @@ umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star, blather=T,
 	names(nu.trust)<-varcomps.names
 
 	if(debug==TRUE){
-	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql, nu.gen=nu.gen, trust.argpath=trust.argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom)
+	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql, nu.gen=nu.gen, trust.argpath=trust.argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom,m1=m1,m2=m2,m3=m3)
 	}
 	
 	return(structure(list(beta=beta.trust,nu=nu.trust, likelihood.value=trust.out$value, likelihood.gradient=trust.out$gradient, likelihood.hessian=trust.out$hessian,

### 2014-10-20 CompareExamples/glmmexamples.tex
working on CLT proof for gradient

@@
 
-\title{R package glmm}
+\title{R package \texttt{glmm}}
 
@@
 \section{Bacteria example}
-The MASS package contains the command glmmPQL and the bacteria data-set.
+The MASS package contains the command \texttt{glmmPQL} and the bacteria data-set.
 The manual describes the data set as follows: ``Tests of the presence of the bacteria H.
@@
 
-The data were fit using glmmPQL, glmm with $m=10^3$, glmm with $m=10^4$, glmm with $m=10^5$.
-(The data did need to be reformatted, which took only a couple minutes).
+The data were fit using \texttt{glmmPQL},  \texttt{glmer}  from the \texttt{lme4} package (method implemented was Laplace approximation),  \texttt{glmm} with $m=10^4$,  \texttt{glmm} with $m=10^5$, and \texttt{glmm} with $m=10^6$.
+(The data did need to be reformatted in a process described at the end of the document).
 The parameter estimates are summarized in the following table.
@@
 & Intercept & trtdrug & trtdrug+ & I(week$>2$) TRUE & $\nu$ \\ \hline
-glmmPQL & 3.41 & -1.25 & -.75 & -1.61 & 1.99\\
-glmm $m=10^3$ & 3.02 & -1.20 & -.89 & -1.44 & .81 \\
-glmm $m=10^4$ & 3.65 & -1.36 & -.92 & -1.66 & 2.02 \\
-glmm $m=10^5$ & 3.49& -1.65 & -.90 & -1.46 & .90 \\
+\texttt{glmmPQL} & 3.41 & -1.25 & -.75 & -1.61 & 1.99\\
+\texttt{glmer} & 3.55 & -1.37 & -.78 & -1.60 & 1.54\\
+\texttt{glmm} $m=10^4$ & 3.57 & -1.38 & -.80 & -1.62 & 1.66 \\
+\texttt{glmm} $m=10^5$ & 3.60& -1.39 & -.79 & -1.63 & 1.73 \\
+\texttt{glmm} $m=10^6$ & 3.58& -1.37 & -.79 & -1.63 & 1.71 \\
 \end{tabular}
 
-It's safe to say that the bacteria glmm results with a paltry Monte Carlo sample size of $m=10^3$ are not reliable.
-We conclude this because the estimates change quite a bit when $m=10^4$.
-The results using $m=10^4$ are very similar to the glmmPQL results.
-
-The glmmPQL results:
+The \texttt{glmmPQL} results:
 \begin{verbatim}
@@
 
-The bacteria glmm results with a  Monte Carlo sample size of $m=10^3$:
 \begin{verbatim}
-> set.seed(1234)
-> bac.glmm1<-glmm(y2~trt+I(week > 2),list(~0+ID), 
-family=bernoulli.glmm, data=bacteria, m=10^3, varcomps.names=c("ID"))
-
-> summary(bac.glmm1)
-
-Call:
-glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID),  varcomps.names = c("ID"), 
-data = bacteria, family.glmm = bernoulli.glmm,     m = 10^3)
-
-Fixed Effects:
-                Estimate Std. Error z value Pr(>|z|)    
-(Intercept)       3.0199     0.4667   6.471 9.75e-11 ***
-trtdrug          -1.1989     0.4473  -2.680 0.007357 ** 
-trtdrug+         -0.8883     0.4742  -1.873 0.061021 .  
-I(week > 2)TRUE  -1.4407     0.4258  -3.384 0.000716 ***
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-Variance Components for Random Effects (P-values are one-tailed):
-   Estimate Std. Error z value Pr(>|z|)/2    
-ID   0.8175     0.1635   4.998   2.89e-07 ***
----
-Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-\end{verbatim}
-The results from fitting the bacteria dataset with glmm and $m=10^4$:
-
-\begin{verbatim}
-> set.seed(1234)
-> bac.glmm2<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^4, varcomps.names=c("ID"))
+proc.time()-ptm
+user  system elapsed 
+ 16.226   0.044  16.281 
 > summary(bac.glmm2)
@@
                 Estimate Std. Error z value Pr(>|z|)    
-(Intercept)       3.6514     0.6327   5.771  7.9e-09 ***
-trtdrug          -1.3645     0.7461  -1.829 0.067427 .  
-trtdrug+         -0.9186     0.7619  -1.206 0.227915    
-I(week > 2)TRUE  -1.6660     0.4645  -3.587 0.000334 ***
+(Intercept)       3.5745     0.6371   5.611 2.01e-08 ***
+trtdrug          -1.3779     0.6701  -2.056 0.039768 *  
+trtdrug+         -0.8014     0.6878  -1.165 0.243950    
+I(week > 2)TRUE  -1.6224     0.4674  -3.471 0.000518 ***
 ---
@@
    Estimate Std. Error z value Pr(>|z|)/2  
-ID   2.0244     0.7515   2.694    0.00353 **
+ID   1.6635     0.7807   2.131     0.0165 *
 ---
@@
 
-Results  of glmm with $m=10^5$:
+Results from  \texttt{glmm}  with $m=10^5$.
 \begin{verbatim}
-> set.seed(1234)
-> bac.glmm3<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^5, varcomps.names=c("ID"))
-> summary(bac.glmm3)
+> bac.glmm2<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^5, varcomps.names=c("ID"))
+> proc.time()-ptm
+   user  system elapsed 
+175.729   0.150 176.211 
+> summary(bac.glmm2)
 
@@
                 Estimate Std. Error z value Pr(>|z|)    
-(Intercept)       3.4888     0.4861   7.178 7.08e-13 ***
-trtdrug          -1.6526     0.5232  -3.159 0.001586 ** 
-trtdrug+         -0.8968     0.5160  -1.738 0.082196 .  
-I(week > 2)TRUE  -1.4614     0.4334  -3.372 0.000746 ***
+(Intercept)       3.6008     0.7666   4.697 2.64e-06 ***
+trtdrug          -1.3868     0.7126  -1.946 0.051641 .  
+trtdrug+         -0.7940     0.6996  -1.135 0.256348    
+I(week > 2)TRUE  -1.6321     0.4917  -3.319 0.000903 ***
 ---
@@
    Estimate Std. Error z value Pr(>|z|)/2  
-ID   0.8999     0.2471   3.641   0.000136 ***
+ID    1.738      1.317    1.32     0.0935 .
 ---
@@
 
+\end{verbatim}
+
+Results from \texttt{glmm} with $m=10^6$.
+\begin{verbatim}
+ proc.time()-ptm
+    user   system  elapsed 
+1491.511    1.299 1494.133 
+> summary(bac.glmm2)
+
+Call:
+glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID), 
+    varcomps.names = c("ID"), data = bacteria, family.glmm = bernoulli.glmm, 
+    m = 10^6)
+
+Fixed Effects:
+                Estimate Std. Error z value Pr(>|z|)    
+(Intercept)       3.5824     0.6952   5.153 2.57e-07 ***
+trtdrug          -1.3714     0.6935  -1.978 0.047980 *  
+trtdrug+         -0.7929     0.7034  -1.127 0.259644    
+I(week > 2)TRUE  -1.6278     0.4801  -3.391 0.000698 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Variance Components for Random Effects (P-values are one-tailed):
+   Estimate Std. Error z value Pr(>|z|)/2  
+ID    1.707      1.050   1.626      0.052 .
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+\end{verbatim}
+
+Results from \texttt{lme4}.
+\begin{verbatim}
+ summary(bac.glmer)
+Generalized linear mixed model fit by maximum likelihood (Laplace
+  Approximation) [glmerMod]
+ Family: binomial ( logit )
+Formula: y2 ~ trt + I(week > 2) + (1 | ID)
+   Data: bacteria
+
+     AIC      BIC   logLik deviance df.resid 
+   202.3    219.2    -96.1    192.3      215
+
+Scaled residuals: 
+    Min      1Q  Median      3Q     Max 
+-4.5615  0.1359  0.3022  0.4217  1.1276
+
+Random effects:
+ Groups Name        Variance Std.Dev.
+ ID     (Intercept) 1.543    1.242   
+Number of obs: 220, groups: ID, 50
+
+Fixed effects:
+                Estimate Std. Error z value Pr(>|z|)    
+(Intercept)       3.5479     0.6958   5.099 3.41e-07 ***
+trtdrug          -1.3667     0.6770  -2.019 0.043517 *  
+trtdrug+         -0.7826     0.6831  -1.146 0.251926    
+I(week > 2)TRUE  -1.5985     0.4759  -3.359 0.000783 ***
+---
+Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+Correlation of Fixed Effects:
+            (Intr) trtdrg trtdr+
+trtdrug     -0.593              
+trtdrug+    -0.537  0.487       
+I(wk>2)TRUE -0.656  0.126  0.064
+> 
 \end{verbatim}
@@
 The model summary says it was fit with Adaptive Hermite Quadrature with 9 quadrature nodes.
-Then I fit the data using glmm with $m=10^4$.
+Then I fit the data using \texttt{glmm}with $m=10^4$.
 This took 16.75 minutes on my netbook.
@@
 glmer (numerical integration) & -1.40 & -.99 & -1.13 & -1.58 & .41\\
-glmm $m=10^4$ & -1.42 & -.99 & -1.13 & -1.58 & .44 \\
+\texttt{glmm}$m=10^4$ & -1.42 & -.99 & -1.13 & -1.58 & .44 \\
 \end{tabular}
@@
 
-Next, the results using glmm with $m=10^4$:
+Next, the results using \texttt{glmm}with $m=10^4$:
 \begin{verbatim}
@@
 
-I fit the salamander data set using  my glmm package (with $m=10^4$ twice and with $m=10^6$ once)  and compared those results to the MCEM results of Booth and Hobert.
+I fit the salamander data set using  my \texttt{glmm} package (with $m=10^4$ twice and with $m=10^6$ once)  and compared those results to the MCEM results of Booth and Hobert.
 They do not look so great, and the $m=10^6$ results took 41 hours to get.
@@
 MCEM (B+H) & 1.03 & .32 & -1.95 & .99 & 1.4 & 1.25 \\
-glmm $m=10^5$ &   0.964      -0.645      -2.767      -0.008
+\texttt{glmm} $m=10^5$ &   0.964      -0.645      -2.767      -0.008
 \end{tabular}
@@
 The data set is from Agresti's \textit{Categorical Data Analysis} book.
-I found similar answers from my glmm package as from glmmPQL.
+I found similar answers from my glmm package as from \texttt{glmmPQL}.
 
@@
 
-The murder model resulting from glmmPQL:
+The murder model resulting from \texttt{glmmPQL}:
 \begin{verbatim}

### 2014-10-20 glmm/R/glmm.R

@@ -60,6 +60,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	if(!is.numeric(p1))stop("p1 must be a number between 0 and 1")
 	if(p1>1) stop("p1 must be a number between 0 and 1")
 	if(p1<0) stop("p1 must be a number between 0 and 1")
+	if(p1==0) stop("p1 must be nonzero")
 	if(!is.numeric(p2))stop("p2 must be a number between 0 and 1")
 	if(p2>1) stop("p2 must be a number between 0 and 1")
 	if(p2<0) stop("p2 must be a number between 0 and 1")
@@ -177,7 +178,7 @@ umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star, blather=T,
 	names(nu.trust)<-varcomps.names
 
 	if(debug==TRUE){
-	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql, nu.gen=nu.gen, trust.argpath=trust.argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom,m1=m1,m2=m2,m3=m3)
+	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql, nu.gen=nu.gen, trust.argpath=trust.argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom,m1=m1,m2=m2,m3=m3,trust.argtry=trust.out$argtry, trust.steptype=trust.out$steptype, trust.accept=trust.out$accept, trust.r=trust.out$r, trust.rho=trust.out$rho, trust.valpath=trust.out$valpath, trust.valtry=trust.out$valtry, trust.preddif=trust.out$preddif, trust.stepnorm=trust.out$stepnorm)
 	}
 	
 	return(structure(list(beta=beta.trust,nu=nu.trust, likelihood.value=trust.out$value, likelihood.gradient=trust.out$gradient, likelihood.hessian=trust.out$hessian,

### 2014-11-05 NewDesignDoc/DesignDocGLMM.tex
Added design document for glmm package.

\documentclass{article}

 \usepackage{url} 
\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}
\DeclareMathOperator{\logit}{logit}
    \DeclareMathOperator{\var}{Var}
   %  \DeclareMathOperator{\det}{det}
     \DeclareMathOperator{\diag}{diag}

\usepackage{geometry}
%\geometry{hmargin=1.025in,vmargin={1.25in,2.5in},nohead,footskip=0.5in} 
%\geometry{hmargin=1.025in,vmargin={1.25in,0.75in},nohead,footskip=0.5in} 
%\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}

\renewcommand{\baselinestretch}{1.25}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} 
\DeclareMathOperator{\cov}{Cov}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\title{Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models}

\author{Christina Knudson}

\begin{document}
\maketitle{}

\begin{abstract}
This design document describes the process of performing Monte Carlo maximum likelihood (MCML) for generalized linear mixed models.  First, penalized quasi-likelihood (PQL) estimates are calculated, which help generate simulated random effects. Then, the Monte Carlo likelihood approximation (MCLA)  is calculated using the simulated random effects. Next, the MCLA is maximized to find Monte Carlo maximum likelihood estimates, the corresponding Fisher Information, and other statistics. Additional inference is then possible, including confidence intervals for the parameters and likelihood ratio tests for comparing nested models.
\end{abstract}

\section{Theory}

Let $y=(y_1, \ldots, y_n)^T$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects. Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects. Let $\theta$ be a vector of length $p+T$ containing all unknown parameters. Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$. Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write the density like this to keep notation simple in future equations.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}
\end{align}
the gradient vector of the MCLA is
\begin{align}
\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
\end{align}

Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.

%Let $y=(y_1, \ldots, y_n)^T$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)^T$ be a vector of unobserved random effects. Let $\theta$ denote a vector of unknown parameters.  The data are distributed conditionally on the random effects according to $f_\theta(y|u)$.

EDITED TO HERE.

 Let $X$ be an $n \times p$ design matrix for the fixed effects.  Let $U$ be an unobservable normal random vector with length $q$, mean $0$ and variance matrix $D$. Let $Z$ be a $n \times q$ model matrix for the random effects.

There are $T$ distinct variance components  $\nu_t$, $t = 1,$ $\ldots,$ $T$.  For the first version of the package, $D$ is assumed diagonal. Let $E_t$ be a diagonal matrix with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. (Learn more about $E_t$ in section \ref{sec:getEk}.) Then $D= \sum_{t=1}^T \nu_t E_t$.  Later versions of this package will allow  more general covariance structures. With other structures, $D$ will be constructed by $D= \sum_{t=1}^T h_t (\nu_t) \; E_t$, but $E_t$ will no longer be diagonal and $h_t$ will be some function. We will have a D for distance (such as in the car theft data) and for autoregressive 1.

 Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},\ldots,\nu_T)$. Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple. Since $U$ is normally distributed, we can write its log density as
\begin{align}
\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} U' D^{-1} U
\end{align}
where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z U
\end{align}
The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. More on these two families can be found in section \ref{sec:fam}. For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
\end{align}

Since $U$ is unobservable, the likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  Then the MCLA of $l(\theta)$ is:
\begin{align}
l_{m}(\theta) = \sum_{k=1}^m \dfrac{ f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}
\end{align}
Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.

%
%Let $\beta$ be a vector of $p$ fixed effect parameters. Let $X$ be an $n \times p$ design matrix for the fixed effects.  Let $U$ be an unobservable normal random vector with length $q$, mean $0$ and variance matrix $D$. Let $Z$ be a $n \times q$ model matrix for the random effects. 
%
%There are $T$ distinct variance components  $\nu_t$, $t = 1,$ $\ldots,$ $T$.  For the first version of the package, $D$ is assumed diagonal. Let $E_t$ be a diagonal matrix with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. (Learn more about $E_t$ in section \ref{sec:getEk}.) Then $D= \sum_{t=1}^T \nu_t E_t$.  Later versions of this package will allow  more general covariance structures. With other structures, $D$ will be constructed by $D= \sum_{t=1}^T h_t (\nu_t) \; E_t$, but $E_t$ will no longer be diagonal and $h_t$ will be some function. We will have a D for distance (such as in the car theft data) and for autoregressive 1.
%
% Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},\ldots,\nu_T)$. Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple. Since $U$ is normally distributed, we can write its log density as
%\begin{align}
%\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} U' D^{-1} U
%\end{align}
%where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.
%
%Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
%\begin{align}
%g(\mu) = X \beta + Z U
%\end{align}
%The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. More on these two families can be found in section \ref{sec:fam}. For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
%\begin{align}
%Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
%\end{align}
%
%Since $U$ is unobservable, the likelihood must be expressed by integrating out the random effects:
%\begin{align}
%l(\theta)=\int f_\theta(y|u) f_\theta(u) \; du
%\end{align}
%For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  Then the MCLA of $l(\theta)$ is:
%\begin{align}
%l_{m}(\theta) = \sum_{k=1}^m \dfrac{ f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}
%\end{align}
%Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
%

\section{Model fitting function} 
This will be the main function the user will use. The users will need to specify the response and the predictors using the R formula mini-language as interpreted by model.matrix. They'll need to specify the  family (either binomial or Poisson, though  really any exponential family would work and I could add more later). The user will specify the random effects in the same way as for the R function \texttt{reaster} in the R package
 \texttt{aster} \citep{aster-package}. That is, random effects will be expressed using the R formula mini-language. Thus, a sample command with fixed predictors $x_1$ and $x_2$ and with random effects $school$ and $classroom$ (in data set as categorical variables ) would look like
\begin{verbatim}
glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="binomial.glmm", 
data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
debug=FALSE )
 \end{verbatim} 
Section \ref{sec:fam} contains more information on the family.

%$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+? Most models require this, so please check if you meant to include it.''

It's possible that the user could want some variance components to be set equal. For example, in the Coull and Agresti flu problem, there are 4 years and random effects for each year. The authors want the within-year variance components to be equal.  There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since year 4's virus was a repeat). Suppose year is a categorical variable with four levels, and year1 through year4 are dummy variables. Thus, the call could contain these arguments
\begin{verbatim}
glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeet,~0+year3,~0+year4),
 varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeet"),
data=flu,family.glmm=bernoulli.glmm)
\end{verbatim}

Thus model.matrix will make Z into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. What we do is we take the design matrices that share a variance component and then cbind them together. Thus we'll have 3 model matrices in our Z list, one for each random effect. We will want to put them in order (1,2,3 according to varcomps.equal) so that the names for each matrix line up ("subject","year","repeet").  Cbinding the design matrices that share a variance component will not effect the way that we compute $\eta$.  Also, the variance estimates should come out in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates  will help determine the means and variances for the normal distributions from which random effects will be generated. After  sampling random effects from a standard normal, they will be scaled and centered to have the desired mean and variance. More information on generated the random effects is in section \ref{sec:genRand}. Next, I'll implement trust on the objective function (details on the objective function are in section \ref{sec:objfun}. Finally, the function will return parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood, and the Hessian of the log likelihood at those estimates.

%%Families%%
\section{Families} \label{sec:fam}
This function will be hidden from the user.  These functions (along with the distribution of random effects) are necessary to approximate the log likelihood.

I'm going to have an S3 class called ``glmm.family'' with a few options (for now, binomial and poisson). Each family  function will output a list including the family name (a character string such as ``binomial''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.

The user will provide the family in the model-fitting function. They can either enter the character string (``bernoulli.glmm''), the function (bernoulli.glmm()), or the result of invoking it.  The following code for using the input to determine the family is adapted from glm.

\begin{verbatim}
logDensity<-function(family.glmm)
{
	if(is.character(family.glmm))
		family.glmm<-get(family.glmm,mode="function",envir=parent.frame())
	if(is.function(family.glmm))
		family.glmm<-family.glmm()
	if(!inherits(family.glmm,"glmm.family")) 
		stop(" 'family.glmm' not recognized") 
	return(family.glmm)
}
\end{verbatim}
We interpret this as follows.  If the user has entered the family as a string, go get the R object with that family name, either from the immediate environment or the parent environment.  If this has happened, \texttt{family.glmm} is now a function.  If \texttt{family.glmm} is a function (either because the user entered it as a function or because of the preceding step), invoke that function.  At this point, \texttt{family.glmm} should have  class ``\texttt{glmm.family}.'' If this is not the case (maybe because of a typo or maybe because they entered ``\texttt{poisson}'' rather than ``\texttt{poisson.glmm}''), then stop and return an error.

For example, one of the family functions will look like this:
\begin{verbatim}
poisson.glmm<-function()
{
	family.glmm<- "poisson.glmm"
	c <- function(eta) exp(eta)
	cp <- function(eta) exp(eta)
	cpp<-function(eta) exp(eta)
	out<-list(family.glmm=family.glmm, c=c,cp=cp,cpp=cpp)
	class(out)<-"glmm.family"
	return(out)
}
\end{verbatim}

Then, to use these functions in order to calculate $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$, I can just call 
\begin{verbatim}
family.glmm$c(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For  Bernoulli, we calculate these values (c, cp, and cpp) with careful computer arithmetic as follows. We also use the \texttt{log1p} function in R for $c(\eta_i)$. 
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
We use the \texttt{log1p} function in R for $c(\eta_i)$ for the Bernoulli family. For poisson, these values (c,cp, and cpp) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

Then we use these pieces to create the scalar $c(\eta)$, the vector $c'(\eta)$ and the matrix $c''(\eta)$. We calculate

\begin{align}
c(\eta)= \sum_i c(\eta_i).
\end{align}
 The vector $c'(\eta)$ has components $c'(\eta_i)$. The matrix $c''(\eta)$ is diagonal with diagonal elements $c''(\eta_i)$.

In the R function \texttt{glm} binomial, the user can choose the link. The canonical link must be used in the \texttt{glmm} package so that we have an exponential family. I'm going to include the link in the value of \texttt{family.glmm} just in case the user doesn't know it already.

Also, I'll have a function to check that the data are valid given the family type. So I'll make sure that if \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0's or 1's. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.  If that's not true, the check returns an error message.

\subsection{Redone in C}
When I redo this in C, I will have a separate function for \texttt{cum} ,\texttt{cp}, and \texttt{cpp}. The inputs will be \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family), and an array of doubles to contain the result. These will be passed in as pointers. The functions will calculate  the cumulant function or one of its first two derivatives. Each function will contain a switch statement for the glmm family. The calculations for each of these functions have  been shown earlier in this section. This function will be type void: rather than returning the cumulant or its derivatives, the pointers will be changed to contain the results. This function will be invoked by \texttt{el}, descriped in section \ref{sec:el}.

\section{PQL}\label{sec:pql}
 Before we get started on describing PQL, we need to change notation to avoid constrained optimization.  Recall that $D=\var(u)$ and is (for now) assumed to be diagonal. Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative. Using A rather than D enables unconstrained optimization.  If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
\begin{align*}
A= \sum_t E_t \sigma_t.
\end{align*}
Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix. PQL will estimate the components contained on the diagonal of A. Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).

In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.

There are two ways to do PQL, both of which are described in the vignette
 \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014). In either case, there is an inner optimization and an outer optimization. The inner optimization is well behaved while the other optimization is a little tougher. I will be using the method that is not quite PQL but is pretty close and is better behaved.  In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$. Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.

The {\bf inner} optimization will be done with the trust function in R. We elect to use trust because it requires two derivatives, which will make the optimization more precise. We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.

The inner optimization maximizes the penalized log likelihood. After defining
\begin{align}
\eta=X\beta +ZAs
\end{align}
we calculate the  log likelihood as 
\begin{align}
l(\eta)= Y' \eta - c(\eta) 
\end{align}
and the penalized likelihood as:
\begin{align}
 l(\eta)- \dfrac{1}{2} s's.
\end{align}
This function will be maximized using the \texttt{trust} package. We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$. We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.

Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
\begin{align}
l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
\end{align}
then 
\begin{align}
\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i.
\end{align}
 Now we can write the following expression:

\begin{align}
%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
&=\sum_i (Y_i-\mu_i) X_{ik}
\end{align}

We find the function's derivative with respect to $s$ as follows:
\begin{align}
\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
&=(Y-\mu)'ZA -s
\end{align}

This gives us the following derivatives of the penalized log likelihood:
\begin{align}
\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
\end{align}

Lastly, we need the Hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
\begin{enumerate}
\item $ \dfrac{\partial^2}{\partial s^2}$
\item $ \dfrac{\partial^2}{\partial \beta^2}$
\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
\end{enumerate}

We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
\begin{align}
  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
&= -AZ' \left[ c''(\eta) \right] ZA - I_q
\end{align}
Note that $c''(\eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.  $I_q$ is the identity matrix of dimension $q$.  This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.

Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
\begin{align}
  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
&=-X'  \left[ c''(\eta)\right]  X   
\end{align}

Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
\begin{align}
\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
&=-   \left[ c''(\eta) X  \right]' ZA   \\
&=-X'   \left[ c''(\eta)   \right] ZA  
\end{align}

And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood. Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$. Trust will maximize the penalized likelihood.

The {\bf outer} optimization is done using \texttt{optim} with the default method of ``\texttt{Nelder-Mead.}'' This requires just the function value and no derivatives.  This optimization method was chosen because the optimization function already contains second derivatives of the cumulant function; requiring derivatives of the optimization function would in turn require higher-order derivatives of the cumulant.

The default of \texttt{optim} is to minimize, but we'd like to  do maximization. Reversing the sign of the optimization function will turn the maximization into minimization.

If $\tilde{\beta}$ and $\tilde{s}$ are available from previous calls to the inner optimization function, then they are used here. Otherwise, the values are taken to be 0 and 1. The outer optimization's function is evaluated by first defining
\begin{align}
\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
A&= \sum_k E_k \sigma_k.
\end{align}
Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
\begin{align}
 l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
\end{align}
Again, \texttt{optim} minimizes, so we have to minimize the negative value of the penalized quasi-likelihood in order to maximize the value of it.

We do need to be careful about the determinant. Let $AZ' W ZA +I= LL'$ where $L$ is the lower triangular matrix resulting from a Cholesky decomposition. Then
\begin{align}
\dfrac{1}{2} \log | AZ' W ZA +I | &= \dfrac{1}{2} \log |LL'|\\
&=\dfrac{1}{2} \log (|L|)^2 \\
&= \log |L|
\end{align}
Since $L$ is triangular, the determinant is just the product of the diagonal elements. Let $l_i$ be the diagonal elements of $L$. Then
\begin{align}
\dfrac{1}{2} \log | AZ' W ZA +I | &= \log \prod l_i \\
&= \sum \log l_i
\end{align}

 If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}. This would be more numerically stable, but is slower. Let $O$ be the matrix containing the eigenvectors and let $\Lambda$ be a diagonal matrix with the eigenvalues $\lambda_i$ on the diagonal. Then
\begin{align}
 &AZ' W ZA = O \Lambda O' 
\end{align}
Then we can rewrite the argument of the determinant as follows:
\begin{align}
& AZ' W ZA +I  = O \Lambda O' + I = O \Lambda O + OO'=O (I+\Lambda) O' 
\end{align}
This leads to the careful calculation of our determinant as follows:
\begin{align}
&   \left| AZ' W ZA +I \right| = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
&\Rightarrow \log   | AZ' W ZA +I | = \sum \log(1+\lambda_i) 
\end{align}
The last quantity can be accurately calculated using the \texttt{log1p} function in R.

The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them. To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold. First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.  Second, after PQL is finished, $\tilde{s}$ is used to help center the generated random effects.

The point of doing PQL is to construct a decent importance sampling distribution. Thus, the estimates don't have to be perfect.  It is possible that one of the $\sigma_k$ will be 0 according to PQL. If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.

\section{Log density of the data (\texttt{el})}\label{sec:el}
This section provides details for the log density of the data and two of its derivatives.

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= Y' \eta +c(\eta) \\
&= \sum_{i} Y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+ZU.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = Y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

We'd like the derivative of the log of the data density with respect to $\beta$. This can be written using the chain rule as follows:
\begin{align}
\dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
&= \left[ Y_j-c'(\eta_j) \right]  X_{j{l}}
\end{align}

The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
\begin{align}
\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
 \end{align}

Letting $c'(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
\begin{align}
\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ Y-  c'(\eta)  \right].
\end{align}

Letting $ c(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
\begin{align}
   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ -  c''(\eta) ] X
\end{align}

\subsection{Redone in C}
The C function to calculate the value of the log data density and its two derivatives will be called by reference. The following pointers will be passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}. The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} will be zeros before \texttt{el.C} is invoked. Invoking \texttt{el.C} will then place the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

%%getEk
\section{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
Since $D$ is a function of $E_t$, we need to find a way to calculate $E_t$. This depends on whether $D$ is diagonal or not.

\subsection{D diagonal}
When $D$ is diagonal, recall $E_t$ are diagonal with either 1 or 0 on the diagonal (1 if that random effect has $\nu_t$ as a variance component and 0 otherwise). Let $q_t$ be the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$. When D is diagonal, then $\sum_{t=1}^T E_t = I_q$, the identity matrix with $q$ rows. Finally, recall that $D=\sum_{t=1}^T \nu_t E_t$.

At this point,  \texttt{mod.mcml\$z} is a list with $T$ design matrices (one for each distinct variance component).  We need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have that many rows (and columns). Thus, each $E_t$ will have q rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{D in general (not necessarily diagonal)}
In progress.

\section{Generating random effects (genRand)}\label{sec:genRand}

This will be hidden from the user. It will be called by the model fitting function.

To approximate the log likelihood, we need simulated random effects.  PQL will give us estimates for $\beta$, $\sigma$ and  $s$. Let $\beta^*$, $s^*$ and $\sigma^*$ be the vectors of PQL estimates. Then 
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*.
\end{align}
Also, we can ``unstandardize'' our random effects this way:
\begin{align}
u^*&=A^*s^*
\end{align}

I'll be generating random effects $u$ from these distributions:
\begin{enumerate}
\item $N(0,I)$
\item $N(u^*,D^*)$
\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$
\end{enumerate}
I'll generate them in these proportions: $p_1,p_2,p_3$ where $p_1+p_2+p_3=1$.

The first distribution is centered at 0, just where we expect the random effects to truly be centered. The random effects will be generated with variance 1, but will be scaled in the trust objective function so that they have variance $\nu$, where $\nu$ is given by the iteration of trust.   The second distribution is centered at the PQL best guess of the random effect values and  has the PQL guess of the variance. The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood hessian.  The idea of this last distribution is to generate random effects from a distribution whose first two derivatives are equal to those of the target distribution $f_\theta(u,y)$.

Let $u_k$ be a vector with length q (the number of random effects in the model). Suppose we want to generate random effects $u_k,k=1,...,m$ from a generic distribution $N(\mu,\Sigma)$.  We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ with diagonal entries being $\Sigma$'s eigenvalues $\lambda$. Eigendecompositions take a little bit more time than Cholesky decompositions, but are more stable. Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\
&=O \Lambda O' \\
&= \Sigma.
\end{align}

For each k, draw a sample of size q from the standard normal distribution. Call this sample $\breve{u}_k$. Then we shift and scale it:
\begin{align}
u_k = \breve{u}_k \Sigma^{1/2} + \mu.
\end{align}
This is the process of drawing our sample from $N(\mu,\Sigma^{1/2}\Sigma^{1/2})$ for any mean vector $\mu$ and variance matrix $\Sigma$.

%%distribution of the random effects%%
\section{Distribution of the random effects }
This will be hidden from the user.  This section provides equations for evaluating the log of a normal distribution with a specified mean and variance. When necessary to emphasize and clarify the choice of mean $\mu$ and variance $\Sigma$ , I use the notation $ f(u|\mu,\Sigma)$ rather than $f_\theta(u)$.

This section is useful for calculating $\tilde{f}(u_k)$.  This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives. These are needed for the calculation in \eqref{eq:MCLA}.

Section \ref{sec:tee} contains details that are not currently implemented, but could be. In this version, all random effects are assumed to be normally distributed and all simulated random effects are generated from a normal distribution.

\subsection{Distribution of random effects is normal (\texttt{distRand})}\label{sec:distRand}
In this subsection, I discuss both the assumed distribution of the unobserved random effects ($N(0,D)$) and the distribution used to generate the simulated random effects.  The equations in this section will provide $\tilde{f}(u)$, $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$ for equation \ref{eq:MCLA}.

We assume the random effects are $N(\mu, \Sigma)$. For $f_\theta(u)$, $\mu=0$. This is written generally with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well. Let $\Sigma$ be the variance matrix. We can then write the distribution of the random effects as:
\begin{align}
\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
\end{align}

For now, I  assume that the true variance of the random effects is $\Sigma=D$ where D is diagonal. Formulas for this are in section \ref{sec:Ddiag}.  I generate random effects both from a normal distribution with $\Sigma=D$ diagonal (details in \ref{sec:Ddiag}) and from a normal distribution with $\Sigma$ general (details in \ref{sec:Dgeneral}).

\subsubsection{$\Sigma=D$ Diagonal}\label{sec:Ddiag}
These equations will be used for both $\log f_\theta(u)$ and for pieces of $\tilde{f}(u)$.  This section provides the equations for finding the value  of a normal density with $D$ diagonal, the gradient vector, and the hessian matrix.

In the simplest case, $\Sigma=D$ is diagonal with variance components $\nu_t$ on the diagonal and 0 off diagonal. This is what I assume to be the true form of D for now.  Remember that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s everywhere else.

We can partition the random effects according to their variance components: $U=(U_1',...,U_T')'$.  Let $D_t$ be the variance matrix for $U_t$ . $D_t$ has $q_t$ rows (and also columns). Thus $D$ is diagonal:
\begin{align}
D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
\end{align}

Since $D$ is diagonal, it follows that $D^{-1}$ is also diagonal with diagonal entries $\dfrac{1}{\nu_1}$,...,$\dfrac{1}{\nu_T}$.  Also, the assumption that $D$ is diagonal makes calculating the determinant of $D$  easy:
\begin{align}
|D|= \nu_1^{q_1}...\nu_T^{q_T}
\end{align}

 Taking these two pieces of information into account allows us to write the log density for $U$ as follows:
\begin{align}
\log f_\theta(u) &= (-1/2) \log |D| - (1/2) (U-\mu)' D^{-1} (U-\mu)\\
&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} (U_t-\mu_t)'(U_t-\mu_t)   \right]
\end{align}

%Letting $I_{q_t}$ be the identity matrix with $q_t$ rows (and columns), we can write this as a sum as follows:
%\begin{align}
%\log f_\theta(u)&= \sum_{t=1}^T \log f_\theta(u_t)\\
%&=\sum_{t=1}^T - \dfrac{1}{2} \log |D_t| - \dfrac{1}{2} (U_t-\mu_t)' D_t^{-1} (U_t-\mu_t) \\ 
%&=\sum_{t=1}^T - \dfrac{1}{2} \log |\nu_t^{q_t}| - \dfrac{1}{2 \nu_t} (U_t-\mu_t)' I_{q_t} (U_t-\mu_t) \\ 
%&=\sum_{t=1}^T - \dfrac{q_t}{2} \log \nu_t - \dfrac{1}{2 \nu_t}(U_t-\mu_t) '(U_t-\mu_t)
%\end{align}

The first and second derivatives of each summand with respect to its associated $\nu_t$ are:
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}(U_t-\mu_t)'(U_t-\mu_t)
\end{align}
and 
\begin{align}
\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} (U_t-\mu_t)'(U_t-\mu_t).
\end{align}
Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
\begin{align}
\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
\end{align}
Also,
\begin{align}
\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
\end{align}
Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is the following vector of length $T$:
\begin{align}
\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} (U_1-\mu_1) ' (U_1-\mu_1) & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} (U_T-\mu_T) '(U_T-\mu_T)   \end{bmatrix} 
\end{align}
To calculate this vector, I will create a function to take $q_t$, $U_t$, and $\nu_t$ and  output $\dfrac{-q_t}{2\nu_t} +\dfrac{1}{2 \nu_t^2} U_t'U_t$. I can then do a loop for t in 1 through T to calculate each entry in the vector.

The Hessian matrix is the following diagonal matrix:
\begin{align}
\dfrac{\partial^2}{\partial \nu^2} \log f_\theta(u) = \begin{bmatrix} \dfrac{q_1}{2 \nu_1^2}- \dfrac{1}{\nu_1^3} U_1'U_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & \dfrac{q_T}{2 \nu_T^2}- \dfrac{1}{\nu_T^3 U_T'U_T} \end{bmatrix}
\end{align}
I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$. Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.

The inputs of the function are the vector $\nu$ of variance components, U (the vector of random effects), and the list \texttt{z} (from \texttt{mod.mcml}).  The list \texttt{z} has $T$ matrices, each with the number of columns equal to $q_t$. We need $q_t,t=1,\ldots,T$ to calculate the log density and its derivatives.

The last thing we need to discuss is how to split $U$ into $U_1,...,U_T$ and $\mu$ into $\mu_1,...,\mu_T$. The same process should work for both, so let's just look at how we'd do it for $U$.  We know that the first $q_1$ items of U are $U_1$, the next $q_2$ items are $U_2$, etc.  In other words, entries 1 through $q_1$ are $U_1$. Items $q_1+1$ through $q_1+q_2$ are $U_2$, etc. In R pseudo code, this means 
\begin{verbatim}
nrand<-lapply(z,ncol) #get the number of columns from each matrix in list z
unlist(nrand) #chance the list into a vector, since each entry in the list is just a scalar

U1<- U[1 through nrand[1]] #first rule is use 1:nrand[1]
Ut<-U[sum(nrand[1:t-1]+1 through sum(nrand[1:t])]
\end{verbatim}

Note that we don't need to actually calculate either A or D. We work only through $\nu$.

\subsubsection{Newer version of $\Sigma=D$ Diagonal}
The newer version of \texttt{distRand} (called \texttt{distRand3}) uses the same equations as those listed in \ref{sec:Ddiag}. The only difference is that the newer version is more efficient. Rather than having each call to \texttt{distRand3} recalculate which elements of $U$ belong to each $t$, this is done ahead of time in the objective function. This information is contained in vector \texttt{meow}. This is then an argument in \texttt{distRand3}.

Rewritten in C, this function will take as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{U} up based on the variance components, the double array \texttt{drgradient} that will contain the resulting gradient, and the double array \texttt{drhessian} that will contain the resulting hessian.

The result of invoking the C function will be the calculation of the gradient and hessian for the distribution of the random effects. The value will be evaluated by the C function in \ref{sec:Dgeneral}.

%%distRandGeneral
\subsubsection{$\Sigma$ in general (\texttt{distRandGeneral})}\label{sec:Dgeneral}
For now, this is only used for $\tilde{f}(u)$. (When $\Sigma$ is general, calculating the derivatives of $\log f_\theta(u)$ is not easy.) Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
\begin{align}
\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
\end{align}
 The only part of this to discuss is $|\Sigma^{-1}|$. We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues. Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$. Thus  
\begin{align}
|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
&= |O'| \; |O| \; |\Lambda| \\
&=|\Lambda|,
\end{align}
which is just the product of the eigenvalues. The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times each optimization iteration, where $m$ is again the Monte Carlo sample size.

This function is also rewritten in C with the following  passed in as pointers: double \texttt{Sigma.inv} $\Sigma^{-1}$, double \texttt{logdet} $\log |\Sigma^{-1}|$, int \texttt{nrow}, double \texttt{uvec} a vector of random effects, double {mu} $\mu$, and double \texttt{distRandGenVal}.

\subsection{$\tilde{f}(u_k)$}\label{sec:ftwiddle}
I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust. I can make a function using the general form for a normal distribution's density as described in the previous subsection, then use that function for the three distributions that I sampled from. Since I generated the random effects from the three distributions in proportions $p_1,p_2,p_3$, I can write 
\begin{align}
\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
\end{align}

\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL). These can be drawn using the package mvtnorm. This package can also give the density evaluated at certain points. I'm not sure if I'll use this or just stick to normals.

\subsection{Central Limit Theorem for MCLA}
Define
\begin{align}
\gamma_1&= \int f_\theta(u,y) du\\
\gamma_2 &=\int \tilde{f}(u)du.
\end{align}
Recall the calculation for the MCLA gradient in  \eqref{eq:MCLAgradient}. We can rewrite it as
\begin{align}
\nabla l_m(\theta)=  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
\end{align}
to more obviously show that both the numerator and denominator are sample means. 
The law of large numbers says that the numerator and denominator  each converge to their true means. That is,
\begin{align}
\dfrac{1}{m}\sum_{k=1}^m \nabla \log f_\theta (u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
&= \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right]
\end{align}
and
\begin{align}
\dfrac{1}{m}\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow E_{\tilde{f}} \left[ \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
&= \int \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
&= \dfrac{\gamma_1}{\gamma_2} \int \dfrac{ f_\theta(u,y)}{\gamma_2} du \\
&= \dfrac{\gamma_1}{\gamma_2}
\end{align}
Then, by Slutsky's,
\begin{align}
\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
&=  E_f \left[ \nabla \log f_\theta(u,y)  \right]
\end{align}
In addition to a law of large numbers, we would like a Central Limit Theorem for $\nabla l_m(\theta)$. In other words, the quantity
\begin{align}
&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- E_f \left[ \nabla \log f_\theta(u,y)  \right]  \right] \\
&=
\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m} E_f \left[ \nabla \log f_\theta(u,y)  \right]  \\
&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)
 - E_f \left[ \nabla \log f_\theta(u,y)  \right] \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
\end{align}
will have a normal distribution if and only if the variances of
\begin{align}
\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
\end{align}
and
\begin{align}
\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
\end{align}
are each finite.
First, we want to show that
\begin{align}
\sum_{k=1}^m \dfrac{f_\theta(u_k)}{\tilde{f}(u_k)}
\end{align}
has finite variance. This is true if and only if 
\begin{align}
\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du < \infty.
\end{align}
We see:
\begin{align}
\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du &= \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
& \leq \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)} du \\
&= \dfrac{1}{p_1} \int N(u|0,D) du \\
&= \dfrac{1}{p_1}.
\end{align}
Therefore $\sum_{k=1}^m f_\theta(u_k)$ has finite variance as long as $p_1\neq 0$.

Second, we want to show that the gradient of the MCLA has finite variance. That is, we want to show that
\begin{align}
\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty.
\end{align}
That is, we want to show 
\begin{align}
\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du  \\
&< \infty.
\end{align}
By the Cauchy-Schwartz inequality, 
\begin{align}
  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
\end{align}
Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
\begin{align}
\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
\end{align}
Recall the notation from section \ref{sec:Ddiag}.  If
\begin{align}
\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
\end{align}
for every $t=1,...,T$, then
\begin{align}
\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
\end{align}
by Cauchy Schwartz.  Therefore, I will prove the integral is finite for the general case of $u_t$ and $\nu_t$. First, we recognize that our integral is actually an expectation of a function of a normal random variable:
\begin{align}
\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
\end{align}
Normal distributions have finite moments of all orders, so we know that\begin{align}
\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
\end{align}

%
%\begin{align}
%&=
%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ \dfrac{q_t^2}{4 \nu_t^2} + \dfrac{(u_t'u_t)^2}{4 \nu_t^4} -2\dfrac{ q_t u_t'u_t  }{4 \nu_t^3} \right] du_t \\
%&=
%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       \dfrac{q_t^2}{4 \nu_t^2} du_t +
% \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{(u_t'u_t)^2}{4 \nu_t^4} du_t \nonumber \\
%&-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{ q_t u_t'u_t  }{2 \nu_t^3}     du_t \\
%&=
% \dfrac{q_t^2}{4 \nu_t^2} \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       du_t +
%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t \\
%&=
% \dfrac{q_t^2}{4 \nu_t}+
%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t 
%\end{align}
%
%Define $erf$ as the ``error function,'' the function from integrating a normal distribution. Evaluating the integral and using the fact that $erf$ is bounded between $-1$ and $1$, we see 
%\begin{align}
%\int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t
%&\propto
%    \nu^{3/2} erf \left(\dfrac{u_t}{\sqrt{2 \nu_t} }   \right ) - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)}   \\
%& \geq -  \nu^{3/2}  - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)} \\
%& \geq - \infty
%\end{align}
%for all $u_t$ and for all $\nu_t.$ Lastly, we need to make sure 
% \begin{align}
% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t < \infty
%\end{align}
%Let $c_1$, $c_2$, and $c_3$  be constants with respect to $u_t$. Then 
%\begin{align}
% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t &= c_1 erf \left( \dfrac{u_t}{\sqrt{2 \nu_t}} \right) - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
%& \leq  c_1  - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
%& < \infty
%\end{align}
%since
%\begin{align}
%e^{-u_t'u_t / (2 \nu_t)} > \infty
%\end{align}
%for all $u_t$ and all $\nu_t$.

Next, we want to show 
\begin{align}
\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
\end{align}
 Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
\begin{align}
\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
\end{align}
We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
\begin{align}
\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
\end{align}
is finite.
When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
\begin{align}
\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
\end{align}
Then
\begin{align}
\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
&\propto \int N(u|0,D) e^{2Zu} du 
\end{align}
is finite.

Therefore, it is proven that 
\begin{align}
\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
\end{align}
exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.

%
%Next, we need to show that the hessian of the MCLA has finite variance. Again, it is enough to show that
%\begin{align}
%\int \dfrac{N(u|0,D)^2 h(u)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty
%\end{align}

%\subsection{Boundedness of importance weights}
%We want to make sure that the importance weights will be bounded when using this distribution. The importance weight for a vector $u_k$ is $\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)}$. Since $f_\theta(y,u_k)< \infty$, 
%\begin{align}
%\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} < \infty \Longleftrightarrow \tilde{f}(u_k) >0.
%\end{align}
%
%We start with
%\begin{align}
% \tilde{f}(u_k) &= p_1  f(u_k \, | \, 0, D)+p_2  f(u_k \, | \, u^*, D^*)+p_3  f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\\
%&\geq p_1  f(u_k \, | \, 0, D) \\
%& \geq p_1 (2 \pi)^{-q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u}
%\end{align}
%In order to show that $\tilde{f}(u_k)>0$, we need to show that both $ |(D^*)^{-1/2}|$ and  $e^{- u' (D^{*})^{-1}u}$ are positive. 
%
% First, I will show that $ |(D^*)^{-1/2}|>0$.Since  $(D^*)^{-1/2}$ is invertible, its determinant must be nonzero.  Since $D^*$ is positive semi-definite, then $(D^*)^{-1/2}$ is also positive semi-definite. Thus the eigenvalues of $(D^*)^{-1/2}$ are neither zero nor negative; they must be positive. An eigendecomposition tells us $|(D^*)^{-1/2}|$ is the product of its eigenvalues. Therefore, $|(D^*)^{-1/2}| >0$. 
%
%Next, I will show that $e^{- u' (D^{*})^{-1}u}>0$.  Since $D^*$ is a covariance matrix, it is positive semi-definite.  By definition, for any vector $a_1$, 
%\begin{align}
%a_1' D^* a_1 \geq 0
%\end{align}
%For every vector $a_2$, there exists an $a_1$ such that $a_2=D^*a_1$. Then
%\begin{align}
%a_2' (D^*)^{-1} a_2 &= (D^* a_1)' (D^*)^{-1} D^* a_1\\
%&= a_1' D^* (D^*)^{-1} D^* a_1\\
%&= a_1' D^* a_1\\
%& \geq 0
%\end{align}
%Therefore,   $(D^*)^{-1}$ is also positive semi-definite. Then for any vector of random effects $u$,
%\begin{align}
% u' (D^*)^{-1} u \geq 0 
%\end{align}
%This means that
%\begin{align}   e^{- u' (D^{*})^{-1}u} >0
%\end{align}
%Therefore,
%\begin{align}
% \tilde{f}(u_k) &\geq p_1 (2 \pi)^{q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u} >0
%\end{align}
%

%%%%%%%%%%%%%%%%%
%\section{t distribution for generating random effects}
%Generating random effects from a normal distribution can give us some random effects from the tails of the distribution, which results in importance sampling weights that are problematically big.  To try to fix this, we can change our importance sampling distribution to a multivariate t-distribution with location parameter $\mu$, scale parameter $D$, and $\gamma$ degrees of freedom.  The pdf is proportional to 
%\begin{align}
%f(U |\theta)=  |D|^{-.5} \left[ 1+\dfrac{1}{\gamma} (U-\mu)' D^{-1}(U-\mu)   \right]^{(\gamma+p)/2}
%\end{align}
%
%Then the log density is
%\begin{align}
%\log f(U |\theta) &\propto -.5 \log |D| - \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} (U-\mu)' D^{-1} (U-\mu)  \right] \\
%& \propto \sum_{t=1}^T \left[- .5  q_t \log \nu_t \right]- \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} (U-\mu)' D^{-1} (U-\mu)  \right]  \\
%& \propto \sum_{t=1}^T \left[- .5  q_t \log \nu_t \right]- \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} \sum_{t=1}^T \dfrac{1}{\nu_t} (U_t-\mu_t)' (U_t-\mu_t)  \right]  \\\end{align}
%Its first derivative with respect to $\nu_t$ is
%\begin{align}
%\dfrac{\partial}{\partial \nu_t} \log f(U |\theta) &= \dfrac{- q_t}{2 \nu_t} + \dfrac{\gamma+q}{2} \dfrac{\dfrac{1}{\gamma \nu_t^2}(U_t-\mu_t)' (U_t-\mu_t)}{1+ \dfrac{1}{\gamma} \sum_{t=1}^T \dfrac{1}{\nu_t} (U_t-\mu_t)' (U_t-\mu_t)}
%\end{align}
%Its second derivative with respect to $\nu_t$ is
%
%\begin{align}
%\dfrac{\partial^2}{\partial \nu_t^2} \log f(U |\theta) &= \dfrac{q_t}{2 \nu_t^2} - \dfrac{\gamma+q}{2} \left[ \dfrac{\dfrac{1}{\nu_t^2 \gamma}(U_t-\mu_t)'(U_t-\mu_t)}{1+ \sum_{t=1}^T \dfrac{1}{\nu_t \gamma }(U_t-\mu_t)'(U_t-\mu_t) }  \right]^2 - \dfrac{\gamma+q}{2}  \left[ \dfrac{\dfrac{2}{\nu_t^3 \gamma}(U_t-\mu_t)'(U_t-\mu_t)}{1+ \sum_{t=1}^T \dfrac{1}{\nu_t \gamma}(U_t-\mu_t)'(U_t-\mu_t) }  \right]
%\end{align}

\section{Objective function: approximated log likelihood}\label{sec:objfun}
This will be hidden from the user. It will be called by the model fitting function.

In order to maximize the approximated log likelihood using trust, I will need an objective function that returns the value, the gradient vector, and the Hessian matrix of the log likelihood.  This objective function will get $\log f_\theta (u_k)$, $c(\eta_i)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.    Denote\\
\begin{align}
b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \label{eq:MCLA}
\end{align}

For computational stability, we'll set $a=max(b_k)$. Then the value is expressed as:\\
\begin{align}
\log f_{\theta,m} = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

Define the weights as:\\
\begin{align}
v_\theta(u_k,y) = \dfrac{e^{b_k-a}}{ \sum_{k=1}^m e^{b_k-a}} 
\end{align}

The gradient vector is: \\
\begin{align}
G &= \sum_{k=1}^m \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] v_\theta(u_k,y)+ \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)
\end{align}

Then the Hessian matrix is: \\
\begin{align}
H = \sum_{k=1}^m \nabla^2 \log f_\theta (u_k,y) v_\theta(u_k,y) +  \sum_{k=1}^m \left[ \nabla  \log f_\theta (u_k,y)  \right] \left[ \nabla  \log f_\theta (u_k,y)  \right]^T  v_\theta(u_k,y) -G G^T
\end{align}
Everything in this has already been defined except:
\begin{align}
\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
\end{align}

\section{Summary of model}
Typing \texttt{summary(mod)} will give more detailed info of the model. This will be broken into two pieces (as is usually done for summaries): there will be the summary.mcml and print.summary.mcml. We split this up because we might have a lot of extra stuff in the summary that we don't necessarily want printed each time. So, when a user types \texttt{summary(mod)}, only the  basic information is automatically printed. More information can be found in the summary list.

The summary will do all the calculations and its value will be a list of the following:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix with the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column. All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
\item the evaluated Monte Carlo log likelihood along with its first and second derivative
\item maybe other things that I haven't thought of
\end{itemize}

A note on the standard errors: to calculate the standard errors, we take the MCLA Hessian matrix, invert it, take the diagonal elements, and square root them. It's possible that the Hessian will be noninvertible if it is close enough to singular that the computer thinks it's singular.  Then the standard errors will all be infinite. We also need to warn the user why this is happening.

Then print.summary.glmm will print these things:
\begin{itemize}
\item call
\item the variance estimate(s)
\item a matrix similar to the output of summary.glm, which will be put through printCoefmat in order to get the significance stars that we're used to. We'll have the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column, the significance stars, and the significance legend.
\item maybe other things that I haven't thought of. I'll probably realize more once I get coding.
\end{itemize}

I think for the printCoefmat to work, the fixed effects matrix needs to look like
\begin{verbatim}
            Estimate Std. Error z value     Pr(>|z|)
(Intercept)        2       0.50       4 6.334248e-05
x1                 2       0.25       8 1.244192e-15
\end{verbatim}
(or at least it worked when I had these as the column names and didn't work before that).

Note that the summary and print.summary functions will be S3 generic functions.  What this means is the user will type ``summary(mod)'' and summary is a generic function. R checks the class of the model $mod$ and then automatically uses the summary and print.summary functions for that class of objects. This also might affect the way that I export the function and document it in the manual.  I think that I'll be ok if I copy parts of his summary.aster and print.summary.aster. I need to read about generics more and will check out the hints library.

\section{Checks}
\subsection{Checking the MCLA finite differences}
To check the function that calculates the MCLA $l_m(\theta)$, I use finite differences on the \citet{booth:hobert:1999} example. To do this, I chose a value of $\theta=(\beta,\sigma)$ and a relatively small value of $\delta$, where $\delta$ is a vector of length $2$. We can check that the value and first derivative of the MCLA function are calculated correctly by making sure the following approximation holds
\begin{align}
\nabla l_m (\theta)  \cdot \delta \approx l_m(\theta+\delta)-l_m(\theta).
\end{align} 
Then, we can check the first and second derivatives of the MCLA are calculated correctly by checking for the following approximation:
\begin{align}
\nabla^2 l_m (\theta) \delta \approx \nabla l_m (\theta+\delta)-\nabla l_m (\theta)
\end{align}

\subsection{Checking  functions using the Booth and Hobert example}
I also test the objective function by taking the  \citet{booth:hobert:1999} example and rewriting the functions for this specific example. I then compare the values produced by the check functions and the original functions. Functions checked this way are:
\begin{enumerate}
\item log of the data density (\texttt{el})
\item log of the density for the random effects (both \texttt{distRand} and \texttt{distRandGeneral})
\item the Monte Carlo likelihood approximation (\texttt{objfun})
\end{enumerate}

\subsection{Checking a putative MLE using MCMC}

Suppose $\hat{\theta}$ is claimed to be the MLE. For example, $\hat{\theta}$ could be the MCMLE.  Ideally, we would like to check whether the true likelihood $l(\theta)$ achieves a local max at $\hat{\theta}$.  Due to the intractible integral in the likelihood expression, the best we can do is make sure that  $\nabla l_m(\theta)$ evaluated at $\hat{\theta}$ is very close to $0$.

Suppose $u_k,k=1,\ldots,m$ are sampled from importance sampling distribution $\tilde{f} (u_k)$. Recall that the gradient of the MCLA is:
\begin{align}
\nabla l_{m}(\theta) = \dfrac{\sum_{k=1}^m\frac{\nabla f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}}{\sum_{k=1}^m \frac{  f_\theta(u_k)   }{\tilde{f}(u_k)}}
\end{align}
We can choose $\tilde{f}(u_k)=f_{\hat{\theta}}(u_k,y)$. If the putative MLE is truly the MLE, then $\theta=\hat{theta}$, which in turn implies that $\tilde{f}(u_k)=f_{{\theta}}(u_k,y)$.   To be clear, in this check  we no longer use a mixture of normals for $\tilde{f}(u_k)$.  With the selected importance sampling distribution, each of the importance sampling weights is equal to $1$ and the sum of the weights is $m$.  Therefore, the gradient of the MCLA simplifies as follows:
\begin{align}
\nabla l_{m}(\theta) &=\dfrac{1}{m} \; {\sum_{k=1}^m\dfrac{\nabla f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k,y)}}\\
&= \dfrac{1}{m} \; {\sum_{k=1}^m\dfrac{\nabla f_\theta(u_k,y)   }{\tilde{f}(u_k,y)}} \\
&= \dfrac{1}{m} \; \sum_{k=1}^m \nabla \log f_\theta(u_k,y)
\end{align}
This shows that the gradient of the MCLA is the average of the gradient of the complete data log likelihood, as long as $\hat{\theta}$ is truly an MLE. We can produce $u_k,k=1,\ldots,m$, using Markov chain Monte Carlo. Using $u$ as the variable, w run the Markov chain (perhaps \texttt{metrop} from the R package \texttt{mcmc}) on the complete data log likelihood.  We then use these samples to calculate the gradient of the complete data log likelihood, which in turn calculates the gradient of the MCLA.

 If we split the MCMC runs into batches, we can calculate the batch means of the MCLA gradient, the grand mean of the MCLA gradient, and the corresponding Monte Carlo standard error.  If the putative MLE truly maximizes the likelihood, then the MCLA gradient's components should be close to $0$. We can check that they are close enough to $0$ by comparing them to the Monte Carlo standard error.

\section{Likelihood ratio test}
This is another S3 generic function that the user can choose to implement if they'd  like  to do likelihood ratio tests for nested models. Eventually I'll want this to handle an arbitrary number of models, but for now we'll stick to comparing two models: $mod2$ nested in $mod1$.  I will assume these models have already been fit.

 There are a few ways that the two models could differ. In other words, we could be testing:
 \begin{enumerate}
 \item whether one or more fixed effects are zero.
 \item whether one variance component is zero.
 \item whether multiple variance components are zero.
 \item whether one or more fixed effects is zero and one or more variance components are zero. 
\end{enumerate}
The hypothesis testing procedure and method for calculating pvalues are different for each of these cases, so I'll go through each one in the following subsubsections.

\subsection{Testing whether one or more fixed effects are zero}
Consider the  case of testing whether one or more fixed effects are zero and the random effects are the same between the two models. I think we just do a likelihood ratio test as we're familiar with for linear models.   Let the number of fixed effect parameters in the larger model be $p_1$  and the number in the nested model be $p_2$. I will need the log likelihood values of the models outputted (from the main function at the same time as when I output the MCMLEs and the Fisher information). The log likelihood is simply the ``value'' from the objective function that trust uses. Let's call the log likelihood from the larger model  $l_1$ and the log likelihood from the nested model  $l_2$.

Then the likelihood ratio test statistic is
\begin{align}
t_{LRT}= -2l_1+2l_2
\end{align}
and this follows a $\chi^2$ distribution with $p_1-p_2$ degrees of freedom. The calculation
\begin{align}
P(t_{\chi^2_{p_1-p_2}>LRT} ).
\end{align}
gives us a two-sided pvalue to fit with the two-sided alternative hypothesis.

 \subsection{Testing whether one variance component is zero}
  Hypothesis testing for one variance component is easy but not what most people expect.
  In this setup, the two models have the exact same fixed effects. The only difference is the larger model has one more random effect $\nu_{t_1}$. This means we want to test whether $\nu_{t_1}=0$. A variance component must be nonnegative, meaning the alternative hypothesis is that $\nu_{t_1}>0$. In other words, the hypotheses are:
 \begin{align}
 H_0: \nu_{t_1}=0\\
 H_1: \nu_{t_1}>0.
 \end{align}
 Note the alternative hypothesis is one-sided. This means we need to calculate a one-sided pvalue. Let $t_{LRT}=2l_2-2l_1$. If we naively follow the pvalue calculation of
 \begin{align}
P( \chi^2_{1}>t_{LRT})
\end{align}
 we will end up calculating a two-sided pvalue. To fix this, we cut this pvalue in half. In other words, use the standard normal distribution $Z$ and think of calculating the one-sided pvalue this way:
 \begin{align}
 P( Z>\sqrt{|  t_{LRT} |})
 \end{align}

 However, this is not obvious; I never would have thought about having a test statistic of
 \begin{align}
 \sqrt{2 | l_2-l_1   |   }.
 \end{align}

 \subsection{Testing whether multiple variance components are zero}
 In this case, the fixed effects are identical between the larger and nested model. The only difference is that the larger model has more than one additional variance components. This gets a lot more complicated (for example, there is no clear way to count parameters so calculating the degrees of freedom is out the window). Luckily, someone's worked out the details already. Charlie suggested a few papers to look at. I'll read about and add this a bit later.  Charlie is hoping that we'll have a mixture of $\chi^2$ distributions, such as
 \begin{align}
 \dfrac{1}{2}P (\chi^2_1> t_{LRT}) +\dfrac{1}{4}P (\chi^2_2> t_{LRT}).
 \end{align}
 The reason this is complicated is because the variance components are restricted to be nonnegative. This means that the likelihood is only defined for nonnegative variance components. Then differentiating the likelihood at $0$ becomes tricky because that's the boundary.

 \subsection{Testing whether one or more fixed effects is zero and one or more variance components are zero}\label{sec:combotest}
 This seems very complicated. I don't know if it's any more complicated than the previous case of testing multiple variance components. I'll have to think about it later when I have time.

 \subsection{Determining the hypothesis test}
 To follow convention, this command will be called ``anova'' and the two arguments will be the two models to be compared.  The command would look like
\begin{verbatim}
anova(mod1,mod2)
\end{verbatim}

 R will first need to figure out if the fixed effects are different and, if they are, which model is the larger model.
\begin{verbatim}
if(length(coef(mod1))>length(coef(mod2)))
      {bigmod<-mod1;  smallmod <-mod2} 
if(length(coef(mod1))<length(coef(mod2)))
      {bigmod<-mod2; smallmod <-mod1}
\end{verbatim}
Next, if there is a difference in the fixed effects, I'm going to make sure the fixed effects of the small model are nested in those of the big model. Otherwise, produce an error. Then, continue with the testing. Note: this check for nesting isn't perfect, but no other anova checks which model is bigger and whether they're nested.
\begin{verbatim}
if(big mod is defined) {
      pnames1<-names(coef(bigmod))
      pnames2<-names(coef(smallmod))
      if(sum(pnames2 %in% pnames1) != length(pnames2)) {
            stop("The models you provided are not nested.")}

     if(the variance components differ between the two models){
             check big model has more variance components (ow: error)
             check variance components of the small model are nested (ow: error)
             calculate pvalue according to section 3.3.4. return it.
      }
      if(the variance components do not differ between the two models){
            calculate pvalue according to 3.3.1. return it.
      }
}
\end{verbatim}
If and only if we've gotten to the end of this chunk of code without returning anything, bigmod has not been defined, meaning the fixed effects are same. Therefore, we now need to figure out whether the variance components differ by one or by more than one. In order to compare the number of random effects, we need to figure out how to get the number of variance components from each model. I  don't yet have a solid grasp of how formula works its magic, but I have the impression that I can count the number of model matrices returned in order to figure out the number of variance components because there is a model matrix for each one.

\begin{verbatim}
T1<- number of variance components for mod1
T2<- number of variance components for mod2
if(T1>T2){call mod1 the bigmod and mod2 the smallmod}
if(T1<T2){call mod2 the bigmod and mod1 the smallmod}
if(bigmod is defined){
     check  small model is nested in the big model (ow: error)
     (maybe do this by looking at the call?)
     if(T1==T2+1 || T2==T1+1) 
           {calculate and return pvalue according to sec 3.3.2}     
     calculate and return pvalue according to section 3.3.3
     }
if still haven't returned anything, produce error.     
\end{verbatim}

Why the last error? If we have gotten to this point in the code with nothing returned, it ends up that the number of variance components in each model are equal, meaning the user has made some kind of mistake. Either the user provided two identical models or they provided non-nested models (e.g. they have the same number of variance components, but different components in each model). Either way, we can't help them.

This command would return something that first reminds the user what predictors are in each model, then has a table. Each model would have one row of the table. The columns would contain the model name,  the log likelihood for each model, the number of parameters in the model. Then the next columns would have one entry each: the calculated difference between the log likelihoods, the calculated difference between the number of parameters, and the pvalue of the test.

%\section{AIC}
%Typing 
%\begin{verbatim}
%AIC(mod)
%\end{verbatim} 
%will return the AIC of the model. To calculate AIC, I'll need the total number of parameters ($p+T$)  and the Monte Carlo log likelihood evaluated at the MCMLEs $l$. Then the AIC is
%\begin{align}
%2(p+T) - 2 l 
%\end{align}

\section{Confidence intervals}
The user can implement this command to create confidence intervals after  fitting a model $mod$ using the main function.  This is an S3 generic.The command would look like
\begin{verbatim}
confint(mod,parm,level=.95)
\end{verbatim}
The only required argument would be the fitted model (the first argument). The third argument (the confidence level) has a default of $.95$.

If the second argument is omitted, confidence intervals would be created for all of the parameters. There are two options to calculate confidence intervals for a subset of the parameters. The user can provide either  the names of the coefficients or a vector with length equal to the number of parameters (entries being 1 if they would like that intervals for that parameter and 0 otherwise).  The code to do this is taken from ``confint.lm'' and is:
\begin{verbatim}
function (object, parm, level = 0.95, ...) 
{
    cf <- coef(object)
    pnames <- names(cf)
    if (missing(parm)) 
        parm <- pnames
    else if (is.numeric(parm)) 
        parm <- pnames[parm]
        stopifnot(parm %in% pnames)

        rest of the code to actually do the confidence intervals goes here
        }
\end{verbatim}
What this says: write down the coefficient names of the object we're given. If parm is missing, we're going to assume they want to create intervals for all the parameters.  If parm is numeric (a vector of 0s and 1s), use that to select the parameter names we want and call that selection parm.  Finally, stop the whole process if parm (the parameter names we want intervals for) don't appear in the model.

The form for a confidence interval is the point estimate plus or minus the standard error of the point estimate times some cutoff.  The point estimate and the standard error are from the summary of the model. The reference distribution used for the cutoff is the standard normal. This will produce an asymptotic confidence interval.

The output would either be a vector of length 2 (if creating confidence interval for only one parameter) or a matrix with 2 columns (one for the lower bound of the confidence interval, one for the upper bond). The column names will be ``$(100-level)/2$'' and ``$50+level/2 $.'' (in the default case, this is``$2.5\%$'' and ``$97.5\%$'').

The only potential hangup is when creating confidence intervals for the variance components $\nu_t, t=1,...,T$.  We know that $\nu_t >0$ for all $t=1,...,T$. We'll find this boundary again makes things complicated. I don't yet know how to deal with this.

To illustrate the problem, consider the scenario that the margin of error for $\nu_t$ is greater than the estimate of $\nu_t$ itself. It wouldn't make sense to produce a confidence interval with a negative lower bound. It could make sense to truncate the interval so that the lower bound starts at 0 and the upper bound remains untouched.

I thought for three or four minutes about a one sided confidence interval, but I don't think that captures what we want. We want the range of all likely values for the variance component, not just an upper bound.

%
%\section{Prediction intervals}
%This is another S3 generic. The user can implement this command to create prediction intervals after  fitting a model $mod$ using the main function. Still learning about these. I currently know almost nothing. What I do know is that I will need the Monte Carlo log likelihood along with its first and second derivatives. I'll get to practice my first-year-theory delta method skills.
%
%Charlie says that neither predict.glm nor predict.glmmPQL do prediction intervals. The former produces parameter estimates with standard errors. The latter does predictions but does not produce a standard error, so no interval can be formed.
%
%One wrong thing to do: once you find the MCMLEs, you simply plug those into the distribution for $U|Y$ and pretend those parameter estimates have no variability and are  correct. Then the random effect has a normal distribution and it'd be easy to find quantiles. The only problem is this is wrong.
%
%We are hoping to find a way to use Fisher information for the Monte Carlo MLE. Hopefully we can do something more similar to the usual asymptotics.
%
%A last resort is a parametric bootstrap. I haven't spent much time thinking about this, but I think the way this involves plugging parameter estimates  into $\log f_\theta(y|u)$ and $\log f_\theta(u)$. Then you repeat: draw a value of $u$, plug that into $\log f_\theta(y|u)$, draw a value of $y$. I say this is a last resort  because having two Monte Carlo calculations will take a long time.

\bibliographystyle{apalike}
\bibliography{brref}

\end{document}

### 2014-11-20 NewDesignDoc/DesignDocGLMM.tex
Updating design doc MCLA derivs
Erroneously did not take into account that ftwiddle depended on theta. Thus, terms were missing.

@@
 l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
-&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}
-\end{align}
-the gradient vector of the MCLA is
-\begin{align}
-\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
+&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
+\end{align}
+When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
+\begin{align}
+\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
 \end{align}
@@
 
+\appendix
+\section{MCLA calculations}
+The Monte Carlo log likelihood approximation is
+\begin{align}
+l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
+&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
+\end{align}
+
+\subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$}
+When $\tilde{f}$ contains $\theta$, the gradient of the MCLA is
+\begin{align}
+\nabla l_m(\theta) &= \dfrac{ \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}  \right] }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
+&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)   }{\tilde{f}(u_k)} - \dfrac{ f_\theta(y,u_k)  \nabla \tilde{f}(u_k) }{\left(\tilde{f}(u_k) \right)^2 } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
+&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)}{f_\theta(y,u_k)} \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
+ \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \dfrac{ \nabla \tilde{f}(u_k) }{\tilde{f}(u_k) } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
+&= \dfrac{   \sum_{k=1}^m  \nabla \log f_\theta(y,u_k) \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
+ \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  \nabla \log \tilde{f}(u_k)  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
+&= \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} 
+\end{align}
+
+Then the Hessian of the MCLA is
+\begin{align}
+\nabla^2 l_m (\theta) &= \nabla \left[ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \right] \\
+&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]  \nabla \left[ \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \right]  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right]\\
+&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \left[  \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)     \right]'\\
+&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&- \left[ \nabla l_m(\theta)  \right] \left[ \nabla l_m(\theta)  \right]'
+\end{align}
+
+\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}
+
+When $\tilde{f}$ is independent of $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
+\begin{align}
+\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) } \label{eq:MCLAgradient}
+\end{align}
+and the Hessian matrix of the MCLA is
+\begin{align}
+\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
+&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
+\end{align}
+
 \end{document}

### 2014-11-20 NewDesignDoc/DesignDocGLMM.tex
Updating design doc equations

@@
 
-Let $y=(y_1, \ldots, y_n)^T$ be a vector of observed data.
+Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data.
 Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects.
 Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects.
-Let $\theta$ be a vector of length $p+T$ containing all unknown parameters.
+Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters.
 Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$.
-Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write the density like this to keep notation simple in future equations.
+Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write both densities with $\theta$ to keep notation simple in future equations.
 
@@
 In these cases, performing even basic inference on the likelihood is not possible.
-Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood.
+Rather than evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood.
 Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.
+MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
 Then the Monte Carlo log likelihood approximation is
@@
 \begin{align}
-\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
+\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
 \end{align}
@@
 \begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
-&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
-\end{align}
-
-Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.
-MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
+&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}.
+\end{align}
+Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.
+
+Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.
 
@@
 
-EDITED TO HERE.
+\section{Notation}
 
@@
 \end{align}
-
-Since $U$ is unobservable, the likelihood must be expressed by integrating out the random effects:
-\begin{align}
-l(\theta)=\int f_\theta(y|u) f_\theta(u) \; du
-\end{align}
-For most datasets, this integral is intractible.
-In these cases, performing even basic inference on the likelihood is not possible.
-Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood.
-Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.
-Then the MCLA of $l(\theta)$ is:
-\begin{align}
-l_{m}(\theta) = \sum_{k=1}^m \dfrac{ f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}
-\end{align}
-Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.
-MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
-
-%
-%Let $\beta$ be a vector of $p$ fixed effect parameters.
-Let $X$ be an $n \times p$ design matrix for the fixed effects.
-Let $U$ be an unobservable normal random vector with length $q$, mean $0$ and variance matrix $D$.
-Let $Z$ be a $n \times q$ model matrix for the random effects.
-%
-%There are $T$ distinct variance components  $\nu_t$, $t = 1,$ $\ldots,$ $T$.
-For the first version of the package, $D$ is assumed diagonal.
-Let $E_t$ be a diagonal matrix with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$.
-(Learn more about $E_t$ in section \ref{sec:getEk}.) Then $D= \sum_{t=1}^T \nu_t E_t$.
-Later versions of this package will allow  more general covariance structures.
-With other structures, $D$ will be constructed by $D= \sum_{t=1}^T h_t (\nu_t) \; E_t$, but $E_t$ will no longer be diagonal and $h_t$ will be some function.
-We will have a D for distance (such as in the car theft data) and for autoregressive 1.
-%
-% Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},\ldots,\nu_T)$.
-Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple.
-Since $U$ is normally distributed, we can write its log density as
-%\begin{align}
-%\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} U' D^{-1} U
-%\end{align}
-%where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.
-%
-%Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
-%\begin{align}
-%g(\mu) = X \beta + Z U
-%\end{align}
-%The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$.
-If the data have a  Bernoulli distribution, the link is $\logit(\mu)$.
-If the data have a Poisson distribution, the link is $\log (\mu)$.
-More on these two families can be found in section \ref{sec:fam}.
-For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$.
-Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
-%\begin{align}
-%Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
-%\end{align}
-%
-%Since $U$ is unobservable, the likelihood must be expressed by integrating out the random effects:
-%\begin{align}
-%l(\theta)=\int f_\theta(y|u) f_\theta(u) \; du
-%\end{align}
-%For most datasets, this integral is intractible.
-In these cases, performing even basic inference on the likelihood is not possible.
-Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood.
-Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.
-Then the MCLA of $l(\theta)$ is:
-%\begin{align}
-%l_{m}(\theta) = \sum_{k=1}^m \dfrac{ f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}
-%\end{align}
-%Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$.
-MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
-%
 
@@
 
-\subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$}
+\subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$} \label{sec:calcs}
 When $\tilde{f}$ contains $\theta$, the gradient of the MCLA is
@@
 \end{align}
-
-\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}
+To reduce the risk of catastrophic cancellation, we can combine the last two terms of the Hessian:
+\begin{align}
+\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
+\end{align}
+We are able to combine the last two terms because $\nabla l_m(\theta)$ is a weighted mean of $\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)$. Letting
+\begin{align}
+Z&=\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)
+\end{align}
+ we can use the following equality:
+\begin{align}
+E(ZZ')-E(Z)E(Z)' = \left[ E(Z-EZ)  \right]\left[ E(Z-EZ)  \right]'.
+\end{align}
+
+\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}\label{sec:calcsindep}
 

### 2014-11-24 NewDesignDoc/DesignDocGLMM.tex
Editing, rearranging design doc

@@
 Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data.
-Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects.
-Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects.
+Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects centered at 0 with variance matrix $D$.
+Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$.
 Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters.
@@
 Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.
-MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
+MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is the mixture distribution specified in section \ref{sec:genRand}.
+
 Then the Monte Carlo log likelihood approximation is
@@
 
-%Let $y=(y_1, \ldots, y_n)^T$ be a vector of observed data.
-Let $u=(u_1,\ldots,u_q)^T$ be a vector of unobserved random effects.
-Let $\theta$ denote a vector of unknown parameters.
-The data are distributed conditionally on the random effects according to $f_\theta(y|u)$.
-
-\section{Notation}
-
- Let $X$ be an $n \times p$ design matrix for the fixed effects.
-Let $U$ be an unobservable normal random vector with length $q$, mean $0$ and variance matrix $D$.
-Let $Z$ be a $n \times q$ model matrix for the random effects.
-
-There are $T$ distinct variance components  $\nu_t$, $t = 1,$ $\ldots,$ $T$.
-For the first version of the package, $D$ is assumed diagonal.
-Let $E_t$ be a diagonal matrix with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$.
-(Learn more about $E_t$ in section \ref{sec:getEk}.) Then $D= \sum_{t=1}^T \nu_t E_t$.
-Later versions of this package will allow  more general covariance structures.
-With other structures, $D$ will be constructed by $D= \sum_{t=1}^T h_t (\nu_t) \; E_t$, but $E_t$ will no longer be diagonal and $h_t$ will be some function.
-We will have a D for distance (such as in the car theft data) and for autoregressive 1.
-
- Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},\ldots,\nu_T)$.
-Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple.
-Since $U$ is normally distributed, we can write its log density as
-\begin{align}
-\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} U' D^{-1} U
-\end{align}
-where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.
+\section{Model fitting function} 
+This is be the primary function the user  uses.
+The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}.
+Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.
+When the fixed effects are specified, the model matrix $X$  is created, where $X$ has dimensions $n \times p$.
+
+The user  specifies the random effects in the same way as for the R function \texttt{reaster} in the R package
+ \texttt{aster} \citep{aster-package}.
+That is, random effects are expressed using the R formula mini-language.
+Let $q_t$ be the number of random effects associated with variance component $\nu_t$.
+When the random effects are specified, a list of T model matrices are created and the $t$th model matrix $Z_t$ has dimensions $n \times q_t$.
+
+The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the name of the variance components.
+
+Thus, the following is a sample command with fixed predictors $x_1$ and $x_2$ and with random effects created by categorical variables $school$ and $classroom$:
+\begin{verbatim}
+glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="bernoulli.glmm", 
+data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
+debug=FALSE)
+ \end{verbatim}
+
+%$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+?
+Most models require this, so please check if you meant to include it.''
+
+It is possible that the user could want some variance components to be set equal.
+For example, \pcite{coull:agresti:2000} influenza dataset contains four years with random effects for each year.
+The authors want the within-year variance components to be equal.
+There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since the strain of flu during year 4 was a repeat of a previous year).
+Suppose \texttt{year} is a categorical variable with four levels, and \texttt{year1} through \texttt{year4} are dummy variables.
+Thus, the call  contains these arguments:
+\begin{verbatim}
+glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeat,~0+year3,~0+year4),
+ varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeat"),
+data=flu,family.glmm=bernoulli.glmm)
+\end{verbatim}
+
+As a result of this command, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices.
+Since we have 3 distinct variance components, we want 3 design matrices.
+We take the design matrices that share a variance component and \texttt{cbind} them together.
+The result is 3 model matrices in the $Z$ list, one for each variance component.
+We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  ("subject","year","repeat") are correctly assigned.
+Also, the variance estimates  come out in this same order as well.
+
+  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.
+The process of finding these estimates is detailed in section \ref{sec:pql}.
+The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.
+More information on generated the random effects is in section \ref{sec:genRand}.
+
+Next, \texttt{trust} is implemented to maximize  the MCLA objective function (details on the objective function are in section \ref{sec:objfun}).
+Finally, the function  returns parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood,  the Hessian of the log likelihood at those estimates, information from the \texttt{trust} optimization, and other information.
+
+%%Families%%
+\section{Families} \label{sec:fam}
 
@@
 \end{align}
-The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$.
+The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$.
 If the data have a  Bernoulli distribution, the link is $\logit(\mu)$.
 If the data have a Poisson distribution, the link is $\log (\mu)$.
-More on these two families can be found in section \ref{sec:fam}.
-For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$.
-Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
+Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.
+
+ For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
 \begin{align}
@@
 
-\section{Model fitting function} 
-This will be the main function the user will use.
-The users will need to specify the response and the predictors using the R formula mini-language as interpreted by model.matrix.
-They'll need to specify the  family (either binomial or Poisson, though  really any exponential family would work and I could add more later).
-The user will specify the random effects in the same way as for the R function \texttt{reaster} in the R package
- \texttt{aster} \citep{aster-package}.
-That is, random effects will be expressed using the R formula mini-language.
-Thus, a sample command with fixed predictors $x_1$ and $x_2$ and with random effects $school$ and $classroom$ (in data set as categorical variables ) would look like
-\begin{verbatim}
-glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="binomial.glmm", 
-data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
-debug=FALSE )
- \end{verbatim} 
-Section \ref{sec:fam} contains more information on the family.
-
-%$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+?
-Most models require this, so please check if you meant to include it.''
-
-It's possible that the user could want some variance components to be set equal.
-For example, in the Coull and Agresti flu problem, there are 4 years and random effects for each year.
-The authors want the within-year variance components to be equal.
-There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since year 4's virus was a repeat).
-Suppose year is a categorical variable with four levels, and year1 through year4 are dummy variables.
-Thus, the call could contain these arguments
-\begin{verbatim}
-glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeet,~0+year3,~0+year4),
- varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeet"),
-data=flu,family.glmm=bernoulli.glmm)
-\end{verbatim}
-
-Thus model.matrix will make Z into a list of 6 design matrices.
-Since we have 3 distinct variance components, we want 3 design matrices.
-What we do is we take the design matrices that share a variance component and then cbind them together.
-Thus we'll have 3 model matrices in our Z list, one for each random effect.
-We will want to put them in order (1,2,3 according to varcomps.equal) so that the names for each matrix line up ("subject","year","repeet").
-Cbinding the design matrices that share a variance component will not effect the way that we compute $\eta$.
-Also, the variance estimates should come out in this same order as well.
-
-  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.
-The process of finding these estimates is detailed in section \ref{sec:pql}.
-The PQL estimates  will help determine the means and variances for the normal distributions from which random effects will be generated.
-After  sampling random effects from a standard normal, they will be scaled and centered to have the desired mean and variance.
-More information on generated the random effects is in section \ref{sec:genRand}.
-Next, I'll implement trust on the objective function (details on the objective function are in section \ref{sec:objfun}.
-Finally, the function will return parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood, and the Hessian of the log likelihood at those estimates.
-
-%%Families%%
-\section{Families} \label{sec:fam}
-This function will be hidden from the user.  These functions (along with the distribution of random effects) are necessary to approximate the log likelihood.
-
-I'm going to have an S3 class called ``glmm.family'' with a few options (for now, binomial and poisson).
-Each family  function will output a list including the family name (a character string such as ``binomial''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.
-
-The user will provide the family in the model-fitting function.
-They can either enter the character string (``bernoulli.glmm''), the function (bernoulli.glmm()), or the result of invoking it.
-The following code for using the input to determine the family is adapted from glm.
+The user is required to specify the family in the model-fitting function.
+Once the family is specified, many family-specific functions are called.
+They are contained in an S3 class called ``glmm.family''.
+Each family function outputs a list including the family name (a character string such as ``bernoulli.glmm''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.
+
+The users provide the family in the model-fitting function by either enter the character string (``bernoulli.glmm''), the function (\texttt{bernoulli.glmm()}), or the result of invoking the function.
+The following code for using the input to determine the family is adapted from \texttt{glm}.
 
@@
 
-For example, one of the family functions will look like this:
-\begin{verbatim}
-poisson.glmm<-function()
-{
-	family.glmm<- "poisson.glmm"
-	c <- function(eta) exp(eta)
-	cp <- function(eta) exp(eta)
-	cpp<-function(eta) exp(eta)
-	out<-list(family.glmm=family.glmm, c=c,cp=cp,cpp=cpp)
-	class(out)<-"glmm.family"
-	return(out)
-}
-\end{verbatim}
-
-Then, to use these functions in order to calculate $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$, I can just call 
+With this, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
 \begin{verbatim}
@@
 
-For  Bernoulli, we calculate these values (c, cp, and cpp) with careful computer arithmetic as follows. We also use the \texttt{log1p} function in R for $c(\eta_i)$. 
+For the Bernoulli distribution, we calculate these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) with careful computer arithmetic as follows.
+We also use the \texttt{log1p} function in R for $c(\eta_i)$.
 \begin{align}
@@
 \end{align}
-We use the \texttt{log1p} function in R for $c(\eta_i)$ for the Bernoulli family. For poisson, these values (c,cp, and cpp) are
+ For poisson, these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) are
 \begin{align}
@@
 
-In the R function \texttt{glm} binomial, the user can choose the link.
+In the R function \texttt{glm}, the user can choose the link.
 The canonical link must be used in the \texttt{glmm} package so that we have an exponential family.
-I'm going to include the link in the value of \texttt{family.glmm} just in case the user doesn't know it already.
-
-Also, I'll have a function to check that the data are valid given the family type.
-So I'll make sure that if \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0's or 1's.
+The canonical link is included in \texttt{family.glmm}  in case the user does not know it already.
+
+Also, I have a function to check that the data are valid given the family type.
+If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0s or 1s.
 If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.
-If that's not true, the check returns an error message.
+If the data does not pass the check, the check returns an error message.
 
 \subsection{Redone in C}
-When I redo this in C, I will have a separate function for \texttt{cum} ,\texttt{cp}, and \texttt{cpp}.
-The inputs will be \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family), and an array of doubles to contain the result.
-These will be passed in as pointers.
-The functions will calculate  the cumulant function or one of its first two derivatives.
-Each function will contain a switch statement for the glmm family.
+Redone in C, I  have  separate functions for calculating $c(\eta)$, $c'(\eta)$, $c''(\eta)$: \texttt{cum3.c}, \texttt{cp3.c}, and \texttt{cpp3.c}.
+The inputs for each function are identical:  \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family: $1$ indicates Bernoulli and $2$ indicates Poisson), and an array of doubles to contain the result.
+These are passed in as pointers.
+The functions  calculate  the cumulant function or one of its first two derivatives.
+Each function  contains a switch statement for the glmm family.
 The calculations for each of these functions have  been shown earlier in this section.
-This function will be type void: rather than returning the cumulant or its derivatives, the pointers will be changed to contain the results.
-This function will be invoked by \texttt{el}, descriped in section \ref{sec:el}.
-
-\section{PQL}\label{sec:pql}
- Before we get started on describing PQL, we need to change notation to avoid constrained optimization.
-Recall that $D=\var(u)$ and is (for now) assumed to be diagonal.
-Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative.
-Using A rather than D enables unconstrained optimization.
-If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
-\begin{align*}
-A= \sum_t E_t \sigma_t.
-\end{align*}
-Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix.
-PQL will estimate the components contained on the diagonal of A.
-Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).
-
-In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.
-
-There are two ways to do PQL, both of which are described in the vignette
- \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014).
-In either case, there is an inner optimization and an outer optimization.
-The inner optimization is well behaved while the other optimization is a little tougher.
-I will be using the method that is not quite PQL but is pretty close and is better behaved.
-In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$.
-Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
-
-The {\bf inner} optimization will be done with the trust function in R.
-We elect to use trust because it requires two derivatives, which will make the optimization more precise.
-We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.
-
-The inner optimization maximizes the penalized log likelihood. After defining
-\begin{align}
-\eta=X\beta +ZAs
-\end{align}
-we calculate the  log likelihood as 
-\begin{align}
-l(\eta)= Y' \eta - c(\eta) 
-\end{align}
-and the penalized likelihood as:
-\begin{align}
- l(\eta)- \dfrac{1}{2} s's.
-\end{align}
-This function will be maximized using the \texttt{trust} package.
-We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$.
-We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.
-
-Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
-\begin{align}
-l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
-\end{align}
-then 
-\begin{align}
-\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i.
-\end{align}
- Now we can write the following expression:
-
-\begin{align}
-%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
-\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
-&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
-&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
-&=\sum_i (Y_i-\mu_i) X_{ik}
-\end{align}
-
-We find the function's derivative with respect to $s$ as follows:
-\begin{align}
-\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
-&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
-&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
-&=(Y-\mu)'ZA -s
-\end{align}
-
-This gives us the following derivatives of the penalized log likelihood:
-\begin{align}
-\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
-\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
-\end{align}
-
-Lastly, we need the Hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
-\begin{enumerate}
-\item $ \dfrac{\partial^2}{\partial s^2}$
-\item $ \dfrac{\partial^2}{\partial \beta^2}$
-\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
-\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
-\end{enumerate}
-
-We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
-\begin{align}
-  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
-&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
-&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
-&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
-&= -AZ' \left[ c''(\eta) \right] ZA - I_q
-\end{align}
-Note that $c''(\eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.
-$I_q$ is the identity matrix of dimension $q$.
-This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.
-
-Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
-\begin{align}
-  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
-&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
-&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
-&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
-&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
-&=-X'  \left[ c''(\eta)\right]  X   
-\end{align}
-
-Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
-\begin{align}
-\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
-&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
-&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
-&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
-&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
-&=-   \left[ c''(\eta) X  \right]' ZA   \\
-&=-X'   \left[ c''(\eta)   \right] ZA  
-\end{align}
-
-And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood.
-Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
-Trust will maximize the penalized likelihood.
-
-The {\bf outer} optimization is done using \texttt{optim} with the default method of ``\texttt{Nelder-Mead.}'' This requires just the function value and no derivatives.
-This optimization method was chosen because the optimization function already contains second derivatives of the cumulant function; requiring derivatives of the optimization function would in turn require higher-order derivatives of the cumulant.
-
-The default of \texttt{optim} is to minimize, but we'd like to  do maximization. Reversing the sign of the optimization function will turn the maximization into minimization.
-
-If $\tilde{\beta}$ and $\tilde{s}$ are available from previous calls to the inner optimization function, then they are used here.
-Otherwise, the values are taken to be 0 and 1.
-The outer optimization's function is evaluated by first defining
-\begin{align}
-\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
-l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
-A&= \sum_k E_k \sigma_k.
-\end{align}
-Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
-\begin{align}
- l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
-\end{align}
-Again, \texttt{optim} minimizes, so we have to minimize the negative value of the penalized quasi-likelihood in order to maximize the value of it.
-
-We do need to be careful about the determinant. Let $AZ' W ZA +I= LL'$ where $L$ is the lower triangular matrix resulting from a Cholesky decomposition. Then
-\begin{align}
-\dfrac{1}{2} \log | AZ' W ZA +I | &= \dfrac{1}{2} \log |LL'|\\
-&=\dfrac{1}{2} \log (|L|)^2 \\
-&= \log |L|
-\end{align}
-Since $L$ is triangular, the determinant is just the product of the diagonal elements. Let $l_i$ be the diagonal elements of $L$. Then
-\begin{align}
-\dfrac{1}{2} \log | AZ' W ZA +I | &= \log \prod l_i \\
-&= \sum \log l_i
-\end{align}
-
- If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
-This would be more numerically stable, but is slower.
-Let $O$ be the matrix containing the eigenvectors and let $\Lambda$ be a diagonal matrix with the eigenvalues $\lambda_i$ on the diagonal.
-Then
-\begin{align}
- &AZ' W ZA = O \Lambda O' 
-\end{align}
-Then we can rewrite the argument of the determinant as follows:
-\begin{align}
-& AZ' W ZA +I  = O \Lambda O' + I = O \Lambda O + OO'=O (I+\Lambda) O' 
-\end{align}
-This leads to the careful calculation of our determinant as follows:
-\begin{align}
-&   \left| AZ' W ZA +I \right| = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
-&\Rightarrow \log   | AZ' W ZA +I | = \sum \log(1+\lambda_i) 
-\end{align}
-The last quantity can be accurately calculated using the \texttt{log1p} function in R.
-
-The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them.
-To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
-First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.
-Second, after PQL is finished, $\tilde{s}$ is used to help center the generated random effects.
-
-The point of doing PQL is to construct a decent importance sampling distribution.
-Thus, the estimates don't have to be perfect.
-It is possible that one of the $\sigma_k$ will be 0 according to PQL.
-If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.
+This function is type void: rather than returning the cumulant or its derivatives, the pointers are changed to contain the results.
+This function is invoked by \texttt{el}, described in section \ref{sec:el}.
 
@@
 \subsection{Redone in C}
-The C function to calculate the value of the log data density and its two derivatives will be called by reference.
-The following pointers will be passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}.
-The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} will be zeros before \texttt{el.C} is invoked.
-Invoking \texttt{el.C} will then place the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\
-
-The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.
+The C function to calculate the value of the log data density and its two derivatives are called by reference.
+The following pointers are passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}.
+The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} are zeros before \texttt{el.C} is invoked.
+Invoking \texttt{el.C}  then places the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\
+
+The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.
+
+\section{Random effect generation and calculations}
+This section is focused on a vector of random effects $u$ with length $q$.
+We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$.
+Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit.
+Section \ref{sec:genRand} describes the process of generating the random effects.
+Section \ref{sec:distRandsec} explains how to evaluate the distribution of the random effects and its first two derivatives.
 
 %%getEk
-\section{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
-Since $D$ is a function of $E_t$, we need to find a way to calculate $E_t$. This depends on whether $D$ is diagonal or not.
-
-\subsection{D diagonal}
-When $D$ is diagonal, recall $E_t$ are diagonal with either 1 or 0 on the diagonal (1 if that random effect has $\nu_t$ as a variance component and 0 otherwise).
-Let $q_t$ be the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$.
-When D is diagonal, then $\sum_{t=1}^T E_t = I_q$, the identity matrix with $q$ rows.
-Finally, recall that $D=\sum_{t=1}^T \nu_t E_t$.
-
-At this point,  \texttt{mod.mcml\$z} is a list with $T$ design matrices (one for each distinct variance component).
-We need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
- Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have that many rows (and columns).
-Thus, each $E_t$ will have q rows and columns as well.
+\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
+For this version of the package, we that assume that $D$ is diagonal.
+Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$.
+That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as a variance component.
+Then $D=\sum_{t=1}^T\nu_t  E_t $.
+
+If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.
+The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).
+In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
+ Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have $q$ rows (and columns).
+Thus, each $E_t$ will have $q$ rows and columns as well.
 Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
@@
 
-\subsection{D in general (not necessarily diagonal)}
-In progress.
-
-\section{Generating random effects (genRand)}\label{sec:genRand}
-
-This will be hidden from the user. It will be called by the model fitting function.
-
-To approximate the log likelihood, we need simulated random effects.
-PQL will give us estimates for $\beta$, $\sigma$ and  $s$.
-Let $\beta^*$, $s^*$ and $\sigma^*$ be the vectors of PQL estimates.
-Then
+\subsection{Generating random effects (genRand)}\label{sec:genRand}
+The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations.
+Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
+That is, $u= D^{1/2}s$.
+Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.
+The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$.
+Then let
 \begin{align}
@@
 \end{align}
-Also, we can ``unstandardize'' our random effects this way:
-\begin{align}
-u^*&=A^*s^*
-\end{align}
-
-I'll be generating random effects $u$ from these distributions:
+Also, we can ``unstandardize'' our random effects:
+\begin{align}
+u^*&=A^*s^*.
+\end{align}
+
+Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$.
+Let $D$ be determined by the current iteration of \texttt{trust}.
+The importance sampling distribution $\tilde{f}$ is the following mixture distribution:
+\begin{align}
+ \tilde{f}(u) = p_1  f(u \, | \, 0, D)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
+\end{align}
+
+Therefore, the random effects are generated from the following distributions:
 \begin{enumerate}
@@
 \item $N(u^*,D^*)$
-\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$
+\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
 \end{enumerate}
-I'll generate them in these proportions: $p_1,p_2,p_3$ where $p_1+p_2+p_3=1$.
-
-The first distribution is centered at 0, just where we expect the random effects to truly be centered.
-The random effects will be generated with variance 1, but will be scaled in the trust objective function so that they have variance $\nu$, where $\nu$ is given by the iteration of trust.
-The second distribution is centered at the PQL best guess of the random effect values and  has the PQL guess of the variance.
-The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood hessian.
-The idea of this last distribution is to generate random effects from a distribution whose first two derivatives are equal to those of the target distribution $f_\theta(u,y)$.
-
-Let $u_k$ be a vector with length q (the number of random effects in the model).
-Suppose we want to generate random effects $u_k,k=1,...,m$ from a generic distribution $N(\mu,\Sigma)$.
-We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ with diagonal entries being $\Sigma$'s eigenvalues $\lambda$.
-Eigendecompositions take a little bit more time than Cholesky decompositions, but are more stable.
+
+The first distribution is the multivariate standard normal, but the random effects drawn from this are scaled in the trust objective function so that their variance matrix $D$ is determined by the current iteration of trust.
+The second distribution is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
+The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood Hessian.
+The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.
+
+The random effects are generated by shifting and scaling a sample from a  standard normal distribution.
+Suppose we want to generate a sample from a general normal distribution $N(\mu,\Sigma)$.
+Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.
+For each k, draw a sample of size $q$ from the standard normal distribution.
+Call this sample $\breve{u}_k$.
+Then shift and scale it:
+\begin{align}
+u_k = \breve{u}_k \Sigma^{1/2} + \mu.
+\end{align}
+so that is a sample from $N(\mu,\Sigma)$.
+
+One way to find $\Sigma^{1/2}$ is through eigendecomposition.
+Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.
+We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).
 Then
 \begin{align}
-\Sigma^{1/2}= O \Lambda^{1/2} O'
+\Sigma^{1/2}= O \Lambda^{1/2} O'.
 \end{align}
@@
 
-For each k, draw a sample of size q from the standard normal distribution. Call this sample $\breve{u}_k$. Then we shift and scale it:
-\begin{align}
-u_k = \breve{u}_k \Sigma^{1/2} + \mu.
-\end{align}
-This is the process of drawing our sample from $N(\mu,\Sigma^{1/2}\Sigma^{1/2})$ for any mean vector $\mu$ and variance matrix $\Sigma$.
-
-%%distribution of the random effects%%
-\section{Distribution of the random effects }
-This will be hidden from the user.
+%
+%%%distribution of the random effects%%
+%\subsection{Distribution of the random effects }\label{sec:distRandsec}
+%This will be hidden from the user.
 This section provides equations for evaluating the log of a normal distribution with a specified mean and variance.
 When necessary to emphasize and clarify the choice of mean $\mu$ and variance $\Sigma$ , I use the notation $ f(u|\mu,\Sigma)$ rather than $f_\theta(u)$.
-
-This section is useful for calculating $\tilde{f}(u_k)$.
+%
+%This section is useful for calculating $\tilde{f}(u_k)$.
 This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives.
 These are needed for the calculation in \eqref{eq:MCLA}.
-
-Section \ref{sec:tee} contains details that are not currently implemented, but could be.
+%
+%Section \ref{sec:tee} contains details that are not currently implemented, but could be.
 In this version, all random effects are assumed to be normally distributed and all simulated random effects are generated from a normal distribution.
@@
 \subsection{Distribution of random effects is normal (\texttt{distRand})}\label{sec:distRand}
-In this subsection, I discuss both the assumed distribution of the unobserved random effects ($N(0,D)$) and the distribution used to generate the simulated random effects.
-The equations in this section will provide $\tilde{f}(u)$, $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$ for equation \ref{eq:MCLA}.
+%
+%  Recall that the first version of the package assumes $D$ is  diagonal. Since $U$ is normally distributed, we can write its log density as
+%\begin{align}
+%\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} u' D^{-1} u
+%\end{align}
+%where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.
+%
+
+In this subsection, I discuss both the assumed distribution of the unobserved random effects $N(0,D)$ and the importance sampling distribution used to generate the simulated random effects.
+The equations in this section provide  $\log f_\theta(u)$, $\nabla \log f_\theta(u)$,  $\nabla^2 \log f_\theta(u)$, $\tilde{f}(u)$, $ \nabla \tilde{f}(u)$, and $\nabla^2 \tilde{f}(u)$  for equation \ref{eq:MCLA}.
 
 We assume the random effects are $N(\mu, \Sigma)$.
-For $f_\theta(u)$, $\mu=0$.
-This is written generally with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.
-Let $\Sigma$ be the variance matrix.
-We can then write the distribution of the random effects as:
+This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.
+We can  write the distribution of the random effects as:
 \begin{align}
@@
 
-For now, I  assume that the true variance of the random effects is $\Sigma=D$ where D is diagonal.
+For now, I  assume that  D is diagonal.
 Formulas for this are in section \ref{sec:Ddiag}.
@@
 
-\subsection{$\tilde{f}(u_k)$}\label{sec:ftwiddle}
-I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust.
+%\subsection{$\tilde{f}(u_k)$}\label{sec:ftwiddle}
+%I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust.
 I can make a function using the general form for a normal distribution's density as described in the previous subsection, then use that function for the three distributions that I sampled from.
 Since I generated the random effects from the three distributions in proportions $p_1,p_2,p_3$, I can write
-\begin{align}
-\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
-\end{align}
-
-\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
-If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL).
+%\begin{align}
+%\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
+%\end{align}
+%
+%\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
+%If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL).
 These can be drawn using the package mvtnorm.
@@
 I'm not sure if I'll use this or just stick to normals.
-
-\subsection{Central Limit Theorem for MCLA}
-Define
-\begin{align}
-\gamma_1&= \int f_\theta(u,y) du\\
-\gamma_2 &=\int \tilde{f}(u)du.
-\end{align}
-Recall the calculation for the MCLA gradient in  \eqref{eq:MCLAgradient}. We can rewrite it as
-\begin{align}
-\nabla l_m(\theta)=  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
-\end{align}
-to more obviously show that both the numerator and denominator are sample means. 
-The law of large numbers says that the numerator and denominator  each converge to their true means. That is,
-\begin{align}
-\dfrac{1}{m}\sum_{k=1}^m \nabla \log f_\theta (u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-&= \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
-&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right]
-\end{align}
-and
-\begin{align}
-\dfrac{1}{m}\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow E_{\tilde{f}} \left[ \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-&= \int \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-&= \dfrac{\gamma_1}{\gamma_2} \int \dfrac{ f_\theta(u,y)}{\gamma_2} du \\
-&= \dfrac{\gamma_1}{\gamma_2}
-\end{align}
-Then, by Slutsky's,
-\begin{align}
-\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
-&=  E_f \left[ \nabla \log f_\theta(u,y)  \right]
-\end{align}
-In addition to a law of large numbers, we would like a Central Limit Theorem for $\nabla l_m(\theta)$. In other words, the quantity
-\begin{align}
-&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- E_f \left[ \nabla \log f_\theta(u,y)  \right]  \right] \\
-&=
-\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m} E_f \left[ \nabla \log f_\theta(u,y)  \right]  \\
-&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)
- - E_f \left[ \nabla \log f_\theta(u,y)  \right] \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
-\end{align}
-will have a normal distribution if and only if the variances of
-\begin{align}
-\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
-and
-\begin{align}
-\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
-\end{align}
-are each finite.
-First, we want to show that
-\begin{align}
-\sum_{k=1}^m \dfrac{f_\theta(u_k)}{\tilde{f}(u_k)}
-\end{align}
-has finite variance. This is true if and only if 
-\begin{align}
-\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du < \infty.
-\end{align}
-We see:
-\begin{align}
-\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du &= \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
-& \leq \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)} du \\
-&= \dfrac{1}{p_1} \int N(u|0,D) du \\
-&= \dfrac{1}{p_1}.
-\end{align}
-Therefore $\sum_{k=1}^m f_\theta(u_k)$ has finite variance as long as $p_1\neq 0$.
-
-Second, we want to show that the gradient of the MCLA has finite variance. That is, we want to show that
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty.
-\end{align}
-That is, we want to show 
-\begin{align}
-\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
-&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
-&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
-&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du  \\
-&< \infty.
-\end{align}
-By the Cauchy-Schwartz inequality, 
-\begin{align}
-  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
-& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
-\end{align}
-Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
-\begin{align}
-\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
-\end{align}
-Recall the notation from section \ref{sec:Ddiag}.  If
-\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
-\end{align}
-for every $t=1,...,T$, then
-\begin{align}
-\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
-\end{align}
-by Cauchy Schwartz.
-Therefore, I will prove the integral is finite for the general case of $u_t$ and $\nu_t$.
-First, we recognize that our integral is actually an expectation of a function of a normal random variable:
-\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
-\end{align}
-Normal distributions have finite moments of all orders, so we know that\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
-\end{align}
-
-%
-%\begin{align}
-%&=
-%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ \dfrac{q_t^2}{4 \nu_t^2} + \dfrac{(u_t'u_t)^2}{4 \nu_t^4} -2\dfrac{ q_t u_t'u_t  }{4 \nu_t^3} \right] du_t \\
-%&=
-%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       \dfrac{q_t^2}{4 \nu_t^2} du_t +
-% \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{(u_t'u_t)^2}{4 \nu_t^4} du_t \nonumber \\
-%&-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{ q_t u_t'u_t  }{2 \nu_t^3}     du_t \\
-%&=
-% \dfrac{q_t^2}{4 \nu_t^2} \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       du_t +
-%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
-%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t \\
-%&=
-% \dfrac{q_t^2}{4 \nu_t}+
-%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
-%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t 
-%\end{align}
-%
-%Define $erf$ as the ``error function,'' the function from integrating a normal distribution. Evaluating the integral and using the fact that $erf$ is bounded between $-1$ and $1$, we see 
-%\begin{align}
-%\int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t
-%&\propto
-%    \nu^{3/2} erf \left(\dfrac{u_t}{\sqrt{2 \nu_t} }   \right ) - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)}   \\
-%& \geq -  \nu^{3/2}  - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)} \\
-%& \geq - \infty
-%\end{align}
-%for all $u_t$ and for all $\nu_t.$ Lastly, we need to make sure 
-% \begin{align}
-% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t < \infty
-%\end{align}
-%Let $c_1$, $c_2$, and $c_3$  be constants with respect to $u_t$. Then 
-%\begin{align}
-% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t &= c_1 erf \left( \dfrac{u_t}{\sqrt{2 \nu_t}} \right) - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
-%& \leq  c_1  - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
-%& < \infty
-%\end{align}
-%since
-%\begin{align}
-%e^{-u_t'u_t / (2 \nu_t)} > \infty
-%\end{align}
-%for all $u_t$ and all $\nu_t$.
-
-Next, we want to show 
-\begin{align}
-\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
-\end{align}
- Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
-\begin{align}
-\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
-\end{align}
-We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
-\begin{align}
-\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-\end{align}
-is finite.
-When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
-\begin{align}
-\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
-\end{align}
-Then
-\begin{align}
-\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
-&\propto \int N(u|0,D) e^{2Zu} du 
-\end{align}
-is finite.
-
-Therefore, it is proven that 
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
-\end{align}
-exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.
-
-%
-%Next, we need to show that the hessian of the MCLA has finite variance. Again, it is enough to show that
-%\begin{align}
-%\int \dfrac{N(u|0,D)^2 h(u)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty
-%\end{align}
-
-%\subsection{Boundedness of importance weights}
-%We want to make sure that the importance weights will be bounded when using this distribution.
-The importance weight for a vector $u_k$ is $\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)}$.
-Since $f_\theta(y,u_k)< \infty$,
-%\begin{align}
-%\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} < \infty \Longleftrightarrow \tilde{f}(u_k) >0.
-%\end{align}
-%
-%We start with
-%\begin{align}
-% \tilde{f}(u_k) &= p_1  f(u_k \, | \, 0, D)+p_2  f(u_k \, | \, u^*, D^*)+p_3  f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\\
-%&\geq p_1  f(u_k \, | \, 0, D) \\
-%& \geq p_1 (2 \pi)^{-q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u}
-%\end{align}
-%In order to show that $\tilde{f}(u_k)>0$, we need to show that both $ |(D^*)^{-1/2}|$ and  $e^{- u' (D^{*})^{-1}u}$ are positive. 
-%
-% First, I will show that $ |(D^*)^{-1/2}|>0$.Since  $(D^*)^{-1/2}$ is invertible, its determinant must be nonzero.
-Since $D^*$ is positive semi-definite, then $(D^*)^{-1/2}$ is also positive semi-definite.
-Thus the eigenvalues of $(D^*)^{-1/2}$ are neither zero nor negative; they must be positive.
-An eigendecomposition tells us $|(D^*)^{-1/2}|$ is the product of its eigenvalues.
-Therefore, $|(D^*)^{-1/2}| >0$.
-%
-%Next, I will show that $e^{- u' (D^{*})^{-1}u}>0$.  Since $D^*$ is a covariance matrix, it is positive semi-definite.  By definition, for any vector $a_1$, 
-%\begin{align}
-%a_1' D^* a_1 \geq 0
-%\end{align}
-%For every vector $a_2$, there exists an $a_1$ such that $a_2=D^*a_1$. Then
-%\begin{align}
-%a_2' (D^*)^{-1} a_2 &= (D^* a_1)' (D^*)^{-1} D^* a_1\\
-%&= a_1' D^* (D^*)^{-1} D^* a_1\\
-%&= a_1' D^* a_1\\
-%& \geq 0
-%\end{align}
-%Therefore,   $(D^*)^{-1}$ is also positive semi-definite. Then for any vector of random effects $u$,
-%\begin{align}
-% u' (D^*)^{-1} u \geq 0 
-%\end{align}
-%This means that
-%\begin{align}   e^{- u' (D^{*})^{-1}u} >0
-%\end{align}
-%Therefore,
-%\begin{align}
-% \tilde{f}(u_k) &\geq p_1 (2 \pi)^{q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u} >0
-%\end{align}
-%
 
@@
 I say this is a last resort  because having two Monte Carlo calculations will take a long time.
+
+\section{PQL}\label{sec:pql}
+ Before we get started on describing PQL, we need to change notation to avoid constrained optimization.
+Recall that $D=\var(u)$ and is (for now) assumed to be diagonal.
+Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative.
+Using A rather than D enables unconstrained optimization.
+If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
+\begin{align*}
+A= \sum_t E_t \sigma_t.
+\end{align*}
+Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix.
+PQL will estimate the components contained on the diagonal of A.
+Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).
+
+In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.
+
+There are two ways to do PQL, both of which are described in the vignette
+ \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014).
+In either case, there is an inner optimization and an outer optimization.
+The inner optimization is well behaved while the other optimization is a little tougher.
+I will be using the method that is not quite PQL but is pretty close and is better behaved.
+In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$.
+Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
+
+The {\bf inner} optimization will be done with the trust function in R.
+We elect to use trust because it requires two derivatives, which will make the optimization more precise.
+We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.
+
+The inner optimization maximizes the penalized log likelihood. After defining
+\begin{align}
+\eta=X\beta +ZAs
+\end{align}
+we calculate the  log likelihood as 
+\begin{align}
+l(\eta)= Y' \eta - c(\eta) 
+\end{align}
+and the penalized likelihood as:
+\begin{align}
+ l(\eta)- \dfrac{1}{2} s's.
+\end{align}
+This function will be maximized using the \texttt{trust} package.
+We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$.
+We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.
+
+Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
+\begin{align}
+l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
+\end{align}
+then 
+\begin{align}
+\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i.
+\end{align}
+ Now we can write the following expression:
+
+\begin{align}
+%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
+\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
+&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
+&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
+&=\sum_i (Y_i-\mu_i) X_{ik}
+\end{align}
+
+We find the function's derivative with respect to $s$ as follows:
+\begin{align}
+\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
+&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
+&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
+&=(Y-\mu)'ZA -s
+\end{align}
+
+This gives us the following derivatives of the penalized log likelihood:
+\begin{align}
+\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
+\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
+\end{align}
+
+Lastly, we need the Hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
+\begin{enumerate}
+\item $ \dfrac{\partial^2}{\partial s^2}$
+\item $ \dfrac{\partial^2}{\partial \beta^2}$
+\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
+\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
+\end{enumerate}
+
+We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
+\begin{align}
+  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
+&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
+&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
+&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
+&= -AZ' \left[ c''(\eta) \right] ZA - I_q
+\end{align}
+Note that $c''(\eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.
+$I_q$ is the identity matrix of dimension $q$.
+This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.
+
+Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
+\begin{align}
+  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
+&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
+&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
+&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
+&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
+&=-X'  \left[ c''(\eta)\right]  X   
+\end{align}
+
+Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
+\begin{align}
+\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
+&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
+&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
+&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
+&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
+&=-   \left[ c''(\eta) X  \right]' ZA   \\
+&=-X'   \left[ c''(\eta)   \right] ZA  
+\end{align}
+
+And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood.
+Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
+Trust will maximize the penalized likelihood.
+
+The {\bf outer} optimization is done using \texttt{optim} with the default method of ``\texttt{Nelder-Mead.}'' This requires just the function value and no derivatives.
+This optimization method was chosen because the optimization function already contains second derivatives of the cumulant function; requiring derivatives of the optimization function would in turn require higher-order derivatives of the cumulant.
+
+The default of \texttt{optim} is to minimize, but we'd like to  do maximization. Reversing the sign of the optimization function will turn the maximization into minimization.
+
+If $\tilde{\beta}$ and $\tilde{s}$ are available from previous calls to the inner optimization function, then they are used here.
+Otherwise, the values are taken to be 0 and 1.
+The outer optimization's function is evaluated by first defining
+\begin{align}
+\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
+l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
+A&= \sum_k E_k \sigma_k.
+\end{align}
+Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
+\begin{align}
+ l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
+\end{align}
+Again, \texttt{optim} minimizes, so we have to minimize the negative value of the penalized quasi-likelihood in order to maximize the value of it.
+
+We do need to be careful about the determinant. Let $AZ' W ZA +I= LL'$ where $L$ is the lower triangular matrix resulting from a Cholesky decomposition. Then
+\begin{align}
+\dfrac{1}{2} \log | AZ' W ZA +I | &= \dfrac{1}{2} \log |LL'|\\
+&=\dfrac{1}{2} \log (|L|)^2 \\
+&= \log |L|
+\end{align}
+Since $L$ is triangular, the determinant is just the product of the diagonal elements. Let $l_i$ be the diagonal elements of $L$. Then
+\begin{align}
+\dfrac{1}{2} \log | AZ' W ZA +I | &= \log \prod l_i \\
+&= \sum \log l_i
+\end{align}
+
+ If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
+This would be more numerically stable, but is slower.
+Let $O$ be the matrix containing the eigenvectors and let $\Lambda$ be a diagonal matrix with the eigenvalues $\lambda_i$ on the diagonal.
+Then
+\begin{align}
+ &AZ' W ZA = O \Lambda O' 
+\end{align}
+Then we can rewrite the argument of the determinant as follows:
+\begin{align}
+& AZ' W ZA +I  = O \Lambda O' + I = O \Lambda O + OO'=O (I+\Lambda) O' 
+\end{align}
+This leads to the careful calculation of our determinant as follows:
+\begin{align}
+&   \left| AZ' W ZA +I \right| = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
+&\Rightarrow \log   | AZ' W ZA +I | = \sum \log(1+\lambda_i) 
+\end{align}
+The last quantity can be accurately calculated using the \texttt{log1p} function in R.
+
+The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them.
+To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
+First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.
+Second, after PQL is finished, $\tilde{s}$ is used to help center the generated random effects.
+
+The point of doing PQL is to construct a decent importance sampling distribution.
+Thus, the estimates don't have to be perfect.
+It is possible that one of the $\sigma_k$ will be 0 according to PQL.
+If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.
 
@@
 
+\section{Central Limit Theorem for MCLA}
+Define
+\begin{align}
+\gamma_1&= \int f_\theta(u,y) du\\
+\gamma_2 &=\int \tilde{f}(u)du.
+\end{align}
+Recall the calculation for the MCLA gradient in  \eqref{eq:MCLAgradient}. We can rewrite it as
+\begin{align}
+\nabla l_m(\theta)=  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
+\end{align}
+to more obviously show that both the numerator and denominator are sample means. 
+The law of large numbers says that the numerator and denominator  each converge to their true means. That is,
+\begin{align}
+\dfrac{1}{m}\sum_{k=1}^m \nabla \log f_\theta (u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
+&= \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
+&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
+&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right]
+\end{align}
+and
+\begin{align}
+\dfrac{1}{m}\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow E_{\tilde{f}} \left[ \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
+&= \int \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
+&= \dfrac{\gamma_1}{\gamma_2} \int \dfrac{ f_\theta(u,y)}{\gamma_2} du \\
+&= \dfrac{\gamma_1}{\gamma_2}
+\end{align}
+Then, by Slutsky's,
+\begin{align}
+\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
+&=  E_f \left[ \nabla \log f_\theta(u,y)  \right]
+\end{align}
+In addition to a law of large numbers, we would like a Central Limit Theorem for $\nabla l_m(\theta)$. In other words, the quantity
+\begin{align}
+&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- E_f \left[ \nabla \log f_\theta(u,y)  \right]  \right] \\
+&=
+\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m} E_f \left[ \nabla \log f_\theta(u,y)  \right]  \\
+&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)
+ - E_f \left[ \nabla \log f_\theta(u,y)  \right] \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
+\end{align}
+will have a normal distribution if and only if the variances of
+\begin{align}
+\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+and
+\begin{align}
+\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
+\end{align}
+are each finite.
+First, we want to show that
+\begin{align}
+\sum_{k=1}^m \dfrac{f_\theta(u_k)}{\tilde{f}(u_k)}
+\end{align}
+has finite variance. This is true if and only if 
+\begin{align}
+\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du < \infty.
+\end{align}
+We see:
+\begin{align}
+\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du &= \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
+& \leq \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)} du \\
+&= \dfrac{1}{p_1} \int N(u|0,D) du \\
+&= \dfrac{1}{p_1}.
+\end{align}
+Therefore $\sum_{k=1}^m f_\theta(u_k)$ has finite variance as long as $p_1\neq 0$.
+
+Second, we want to show that the gradient of the MCLA has finite variance. That is, we want to show that
+\begin{align}
+\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty.
+\end{align}
+That is, we want to show 
+\begin{align}
+\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
+&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
+&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
+&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du  \\
+&< \infty.
+\end{align}
+By the Cauchy-Schwartz inequality, 
+\begin{align}
+  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
+& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
+\end{align}
+Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
+\begin{align}
+\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
+\end{align}
+Recall the notation from section \ref{sec:Ddiag}.  If
+\begin{align}
+\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
+\end{align}
+for every $t=1,...,T$, then
+\begin{align}
+\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
+\end{align}
+by Cauchy Schwartz.
+Therefore, I will prove the integral is finite for the general case of $u_t$ and $\nu_t$.
+First, we recognize that our integral is actually an expectation of a function of a normal random variable:
+\begin{align}
+\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
+\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
+\end{align}
+Normal distributions have finite moments of all orders, so we know that\begin{align}
+\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
+\end{align}
+
+%
+%\begin{align}
+%&=
+%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ \dfrac{q_t^2}{4 \nu_t^2} + \dfrac{(u_t'u_t)^2}{4 \nu_t^4} -2\dfrac{ q_t u_t'u_t  }{4 \nu_t^3} \right] du_t \\
+%&=
+%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       \dfrac{q_t^2}{4 \nu_t^2} du_t +
+% \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{(u_t'u_t)^2}{4 \nu_t^4} du_t \nonumber \\
+%&-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{ q_t u_t'u_t  }{2 \nu_t^3}     du_t \\
+%&=
+% \dfrac{q_t^2}{4 \nu_t^2} \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       du_t +
+%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
+%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t \\
+%&=
+% \dfrac{q_t^2}{4 \nu_t}+
+%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
+%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t 
+%\end{align}
+%
+%Define $erf$ as the ``error function,'' the function from integrating a normal distribution. Evaluating the integral and using the fact that $erf$ is bounded between $-1$ and $1$, we see 
+%\begin{align}
+%\int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t
+%&\propto
+%    \nu^{3/2} erf \left(\dfrac{u_t}{\sqrt{2 \nu_t} }   \right ) - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)}   \\
+%& \geq -  \nu^{3/2}  - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)} \\
+%& \geq - \infty
+%\end{align}
+%for all $u_t$ and for all $\nu_t.$ Lastly, we need to make sure 
+% \begin{align}
+% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t < \infty
+%\end{align}
+%Let $c_1$, $c_2$, and $c_3$  be constants with respect to $u_t$. Then 
+%\begin{align}
+% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t &= c_1 erf \left( \dfrac{u_t}{\sqrt{2 \nu_t}} \right) - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
+%& \leq  c_1  - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
+%& < \infty
+%\end{align}
+%since
+%\begin{align}
+%e^{-u_t'u_t / (2 \nu_t)} > \infty
+%\end{align}
+%for all $u_t$ and all $\nu_t$.
+
+Next, we want to show 
+\begin{align}
+\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
+\end{align}
+ Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
+\begin{align}
+\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
+%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
+\end{align}
+We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
+\begin{align}
+\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
+\end{align}
+is finite.
+When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
+\begin{align}
+\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
+\end{align}
+Then
+\begin{align}
+\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
+&\propto \int N(u|0,D) e^{2Zu} du 
+\end{align}
+is finite.
+
+Therefore, it is proven that 
+\begin{align}
+\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
+\end{align}
+exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.
+
+%
+%Next, we need to show that the hessian of the MCLA has finite variance. Again, it is enough to show that
+%\begin{align}
+%\int \dfrac{N(u|0,D)^2 h(u)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty
+%\end{align}
+
+%\subsection{Boundedness of importance weights}
+%We want to make sure that the importance weights will be bounded when using this distribution.
+The importance weight for a vector $u_k$ is $\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)}$.
+Since $f_\theta(y,u_k)< \infty$,
+%\begin{align}
+%\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} < \infty \Longleftrightarrow \tilde{f}(u_k) >0.
+%\end{align}
+%
+%We start with
+%\begin{align}
+% \tilde{f}(u_k) &= p_1  f(u_k \, | \, 0, D)+p_2  f(u_k \, | \, u^*, D^*)+p_3  f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\\
+%&\geq p_1  f(u_k \, | \, 0, D) \\
+%& \geq p_1 (2 \pi)^{-q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u}
+%\end{align}
+%In order to show that $\tilde{f}(u_k)>0$, we need to show that both $ |(D^*)^{-1/2}|$ and  $e^{- u' (D^{*})^{-1}u}$ are positive. 
+%
+% First, I will show that $ |(D^*)^{-1/2}|>0$.Since  $(D^*)^{-1/2}$ is invertible, its determinant must be nonzero.
+Since $D^*$ is positive semi-definite, then $(D^*)^{-1/2}$ is also positive semi-definite.
+Thus the eigenvalues of $(D^*)^{-1/2}$ are neither zero nor negative; they must be positive.
+An eigendecomposition tells us $|(D^*)^{-1/2}|$ is the product of its eigenvalues.
+Therefore, $|(D^*)^{-1/2}| >0$.
+%
+%Next, I will show that $e^{- u' (D^{*})^{-1}u}>0$.  Since $D^*$ is a covariance matrix, it is positive semi-definite.  By definition, for any vector $a_1$, 
+%\begin{align}
+%a_1' D^* a_1 \geq 0
+%\end{align}
+%For every vector $a_2$, there exists an $a_1$ such that $a_2=D^*a_1$. Then
+%\begin{align}
+%a_2' (D^*)^{-1} a_2 &= (D^* a_1)' (D^*)^{-1} D^* a_1\\
+%&= a_1' D^* (D^*)^{-1} D^* a_1\\
+%&= a_1' D^* a_1\\
+%& \geq 0
+%\end{align}
+%Therefore,   $(D^*)^{-1}$ is also positive semi-definite. Then for any vector of random effects $u$,
+%\begin{align}
+% u' (D^*)^{-1} u \geq 0 
+%\end{align}
+%This means that
+%\begin{align}   e^{- u' (D^{*})^{-1}u} >0
+%\end{align}
+%Therefore,
+%\begin{align}
+% \tilde{f}(u_k) &\geq p_1 (2 \pi)^{q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u} >0
+%\end{align}
+%
+
 \end{document}

### 2014-11-24 NewDesignDoc/DesignDocGLMM.tex
Editing design doc

@@
 l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
-&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
+&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
 \end{align}
@@
    \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}.
+   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}. \label{eq:MCLAhessian}
 \end{align}
@@
 Section \ref{sec:genRand} describes the process of generating the random effects.
-Section \ref{sec:distRandsec} explains how to evaluate the distribution of the random effects and its first two derivatives.
+Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.
 
@@
 In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
- Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have $q$ rows (and columns).
-Thus, each $E_t$ will have $q$ rows and columns as well.
+ Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  has $q$ rows (and columns).
+Thus, each $E_t$  has $q$ rows and columns as well.
 Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
@@
 \begin{align}
- D^* =A^*A^*.
-\end{align}
-Also, we can ``unstandardize'' our random effects:
+ D^* =A^*A^*
+\end{align}
+be matrices based on PQL estimates. Also, we can ``unstandardize'' our PQL-based random effects:
 \begin{align}
@@
 \begin{align}
- \tilde{f}(u) = p_1  f(u \, | \, 0, D)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
+ \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
 \end{align}
@@
 Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.
-For each k, draw a sample of size $q$ from the standard normal distribution.
+For each $k$, $k=1,\ldots,m$, draw a sample of size $q$ from the standard normal distribution.
 Call this sample $\breve{u}_k$.
@@
 \begin{align}
-u_k = \breve{u}_k \Sigma^{1/2} + \mu.
-\end{align}
-so that is a sample from $N(\mu,\Sigma)$.
-
+u_k = \breve{u}_k \Sigma^{1/2} + \mu
+\end{align}
+so that it is a sample from $N(\mu,\Sigma)$.
+
+This method requires $\Sigma^{1/2}$.
 One way to find $\Sigma^{1/2}$ is through eigendecomposition.
@@
 
-We assume the random effects are $N(\mu, \Sigma)$.
-This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.
-We can  write the distribution of the random effects as:
-\begin{align}
-\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
-\end{align}
-
-For now, I  assume that  D is diagonal.
-Formulas for this are in section \ref{sec:Ddiag}.
-I generate random effects both from a normal distribution with $\Sigma=D$ diagonal (details in \ref{sec:Ddiag}) and from a normal distribution with $\Sigma$ general (details in \ref{sec:Dgeneral}).
+Consider $N(\mu, \Sigma)$.  This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.  We can  write the distribution  as:
+\begin{align}
+\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
+\end{align}
+
+Since the first version of the package assumes $D$ is diagonal, formulas involving $D$ are found in section \ref{sec:Ddiag}.  
+Since random effects are also generated from a normal with nondiagonal variance matrix, we also need to evaluate $f(u|\mu,\Sigma)$ for general covariance matrix $\Sigma$.
+Information on this calculation for general covariance matrix $\Sigma$ are in \ref{sec:Dgeneral}).
 
 \subsubsection{$\Sigma=D$ Diagonal}\label{sec:Ddiag}
-These equations will be used for both $\log f_\theta(u)$ and for pieces of $\tilde{f}(u)$.
-This section provides the equations for finding the value  of a normal density with $D$ diagonal, the gradient vector, and the hessian matrix.
-
-In the simplest case, $\Sigma=D$ is diagonal with variance components $\nu_t$ on the diagonal and 0 off diagonal.
-This is what I assume to be the true form of D for now.
-Remember that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s everywhere else.
-
-We can partition the random effects according to their variance components: $U=(U_1',...,U_T')'$.
-Let $D_t$ be the variance matrix for $U_t$ .
+This section provides the equations for $\log f_\theta(u)$ and  $\log \tilde{f}(u)$ and their first two derivatives.
+
+ Recall that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s  elsewhere.
+
+We can partition the random effects according to their variance components: $u=(U_1',...,U_T')'$.
+Let $D_t$ be the variance matrix for $U_t$.
 $D_t$ has $q_t$ rows (and also columns).
-Thus $D$ is diagonal:
+Thus $D$ can be expressed as:
 \begin{align}
@@
 \end{align}
-To calculate this vector, I will create a function to take $q_t$, $U_t$, and $\nu_t$ and  output $\dfrac{-q_t}{2\nu_t} +\dfrac{1}{2 \nu_t^2} U_t'U_t$.
-I can then do a loop for t in 1 through T to calculate each entry in the vector.
 
@@
 \end{align}
-I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$.
+%I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$.
 Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.
 
-The inputs of the function are the vector $\nu$ of variance components, U (the vector of random effects), and the list \texttt{z} (from \texttt{mod.mcml}).
+To calculate the value, gradient, and Hessian, we need to provide  $\nu$, $u$, and the list \texttt{z} (from \texttt{mod.mcml}).
 The list \texttt{z} has $T$ matrices, each with the number of columns equal to $q_t$.
@@
 
-Rewritten in C, this function will take as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{U} up based on the variance components, the double array \texttt{drgradient} that will contain the resulting gradient, and the double array \texttt{drhessian} that will contain the resulting hessian.
-
-The result of invoking the C function will be the calculation of the gradient and hessian for the distribution of the random effects.
-The value will be evaluated by the C function in \ref{sec:Dgeneral}.
+Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting hessian.
+
+The result of invoking the C function is  the calculation of the gradient and hessian for the distribution of the random effects. The value is  evaluated by the C function in \ref{sec:Dgeneral}.
 
@@
 \subsubsection{$\Sigma$ in general (\texttt{distRandGeneral})}\label{sec:Dgeneral}
-For now, this is only used for $\tilde{f}(u)$.
-(When $\Sigma$ is general, calculating the derivatives of $\log f_\theta(u)$ is not easy.) Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
-\begin{align}
-\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
+ Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
+\begin{align}
+\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
 \end{align}
@@
 
-%\subsection{$\tilde{f}(u_k)$}\label{sec:ftwiddle}
+%\subsection{$\tilde{f}(u_k)$}
 %I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust.
@@
 
+\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
+Recall that the importance sampling distribution, first stated explicitly in equation \ref{eq:ftwiddle}, is:
+\begin{align}
+ \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) 
+\end{align}
+Then
+\begin{align}
+ \log \tilde{f}(u) = \log \left( p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
+\end{align}
+Because only the first component of the mixture distribution contains any parameters,
+\begin{align}
+\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
+\end{align}
+and 
+\begin{align}
+\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
+\end{align}
+These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
+\begin{align}
+\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
+&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
+\end{align}
+and
+\begin{align}
+\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
+\end{align}
+
 \section{Objective function: approximated log likelihood}\label{sec:objfun}
-This will be hidden from the user. It will be called by the model fitting function.
-
-In order to maximize the approximated log likelihood using trust, I will need an objective function that returns the value, the gradient vector, and the Hessian matrix of the log likelihood.
-This objective function will get $\log f_\theta (u_k)$, $c(\eta_i)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.
-Denote\\
+The objective function is optimized by \texttt{trust} within the the model fitting function.
+In order to evaluate and maximize the approximated log likelihood, I need an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.
+Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, \ref{eq:MCLAhessian}   for calculating these quantities.
+This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.
+
+Denote
 \begin{align}
@@
 
-For computational stability, we'll set $a=max(b_k)$. Then the value is expressed as:\\
-\begin{align}
-\log f_{\theta,m} = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
-\end{align}
-
-Define the weights as:\\
+For computational stability, we'll set $a=max(b_k)$. Then the value of the MCLA is expressed as:\\
+\begin{align}
+\l_m (\theta) = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
+\end{align}
+
+Define the weights as:
 \begin{align}
@@
 
-The gradient vector is: \\
-\begin{align}
-G &= \sum_{k=1}^m \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
-&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
+Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
+\begin{align}
+\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
+&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
+&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
 &= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] v_\theta(u_k,y)+ \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)
-\end{align}
-
-Then the Hessian matrix is: \\
-\begin{align}
-H = \sum_{k=1}^m \nabla^2 \log f_\theta (u_k,y) v_\theta(u_k,y) +  \sum_{k=1}^m \left[ \nabla  \log f_\theta (u_k,y)  \right] \left[ \nabla  \log f_\theta (u_k,y)  \right]^T  v_\theta(u_k,y) -G G^T
+&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
+&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
+&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
+\end{align}
+
+Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
+\begin{align}
+&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
+&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
 \end{align}
@@
 \section{Summary of model}
-Typing \texttt{summary(mod)} will give more detailed info of the model.
-This will be broken into two pieces (as is usually done for summaries): there will be the summary.mcml and print.summary.mcml.
-We split this up because we might have a lot of extra stuff in the summary that we don't necessarily want printed each time.
-So, when a user types \texttt{summary(mod)}, only the  basic information is automatically printed.
+Typing \texttt{summary(mod)}  gives more detailed info of the model.
+This is broken into two pieces (as is usually done for summaries): there will be the \texttt{summary.glmm} and \texttt{print.summary.glmm}.
+We split this up because we have a lot of information in the summary that we do not necessarily want printed each time.
+When a user types \texttt{summary(mod)}, only the  basic information is automatically printed.
 More information can be found in the summary list.
@@
 \begin{align}
-\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) } \label{eq:MCLAgradient}
+\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
 \end{align}
@@
 \end{align}
-Recall the calculation for the MCLA gradient in  \eqref{eq:MCLAgradient}. We can rewrite it as
-\begin{align}
-\nabla l_m(\theta)=  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
-\end{align}
-to more obviously show that both the numerator and denominator are sample means. 
-The law of large numbers says that the numerator and denominator  each converge to their true means. That is,
+Recall the calculation for the MCLA gradient originally stated in  \eqref{eq:MCLAgradient}:
+\begin{align}
+\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }\\
+&=\dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }   -\dfrac{\sum_{k=1}^m     \nabla \log \tilde{f}(u_k)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
+\end{align}
+
+ Note that both the numerator and denominator are sample means. By the law of large numbers, the numerators and denominator  each converge to their true means. That is,
 \begin{align}
@@
 \begin{align}
+\dfrac{1}{m}\sum_{k=1}^m \nabla \log \tilde{f}(u_k)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
+&= \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
+&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
+&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right]
+\end{align}
+and
+\begin{align}
 \dfrac{1}{m}\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow E_{\tilde{f}} \left[ \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
@@
 \begin{align}
-\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
-&=  E_f \left[ \nabla \log f_\theta(u,y)  \right]
+\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}-\dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
+&=  E_f \left[ \nabla \log f_\theta(u,y) - \nabla \log \tilde{f}(u)  \right] \\
+&=\nabla l(\theta)
 \end{align}
@@
 \begin{align}
-&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- E_f \left[ \nabla \log f_\theta(u,y)  \right]  \right] \\
+&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f}(u_k) \right)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- l(\theta)  \right] \\
 &=
-\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m} E_f \left[ \nabla \log f_\theta(u,y)  \right]  \\
-&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)
- - E_f \left[ \nabla \log f_\theta(u,y)  \right] \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
+\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  - \nabla \log \tilde{f}(u_k) \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m}\;  l(\theta)  \\
+&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f}(u_k) \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
+ - l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
 \end{align}
@@
 \sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
+\end{align}
+and
+\begin{align}
+\sum_{k=1}^m     \nabla \log \tilde{f}(u_k)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
 \end{align}
@@
 
-Second, we want to show that the gradient of the MCLA has finite variance. That is, we want to show that
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty.
-\end{align}
-That is, we want to show 
+Second, we want to show that
+\begin{align}
+\sum_{k=1}^m \nabla \log f_\theta (u_k,y) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+ has finite variance. That is, we want to show the existence of
+\begin{align}
+\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
+\end{align}
+That is, we want to show the existence of
 \begin{align}
@@
 &\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
-&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du  \\
-&< \infty.
+&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du.
 \end{align}
@@
 
-%
-%\begin{align}
-%&=
-%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ \dfrac{q_t^2}{4 \nu_t^2} + \dfrac{(u_t'u_t)^2}{4 \nu_t^4} -2\dfrac{ q_t u_t'u_t  }{4 \nu_t^3} \right] du_t \\
-%&=
-%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       \dfrac{q_t^2}{4 \nu_t^2} du_t +
-% \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{(u_t'u_t)^2}{4 \nu_t^4} du_t \nonumber \\
-%&-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}  \dfrac{ q_t u_t'u_t  }{2 \nu_t^3}     du_t \\
-%&=
-% \dfrac{q_t^2}{4 \nu_t^2} \int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)}       du_t +
-%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
-%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t \\
-%&=
-% \dfrac{q_t^2}{4 \nu_t}+
-%\dfrac{1}{4 \nu_t^{4+q_t/2}}  \int  e^{-u_t'u_t/(2 \nu_t)}  (u_t'u_t)^2 du_t \nonumber \\
-%&-  \dfrac{q_t}{2 \nu_t^{3+q_t/2}} \int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t 
-%\end{align}
-%
-%Define $erf$ as the ``error function,'' the function from integrating a normal distribution. Evaluating the integral and using the fact that $erf$ is bounded between $-1$ and $1$, we see 
-%\begin{align}
-%\int e^{-u_t'u_t/(2 \nu_t)}  {  u_t'u_t  }  du_t
-%&\propto
-%    \nu^{3/2} erf \left(\dfrac{u_t}{\sqrt{2 \nu_t} }   \right ) - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)}   \\
-%& \geq -  \nu^{3/2}  - \nu_t u_t e^{- u_t'u_t/(2 \nu_t)} \\
-%& \geq - \infty
-%\end{align}
-%for all $u_t$ and for all $\nu_t.$ Lastly, we need to make sure 
-% \begin{align}
-% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t < \infty
-%\end{align}
-%Let $c_1$, $c_2$, and $c_3$  be constants with respect to $u_t$. Then 
-%\begin{align}
-% \int e^{-u_t'u_t/(2 \nu_t)}  { ( u_t'u_t)^2  }  du_t &= c_1 erf \left( \dfrac{u_t}{\sqrt{2 \nu_t}} \right) - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
-%& \leq  c_1  - c_2 u_t e^{-u_t'u_t / (2 \nu_t)} (3 \nu_t + u_t'u_t) +c_3\\
-%& < \infty
-%\end{align}
-%since
-%\begin{align}
-%e^{-u_t'u_t / (2 \nu_t)} > \infty
-%\end{align}
-%for all $u_t$ and all $\nu_t$.
-
 Next, we want to show 
@@
 \end{align}
-is finite.
+exists and is finite.
 When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
@@
 \end{align}
-is finite.
+exists and is finite.
+
+Third and lastly, we want to show that 
+\begin{align}
+\sum_{k=1}^m \nabla \log \tilde{f} (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+ has finite variance. That is, we want to show the existence of
+\begin{align}
+\int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
+\end{align}
+We see that
+\begin{align}
+\int& \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du\\
+&\leq  \int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)} du\\
+&= \dfrac{1}{p_1} \int N(u|0,D) \left[\nabla \log \tilde{f}u) \right]^2 du\\
+&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)+p_2 N(u|u^*,D^*)+p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}   \right]^2 du\\
+&\leq\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)}   \right]^2 du\\
+&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{ \dfrac{\partial}{\partial \nu} N(u|0,D)}{ N(u|0,D)}   \right]^2 du\\
+&= \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du \\
+\end{align}
 

### 2014-11-24 NewDesignDoc/DesignDocGLMM.tex
Completed update to CLT proof

@@
 Typing \texttt{summary(mod)}  gives more detailed info of the model.
-This is broken into two pieces (as is usually done for summaries): there will be the \texttt{summary.glmm} and \texttt{print.summary.glmm}.
+This is broken into two pieces (as is usually done for summaries):  \texttt{summary.glmm} and \texttt{print.summary.glmm}.
 We split this up because we have a lot of information in the summary that we do not necessarily want printed each time.
@@
 
-The summary will do all the calculations and its value will be a list of the following:
+The summary performs all calculations and its value is a list of the following:
 \begin{itemize}
@@
 \end{align}
+Now we use
+\begin{align}
+\dfrac{\partial}{\partial \nu} \log N(u|0,D) = \dfrac{\dfrac{\partial}{\partial \nu}  N(u|0,D)}{N(u|0,D)}
+\end{align}
+to see
+\begin{align}
+ \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du &=\dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du \\
+&= \dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du 
+\end{align}
+which exists and is finite.
 

### 2014-12-07 NewDesignDoc/DesignDocGLMM.tex
Editing design doc w ftwiddle incl N(u|0,D)
About to change directions though to instead use t

@@
 
-Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$.
-Let $D$ be determined by the current iteration of \texttt{trust}.
-The importance sampling distribution $\tilde{f}$ is the following mixture distribution:
-\begin{align}
- \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
-\end{align}
-
-Therefore, the random effects are generated from the following distributions:
+ Recall that $m$ is the overall Monte Carlo sample size.
+That is, $\breve{u}_k,k=1\ldots,m$ are vectors of random effects are drawn from a multivariate standard normal.
+Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$ and let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$.
+Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
+Construct $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.
+Then we center and scale $\breve{u}_k$ as follows:
+\begin{align}
+u_k&= A\, \breve{u}_k, k=1,\ldots,m_1\\
+u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
+u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
+\end{align}
+ The result or centering and scaling is that
 \begin{enumerate}
-\item $N(0,I)$
-\item $N(u^*,D^*)$
-\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
+\item $u_k,k=1,\ldots,m_1 \sim N(0,D)$
+\item $u_k,k=m_1+1,\ldots,m_1+m_2 \sim N(u^*,D^*)$
+\item $u_k,k=m_1+m2+1,\ldots,m \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
 \end{enumerate}
-
-The first distribution is the multivariate standard normal, but the random effects drawn from this are scaled in the trust objective function so that their variance matrix $D$ is determined by the current iteration of trust.
-The second distribution is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
+ Let $f(u_k|\mu,\Sigma)$ denote the pdf for $u_k \sim N(\mu,\Sigma)$. Then the importance sampling distribution $\tilde{f}$ is the following mixture distribution:
+\begin{align}
+ \tilde{f}(u) = p_1 \,  f(u \, | \, 0, D)+p_2 \, f(u \, | \, u^*, D^*)+p_3 \, f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
+\end{align}
+
+\subsubsection{Rationale}
+The first distribution $N(0,D)$ was chosen to ensure a Central Limit Theorem, which is proven in  section \ref{sec:CLT}.
+The second distribution is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
 The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood Hessian.
@@
 
-The random effects are generated by shifting and scaling a sample from a  standard normal distribution.
-Suppose we want to generate a sample from a general normal distribution $N(\mu,\Sigma)$.
-Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.
-For each $k$, $k=1,\ldots,m$, draw a sample of size $q$ from the standard normal distribution.
-Call this sample $\breve{u}_k$.
-Then shift and scale it:
-\begin{align}
-u_k = \breve{u}_k \Sigma^{1/2} + \mu
-\end{align}
-so that it is a sample from $N(\mu,\Sigma)$.
-
-This method requires $\Sigma^{1/2}$.
-One way to find $\Sigma^{1/2}$ is through eigendecomposition.
+\subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
+The shifting and scaling of $\breve{u})_k$ relies on finding the square root of a variance matrix.
+In the first two distributions, the variance matrices $D$ and $D^*$ are diagonal, and therefore we simply take the root of the diagonal elements.
+For the third distribution, we can use eigendecomposition.
 Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.
-We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).
+Let $\Sigma$ denote the nondiagonal variance matrix.
+Eigendecomposition of $\Sigma$ provides orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).
 Then
@@
 \begin{align}
-\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\
-&= O \Lambda^{1/2}\Lambda^{1/2} O'\\
-&=O \Lambda O' \\
+\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\ \nonumber
+&= O \Lambda^{1/2}\Lambda^{1/2} O'\\\nonumber
+&=O \Lambda O' \\\nonumber
 &= \Sigma.
+\end{align}
+
+\subsubsection{NEED TO PUT THIS SOMEWHERE}
+
+Recall that $D$ is a diagonal with dimension $q$.
+$D$ has $q_1$ entries being $\nu_1$, $q_2$ entries being $\nu_2$, ..., and $q_T$ entries being $\nu_T$.
+Since the distribution of $u_k=A \breve{u}_k$ is $N(0,D)$, we write its pdf as:
+\begin{align}
+f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
+&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
+&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
+\end{align}
+So that 
+\begin{align}
+\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
+&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
+\end{align}
+Then 
+\begin{align}
+\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
+\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
 \end{align}
@@
 
-\subsection{Distribution of random effects is normal (\texttt{distRand})}\label{sec:distRand}
-%
-%  Recall that the first version of the package assumes $D$ is  diagonal. Since $U$ is normally distributed, we can write its log density as
-%\begin{align}
-%\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} u' D^{-1} u
-%\end{align}
-%where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.
-%
+\subsection{Distribution of random effects (\texttt{distRand})}\label{sec:distRand}
 
@@
 
-Then print.summary.glmm will print these things:
+Then \texttt{print.summary.glmm}  prints the following:
 \begin{itemize}
@@
 Note that the summary and print.summary functions will be S3 generic functions.
-What this means is the user will type ``summary(mod)'' and summary is a generic function.
-R checks the class of the model $mod$ and then automatically uses the summary and print.summary functions for that class of objects.
-This also might affect the way that I export the function and document it in the manual.
-I think that I'll be ok if I copy parts of his summary.aster and print.summary.aster.
-I need to read about generics more and will check out the hints library.
+This means is the user  types \texttt{summary(mod)} and \texttt{summary} is a generic function.
+R checks the class of the model \texttt{mod} and then automatically uses the \texttt{summary} and \texttt{print.summary} functions for that class of objects.
+
+\section{PQL}\label{sec:pql}
+ Before we get started on describing PQL, we need to change notation to avoid constrained optimization.
+Recall that $D=\var(u)$ and is (for now) assumed to be diagonal.
+Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative.
+Using A rather than D enables unconstrained optimization.
+If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
+\begin{align*}
+A= \sum_t E_t \sigma_t.
+\end{align*}
+Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix.
+PQL will estimate the components contained on the diagonal of A.
+Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).
+
+In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.
+
+There are two ways to do PQL, both of which are described in the vignette
+ \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014).
+In either case, there is an inner optimization and an outer optimization.
+The inner optimization is well behaved while the other optimization is a little tougher.
+I will be using the method that is not quite PQL but is pretty close and is better behaved.
+In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$.
+Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
+
+The {\bf inner} optimization will be done with the trust function in R.
+We elect to use trust because it requires two derivatives, which will make the optimization more precise.
+We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.
+
+The inner optimization maximizes the penalized log likelihood. After defining
+\begin{align}
+\eta=X\beta +ZAs
+\end{align}
+we calculate the  log likelihood as 
+\begin{align}
+l(\eta)= Y' \eta - c(\eta) 
+\end{align}
+and the penalized likelihood as:
+\begin{align}
+ l(\eta)- \dfrac{1}{2} s's.
+\end{align}
+This function will be maximized using the \texttt{trust} package.
+We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$.
+We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.
+
+Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
+\begin{align}
+l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
+\end{align}
+then 
+\begin{align}
+\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i.
+\end{align}
+ Now we can write the following expression:
+
+\begin{align}
+%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
+\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
+&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
+&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
+&=\sum_i (Y_i-\mu_i) X_{ik}
+\end{align}
+
+We find the function's derivative with respect to $s$ as follows:
+\begin{align}
+\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
+&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
+&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
+&=(Y-\mu)'ZA -s
+\end{align}
+
+This gives us the following derivatives of the penalized log likelihood:
+\begin{align}
+\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
+\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
+\end{align}
+
+Lastly, we need the Hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
+\begin{enumerate}
+\item $ \dfrac{\partial^2}{\partial s^2}$
+\item $ \dfrac{\partial^2}{\partial \beta^2}$
+\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
+\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
+\end{enumerate}
+
+We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
+\begin{align}
+  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
+&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
+&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
+&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
+&= -AZ' \left[ c''(\eta) \right] ZA - I_q
+\end{align}
+Note that $c''(\eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.
+$I_q$ is the identity matrix of dimension $q$.
+This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.
+
+Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
+\begin{align}
+  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
+&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
+&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
+&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
+&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
+&=-X'  \left[ c''(\eta)\right]  X   
+\end{align}
+
+Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
+\begin{align}
+\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
+&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
+&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
+&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
+&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
+&=-   \left[ c''(\eta) X  \right]' ZA   \\
+&=-X'   \left[ c''(\eta)   \right] ZA  
+\end{align}
+
+And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood.
+Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
+Trust will maximize the penalized likelihood.
+
+The {\bf outer} optimization is done using \texttt{optim} with the default method of ``\texttt{Nelder-Mead.}'' This requires just the function value and no derivatives.
+This optimization method was chosen because the optimization function already contains second derivatives of the cumulant function; requiring derivatives of the optimization function would in turn require higher-order derivatives of the cumulant.
+
+The default of \texttt{optim} is to minimize, but we'd like to  do maximization. Reversing the sign of the optimization function will turn the maximization into minimization.
+
+If $\tilde{\beta}$ and $\tilde{s}$ are available from previous calls to the inner optimization function, then they are used here.
+Otherwise, the values are taken to be 0 and 1.
+The outer optimization's function is evaluated by first defining
+\begin{align}
+\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
+l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
+A&= \sum_k E_k \sigma_k.
+\end{align}
+Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
+\begin{align}
+ l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
+\end{align}
+Again, \texttt{optim} minimizes, so we have to minimize the negative value of the penalized quasi-likelihood in order to maximize the value of it.
+
+We do need to be careful about the determinant. Let $AZ' W ZA +I= LL'$ where $L$ is the lower triangular matrix resulting from a Cholesky decomposition. Then
+\begin{align}
+\dfrac{1}{2} \log | AZ' W ZA +I | &= \dfrac{1}{2} \log |LL'|\\
+&=\dfrac{1}{2} \log (|L|)^2 \\
+&= \log |L|
+\end{align}
+Since $L$ is triangular, the determinant is just the product of the diagonal elements. Let $l_i$ be the diagonal elements of $L$. Then
+\begin{align}
+\dfrac{1}{2} \log | AZ' W ZA +I | &= \log \prod l_i \\
+&= \sum \log l_i
+\end{align}
+
+ If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
+This would be more numerically stable, but is slower.
+Let $O$ be the matrix containing the eigenvectors and let $\Lambda$ be a diagonal matrix with the eigenvalues $\lambda_i$ on the diagonal.
+Then
+\begin{align}
+ &AZ' W ZA = O \Lambda O' 
+\end{align}
+Then we can rewrite the argument of the determinant as follows:
+\begin{align}
+& AZ' W ZA +I  = O \Lambda O' + I = O \Lambda O + OO'=O (I+\Lambda) O' 
+\end{align}
+This leads to the careful calculation of our determinant as follows:
+\begin{align}
+&   \left| AZ' W ZA +I \right| = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
+&\Rightarrow \log   | AZ' W ZA +I | = \sum \log(1+\lambda_i) 
+\end{align}
+The last quantity can be accurately calculated using the \texttt{log1p} function in R.
+
+The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them.
+To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
+First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.
+Second, after PQL is finished, $\tilde{s}$ is used to help center the generated random effects.
+
+The point of doing PQL is to construct a decent importance sampling distribution.
+Thus, the estimates don't have to be perfect.
+It is possible that one of the $\sigma_k$ will be 0 according to PQL.
+If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.
 
@@
 
-\section{PQL}\label{sec:pql}
- Before we get started on describing PQL, we need to change notation to avoid constrained optimization.
-Recall that $D=\var(u)$ and is (for now) assumed to be diagonal.
-Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative.
-Using A rather than D enables unconstrained optimization.
-If $\sigma$ is a vector of the distinct standard deviations with components $\sigma_t$, we can write A as a function of $\sigma$ by
-\begin{align*}
-A= \sum_t E_t \sigma_t.
-\end{align*}
-Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix.
-PQL will estimate the components contained on the diagonal of A.
-Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).
-
-In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.
-
-There are two ways to do PQL, both of which are described in the vignette
- \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014).
-In either case, there is an inner optimization and an outer optimization.
-The inner optimization is well behaved while the other optimization is a little tougher.
-I will be using the method that is not quite PQL but is pretty close and is better behaved.
-In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$.
-Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
-
-The {\bf inner} optimization will be done with the trust function in R.
-We elect to use trust because it requires two derivatives, which will make the optimization more precise.
-We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.
-
-The inner optimization maximizes the penalized log likelihood. After defining
-\begin{align}
-\eta=X\beta +ZAs
-\end{align}
-we calculate the  log likelihood as 
-\begin{align}
-l(\eta)= Y' \eta - c(\eta) 
-\end{align}
-and the penalized likelihood as:
-\begin{align}
- l(\eta)- \dfrac{1}{2} s's.
-\end{align}
-This function will be maximized using the \texttt{trust} package.
-We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$.
-We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.
-
-Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
-\begin{align}
-l(\eta) = \sum_i Y_i \eta_i - c(\eta_i)
-\end{align}
-then 
-\begin{align}
-\dfrac{\partial \l(\eta)}{\partial \eta_i} = Y_i-c'(\eta_i) = Y_i -\mu_i.
-\end{align}
- Now we can write the following expression:
-
-\begin{align}
-%\dfrac{\partial}{\partial s_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right]= \sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial s_k} \\
-\dfrac{\partial}{\partial \beta_k} \left[ l(\eta)-\dfrac{1}{2} s's  \right] &= \dfrac{\partial l(\eta)}{\partial \beta_k}    \\
-&= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
-&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
-&=\sum_i (Y_i-\mu_i) X_{ik}
-\end{align}
-
-We find the function's derivative with respect to $s$ as follows:
-\begin{align}
-\dfrac{\partial }{\partial s} \left[ l(\eta) -\dfrac{1}{2} s's   \right] &= \dfrac{\partial l(\eta)}{\partial s} -\dfrac{1}{2} \dfrac{\partial s's}{\partial s}\\
-&= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
-&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
-&=(Y-\mu)'ZA -s
-\end{align}
-
-This gives us the following derivatives of the penalized log likelihood:
-\begin{align}
-\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
-\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
-\end{align}
-
-Lastly, we need the Hessian of the penalized likelihood. This matrix can be broken down into four pieces: 
-\begin{enumerate}
-\item $ \dfrac{\partial^2}{\partial s^2}$
-\item $ \dfrac{\partial^2}{\partial \beta^2}$
-\item $\dfrac{\partial^2}{ \partial s \; \partial \beta}$
-\item $\left(\dfrac{\partial^2}{ \partial s \; \partial \beta}\right) ' =\dfrac{\partial^2}{ \partial \beta \; \partial s}$
-\end{enumerate}
-
-We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
-\begin{align}
-  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
-&=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
-&= \left[-\dfrac{\partial c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s}   \right] 'ZA - I_q \\
-&= \left[ -c''(\eta) ZA  \right]'ZA - I_q \\
-&= -AZ' \left[ c''(\eta) \right] ZA - I_q
-\end{align}
-Note that $c''(\eta)$ is a $q\times q$ diagonal matrix with diagonal elements $c''(\eta_i)$.
-$I_q$ is the identity matrix of dimension $q$.
-This makes  $ \dfrac{\partial^2}{\partial s^2}$ a $q \times q$ matrix.
-
-Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
-\begin{align}
-  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
-&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
-&= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
-&=-X  \left[' \dfrac{\partial}{\partial \beta} c'(\eta)  \right] \\
-&=-X'  \left[ \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right] \\
-&=-X'  \left[ c''(\eta)\right]  X   
-\end{align}
-
-Next is the $p \times q$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}$.
-\begin{align}
-\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
-&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
-&=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
-&=- \left[  \dfrac{\partial  c'(\eta)}{\partial \beta}  \right]' ZA  \\
-&=-   \left[  \dfrac{\partial  c'(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial \beta}  \right]' ZA   \\
-&=-   \left[ c''(\eta) X  \right]' ZA   \\
-&=-X'   \left[ c''(\eta)   \right] ZA  
-\end{align}
-
-And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood.
-Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
-Trust will maximize the penalized likelihood.
-
-The {\bf outer} optimization is done using \texttt{optim} with the default method of ``\texttt{Nelder-Mead.}'' This requires just the function value and no derivatives.
-This optimization method was chosen because the optimization function already contains second derivatives of the cumulant function; requiring derivatives of the optimization function would in turn require higher-order derivatives of the cumulant.
-
-The default of \texttt{optim} is to minimize, but we'd like to  do maximization. Reversing the sign of the optimization function will turn the maximization into minimization.
-
-If $\tilde{\beta}$ and $\tilde{s}$ are available from previous calls to the inner optimization function, then they are used here.
-Otherwise, the values are taken to be 0 and 1.
-The outer optimization's function is evaluated by first defining
-\begin{align}
-\tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
-l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
-A&= \sum_k E_k \sigma_k.
-\end{align}
-Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
-\begin{align}
- l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
-\end{align}
-Again, \texttt{optim} minimizes, so we have to minimize the negative value of the penalized quasi-likelihood in order to maximize the value of it.
-
-We do need to be careful about the determinant. Let $AZ' W ZA +I= LL'$ where $L$ is the lower triangular matrix resulting from a Cholesky decomposition. Then
-\begin{align}
-\dfrac{1}{2} \log | AZ' W ZA +I | &= \dfrac{1}{2} \log |LL'|\\
-&=\dfrac{1}{2} \log (|L|)^2 \\
-&= \log |L|
-\end{align}
-Since $L$ is triangular, the determinant is just the product of the diagonal elements. Let $l_i$ be the diagonal elements of $L$. Then
-\begin{align}
-\dfrac{1}{2} \log | AZ' W ZA +I | &= \log \prod l_i \\
-&= \sum \log l_i
-\end{align}
-
- If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
-This would be more numerically stable, but is slower.
-Let $O$ be the matrix containing the eigenvectors and let $\Lambda$ be a diagonal matrix with the eigenvalues $\lambda_i$ on the diagonal.
-Then
-\begin{align}
- &AZ' W ZA = O \Lambda O' 
-\end{align}
-Then we can rewrite the argument of the determinant as follows:
-\begin{align}
-& AZ' W ZA +I  = O \Lambda O' + I = O \Lambda O + OO'=O (I+\Lambda) O' 
-\end{align}
-This leads to the careful calculation of our determinant as follows:
-\begin{align}
-&   \left| AZ' W ZA +I \right| = 1 *  \prod_{i=1}^n (1+\lambda_i) *1\\
-&\Rightarrow \log   | AZ' W ZA +I | = \sum \log(1+\lambda_i) 
-\end{align}
-The last quantity can be accurately calculated using the \texttt{log1p} function in R.
-
-The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them.
-To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
-First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.
-Second, after PQL is finished, $\tilde{s}$ is used to help center the generated random effects.
-
-The point of doing PQL is to construct a decent importance sampling distribution.
-Thus, the estimates don't have to be perfect.
-It is possible that one of the $\sigma_k$ will be 0 according to PQL.
-If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.
-
 \bibliographystyle{apalike}
@@
 \section{MCLA calculations}
-The Monte Carlo log likelihood approximation is
+Let $u_k,k=1,\ldots,m$ be a sample from $\tilde{f}(u_k)$. The Monte Carlo log likelihood approximation is
 \begin{align}
@@
 \end{align}
+The gradient vector and Hessian matrix calculations depend on whether $\tilde{f}(u_k)$ contains $\theta$.
+The calculations in \ref{sec:calcs} are for $\tilde{f}(u_k)$ containing $\theta$ and the calculations in \ref{sec:calcsindep} are for $\tilde{f}(u_k)$ not containing $\theta$.
 
@@
 
-\section{Central Limit Theorem for MCLA}
+\section{Central Limit Theorem for MCLA}\label{sec:CLT}
 Define

### 2015-01-09 NewDesignDoc/DesignDocGLMM.tex
Continuing proof. About half done.

@@
 \section{Central Limit Theorem for MCLA}\label{sec:CLT}
+In this section, we focus on the gradient of the MCLA because the trust criterion for finding the maximum is based on the gradient.
 Define
@@
 \begin{align}
-\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }\\
-&=\dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }   -\dfrac{\sum_{k=1}^m     \nabla \log \tilde{f}(u_k)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
+\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
 \end{align}
@@
 \end{align}
-and
-\begin{align}
-\dfrac{1}{m}\sum_{k=1}^m \nabla \log \tilde{f}(u_k)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-&= \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
-&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right]
-\end{align}
+%and
+%\begin{align}
+%\dfrac{1}{m}\sum_{k=1}^m \nabla \log \tilde{f}(u_k)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
+%&= \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
+%&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
+%&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right]
+%\end{align}
 and
@@
 \end{align}
-Then, by Slutsky's,
-\begin{align}
-\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}-\dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
-&=  E_f \left[ \nabla \log f_\theta(u,y) - \nabla \log \tilde{f}(u)  \right] \\
+Then, by Slutsky's theorem,
+\begin{align}
+\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
+&=  E_f \left[ \nabla \log f_\theta(u,y) \right] \\
 &=\nabla l(\theta)
@@
 \begin{align}
-&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f}(u_k) \right)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- l(\theta)  \right] \\
+&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- l(\theta)  \right] \\
 &=
-\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  - \nabla \log \tilde{f}(u_k) \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \sqrt{m}\;  l(\theta)  \\
-&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f}(u_k) \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
+\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\;  l(\theta)  \\
+&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
  - l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
@@
 \end{align}
+are finite.
+First, we want to show that
+\begin{align}
+\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+has finite variance. This is true if and only if 
+\begin{align}
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du < \infty.
+\end{align}
+We see:
+\begin{align}
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &=\int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{\tilde{f}(u)} du\\
+&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u)} du\\
+&\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, f_\theta(y|u)^2  }{\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}} du
+\end{align}
+Since
+\begin{align}
+f_\theta(y|u)^2=\left( e^{y'\eta-c(\eta)} \right)^2 = e^{2y'\eta-2c(\eta)} ,
+\end{align}
+\begin{align}
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, \exp(2y'\eta)  }{\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2} \,  \exp(2c(\eta))  } du\\
+&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'\eta)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du \\
+&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du 
+\end{align}
+When $y|u\sim$ Bernoulli, then
+\begin{align}
+c(\eta)=\log(1+e^\eta) \Rightarrow \exp(2c(\eta))&= \exp( 2 \log(1+e^\eta)  )\\
+&= ( \exp( \log(1+e^\eta)  )  )^2\\
+&= (1+e^\eta) ^2
+\end{align}
 and
 \begin{align}
-\sum_{k=1}^m     \nabla \log \tilde{f}(u_k)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
-\end{align}
-are each finite.
-First, we want to show that
-\begin{align}
-\sum_{k=1}^m \dfrac{f_\theta(u_k)}{\tilde{f}(u_k)}
-\end{align}
-has finite variance. This is true if and only if 
-\begin{align}
-\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du < \infty.
-\end{align}
-We see:
-\begin{align}
-\int \dfrac{f_\theta(u)^2}{\tilde{f}(u)} du &= \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
-& \leq \int \dfrac{N(u|0,D)^2}{p_1 N(u|0,D)} du \\
-&= \dfrac{1}{p_1} \int N(u|0,D) du \\
-&= \dfrac{1}{p_1}.
-\end{align}
-Therefore $\sum_{k=1}^m f_\theta(u_k)$ has finite variance as long as $p_1\neq 0$.
-
-Second, we want to show that
-\begin{align}
-\sum_{k=1}^m \nabla \log f_\theta (u_k,y) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
- has finite variance. That is, we want to show the existence of
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  (1+e^\eta) ^2  } du,
+\end{align}
+which converges since $\exp(u'D^{-1}u$ grows faster than any term in the numerator of the integrand.
+
+On the other hand, when $y|u\sim$ Poisson, then 
+\begin{align}
+c(\eta)=\exp(\eta) \Rightarrow \exp(2 c(\eta)) = (\exp(\exp(\eta)))^2
+\end{align}
+and
+\begin{align}
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,   (\exp(\exp(\eta)))^2  } du,
+\end{align}
+which converges since  $(\exp(\exp(\eta)))^2$ grows faster than any term in the numerator of the integrand.
+
+Therefore, whether the data are Bernoulli or Poisson distributed, the importance sampling weights 
+\begin{align}
+\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+have finite variance.
+
+Now, we want to show that
+\begin{align}
+\sum_{k=1}^m \nabla \log f_\theta (u_k,y) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} =\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} +\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+ has finite variance. We need to show the following are finite:
+\begin{align}
+\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}   \right) \\
+\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)\\
+\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)_.
+\end{align}
+
+That is, we want to show the existence of the following:
+\begin{align}
+&\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:first} \\
+&\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:second} \\
+&\int \dfrac{  \nabla \log f_\theta (u)  \nabla \log f_\theta (y|u)  \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \\
+\end{align}
+
+First, we start with \eqref{eq:first}:
+\begin{align}
+\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du 
+&=\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
+&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u)} du \\
+&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}} du \\
+&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
+&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
+\end{align}
+By the Cauchy-Schwartz inequality, the integral in \eqref{eq:soon} exists if
+\begin{align}
+ & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
+&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
+\end{align}
+exists, where $c'(\eta)$ is a vector with components $c'(\eta_i)$.
+This will converge as long as no term in the numerator of the integrand grows more quickly than $\exp(u'D^{-1}u$.
+The second and third terms grow less quickly than $\exp(u'D^{-1}u$, so $\left[ c'(\eta)\right]^2$ is the only term to check.
+If $y|u \sim$ Bernoulli, then
+\begin{align}
+c'(\eta_i)= \dfrac{\exp(\eta_i)}{1+\exp(\eta_i)},
+\end{align}
+which does not grow more quickly than $\exp(u'D^{-1}u)$. If $y|u \sim$ Poisson, then
+\begin{align}
+c'(\eta_i)= e^{\eta_i},
+\end{align}
+which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and the variance of 
+\begin{align}
+\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\end{align}
+is finite.
+
+Next, we move on to showing the existence of \eqref{eq:second}.
+EDITED TO HERE.
+
+ That is, we want to show the existence of
 \begin{align}
@@
 
-Third and lastly, we want to show that 
+Next, we want to show that 
 \begin{align}

### 2015-01-11 NewDesignDoc/DesignDocGLMM.tex
Finished theorem.
About to start editing doc from beginning.

@@
 &\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:second} \\
-&\int \dfrac{  \nabla \log f_\theta (u)  \nabla \log f_\theta (y|u)  \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \\
+&\int \dfrac{  \nabla \log f_\theta (u)  \nabla \log f_\theta (y|u)  \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:third} 
 \end{align}
@@
 \end{align}
-which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and the variance of 
-\begin{align}
-\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and 
+\begin{align}
+\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
 \end{align}
@@
 Next, we move on to showing the existence of \eqref{eq:second}.
-EDITED TO HERE.
-
- That is, we want to show the existence of
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
-\end{align}
-That is, we want to show the existence of
-\begin{align}
-\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
-&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
-&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
-&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du.
-\end{align}
-By the Cauchy-Schwartz inequality, 
-\begin{align}
-  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
-& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
-\end{align}
-Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
-\begin{align}
-\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
-\end{align}
-Recall the notation from section \ref{sec:Ddiag}.  If
-\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
-\end{align}
-for every $t=1,...,T$, then
-\begin{align}
-\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
-\end{align}
-by Cauchy Schwartz.
+\begin{align}
+\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &= 
+\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du\\
+&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du
+\end{align}
+By the Cauchy-Schwartz inequality, the above integral exists as long as  
+\begin{align}
+\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du < \infty
+\end{align}
+for every $t=1,\ldots, T$. We see
+\begin{align}
+&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \\
+&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u)} du \\
+&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
+&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
+\end{align}
+Because every term in the numerator of the integrand grows less quickly than $\exp(u'D^{-1}u)$, the integral exists.  Therefore, \eqref{eq:second} exists and 
+\begin{align}
+\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
+\end{align}
+
+Lastly, we want to show the existence of \eqref{eq:third}. We see
+\begin{align}
+\int \dfrac{  \nabla \log f_\theta (u)\,  \nabla \log f_\theta (y|u) \, \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &=
+\int \dfrac{  \nabla \log f_\theta (u) \, \nabla \log f_\theta (y|u) \, \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
+&\leq
+\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
+\end{align}
+Again, we use the Cauchy-Schwartz inequality to check the existence of \eqref{eq:this1}. If, for every $t=1,\ldots,T$,
+\begin{align}
+\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
+\end{align}
+exists, then \eqref{eq:this1} also exists. Continuing, we see
+\begin{align}
+& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du\\
+ &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u)} du\\
+ &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
+&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
+&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du.
+\label{eq:this2}
+\end{align}
+We have already shown that $-c'(\eta)$ does not grow more quickly than $\exp(u'D^{-1}u)$.
+The other terms in the numerator of the integrand shown in \eqref{eq:this2} also grow more slowly than $\exp(u'D^{-1}u)$.
+We conclude that \eqref{eq:this2} exists, which implies that \eqref{eq:third} exists and
+\begin{align}
+\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
+\end{align}
+This shows that
+\begin{align}
+\var \left( \sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty. 
+\end{align}
+Therefore, the gradient of the MCLA has finite variance and a central limit theorem.
+
+%EDITED TO HERE.
+%
+% That is, we want to show the existence of
+%\begin{align}
+%\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
+%\end{align}
+%That is, we want to show the existence of
+%\begin{align}
+%\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
+%&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
+%&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
+%&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du.
+%\end{align}
+%By the Cauchy-Schwartz inequality, 
+%\begin{align}
+%  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
+%& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
+%\end{align}
+%Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
+%\begin{align}
+%\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
+%\end{align}
+%Recall the notation from section \ref{sec:Ddiag}.  If
+%\begin{align}
+%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
+%\end{align}
+%for every $t=1,...,T$, then
+%\begin{align}
+%\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
+%\end{align}
+%by Cauchy Schwartz.
 Therefore, I will prove the integral is finite for the general case of $u_t$ and $\nu_t$.
 First, we recognize that our integral is actually an expectation of a function of a normal random variable:
-\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
-\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
-\end{align}
-Normal distributions have finite moments of all orders, so we know that\begin{align}
-\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
-\end{align}
-
-Next, we want to show 
-\begin{align}
-\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
-\end{align}
- Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
-\begin{align}
-\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
-\end{align}
-We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
-\begin{align}
-\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-\end{align}
-exists and is finite.
-When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
-\begin{align}
-\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
-\end{align}
-Then
-\begin{align}
-\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
-&\propto \int N(u|0,D) e^{2Zu} du 
-\end{align}
-exists and is finite.
-
-Next, we want to show that 
-\begin{align}
-\sum_{k=1}^m \nabla \log \tilde{f} (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
- has finite variance. That is, we want to show the existence of
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
-\end{align}
-We see that
-\begin{align}
-\int& \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du\\
-&\leq  \int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)} du\\
-&= \dfrac{1}{p_1} \int N(u|0,D) \left[\nabla \log \tilde{f}u) \right]^2 du\\
-&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)+p_2 N(u|u^*,D^*)+p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}   \right]^2 du\\
-&\leq\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)}   \right]^2 du\\
-&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{ \dfrac{\partial}{\partial \nu} N(u|0,D)}{ N(u|0,D)}   \right]^2 du\\
-&= \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du \\
-\end{align}
-Now we use
-\begin{align}
-\dfrac{\partial}{\partial \nu} \log N(u|0,D) = \dfrac{\dfrac{\partial}{\partial \nu}  N(u|0,D)}{N(u|0,D)}
-\end{align}
-to see
-\begin{align}
- \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du &=\dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du \\
-&= \dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du 
-\end{align}
-which exists and is finite.
-
-Therefore, it is proven that 
-\begin{align}
-\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
-\end{align}
-exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.
+%\begin{align}
+%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
+%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
+%\end{align}
+%Normal distributions have finite moments of all orders, so we know that\begin{align}
+%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
+%\end{align}
+%
+%
+%
+%
+%Next, we want to show 
+%\begin{align}
+%\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
+%\end{align}
+% Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
+%\begin{align}
+%\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
+%%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
+%\end{align}
+%We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
+%\begin{align}
+%\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
+%\end{align}
+%exists and is finite.
+%When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
+%\begin{align}
+%\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
+%\end{align}
+%Then
+%\begin{align}
+%\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
+%&\propto \int N(u|0,D) e^{2Zu} du 
+%\end{align}
+%exists and is finite.
+%
+%Next, we want to show that 
+%\begin{align}
+%\sum_{k=1}^m \nabla \log \tilde{f} (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+%\end{align}
+% has finite variance. That is, we want to show the existence of
+%\begin{align}
+%\int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
+%\end{align}
+%We see that
+%\begin{align}
+%\int& \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du\\
+%&\leq  \int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)} du\\
+%&= \dfrac{1}{p_1} \int N(u|0,D) \left[\nabla \log \tilde{f}u) \right]^2 du\\
+%&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)+p_2 N(u|u^*,D^*)+p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}   \right]^2 du\\
+%&\leq\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)}   \right]^2 du\\
+%&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{ \dfrac{\partial}{\partial \nu} N(u|0,D)}{ N(u|0,D)}   \right]^2 du\\
+%&= \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du \\
+%\end{align}
+%Now we use
+%\begin{align}
+%\dfrac{\partial}{\partial \nu} \log N(u|0,D) = \dfrac{\dfrac{\partial}{\partial \nu}  N(u|0,D)}{N(u|0,D)}
+%\end{align}
+%to see
+%\begin{align}
+% \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du &=\dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du \\
+%&= \dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du 
+%\end{align}
+%which exists and is finite.
+%
+%
+%
+%Therefore, it is proven that 
+%\begin{align}
+%\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
+%\end{align}
+%exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.
 

### 2015-01-13 NewDesignDoc/DesignDocGLMM.tex
Updated design doc's first 5 sections

@@
 \end{align}
-When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
-\begin{align}
-\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
+When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
+\begin{align}
+\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
 \end{align}
@@
 \begin{align}
-&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}. \label{eq:MCLAhessian}
-\end{align}
-Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.
-
-Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.
+\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
+\end{align}
+Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.
+
+Now any likelihood-based inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.
 
 \section{Model fitting function} 
-This is be the primary function the user  uses.
+The model fitting function is be the primary function the user  uses.
 The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}.
 Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.
-When the fixed effects are specified, the model matrix $X$  is created, where $X$ has dimensions $n \times p$.
-
-The user  specifies the random effects in the same way as for the R function \texttt{reaster} in the R package
- \texttt{aster} \citep{aster-package}.
+As a result of the user specifying the fixed effects, the model fitting function creates matrix $X$, which has dimensions $n \times p$.
+
+The user  specifies the random effects in the same way as the fixed effects. This is also how users of  \texttt{reaster} (of R package
+ \texttt{aster} \citep{aster-package}) specify the random effects.
 That is, random effects are expressed using the R formula mini-language.
 Let $q_t$ be the number of random effects associated with variance component $\nu_t$.
-When the random effects are specified, a list of T model matrices are created and the $t$th model matrix $Z_t$ has dimensions $n \times q_t$.
-
-The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the name of the variance components.
+When the random effects are specified, a list of $T$ model matrices are created.
+The $t$th model matrix $Z_t$ has dimensions $n \times q_t$.
+
+The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the names of the variance components.
 
@@
  \end{verbatim}
-
-%$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+?
-Most models require this, so please check if you meant to include it.''
 
@@
 
-As a result of this command, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices.
+Initially, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices.
 Since we have 3 distinct variance components, we want 3 design matrices.
-We take the design matrices that share a variance component and \texttt{cbind} them together.
+We take the design matrices that share a variance component and use \texttt{cbind} to bind them together.
 The result is 3 model matrices in the $Z$ list, one for each variance component.
-We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  ("subject","year","repeat") are correctly assigned.
-Also, the variance estimates  come out in this same order as well.
+We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  (``subject'',``year'',``repeat'') are correctly assigned.
+The variance estimates  are eventually listed in the model summary in this same order as well.
 
@@
 The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.
-More information on generated the random effects is in section \ref{sec:genRand}.
+More information on generating the random effects is in section \ref{sec:genRand}.
 
@@
 \begin{align}
-g(\mu) = X \beta + Z U
+g(\mu) = X \beta + Z u
 \end{align}
@@
 
- For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
-\begin{align}
-Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
+ For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z u$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
+\begin{align}
+y' \eta - c(\eta) = \sum_i \left[ y_i \eta_i - c(\eta_i)  \right]
 \end{align}
@@
 
-With this, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
+With this family of functions, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
 \begin{verbatim}
-family.glmm$c(args)
+family.glmm$cum(args)
 family.glmm$cp(args)
@@
 
-For the Bernoulli distribution, we calculate these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) with careful computer arithmetic as follows.
-We also use the \texttt{log1p} function in R for $c(\eta_i)$.
+For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows.
+In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.
 \begin{align}
@@
 \end{align}
- For poisson, these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) are
+ For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
 \begin{align}
@@
 
-Also, I have a function to check that the data are valid given the family type.
-If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0s or 1s.
-If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.
-If the data does not pass the check, the check returns an error message.
+Also, this family of functions contains a check to ensure the data are valid given the family type.
+If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should contain only $0$ and $1$.
+If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative.
+If the data set does not pass the check, the check returns an error message.
 
@@
 \begin{align}
-\log f_\theta(y|u) &= Y' \eta +c(\eta) \\
-&= \sum_{i} Y_{i} {\eta_{i}} - c({\eta_{i}})
+\log f_\theta(y|u) &= y' \eta +c(\eta) \\
+&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
 \end{align}
@@
 \begin{align}
-\eta=X\beta+ZU.
+\eta=X\beta+Zu.
 \end{align}
@@
 \begin{align}
-\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = Y_j-c'(\eta_j).
+\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
 \end{align}
@@
 
-We'd like the derivative of the log of the data density with respect to $\beta$. This can be written using the chain rule as follows:
+We'd like the derivative of the log of the data density with respect to $\beta$. The first step is using the chain rule as follows:
 \begin{align}
 \dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
-&= \left[ Y_j-c'(\eta_j) \right]  X_{j{l}}
+&= \left[ y_j-c'(\eta_j) \right]  X_{j{l}}
 \end{align}
@@
 \begin{align}
-\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
-&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
+\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
+&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
 &= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
@@
 \begin{align}
-\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ Y-  c'(\eta)  \right].
-\end{align}
-
-Letting $ c(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
+\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ y-  c'(\eta)  \right].
+\end{align}
+
+Letting $ c''(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
 \begin{align}
@@
 \section{Random effect generation and calculations}
-This section is focused on a vector of random effects $u$ with length $q$.
-We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$.
-Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit.
+This section is focused on a vector of random effects $u$ with length $q$.  Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$.
+Section \ref{sec:ftwiddle} states the importance sampling distribution.
  Section \ref{sec:genRand} describes the process of generating the random effects.
@@
 \subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
-For this version of the package, we that assume that $D$ is diagonal.
+Recall that, for this version of the package, we assume that $D$ is diagonal.
 Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$.
-That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as a variance component.
+That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as its variance component.
 Then $D=\sum_{t=1}^T\nu_t  E_t $.
 
-If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.
-The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).
+Recall $q_t$ is the number of random effects associated with variance component $\nu_t$.
+Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.
+The model fitting function of \texttt{glmm} has made  $Z$ into a list with $T$ design matrices (one for each distinct variance component).
 In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
- Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  has $q$ rows (and columns).
-Thus, each $E_t$  has $q$ rows and columns as well.
-Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
+ Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  is $q \times q$.
+Thus, each $E_t$  is $q \times q$.
+$E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
 $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.
-
-\subsection{Generating random effects (genRand)}\label{sec:genRand}
-The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations.
+We continue this process to complete the construction of $E_t, t=1,\ldots, T$.
+
+\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
  Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
-That is, $u= D^{1/2}s$.
+That is $s$ is defined such that, $u= D^{1/2}s$.
 Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.
-The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$.
-Then let
+Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$.
+The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$.
+Let
 \begin{align}
@@
 \end{align}
-be matrices based on PQL estimates. Also, we can ``unstandardize'' our PQL-based random effects:
+be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
 \begin{align}
@@
 
- Recall that $m$ is the overall Monte Carlo sample size.
-That is, $\breve{u}_k,k=1\ldots,m$ are vectors of random effects are drawn from a multivariate standard normal.
-Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$ and let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$.
-Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
-Construct $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.
-Then we center and scale $\breve{u}_k$ as follows:
-\begin{align}
-u_k&= A\, \breve{u}_k, k=1,\ldots,m_1\\
-u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
-u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
-\end{align}
- The result or centering and scaling is that
-\begin{enumerate}
-\item $u_k,k=1,\ldots,m_1 \sim N(0,D)$
-\item $u_k,k=m_1+1,\ldots,m_1+m_2 \sim N(u^*,D^*)$
-\item $u_k,k=m_1+m2+1,\ldots,m \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
-\end{enumerate}
- Let $f(u_k|\mu,\Sigma)$ denote the pdf for $u_k \sim N(\mu,\Sigma)$. Then the importance sampling distribution $\tilde{f}$ is the following mixture distribution:
-\begin{align}
- \tilde{f}(u) = p_1 \,  f(u \, | \, 0, D)+p_2 \, f(u \, | \, u^*, D^*)+p_3 \, f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
-\end{align}
-
-\subsubsection{Rationale}
-The first distribution $N(0,D)$ was chosen to ensure a Central Limit Theorem, which is proven in  section \ref{sec:CLT}.
-The second distribution is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
-The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood Hessian.
+Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.
+Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.
+Let $\grave{f}(u)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate standard $t$ with $\zeta$ degrees of freedom.
+Then the importance sampling distribution is:
+\begin{align}
+ \tilde{f}(u) = p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
+\end{align}
+Then
+\begin{align}
+ \log \tilde{f}(u) = \log \left( p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
+\end{align}
+
+The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.
+The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
+The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.
 The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.
 
+\subsection{Generating random effects (genRand)}\label{sec:genRand}
+
+ Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. Let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$. 
+%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
+Then we draw 
+\begin{align}
+u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
+u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
+u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
+\end{align}
+Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}.
+
 \subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
-The shifting and scaling of $\breve{u})_k$ relies on finding the square root of a variance matrix.
-In the first two distributions, the variance matrices $D$ and $D^*$ are diagonal, and therefore we simply take the root of the diagonal elements.
-For the third distribution, we can use eigendecomposition.
+In order to generate draws from a nonstandard normal distribution, we need to calculate the square root of a variance matrix.
+When the variance matrix is diagonal (for example, $D^*$), we simply take the root of the diagonal elements.
+When the variance matrix is not diagonal (for example, in the second component of the mixture distribution), we can use eigendecomposition.
 Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.
+
 Let $\Sigma$ denote the nondiagonal variance matrix.
@@
 
-\subsubsection{NEED TO PUT THIS SOMEWHERE}
-
-Recall that $D$ is a diagonal with dimension $q$.
+\subsubsection{Generating from nonstandard normal distributions}\label{sec:gennonstand}
+ Construct $A^*=(D^*)^{1/2}$ and $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.
+
+Let $\breve{u}_k,k=m_1+1\ldots,m$ be vectors of random effects that are drawn from the standard normal distribution. Then we center and scale $\breve{u}_k$ as follows:
+\begin{align}
+u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
+u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
+\end{align}
+ The result or centering and scaling is that
+\begin{align}
+ &u_k \sim N(u^*,D^*), \, k=m_1+1,\ldots,m_1+m_2\\
+& u_k \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}), \, k=m_1+m2+1,\ldots,m .
+\end{align}
+
+% 
+%\subsubsection{NEED TO PUT THIS SOMEWHERE}
+%
+%
+%Recall that $D$ is a diagonal with dimension $q$.
 $D$ has $q_1$ entries being $\nu_1$, $q_2$ entries being $\nu_2$, ..., and $q_T$ entries being $\nu_T$.
 Since the distribution of $u_k=A \breve{u}_k$ is $N(0,D)$, we write its pdf as:
-\begin{align}
-f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
-&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
-&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
-\end{align}
-So that 
-\begin{align}
-\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
-&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
-\end{align}
-Then 
-\begin{align}
-\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
-\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
-\end{align}
-
-%
-%%%distribution of the random effects%%
-%\subsection{Distribution of the random effects }\label{sec:distRandsec}
-%This will be hidden from the user.
-This section provides equations for evaluating the log of a normal distribution with a specified mean and variance.
-When necessary to emphasize and clarify the choice of mean $\mu$ and variance $\Sigma$ , I use the notation $ f(u|\mu,\Sigma)$ rather than $f_\theta(u)$.
-%
-%This section is useful for calculating $\tilde{f}(u_k)$.
+%\begin{align}
+%f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
+%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
+%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
+%\end{align}
+%So that 
+%\begin{align}
+%\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
+%&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
+%\end{align}
+%Then 
+%\begin{align}
+%\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
+%\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
+%\end{align}
+%
+%Because only the first component of the mixture distribution contains any parameters,
+%\begin{align}
+%\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
+%\end{align}
+%and 
+%\begin{align}
+%\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
+%\end{align}
+%These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
+%\begin{align}
+%\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
+%&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
+%\end{align}
+%and
+%\begin{align}
+%\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
+%\end{align}
+%
+%
+
+%%
+%%This section is useful for calculating $\tilde{f}(u_k)$.
 This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives.
 These are needed for the calculation in \eqref{eq:MCLA}.
-%
-%Section \ref{sec:tee} contains details that are not currently implemented, but could be.
+%%
+%%Section \ref{sec:tee} contains details that are not currently implemented, but could be.
 In this version, all random effects are assumed to be normally distributed and all simulated random effects are generated from a normal distribution.
 
-\subsection{Distribution of random effects (\texttt{distRand})}\label{sec:distRand}
-
-In this subsection, I discuss both the assumed distribution of the unobserved random effects $N(0,D)$ and the importance sampling distribution used to generate the simulated random effects.
-The equations in this section provide  $\log f_\theta(u)$, $\nabla \log f_\theta(u)$,  $\nabla^2 \log f_\theta(u)$, $\tilde{f}(u)$, $ \nabla \tilde{f}(u)$, and $\nabla^2 \tilde{f}(u)$  for equation \ref{eq:MCLA}.
-
-Consider $N(\mu, \Sigma)$.  This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.  We can  write the distribution  as:
-\begin{align}
-\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
-\end{align}
-
-Since the first version of the package assumes $D$ is diagonal, formulas involving $D$ are found in section \ref{sec:Ddiag}.  
-Since random effects are also generated from a normal with nondiagonal variance matrix, we also need to evaluate $f(u|\mu,\Sigma)$ for general covariance matrix $\Sigma$.
-Information on this calculation for general covariance matrix $\Sigma$ are in \ref{sec:Dgeneral}).
-
-\subsubsection{$\Sigma=D$ Diagonal}\label{sec:Ddiag}
-This section provides the equations for $\log f_\theta(u)$ and  $\log \tilde{f}(u)$ and their first two derivatives.
-
+\subsection{Evaluating the distribution of random effects (\texttt{distRand})}\label{sec:distRand}
+This section discusses evaluating the distribution of the random effects, the gradient of the random effects, and the Hessian of the random effects.
+That is, the equations in this section provide  $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$  for equation \ref{eq:MCLA}.
+
+ We can  express the log unnormalized pdf of $N(0, D)$  as:
+\begin{align}
+\log f_\theta(u)= \log f (u| 0, D) = (-1/2) \log |D| - (1/2) u' D^{-1} u
+\end{align}
+
+ To find derivatives of this, we are going to use the diagonal form of $D$.
 Recall that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s  elsewhere.
@@
 Let $D_t$ be the variance matrix for $U_t$.
-$D_t$ has $q_t$ rows (and also columns).
+$D_t$ has $q_t$ rows  and $q_t$ columns.
 Thus $D$ can be expressed as:
@@
 \begin{align}
-\log f_\theta(u) &= (-1/2) \log |D| - (1/2) (U-\mu)' D^{-1} (U-\mu)\\
-&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} (U_t-\mu_t)'(U_t-\mu_t)   \right]
-\end{align}
-
-%Letting $I_{q_t}$ be the identity matrix with $q_t$ rows (and columns), we can write this as a sum as follows:
-%\begin{align}
-%\log f_\theta(u)&= \sum_{t=1}^T \log f_\theta(u_t)\\
-%&=\sum_{t=1}^T - \dfrac{1}{2} \log |D_t| - \dfrac{1}{2} (U_t-\mu_t)' D_t^{-1} (U_t-\mu_t) \\ 
-%&=\sum_{t=1}^T - \dfrac{1}{2} \log |\nu_t^{q_t}| - \dfrac{1}{2 \nu_t} (U_t-\mu_t)' I_{q_t} (U_t-\mu_t) \\ 
-%&=\sum_{t=1}^T - \dfrac{q_t}{2} \log \nu_t - \dfrac{1}{2 \nu_t}(U_t-\mu_t) '(U_t-\mu_t)
-%\end{align}
+\log f_\theta(u) &= -\dfrac{1}{2} \log |D| -\dfrac{1}{2} u' D^{-1} u\\
+&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} U_t'U_t   \right]
+\end{align}
 
@@
 \begin{align}
-\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}(U_t-\mu_t)'(U_t-\mu_t)
+\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}U_t'U_t
 \end{align}
@@
 \begin{align}
-\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} (U_t-\mu_t)'(U_t-\mu_t).
+\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} U_t'U_t.
 \end{align}
@@
 \begin{align}
-\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} (U_1-\mu_1) ' (U_1-\mu_1) & ...
-& - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} (U_T-\mu_T) '(U_T-\mu_T)   \end{bmatrix}
+\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} U_1'U_1 & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} U_T'U_T   \end{bmatrix} 
 \end{align}
@@
 
-The last thing we need to discuss is how to split $U$ into $U_1,...,U_T$ and $\mu$ into $\mu_1,...,\mu_T$.
-The same process should work for both, so let's just look at how we'd do it for $U$.
+The last thing we need to discuss is how to split $U$ into $U_1,...,U_T$.
 We know that the first $q_1$ items of U are $U_1$, the next $q_2$ items are $U_2$, etc.
@@
 Items $q_1+1$ through $q_1+q_2$ are $U_2$, etc.
-In R pseudo code, this means
-\begin{verbatim}
-nrand<-lapply(z,ncol) #get the number of columns from each matrix in list z
-unlist(nrand) #chance the list into a vector, since each entry in the list is just a scalar
-
-U1<- U[1 through nrand[1]] #first rule is use 1:nrand[1]
-Ut<-U[sum(nrand[1:t-1]+1 through sum(nrand[1:t])]
-\end{verbatim}
-
-Note that we don't need to actually calculate either A or D. We work only through $\nu$.
-
-\subsubsection{Newer version of $\Sigma=D$ Diagonal}
-The newer version of \texttt{distRand} (called \texttt{distRand3}) uses the same equations as those listed in \ref{sec:Ddiag}.
-The only difference is that the newer version is more efficient.
-Rather than having each call to \texttt{distRand3} recalculate which elements of $U$ belong to each $t$, this is done ahead of time in the objective function.
-This information is contained in vector \texttt{meow}.
-This is then an argument in \texttt{distRand3}.
+We find these numbers ($q_1$, $q_2$, etc) from the number of columns of the items in the list \texttt{z}.
+This information of how to split $u$ up is contained in vector \texttt{meow} and used by function \texttt{distRand3}.
+This is faster than recalculating \texttt{meow} for every call of \texttt{distRand3}.
 
@@
 
-The result of invoking the C function is  the calculation of the gradient and hessian for the distribution of the random effects. The value is  evaluated by the C function in \ref{sec:Dgeneral}.
+The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and hessian for the distribution of the random effects.
+The value is  evaluated by the C function \texttt{distRandGeneral}, described in section \ref{sec:Dgeneral}.
+
+\subsection{Evaluating the importance sampling distribution}
+This section discusses evaluating the importance sampling distribution shown in \ref{eq:ftwiddle}.
+Section \ref{sec:Dgeneral} describes how to evaluate the pdf for the normal components of $\tilde{f}(u)$.
+Section \ref{sec:tstuff} describes how to evaluate the pdf for the t component of $\tilde{f}(u)$.
+The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.
 
 %%distRandGeneral
-\subsubsection{$\Sigma$ in general (\texttt{distRandGeneral})}\label{sec:Dgeneral}
- Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
-\begin{align}
-\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
+\subsubsection{Normal distribution for general variance matrix $\Sigma$ (\texttt{distRandGeneral})}\label{sec:Dgeneral}
+Let $\Sigma$ be a variance matrix and let $\mu$ be a mean vector. Consider $N(\mu, \Sigma)$.   We can  write the log pdf of this distribution  as:
+\begin{align}
+\log f (u| \mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
 \end{align}
@@
 which is just the product of the eigenvalues.
-The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times each optimization iteration, where $m$ is again the Monte Carlo sample size.
+The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times throughout each trust optimization iteration, where $m$ is again the Monte Carlo sample size.
 
 This function is also rewritten in C with the following  passed in as pointers: double \texttt{Sigma.inv} $\Sigma^{-1}$, double \texttt{logdet} $\log |\Sigma^{-1}|$, int \texttt{nrow}, double \texttt{uvec} a vector of random effects, double {mu} $\mu$, and double \texttt{distRandGenVal}.
+
+\subsubsection{t distribution}\label{sec:tstuff}
+
+Consider $t_\zeta$. We can write the log unnormalized pdf of this distribution as:
+\begin{align}
+\log \grave{f}(u)= \left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}
+\end{align}
 
@@
 %\end{align}
-
-\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
-Recall that the importance sampling distribution, first stated explicitly in equation \ref{eq:ftwiddle}, is:
-\begin{align}
- \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) 
-\end{align}
-Then
-\begin{align}
- \log \tilde{f}(u) = \log \left( p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
-\end{align}
-Because only the first component of the mixture distribution contains any parameters,
-\begin{align}
-\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
-\end{align}
-and 
-\begin{align}
-\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
-\end{align}
-These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
-\begin{align}
-\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
-&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
-\end{align}
-and
-\begin{align}
-\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
-\end{align}
 
@@
 The calculations in \ref{sec:calcs} are for $\tilde{f}(u_k)$ containing $\theta$ and the calculations in \ref{sec:calcsindep} are for $\tilde{f}(u_k)$ not containing $\theta$.
+
+\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}\label{sec:calcsindep}
+
+When $\tilde{f}$ is independent of $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
+\begin{align}
+\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
+\end{align}
+and the Hessian matrix of the MCLA is
+\begin{align}
+\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
+&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
+\end{align}
+To reduce the risk of catastrophic cancellation, we can combine the last two terms of the Hessian:
+\begin{align}
+\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
+\end{align}
+We are able to combine the last two terms because $\nabla l_m(\theta)$ is a weighted mean of $\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)$. Letting
+\begin{align}
+Z&=\nabla \log f_\theta (y,u_k)
+\end{align}
+ we can use the following equality:
+\begin{align}
+E(ZZ')-E(Z)E(Z)' = \left[ E(Z-EZ)  \right]\left[ E(Z-EZ)  \right]'.
+\end{align}
 
@@
 
-\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}\label{sec:calcsindep}
-
-When $\tilde{f}$ is independent of $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
-\begin{align}
-\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
-\end{align}
-and the Hessian matrix of the MCLA is
-\begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
-&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
-\end{align}
-
 \section{Central Limit Theorem for MCLA}\label{sec:CLT}

### 2015-01-14 NewDesignDoc/DesignDocGLMM.tex
Completed first round of design doc edit/revamp

@@
 \nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
-\end{align}
-Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.
+&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \label{eq:MCLAhessian}
+\end{align}
+Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and Hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.
 
@@
 
-The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.
+The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the Hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.
 
@@
 Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.
-The model fitting function of \texttt{glmm} has made  $Z$ into a list with $T$ design matrices (one for each distinct variance component).
-In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
- Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  is $q \times q$.
-Thus, each $E_t$  is $q \times q$.
+The model fitting function of \texttt{glmm} has made  \texttt{z} into a list with $T$ design matrices (one for each distinct variance component).
+In order to create $E_t$, we need to  count the  number of columns ($q_t$) for each design matrix in the list.
+Since $D$  is $q \times q$, each $E_t$  is $q \times q$.
 $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
@@
 That is $s$ is defined such that, $u= D^{1/2}s$.
-Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.
+Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t},t=1,\ldots,T$.
 Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$.
@@
 \end{align}
-Then
-\begin{align}
- \log \tilde{f}(u) = \log \left( p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
-\end{align}
 
@@
 The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.
-The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.
+The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y,u)$.
 
@@
 
- Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. Let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$. 
+ Recall that $m$ is the overall Monte Carlo sample size.
+Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.
+We randomly determine $m_1,m_2,m_3$ according to the proportions $p_1,p_2,p_3$.
+More specifically, we generate $m$ random numbers from a distribution that is uniform between $0$ and $1$.
+Then $m_1$ is the frequency with which the randomly generated numbers are between $0$ and $p_1$,  $m_2$ is the frequency with which the randomly generated numbers are between $p_1$ and $p_1+p_2$, and $m_3=m-p_1-p_2$.
 %Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
-Then we draw 
+
+Once $m_1, m_2$ and $m_3$ are determined, we draw 
 \begin{align}
@@
 \end{align}
-Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}.
+Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}. We use R package \texttt{mvtnorm} to draw from a $t$ distribution.
 
@@
 
-Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting hessian.
-
-The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and hessian for the distribution of the random effects.
+Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting Hessian.
+
+The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and Hessian for the distribution of the random effects.
 The value is  evaluated by the C function \texttt{distRandGeneral}, described in section \ref{sec:Dgeneral}.
@@
 
+Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int, and \texttt{logft} as a pointer.
+
 %\subsection{$\tilde{f}(u_k)$}
@@
 The objective function is optimized by \texttt{trust} within the the model fitting function.
-In order to evaluate and maximize the approximated log likelihood, I need an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.
-Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, \ref{eq:MCLAhessian}   for calculating these quantities.
-This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.
+In order to evaluate and maximize the approximated log likelihood, \texttt{trust} needs an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.
+Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, and \ref{eq:MCLAhessian}   for calculating these quantities.
+%This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.
 
@@
 \begin{align}
-b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
-&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \label{eq:MCLA}
-\end{align}
-
-For computational stability, we'll set $a=max(b_k)$. Then the value of the MCLA is expressed as:\\
+b_k = \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k). \label{eq:MCLA}
+\end{align}
+
+For computational stability, we'll set $a=$max$(b_k)$. Then the value of the MCLA is expressed as:\\
 \begin{align}
@@
 \begin{align}
-\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
-&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
+\nabla l_m(\theta) &= \sum_{k=1}^m  \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
+&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)   \right] v_\theta(u_k,y)\\
+&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
 &= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
-&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
-&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
-&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
+&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]   \right) v_\theta(u_k,y)   \\
+&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(u_k)  \right] v_\theta(u_k,y) \label{eq:MCLAgr}
 \end{align}
@@
 \begin{align}
-&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
+&\nabla^2 l_m(\theta)=  \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)   v_\theta(u_k,y) \label{eq:MCLAh} \\ 
 &+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
+    \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
 \end{align}
@@
 
+%HERE IS WHEN FTWIDDLE DEPENDS ON NU
+%Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
+%\begin{align}
+%\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
+%&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
+%&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
+%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
+%&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
+%&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
+%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
+%\end{align}
+%
+%Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
+%\begin{align}
+%&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
+%   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
+%&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
+%   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
+%   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
+%\end{align}
+%Everything in this has already been defined except:
+%\begin{align}
+%\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
+%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} +
+%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
+%&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
+%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
+%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
+%\end{align}
+%
+%
+
 \section{Summary of model}
-Typing \texttt{summary(mod)}  gives more detailed info of the model.
-This is broken into two pieces (as is usually done for summaries):  \texttt{summary.glmm} and \texttt{print.summary.glmm}.
-We split this up because we have a lot of information in the summary that we do not necessarily want printed each time.
+Typing \texttt{summary(mod)}  provide a summary of the model.
+This is broken into two pieces (as is usually done for summaries in R):  \texttt{summary.glmm} and \texttt{print.summary.glmm}.
+The first contains more calculated results than the second.
 When a user types \texttt{summary(mod)}, only the  basic information is automatically printed.
@@
 \begin{itemize}
-\item call
-\item the variance estimate(s)
-\item a matrix with the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column.
+\item the call.
+\item a matrix with the predictor name in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided p-value  in the fifth column.
 All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
-\item the evaluated Monte Carlo log likelihood along with its first and second derivative
-\item maybe other things that I haven't thought of
+\item a matrix with the variance estimate name  in the first column, the estimated variance component in the second column, the standard error in the third column, the z value in the fourth column, and the one-sided p-value  in the fifth column.
+All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
+\item the evaluated Monte Carlo log likelihood along with its first and second derivative.
+\item $m_1$, $m_2$, $m_3$.
+\item output from texttt{trust}, such as whether it converged. (See \texttt{trust} for more details.)
+\item the generated random effects.
 \end{itemize}
 
-A note on the standard errors: to calculate the standard errors, we take the MCLA Hessian matrix, invert it, take the diagonal elements, and square root them.
-It's possible that the Hessian will be noninvertible if it is close enough to singular that the computer thinks it's singular.
-Then the standard errors will all be infinite.
-We also need to warn the user why this is happening.
+A note on the standard errors: to calculate the standard errors, we take the MCLA Hessian matrix, invert it, take the diagonal elements, and apply the square root to them.
+It is possible that the Hessian will be noninvertible if it is close enough to singular that the computer thinks it is singular.
+Then the standard errors will all be infinite  the user is warned.
 
@@
 \begin{itemize}
-\item call
-\item the variance estimate(s)
-\item a matrix similar to the output of summary.glm, which will be put through printCoefmat in order to get the significance stars that we're used to.
-We'll have the predictor in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided pvalue  in the fifth column, the significance stars, and the significance legend.
-\item maybe other things that I haven't thought of. I'll probably realize more once I get coding.
+\item the call.
+\item a matrix with the predictor name in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided asymptotic p-value  in the fifth column.
+\item a matrix with the variance estimate name  in the first column, the estimated variance component in the second column, the standard error in the third column, the z value in the fourth column, and the one-sided p-value  in the fifth column.
+All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
 \end{itemize}
 
-I think for the printCoefmat to work, the fixed effects matrix needs to look like
-\begin{verbatim}
-            Estimate Std. Error z value     Pr(>|z|)
-(Intercept)        2       0.50       4 6.334248e-05
-x1                 2       0.25       8 1.244192e-15
-\end{verbatim}
-(or at least it worked when I had these as the column names and didn't work before that).
-
-Note that the summary and print.summary functions will be S3 generic functions.
+%I think for the printCoefmat to work, the fixed effects matrix needs to look like
+%\begin{verbatim}
+%            Estimate Std. Error z value     Pr(>|z|)
+%(Intercept)        2       0.50       4 6.334248e-05
+%x1                 2       0.25       8 1.244192e-15
+%\end{verbatim}
+%(or at least it worked when I had these as the column names and didn't work before that).
+
+Note that the summary and print.summary functions are S3 generic functions.
 This means is the user  types \texttt{summary(mod)} and \texttt{summary} is a generic function.
-R checks the class of the model \texttt{mod} and then automatically uses the \texttt{summary} and \texttt{print.summary} functions for that class of objects.
+R checks the class of the model \texttt{mod} and  automatically uses the \texttt{summary} and \texttt{print.summary} functions for that class of objects.
 
 \section{PQL}\label{sec:pql}
- Before we get started on describing PQL, we need to change notation to avoid constrained optimization.
-Recall that $D=\var(u)$ and is (for now) assumed to be diagonal.
+ This section on PQL requires a change in notation so that we can avoid constrained optimization.
+Recall that $D=\var(u)$ and is assumed to be diagonal.
 Let $A=D^{1/2}$ so that A has diagonal components that are positive or negative.
@@
 Recall that $E_t$ has a diagonal of indicators to show which random effects have the same variance components, and $\sum_{t=1}^T E_t$ is the identity matrix.
-PQL will estimate the components contained on the diagonal of A.
-Taking the absolute value of those components will provide the standard deviations (the square root of the variance components).
-
-In addition to using $A$ rather than $D$, the other change is that we'll be using $s$ where $u=As$. The purpose of this is to avoid $D^{-1/2}$ in the functon we're trying to optimize.
-
-There are two ways to do PQL, both of which are described in the vignette
+PQL  estimates the components contained on the diagonal of A.
+Taking the absolute value of those components  provides the standard deviations (where the standard deviations are the square root of the variance components).
+
+We also define $s$ where $u=As$. The purpose of using $s$ rather than $u$ is to avoid $D^{-1/2}$ in the objective functon that we optimize.
+
+There are two variations of PQL, both of which are described in the vignette
  \texttt{re.pdf} in the R package \texttt{aster} (Geyer, 2014).
-In either case, there is an inner optimization and an outer optimization.
-The inner optimization is well behaved while the other optimization is a little tougher.
-I will be using the method that is not quite PQL but is pretty close and is better behaved.
+Both variations have an inner optimization and an outer optimization.
+The inner optimization is a well behaved quadratic while the outer optimization is more challenging.
+I use the variation of PQL that is not the original version, but is pretty close and is better behaved.
 In this version, the inner optimization finds $\tilde{\beta}$ and $\tilde{s}$ given $X$, $Z$ and $A$.
@@
 
-The {\bf inner} optimization will be done with the trust function in R.
-We elect to use trust because it requires two derivatives, which will make the optimization more precise.
-We would like more accuracy in the inner maximization because any sloppiness will carry into the outer optimization.
+The {\bf inner} optimization is done with the \texttt{trust} function in R.
+We chose to use \texttt{trust} because it requires two derivatives, which  make the optimization more precise.
+We would like more accuracy in the inner maximization because any imprecision carries into the outer optimization.
 
@@
 \begin{align}
-l(\eta)= Y' \eta - c(\eta) 
+l(\eta)= y' \eta - c(\eta) 
 \end{align}
@@
 \end{align}
-This function will be maximized using the \texttt{trust} package.
-We need to give \texttt{trust} derivatives with respect to $s$ and $\beta$.
-We're going to express these via the multivariate chain rule, taking advantage of $\eta_i$.
-
-Create  vector $\mu$ from the components $\mu_i=c'(\eta_i)$.  Since
+Since this function is maximized using  \texttt{trust}, we need to provide derivatives with respect to $s$ and $\beta$. We express these via the multivariate chain rule, taking advantage of $\eta_i$.
+
+Create  vector $\mu$ with components $\mu_i=c'(\eta_i)$.  Since
 \begin{align}
@@
 &= \dfrac{\partial l(\eta)}{\partial \eta_i} \dfrac{\partial \eta_i}{\partial \beta_k}    \\
-&=\sum_i (Y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
-&=\sum_i (Y_i-\mu_i) X_{ik}
+&=\sum_i (y_i-\mu_i) \dfrac{\partial \eta_i}{\partial \beta_k} \\
+&=\sum_i (y_i-\mu_i) X_{ik}
 \end{align}
@@
 &= \dfrac{\partial l(\eta)}{\partial \eta} \dfrac{\partial \eta}{\partial s} -s\\
-&= (Y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
-&=(Y-\mu)'ZA -s
+&= (y-\mu)' \left[ \dfrac{\partial}{\partial s} ZAs \right] -s\\
+&=(y-\mu)'ZA -s
 \end{align}
@@
 \begin{align}
-\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (Y-\mu)\\
-\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (Y-\mu)  -s
+\dfrac{\partial}{\partial \beta} \left[ l(\eta)-(1/2)s's \right]&= X' (y-\mu)\\
+\dfrac{\partial}{\partial s} \left[ l(\eta)-(1/2)s's \right]&= AZ' (y-\mu)  -s
 \end{align}
@@
 
-We'll start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
-\begin{align}
-  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (Y-c'(\eta))'ZA -s   \right]\\
+We start at the top with  $ \dfrac{\partial^2}{\partial s^2}$, which is a $q \times q$ matrix.
+\begin{align}
+  \frac{\partial^2}{\partial s^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \frac{\partial}{\partial s} \left[ (y-c'(\eta))'ZA -s   \right]\\
 &=\left[- \frac{\partial}{\partial s} c'(\eta)\right]'ZA - I_q \\
@@
 
-Next up is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
-\begin{align}
-  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (Y-c'(\eta))  \right]\\
-&= \dfrac{\partial}{\partial \beta} \left[ X' Y-X'(c'(\eta))  \right] \\
+Next  is  $ \dfrac{\partial^2}{\partial \beta^2}$, which is a $p \times p$ matrix.
+\begin{align}
+  \dfrac{\partial^2}{\partial \beta^2}  \left[ l(\eta) - (1/2) s's   \right] &=   \dfrac{\partial}{\partial \beta} \left[ X' (y-c'(\eta))  \right]\\
+&= \dfrac{\partial}{\partial \beta} \left[ X' y-X'(c'(\eta))  \right] \\
 &= \dfrac{\partial}{\partial \beta} \left[ -X'(c'(\eta))  \right] \\
@@
 \begin{align}
-\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ Y-c'(\eta)  \right]' ZA  \right\}\\
-&= \dfrac{\partial}{\partial \beta} \left\{Y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
+\dfrac{\partial^2}{ \partial \beta \;  \partial s}  \left[ l(\eta) - (1/2) s's   \right] &= \dfrac{\partial}{\partial \beta} \left\{  \left[ y-c'(\eta)  \right]' ZA  \right\}\\
+&= \dfrac{\partial}{\partial \beta} \left\{y'ZA-  \left[ c'(\eta)  \right]' ZA  \right\} \\
 &=- \dfrac{\partial}{\partial \beta} \left\{  \left[ c'(\eta)  \right]' ZA  \right\} \\
@@
 And last we have the $q \times p$ mixed partial $\dfrac{\partial^2}{  \partial \beta \; \partial s}= -AZ' [c''(\eta)] X.$  These four pieces specify the hessian matrix for the penalized likelihood.
-Trust can now perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
-Trust will maximize the penalized likelihood.
+Now \texttt{trust} can perform the inner optimization to find $\tilde{\beta}$ and $\tilde{s}$.
 
@@
 \tilde{\eta} &= X \tilde{\beta} +ZA \tilde{s}\\
-l(\tilde{\eta}) &= Y' \tilde{\eta} - c(\tilde{\eta})\\
+l(\tilde{\eta}) &= y' \tilde{\eta} - c(\tilde{\eta})\\
 A&= \sum_k E_k \sigma_k.
 \end{align}
-Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we'd like to maximize is the penalized quasi-likelihood:
-\begin{align}
- l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| \\
+Let $W$ be a diagonal matrix with elements $c''(\eta_i)$ on the diagonal. Then the quantity we would like to maximize is the penalized quasi-likelihood:
+\begin{align}
+ l(\tilde{\eta}) - \dfrac{1}{2} \tilde{s}' \tilde{s} - \dfrac{1}{2} \log  \left| AZ' W ZA +I  \right| 
 \end{align}
@@
 
- If I am worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
+ If I become worried about all variance components being zero, I could implement an eigendecomposition using the R function \texttt{eigen}.
 This would be more numerically stable, but is slower.
@@
 The outer optimization  uses $\tilde{\beta}$ and $\tilde{s}$ provided by the inner optimization, but it does not return them.
-To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call ``cache.''  The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
+To keep track of the most recent  $\tilde{s}$, so store them in an environment that I call \texttt{cache}.
+The purpose of  $\tilde{\beta}$ and $\tilde{s}$ is two-fold.
 First, if they are available from a previous iteration of the inner optimization, then they are used in the outer optimization of PQL.
@@
 
-The point of doing PQL is to construct a decent importance sampling distribution.
-Thus, the estimates don't have to be perfect.
+The point of utilizing PQL is to construct a decent importance sampling distribution.
+Thus, the estimates do not have to be perfect.
 It is possible that one of the $\sigma_k$ will be 0 according to PQL.
-If this happens, then I'll just use $\sigma_k = .01$ or something like that for the importance sampling distribution.
+If this happens, then I just use $\sigma_k = .01$ for the importance sampling distribution.
 
@@
 
-\section{Likelihood ratio test}
-This is another S3 generic function that the user can choose to implement if they'd  like  to do likelihood ratio tests for nested models.
+%%%%%%LIKELIHOOD RATIO TEST STUFF
+%\section{Likelihood ratio test}
+%This is another S3 generic function that the user can choose to implement if they'd  like  to do likelihood ratio tests for nested models.
 Eventually I'll want this to handle an arbitrary number of models, but for now we'll stick to comparing two models: $mod2$ nested in $mod1$.
 I will assume these models have already been fit.
-
- There are a few ways that the two models could differ. In other words, we could be testing:
- \begin{enumerate}
- \item whether one or more fixed effects are zero.
- \item whether one variance component is zero.
- \item whether multiple variance components are zero.
- \item whether one or more fixed effects is zero and one or more variance components are zero. 
-\end{enumerate}
-The hypothesis testing procedure and method for calculating pvalues are different for each of these cases, so I'll go through each one in the following subsubsections.
-
-\subsection{Testing whether one or more fixed effects are zero}
-Consider the  case of testing whether one or more fixed effects are zero and the random effects are the same between the two models.
+%
+%
+% There are a few ways that the two models could differ. In other words, we could be testing:
+% \begin{enumerate}
+% \item whether one or more fixed effects are zero.
+% \item whether one variance component is zero.
+% \item whether multiple variance components are zero.
+% \item whether one or more fixed effects is zero and one or more variance components are zero. 
+%\end{enumerate}
+%The hypothesis testing procedure and method for calculating p-values are different for each of these cases, so I'll go through each one in the following subsubsections.
+%
+%\subsection{Testing whether one or more fixed effects are zero}
+%Consider the  case of testing whether one or more fixed effects are zero and the random effects are the same between the two models.
 I think we just do a likelihood ratio test as we're familiar with for linear models.
@@
 Let's call the log likelihood from the larger model  $l_1$ and the log likelihood from the nested model  $l_2$.
-
-Then the likelihood ratio test statistic is
-\begin{align}
-t_{LRT}= -2l_1+2l_2
-\end{align}
-and this follows a $\chi^2$ distribution with $p_1-p_2$ degrees of freedom. The calculation
-\begin{align}
-P(t_{\chi^2_{p_1-p_2}>LRT} ).
-\end{align}
-gives us a two-sided pvalue to fit with the two-sided alternative hypothesis.
-
- \subsection{Testing whether one variance component is zero}
-  Hypothesis testing for one variance component is easy but not what most people expect.
-  In this setup, the two models have the exact same fixed effects.
+%
+%Then the likelihood ratio test statistic is
+%\begin{align}
+%t_{LRT}= -2l_1+2l_2
+%\end{align}
+%and this follows a $\chi^2$ distribution with $p_1-p_2$ degrees of freedom. The calculation
+%\begin{align}
+%P(t_{\chi^2_{p_1-p_2}>LRT} ).
+%\end{align}
+%gives us a two-sided p-value to fit with the two-sided alternative hypothesis.
+%
+% 
+% \subsection{Testing whether one variance component is zero}
+%  Hypothesis testing for one variance component is easy but not what most people expect.
+%  In this setup, the two models have the exact same fixed effects.
 The only difference is the larger model has one more random effect $\nu_{t_1}$.
@@
 In other words, the hypotheses are:
- \begin{align}
- H_0: \nu_{t_1}=0\\
- H_1: \nu_{t_1}>0.
- \end{align}
- Note the alternative hypothesis is one-sided. This means we need to calculate a one-sided pvalue. Let $t_{LRT}=2l_2-2l_1$. If we naively follow the pvalue calculation of
- \begin{align}
-P( \chi^2_{1}>t_{LRT})
-\end{align}
- we will end up calculating a two-sided pvalue.
-To fix this, we cut this pvalue in half.
-In other words, use the standard normal distribution $Z$ and think of calculating the one-sided pvalue this way:
- \begin{align}
- P( Z>\sqrt{|  t_{LRT} |})
- \end{align}
-
- However, this is not obvious; I never would have thought about having a test statistic of
- \begin{align}
- \sqrt{2 | l_2-l_1   |   }.
- \end{align}
-
- \subsection{Testing whether multiple variance components are zero}
- In this case, the fixed effects are identical between the larger and nested model.
+% \begin{align}
+% H_0: \nu_{t_1}=0\\
+% H_1: \nu_{t_1}>0.
+% \end{align}
+% Note the alternative hypothesis is one-sided. This means we need to calculate a one-sided p-value. Let $t_{LRT}=2l_2-2l_1$. If we naively follow the p-value calculation of
+% \begin{align}
+%P( \chi^2_{1}>t_{LRT})
+%\end{align}
+% we will end up calculating a two-sided p-value.
+To fix this, we cut this p-value in half.
+In other words, use the standard normal distribution $Z$ and think of calculating the one-sided p-value this way:
+% \begin{align}
+% P( Z>\sqrt{|  t_{LRT} |})
+% \end{align}
+% 
+% However, this is not obvious; I never would have thought about having a test statistic of
+% \begin{align}
+% \sqrt{2 | l_2-l_1   |   }.
+% \end{align}
+% 
+% \subsection{Testing whether multiple variance components are zero}
+% In this case, the fixed effects are identical between the larger and nested model.
 The only difference is that the larger model has more than one additional variance components.
@@
 Charlie is hoping that we'll have a mixture of $\chi^2$ distributions, such as
- \begin{align}
- \dfrac{1}{2}P (\chi^2_1> t_{LRT}) +\dfrac{1}{4}P (\chi^2_2> t_{LRT}).
- \end{align}
- The reason this is complicated is because the variance components are restricted to be nonnegative.
+% \begin{align}
+% \dfrac{1}{2}P (\chi^2_1> t_{LRT}) +\dfrac{1}{4}P (\chi^2_2> t_{LRT}).
+% \end{align}
+% The reason this is complicated is because the variance components are restricted to be nonnegative.
 This means that the likelihood is only defined for nonnegative variance components.
 Then differentiating the likelihood at $0$ becomes tricky because that's the boundary.
-
- \subsection{Testing whether one or more fixed effects is zero and one or more variance components are zero}\label{sec:combotest}
- This seems very complicated. I don't know if it's any more complicated than the previous case of testing multiple variance components. I'll have to think about it later when I have time.
-
- \subsection{Determining the hypothesis test}
- To follow convention, this command will be called ``anova'' and the two arguments will be the two models to be compared.  The command would look like
-\begin{verbatim}
-anova(mod1,mod2)
-\end{verbatim}
-
- R will first need to figure out if the fixed effects are different and, if they are, which model is the larger model.
-\begin{verbatim}
-if(length(coef(mod1))>length(coef(mod2)))
-      {bigmod<-mod1;  smallmod <-mod2} 
-if(length(coef(mod1))<length(coef(mod2)))
-      {bigmod<-mod2; smallmod <-mod1}
-\end{verbatim}
-Next, if there is a difference in the fixed effects, I'm going to make sure the fixed effects of the small model are nested in those of the big model.
+%
+% \subsection{Testing whether one or more fixed effects is zero and one or more variance components are zero}\label{sec:combotest}
+% This seems very complicated. I don't know if it's any more complicated than the previous case of testing multiple variance components. I'll have to think about it later when I have time.
+% 
+% \subsection{Determining the hypothesis test}
+% To follow convention, this command will be called ``anova'' and the two arguments will be the two models to be compared.  The command would look like
+%\begin{verbatim}
+%anova(mod1,mod2)
+%\end{verbatim} 
+%
+% R will first need to figure out if the fixed effects are different and, if they are, which model is the larger model.
+%\begin{verbatim}
+%if(length(coef(mod1))>length(coef(mod2)))
+%      {bigmod<-mod1;  smallmod <-mod2} 
+%if(length(coef(mod1))<length(coef(mod2)))
+%      {bigmod<-mod2; smallmod <-mod1}
+%\end{verbatim}
+%Next, if there is a difference in the fixed effects, I'm going to make sure the fixed effects of the small model are nested in those of the big model.
 Otherwise, produce an error.
@@
 Note: this check for nesting isn't perfect, but no other anova checks which model is bigger and whether they're nested.
-\begin{verbatim}
-if(big mod is defined) {
-      pnames1<-names(coef(bigmod))
-      pnames2<-names(coef(smallmod))
-      if(sum(pnames2 %in% pnames1) != length(pnames2)) {
-            stop("The models you provided are not nested.")}
-
-     if(the variance components differ between the two models){
-             check big model has more variance components (ow: error)
-             check variance components of the small model are nested (ow: error)
-             calculate pvalue according to section 3.3.4. return it.
-      }
-      if(the variance components do not differ between the two models){
-            calculate pvalue according to 3.3.1. return it.
-      }
-}
-\end{verbatim}
-If and only if we've gotten to the end of this chunk of code without returning anything, bigmod has not been defined, meaning the fixed effects are same.
+%\begin{verbatim}
+%if(big mod is defined) {
+%      pnames1<-names(coef(bigmod))
+%      pnames2<-names(coef(smallmod))
+%      if(sum(pnames2 %in% pnames1) != length(pnames2)) {
+%            stop("The models you provided are not nested.")}
+%
+%     if(the variance components differ between the two models){
+%             check big model has more variance components (ow: error)
+%             check variance components of the small model are nested (ow: error)
+%             calculate p-value according to section 3.3.4. return it.
+%      }
+%      if(the variance components do not differ between the two models){
+%            calculate p-value according to 3.3.1. return it.
+%      }
+%}
+%\end{verbatim}
+%If and only if we've gotten to the end of this chunk of code without returning anything, bigmod has not been defined, meaning the fixed effects are same.
 Therefore, we now need to figure out whether the variance components differ by one or by more than one.
@@
 I  don't yet have a solid grasp of how formula works its magic, but I have the impression that I can count the number of model matrices returned in order to figure out the number of variance components because there is a model matrix for each one.
-
-\begin{verbatim}
-T1<- number of variance components for mod1
-T2<- number of variance components for mod2
-if(T1>T2){call mod1 the bigmod and mod2 the smallmod}
-if(T1<T2){call mod2 the bigmod and mod1 the smallmod}
-if(bigmod is defined){
-     check  small model is nested in the big model (ow: error)
-     (maybe do this by looking at the call?)
-     if(T1==T2+1 || T2==T1+1) 
-           {calculate and return pvalue according to sec 3.3.2}     
-     calculate and return pvalue according to section 3.3.3
-     }
-if still haven't returned anything, produce error.     
-\end{verbatim}
-
-Why the last error?
+%
+%\begin{verbatim}
+%T1<- number of variance components for mod1
+%T2<- number of variance components for mod2
+%if(T1>T2){call mod1 the bigmod and mod2 the smallmod}
+%if(T1<T2){call mod2 the bigmod and mod1 the smallmod}
+%if(bigmod is defined){
+%     check  small model is nested in the big model (ow: error)
+%     (maybe do this by looking at the call?)
+%     if(T1==T2+1 || T2==T1+1) 
+%           {calculate and return p-value according to sec 3.3.2}     
+%     calculate and return p-value according to section 3.3.3
+%     }
+%if still haven't returned anything, produce error.     
+%\end{verbatim}	
+%
+%Why the last error?
 If we have gotten to this point in the code with nothing returned, it ends up that the number of variance components in each model are equal, meaning the user has made some kind of mistake.
@@
 Either way, we can't help them.
-
-This command would return something that first reminds the user what predictors are in each model, then has a table.
+%
+%
+%
+%
+%
+%This command would return something that first reminds the user what predictors are in each model, then has a table.
 Each model would have one row of the table.
 The columns would contain the model name,  the log likelihood for each model, the number of parameters in the model.
-Then the next columns would have one entry each: the calculated difference between the log likelihoods, the calculated difference between the number of parameters, and the pvalue of the test.
+Then the next columns would have one entry each: the calculated difference between the log likelihoods, the calculated difference between the number of parameters, and the p-value of the test.
 
@@
 
-I thought for three or four minutes about a one sided confidence interval, but I don't think that captures what we want.
-We want the range of all likely values for the variance component, not just an upper bound.
-
 %
@@
 \subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$} \label{sec:calcs}
+This section is still under construction.
+
 When $\tilde{f}$ contains $\theta$, the gradient of the MCLA is
@@
 \section{Central Limit Theorem for MCLA}\label{sec:CLT}
-In this section, we focus on the gradient of the MCLA because the trust criterion for finding the maximum is based on the gradient.
+In this section, we focus on the gradient of the MCLA because the trust criterion for finding the maximum is based on the gradient when $\tilde{f}(u)$ is independent of $\theta$. 
 Define
@@
 %
-%Next, we need to show that the hessian of the MCLA has finite variance. Again, it is enough to show that
+%Next, we need to show that the Hessian of the MCLA has finite variance. Again, it is enough to show that
 %\begin{align}

### 2015-01-14 NewDesignDoc/DesignDocGLMM.tex
Changed standard t to t mean 0, scale matrix D*
Where D^* is the PQL guess of the variance matrix

@@
 Next, the MCLA is maximized to find Monte Carlo maximum likelihood estimates, the corresponding Fisher Information, and other statistics.
-Additional inference is then possible, including confidence intervals for the parameters and likelihood ratio tests for comparing nested models.
+Additional inference is then possible, including confidence intervals for the parameters.
 \end{abstract}
@@
 Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data.
-Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects centered at 0 with variance matrix $D$.
+Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved, normally-distributed random effects centered at 0 with variance matrix $D$.
 Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$.
@@
 Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.
-Let $\grave{f}(u)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate standard $t$ with $\zeta$ degrees of freedom.
+Let $\grave{f}(u|0,D^*)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate  $t$ with mean $0$, scale matrix $D^*$, and $\zeta$ degrees of freedom.
 Then the importance sampling distribution is:
 \begin{align}
- \tilde{f}(u) = p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
+ \tilde{f}(u) = p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
 \end{align}
@@
 \begin{align}
-\log \grave{f}(u)= \left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}
-\end{align}
-
-Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int, and \texttt{logft} as a pointer.
+\log \grave{f}(u)&=\log \left[ 1+u'D^* u/\zeta  \right]^{-(\zeta+q)/2}\\
+&=-\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
+&= -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t}\right)
+\end{align}
+
+Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int array, \texttt{meow} as an int array, and \texttt{logft} as a pointer.
 
@@
 &\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u)} du\\
-&\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, f_\theta(y|u)^2  }{\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}} du
+&\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, f_\theta(y|u)^2  }{\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du
 \end{align}
@@
 \begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, \exp(2y'\eta)  }{\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2} \,  \exp(2c(\eta))  } du\\
-&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'\eta)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du \\
-&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du 
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, \exp(2y'\eta)  }{\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2} \,  \exp(2c(\eta))  } du\\
+&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'\eta)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du \\
+&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du 
 \end{align}
@@
 \begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  (1+e^\eta) ^2  } du,
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  (1+e^\eta) ^2  } du,
 \end{align}
@@
 \begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,   (\exp(\exp(\eta)))^2  } du,
+\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,   (\exp(\exp(\eta)))^2  } du,
 \end{align}
@@
 &\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u)} du \\
-&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'u/\zeta  \right]^{-(\zeta+q)/2}} du \\
-&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
-&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
+&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du \\
+&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
+&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
 \end{align}
@@
 \begin{align}
- & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
-&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
+ & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
+&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
 \end{align}
@@
 &= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u)} du \\
-&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
-&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
+&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
+&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
 \end{align}
@@
  &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u)} du\\
- &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
-&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
-&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du.
+ &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
+&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
+&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du.
 \label{eq:this2}

### 2015-01-23 NewDesignDoc/DesignDocGLMM.tex
Added constants in to f-twiddle

@@
 \begin{align}
-\log f_\theta(u)= \log f (u| 0, D) = (-1/2) \log |D| - (1/2) u' D^{-1} u
+\log f_\theta(u)= \log f (u| 0, D) = - q/2 \log (2 \pi) -1/2 \log |D| - (1/2) u' D^{-1} u
 \end{align}
@@
 This section discusses evaluating the importance sampling distribution shown in \ref{eq:ftwiddle}.
+The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.
 Section \ref{sec:Dgeneral} describes how to evaluate the pdf for the normal components of $\tilde{f}(u)$.
 Section \ref{sec:tstuff} describes how to evaluate the pdf for the t component of $\tilde{f}(u)$.
-The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.
+
+To be computationally stable, we need to be careful  with the addition of the terms in \ref{eq:ftwiddle} to prevent overflow. We know that
+\begin{align}
+ \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \\
+&= p_1 \exp( \log  \grave{f}(u|0,D^*) ))+p_2 \exp( \log(  f(u \, | \, u^*, D^*) ))+ \nonumber\\
+&+ p_3 \exp( \log( f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) )).
+\end{align}
+Set
+\begin{align}
+\tilde{a}= \max \left\{ \log  \grave{f}(u|0,D^*) ), \log  f(u \, | \, u^*, D^*) ), \log f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})  \right\}
+\end{align}
+Then
+\begin{align}
+ \tilde{f}(u) 
+&= p_1 \exp\left( \log  \grave{f}(u|0,D^*)  -\tilde{a}\right)+p_2 \exp \left( \log  f(u \, | \, u^*, D^* )-\tilde{a} \right)+ \nonumber\\
+&+ p_3 \exp\left[ \log f \left(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1} \right)  -\tilde{a} \right].
+\end{align}
+is more stable computationally.
 
@@
 \begin{align}
-\log f (u| \mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
+\log f (u| \mu, \Sigma) = -\dfrac{q}{2} \log (2 \pi) + \dfrac{1}{2} \log |\Sigma^{-1}| - \dfrac{1}{2} (u-\mu)' \Sigma^{-1} (u-\mu)
 \end{align}
@@
 \begin{align}
-\log \grave{f}(u)&=\log \left[ 1+u'D^* u/\zeta  \right]^{-(\zeta+q)/2}\\
-&=-\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
-&= -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t}\right)
-\end{align}
-
-Rewritten in C, we need to pass in \texttt{u} as a pointer, \texttt{zeta} as an int, \texttt{myq} as an int array, \texttt{meow} as an int array, and \texttt{logft} as a pointer.
+\log \grave{f}(u|0,D^*)&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
+&- \dfrac{1}{2} \log |D^*|     -\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
+&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
+&- \dfrac{1}{2} \log |D^*|   -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t^*}\right)
+\end{align}
+
+Rewritten in C, we need to pass in the first four terms \texttt{tconstants} as a double, \texttt{u} as an array of doubles, \texttt{zeta} as an int, \texttt{myq} as an array of ints, \texttt{meow} as an array of ints, \texttt{nut} as an array of doubles,  and \texttt{logft} as a double.
+The final calculation will be written into \texttt{logft}.
 
@@
 \int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &=\int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{\tilde{f}(u)} du\\
-&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u)} du\\
+&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u|0,D^*)} du\\
 &\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, f_\theta(y|u)^2  }{\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du
@@
 &=\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
-&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u)} du \\
+&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u|0,D^*)} du \\
 &\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du \\
@@
 \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du\\
-&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du
+&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du
 \end{align}
@@
 \begin{align}
-\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du < \infty
+\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du < \infty
 \end{align}
@@
 \begin{align}
-&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \\
-&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u)} du \\
+&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \\
+&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u|0,D^*)} du \\
 &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
@@
 &\leq
-\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
+\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
 \end{align}
@@
 \begin{align}
-\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du \label{eq:this1}
+\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
 \end{align}
@@
 \begin{align}
-& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u)} du\\
- &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u)} du\\
+& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du\\
+ &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u|0,D^*)} du\\
  &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\

### 2015-01-23 NewDesignDoc/DesignDocGLMM.tex
Changes and additions for t in R, normal stuff
I've coded the t pdf into R (to be used for tests). Also a little function to calculate the t constant in the pdf. Added the constants to the normal distribution stuff, both for R and C. Some small changes in design doc related to T in R and normal in both R and C. About to move onto creating t in C.

@@
 \begin{align}
-\log \grave{f}(u|0,D^*)&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
-&- \dfrac{1}{2} \log |D^*|     -\left[ \dfrac{\zeta+q}{2} \right] \log (1+u'D^*u/\zeta)\\
-&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\nu \pi)\\
-&- \dfrac{1}{2} \log |D^*|   -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\sum_{t=1}^T \dfrac{u_t'u_t}{\zeta \nu_t^*}\right)
-\end{align}
-
-Rewritten in C, we need to pass in the first four terms \texttt{tconstants} as a double, \texttt{u} as an array of doubles, \texttt{zeta} as an int, \texttt{myq} as an array of ints, \texttt{meow} as an array of ints, \texttt{nut} as an array of doubles,  and \texttt{logft} as a double.
+\log \grave{f}(u|0,D^*)&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\zeta \pi)\\
+&- \dfrac{1}{2} \log |D^*|     -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\dfrac{u'D^{* -1}u}{\zeta} \right)\\
+\end{align}
+
+The first four terms are constant with respect to $u$ and can be calculated ahead of time.
+Rewritten in C, we need to pass in  \texttt{tconstants} (the sum of the first four terms) as a double, \texttt{u} as an array of doubles, \texttt{zeta} as an int, \texttt{myq} as an int, the diagonal of \texttt{Dstarinv} as an array of doubles,  and \texttt{logft} as a double.
 The final calculation will be written into \texttt{logft}.
+
+I have another function (written in R) that calculates the constants for the t distribution.
+The arguments of this function are \texttt{zeta}, the diagonal of $D^*$ \texttt{Dstardiag}, and the length of $u$ \texttt{myq}.
 

### 2015-01-23 glmm/R/distRand.R

@@ -11,6 +11,21 @@ tdist <-function(nu,U,z.list,mu,gamm){
 
 }
 
+#note that Dstarinv is just the DIAGONAL of Dstarinv
+tdist2<-function(tconst,u, Dstarinv,zeta,myq){
+	inside<-1+sum(t(u)*Dstarinv*u)/zeta
+	logft<-tconst - ((zeta+myq)/2)*log(inside)
+	logft
+}
+
+tconstant<-function(zeta,myq,Dstarinvdiag){
+	piece1<-log(gamma((zeta+myq)/2))
+	piece1b<--log(gamma(zeta/2))
+	piece1c<--myq*log(zeta*pi)/2
+	piece2<- .5*log(prod(Dstarinvdiag))	
+	piece1+piece1b+piece1c+piece2
+}
+
 # no longer used
 #distRandGeneral2<-function(uvec,mu,Sigma.inv,logDetSigmaInv){
 #	print("still using distRandGeneral2.R")

### 2015-01-23 glmm/R/objfun.R

@@ -40,7 +40,7 @@ function(par,nbeta,nu.pql,umat,u.star,mod.mcml,family.glmm,cache,p1,p2,p3,m1,D.s
 	throwaway<-T+1
 	meow[2:throwaway]<-cumsum(nrandom)
 	
-	pee<-c(p1,p2,p3)
+	pea<-c(p1,p2,p3)
 	n<-nrow(mod.mcml$x)
 
 #need to scale first m1 vectors of generated random effects by multiplying by A
@@ -53,7 +53,7 @@ function(par,nbeta,nu.pql,umat,u.star,mod.mcml,family.glmm,cache,p1,p2,p3,m1,D.s
 		umat[k,]<-u.swoop*Afornu
 		}
 
-	stuff<-.C("objfunc", as.double(mod.mcml$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod.mcml$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(u.star), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pee=as.double(pee), nps=as.integer(length(pee)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), v=double(m),value=double(1),gradient=double(length(par)),hessian=double((length(par))^2),PACKAGE="glmm")
+	stuff<-.C("objfunc", as.double(mod.mcml$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod.mcml$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(u.star), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), v=double(m),value=double(1),gradient=double(length(par)),hessian=double((length(par))^2),PACKAGE="glmm")
 
 
 	if (!missing(cache)) cache$weights<-stuff$v		

### 2015-01-23 glmm/src/distRand.c

@@ -16,6 +16,6 @@ void distRandGenC(double *SigmaInv, double *logdet, int *nrow, double *uvec, dou
 
 	Free(blah);
 	Free(umu);
-	*distRandGenVal=.5*((*logdet)-(*blah2));
+	*distRandGenVal=.5*((-*nrow)*log(2*PI)+ (*logdet)-(*blah2));
 	Free(blah2);
 }

### 2015-01-23 glmm/tests/objfunTest.R

@@ -1,4 +1,3 @@
-
 #check objfun using finite differences
 library(glmm)
 data(BoothHobert)

### 2015-01-23 glmm/tests/testpiecesBH.R

@@ -52,12 +52,12 @@ all.equal(as.numeric(this$hessian),as.numeric(that$hessian))
 #want to check distRand when we use a normal distribution to get our random effects
 #check written for BH
 distRandCheck<-function(nu,uvec,muvec){
-ukmuk<-sum((uvec-muvec)^2)
-value<--5*log(nu)-ukmuk/(2*nu)
-gradient<- -5/nu +ukmuk/(2*nu^2)
-hessian<- 5/(nu^2)-ukmuk/(nu^3)
-hessian<-as.matrix(hessian)
-list(value=value,gradient=gradient,hessian=hessian)
+	ukmuk<-sum((uvec-muvec)^2)
+	value<--length(uvec)*.5*log(2*pi)-5*log(nu)-ukmuk/(2*nu)
+	gradient<- -5/nu +ukmuk/(2*nu^2)
+	hessian<- 5/(nu^2)-ukmuk/(nu^3)
+	hessian<-as.matrix(hessian)
+	list(value=value,gradient=gradient,hessian=hessian)
 }
 
 #function written originally in R
@@ -95,7 +95,7 @@ function(nu,U,z.list,mu){
 		you<-as.vector(U.list[[t]])
 		mew<-as.vector(mu.list[[t]])
 		Umu<-(you-mew)%*%(you-mew)
-		val[t]<- as.numeric(-.5*nrandom[t]*log(nu[t])-Umu/(2*nu[t]))
+		val[t]<--length(you)*.5*log(2*pi)+ as.numeric(-.5*nrandom[t]*log(nu[t])-Umu/(2*nu[t]))
 		
 		gradient[t]<- -nrandom[t]/(2*nu[t])+Umu/(2*(nu[t])^2)
 		
@@ -116,6 +116,13 @@ this<-distRandCheck(2,you,u.pql)
 that<-distRand(2,you,mod.mcml$z,u.pql)
 
 all.equal(this,that)
+
+#use finite diffs to make sure distRandCheck (and distRand) have correct derivs
+del<-.0001
+thisdel<-distRandCheck(2+del,you,u.pql)
+firstthing<-thisdel$value-this$value
+secondthing<-as.vector(this$gradient%*%del)
+all.equal(firstthing,secondthing,tol=10^-5)
 
 #compare the gradient and hessian of the C functions by using these functions
 #(the value is checked in distRandGeneral)
[…]

### 2015-02-16 glmm/R/summary.mcla.R
Added link to summary, edited vignette

@@ -34,9 +34,11 @@ function(object,...){
 	nucoefmat<-cbind(nu,nu.se,nuzval,pnorm(abs(nuzval),lower.tail=F))
 	colnames(nucoefmat)<-c("Estimate","Std. Error", "z value", "Pr(>|z|)/2")
 	rownames(nucoefmat)<-mod.mcml$varcomps.names
+	link<-mod.mcml$family.glmm$link
+
 	
 	return(structure(list(x=x,y=y, z=z, coefmat=coefmat, fixedcall=fixedcall, randcall=randcall, coefficients=coefficients,
-family.mcml=mod.mcml$family.mcml,call=call,nucoefmat=nucoefmat),class="summary.glmm"))
+family.mcml = mod.mcml$family.mcml, call = call, nucoefmat = nucoefmat, link=link),class="summary.glmm"))
 }
 
 print.summary.glmm <-
@@ -47,6 +49,8 @@ print.summary.glmm <-
     stopifnot(inherits(summ, "summary.glmm"))
 
     cat("\nCall:\n", paste(deparse(summ$call), sep = "\n", collapse = "\n"), "\n\n", sep = "")
+
+ cat("\nLink is: ", paste(deparse(summ$link)),"\n\n",sep="")
    
 cat("Fixed Effects:")
    cat("\n")

### 2015-02-23 glmm/R/glmm.R
Editing the vignette 'intro.pdf'

@@ -92,6 +92,8 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 		if(length(par.init)!=nbetaplusT) stop("par.init is not the correct length. It should contain initial values for the fixed effects and variance components.")
 		vcs<-par.init[-(1:nbeta)]
 		if(any(vcs<=10^-9)) stop("Initial values for the variance components in par.init must be positive and sufficiently large (greater than 10^-9).")
+		doPQL<-FALSE
+		#if par.init is given, we want to use those not PQL
 	}
 	
 	#cache will hold some pql estimates and the importance sampling weights that wouldn't otherwise be returned

### 2015-02-23 glmm/R/summary.mcla.R

@@ -3,6 +3,7 @@ function(object,...){
     mod.mcml<-object
     stopifnot(inherits(mod.mcml, "glmm"))
 
+	trust.converged<-mod.mcml$trust.converged
 	fixedcall<-mod.mcml$fixedcall
 	randcall<-mod.mcml$randcall
 	call<-mod.mcml$call
@@ -37,8 +38,8 @@ function(object,...){
 	link<-mod.mcml$family.glmm$link
 
 	
-	return(structure(list(x=x,y=y, z=z, coefmat=coefmat, fixedcall=fixedcall, randcall=randcall, coefficients=coefficients,
-family.mcml = mod.mcml$family.mcml, call = call, nucoefmat = nucoefmat, link=link),class="summary.glmm"))
+	return(structure(list(x=x,y=y, z=z, coefmat=coefmat, fixedcall = fixedcall, randcall = randcall, coefficients = coefficients,
+family.mcml = mod.mcml$family.mcml, call = call, nucoefmat = nucoefmat, link = link, trust.converged = trust.converged),class="summary.glmm"))
 }
 
 print.summary.glmm <-
@@ -47,6 +48,8 @@ print.summary.glmm <-
 {
     summ<-x	
     stopifnot(inherits(summ, "summary.glmm"))
+
+	if(summ$trust.converged==FALSE)  cat("\nWARNING: the optimizer trust has not converged to the MCMLE. The following estimates are not maximum likelihood estimates, but they can be used in the argument par.init when rerunning glmm. ", paste(deparse(summ$link)),"\n\n",sep="")
 
     cat("\nCall:\n", paste(deparse(summ$call), sep = "\n", collapse = "\n"), "\n\n", sep = "")
 

### 2015-02-23 glmm/R/glmm.R
Fixed thing if par.init arg is used

@@ -85,7 +85,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	#end figuring out how to interpret the formula
 
 	#make sure par.init has the right number of parameters and that the variance 
-	#components are positive to start
+	#components are positive to start. also skip pql bc using par.init instead
 	if(!is.null(par.init)){
 		nbeta<-ncol(x)
 		nbetaplusT<-nbeta+length(z)
@@ -118,7 +118,11 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	    nrandom<-unlist(nrand)
 	    totnrandom<-sum(nrandom)
 	    s.pql<-rep(0,totnrandom)
-		#if(!is.null(par.init)) then par.init is already specified by user
+		if(!is.null(par.init)){ #then par.init is already specified by user
+			beta.pql<-par.init[1:nbeta]
+			nu.pql<-par.init[-(1:nbeta)]
+			sigma.pql<-sqrt(nu.pql)
+		}
 		if(is.null(par.init)){
 	        sigma.pql<-nu.pql<-rep(1,length(mod.mcml$z))
 	        beta.pql<-rep(0,ncol(mod.mcml$x))

### 2015-04-27 glmm/R/summary.mcla.R
Added vcov, updated vignette, ready to resubmit

@@ -49,7 +49,7 @@ print.summary.glmm <-
     summ<-x	
     stopifnot(inherits(summ, "summary.glmm"))
 
-	if(summ$trust.converged==FALSE)  cat("\nWARNING: the optimizer trust has not converged to the MCMLE. The following estimates are not maximum likelihood estimates, but they can be used in the argument par.init when rerunning glmm. ", paste(deparse(summ$link)),"\n\n",sep="")
+	if(summ$trust.converged==FALSE)  cat("\nWARNING: the optimizer trust has not converged to the MCMLE. The following estimates are not maximum likelihood estimates, but they can be used in the argument par.init when rerunning glmm. \n\n",sep="")
 
     cat("\nCall:\n", paste(deparse(summ$call), sep = "\n", collapse = "\n"), "\n\n", sep = "")
 
@@ -81,6 +81,22 @@ function(object,...){
 	coefficients<-mod$beta
 	names(coefficients)<-colnames(mod$x)
 	coefficients
+}
+
+
+vcov.glmm <-
+function(object,...){
+	mod<-object
+   	stopifnot(inherits(mod, "glmm"))
+	vcov <- -solve(mod$likelihood.hessian)
+
+	#get names for vcov matrix
+	rownames(vcov)<-colnames(vcov)<-rep(c("blah"),nrow(vcov))
+	nbeta<-length(mod$beta)
+	rownames(vcov)[1:nbeta] <- colnames(vcov)[1:nbeta] <- colnames(mod$x)
+	rownames(vcov)[-(1:nbeta)] <- colnames(vcov)[-(1:nbeta)] <- mod$varcomps.names
+
+	vcov
 }
 
 varcomps<-function(object,...){

### 2016-06-02 glmm/src/cp3.c
Added new arg ntrials to cumul function and derivs
Did not change anything in cum3, cp3, cpp3 besides adding the new argument (ntrials). Changed tests to show updated call. Temp added a few lines in  elc elval elGH to hard code ntrials=1.

@@ -1,5 +1,5 @@
 #include "myheader.h"
-void cp3(double *etain, int *neta, int *typein, double *cpout)
+void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout)
 {
     int leneta=neta[0];
     int type = typein[0];

### 2016-06-02 glmm/src/cpp3.c

@@ -3,7 +3,7 @@
 #include "myheader.h"
 
 
-void cpp3(double *etain, int *neta, int *typein, double *cppout)
+void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout)
 {
     int leneta=neta[0];
     int type = typein[0];

### 2016-06-02 glmm/src/cum3.c

@@ -3,7 +3,7 @@
 #include "myheader.h"
 
 
-void cum3(double *etain, int *neta, int *typein, double *cumout)
+void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout)
 {
     int leneta=neta[0];
     int type = typein[0];

### 2016-06-02 glmm/src/elGH.c

@@ -5,9 +5,11 @@ void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family
 	double *cppout=Calloc(*nrowX,double);
 
 	/*calling cp3, cpp3 will just change the doubles cpout, cppout  */
-
-	cp3(eta,nrowX,family,cpout);
-	cpp3(eta,nrowX,family,cppout);
+	int *ntrials = Calloc(1, int);
+    ntrials[0]=1;
+	cp3(eta,nrowX,family,ntrials,cpout);
+	cpp3(eta,nrowX,family,ntrials,cppout);
+	Free(ntrials);
 
 	/*calculate gradient of el: X^T (Y-c'(eta))  
 	first use loop to calculate Y-c'(eta). then turn c''(eta) into -c''(eta)*/

### 2016-06-02 glmm/src/elc.c

@@ -10,9 +10,12 @@ void elc(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family,
 	memset(cppout,0,*nrowX);
 
 	/*calling cum3, cp3, cpp3 will just change the doubles cumout, cpout, cppout  */
-	cum3(eta,nrowX,family,cumout);
-	cp3(eta,nrowX,family,cpout);
-	cpp3(eta,nrowX,family,cppout);
+    int *ntrials = Calloc(1, int);
+    ntrials[0]=1;
+	cum3(eta,nrowX,family,ntrials, cumout);
+	cp3(eta,nrowX,family, ntrials, cpout);
+	cpp3(eta,nrowX,family,ntrials, cppout);
+    Free(ntrials);
 
 	/*calculate value of el: Y^T eta-c(eta)  */
 	int thing1=1,*ione=&thing1;

### 2016-06-02 glmm/src/elval.c

@@ -1,11 +1,16 @@
 #include "myheader.h"
 void elval(double *Y,  int *nrowX, int *ncolX, double *eta, int *family, double *elval)
 {
+
+	int *ntrials = Calloc(1, int);
+    ntrials[0]=1;
+
 	double *cumout=Calloc(1,double);
 	cumout[0]=0.0;
 
 	/*calling cum3, cp3, cpp3 will just change the doubles cumout, cpout, cppout  */
-	cum3(eta,nrowX,family,cumout);
+	cum3(eta,nrowX,family,ntrials, cumout);
+	Free(ntrials);
 
 	/*calculate value of el: Y^T eta-c(eta)  */
 	int intone=1;

### 2016-06-02 glmm/src/myheader.h

@@ -8,9 +8,9 @@
 #include <R.h>
 #include <R_ext/BLAS.h>
 
-void cum3(double *etain, int *neta, int *typein, double *cumout);
-void cp3(double *etain, int *neta, int *typein, double *cpout);
-void cpp3(double *etain, int *neta, int *typein, double *cppout);
+void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout);
+void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout);
+void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout);
 
 void matTvecmult(double *a, double *b, int *nrow, int *ncol, double *result);
 void diag(double *invec, int *nrow, double *outmat);

### 2016-06-02 glmm/tests/cfamilies.R

@@ -13,15 +13,15 @@ cpp<-function(eta) {(1/(1+exp(-eta)))*(1/(1+exp(eta)))}
 #check cum3 for a vector containing both positive and negative values of eta
 eta<-seq(-1,1,.1)
 neta<-length(eta)
-that<-.C("cum3",eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), cumout=double(1))
+that<-.C("cum3",eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), ntrials=as.integer(1), cumout=double(1))
 all.equal(sum(cum(eta)),that$cumout)
 
 #check cp3
-that<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))
+that<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cpout=double(neta))
 all.equal(cp(eta),that$cpout)
 
 #check cpp3
-that<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cppout=double(neta))
+that<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cppout=double(neta))
 all.equal(cpp(eta),that$cppout)
 
 ######################
@@ -33,14 +33,14 @@ eta<-c(4.5,5,5.5)
 neta<-length(eta)
 
 #check cum3
-that<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cumout=double(1))
+that<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1), cumout=double(1))
 all.equal(sum(cum(eta)),that$cumout)
 
 #check cp3
-that<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cpout=double(neta))
+that<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1), cpout=double(neta))
 all.equal(cp(eta),that$cpout)
 
 #check cpp3
-that<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cppout=double(neta))
[…]

### 2016-06-02 glmm/tests/elTest.R

@@ -31,14 +31,14 @@ function(Y,X,eta,family.mcml){
 	neta<-length(eta)
 
 	if(family.mcml$family.glmm=="bernoulli.glmm"){
-		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cumout=double(1))$cumout
-		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))$cpout
-		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cppout=double(neta))$cppout
+		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cumout=double(1))$cumout
+		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cpout=double(neta))$cpout
+		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cppout=double(neta))$cppout
 	}
 	if(family.mcml$family.glmm=="poisson.glmm"){
-		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cumout=double(1))$cumout
-		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cpout=double(neta))$cpout
-		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cppout=double(neta))$cppout
+		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1), cumout=double(1))$cumout
+		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cpout=double(neta))$cpout
+		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cppout=double(neta))$cppout
 	}
 
 	value<-as.numeric(Y%*%eta-foo)

### 2016-06-02 glmm/tests/familiesFiniteDiffs.R

@@ -4,23 +4,23 @@ neta<-length(eta)
 delta<-rep(.0000000001,neta)
 
 #check derivatives for bernoulli using finite differences
-foo<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))$cpout
+foo<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout
 this<-foo*delta
-that<-.C("cum3",eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),cumout=double(neta))$cumout-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), cumout=double(neta))$cumout
+that<-.C("cum3",eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cumout=double(neta))$cumout-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cumout=double(neta))$cumout
 all.equal(this,that)
 
-this<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cppout=double(neta))$cppout*delta
-that<-.C("cp3",eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))$cpout-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))$cpout
+this<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cppout=double(neta))$cppout*delta
+that<-.C("cp3",eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout
 all.equal(this,that)
 
 #check derivatives for poisson using finite differences
-foo<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cpout=double(neta))$cpout
[…]

### 2016-06-02 glmm/tests/objfunTest.R

@@ -170,14 +170,14 @@ function(Y,X,eta,family.mcml){
 	neta<-length(eta)
 
 	if(family.mcml$family.glmm=="bernoulli.glmm"){
-		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cumout=double(1))$cumout
-		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cpout=double(neta))$cpout
-		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),cppout=double(neta))$cppout
+		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cumout=double(1))$cumout
+		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout
+		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cppout=double(neta))$cppout
 	}
 	if(family.mcml$family.glmm=="poisson.glmm"){
-		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cumout=double(1))$cumout
-		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cpout=double(neta))$cpout
-		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),cppout=double(neta))$cppout
+		foo<-.C("cum3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cumout=double(1))$cumout
+		mu<-.C("cp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cpout=double(neta))$cpout
+		cdub<-.C("cpp3",eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cppout=double(neta))$cppout
 	}
 
 	value<-as.numeric(Y%*%eta-foo)

### 2016-06-02 NewDesignDoc/DesignDocGLMM.tex
Minor changes to design doc and pkg's manual
Expanding ntrials arg to likelihood level
Previously only cumulant function had arg ntrials. Now elc, elval, elGH all take ntrials arg. Tests and calls to these functions updated to reflect this.

@@
 
-\title{Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models}
+\title{Design Document for R Package glmm}
 
@@
 \end{align}
+
  For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
@@
 \begin{align}
-\log f_\theta(y|u) &= y' \eta +c(\eta) \\
+\log f_\theta(y|u) &= y' \eta -c(\eta) \\
 &= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
@@
 It could make sense to truncate the interval so that the lower bound starts at 0 and the upper bound remains untouched.
+
+\section{New Additions (Summer 2016)}
+\subsection{New Family: Binomial}
+One year later, I'm adding binomial.
+For Binomial, the calculations change only in the cumulant function.
+They rely on one more number: the number of trials.
+The inputted response is usually a vector, but for binomial (with the exception of Bernoulli) we need 2 vectors cbinded (cbound?) together.
+The first column is the number of successes and the second is the number of failures.
+Summing across the row gives us the number of trials.
+I call this $ntrials$ from here on out.
+
+The cumulant function and its first two derivatives with respect to the canonical parameter $\eta_i$ are below. 
+\begin{align}
+c(\eta_i) &=ntrials_i \, \log(1+e^{\eta_i}) =
+  \begin{cases}
+   ntrials_i \, \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
+   ntrials_i \,\left(  \eta_i+\log(e^{-\eta_i}+1) \right) & \text{if } \eta_i >0,
+  \end{cases}\\
+c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{ntrials_i}{1+e^{-\eta_i}}\\
+c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{ntrials_i}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
+\end{align}
+
+I will implement this change in steps. Game plan:
+\begin{enumerate}
+\item Start by writing the binomial cumulant in R.
+Check to make sure this calculation is identical to that produced by bernoulli for some parameter value $\eta_i$ and for $ntrials=1$.
+Check for both $\eta$ positive and negative.
+\item Check the three calculations for $ntrials=5$ (or some number other than 1) compared to my by-hand calculation. 
+\item Code it in C and check that the calculation is identical to that produced in R for $ntrials=1$ and for $ntrials \neq 1$.
+\item The second lowest level will then call binomial.glmm rather than bernoulli.glmm as the family.
+I will work my way up the levels incorporating ntrials.
+At the top, I will have a variable to set ntrials = 1 (for now).
+As I do each layer, I will make sure I get all the same answers from my salamanders problem (or some other bernoulli problem).
+\item I will go to the top-most layer, right where the model matrices are being made.
+The new response vector y will be the first column of the user-provided response.
+The sum of the two entries in the row will create the vector ntrials.
+I will check that the salamanders problem gives the same results (when specified as cbind(Mate, 1-Mate)).
+
+\item I will add checks for the model matrix stage.  See below.
+\item I want to allow the response for bernoulli to be a single vector  (rather than two cbinded vectors).
+Therefore, if the family is binomial.glmm and the length of the dimension of the response is 1 (meaning we just have a vector), then create the ntrials vector to be a vector of 1s.
+Test this with salamanders or some other problem to make sure it gives the same results.
+
+\end{enumerate}
+
+Checks for the model matrix stage: 
+\begin{enumerate}
+\item  If  the length of the dimension of the response is 2 dimensions (meaning we have both rows and columns), then the family must be binomial.glmm.
+\item If the family is binomial.glmm, then the response must have 2 columns
+\end{enumerate}
+
+These changes will need to be done in both C and R. The checks will be in R. I'll need to add  documentation for binomial. Bernoulli will be retained.
 
@@
 \begin{align}
-&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- l(\theta)  \right] \\
+&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- \nabla l(\theta)  \right] \\
 &=
-\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\;  l(\theta)  \\
+\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\; \nabla  l(\theta)  \\
 &=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
- - l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
+ -\nabla l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
 \end{align}

### 2016-06-02 glmm/R/glmm.R

@@ -199,8 +199,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	umat<-rbind(genData,genData2,genData3)
 
 	#use trust to max the objfun (monte carlo likelihood)
-	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, 
-umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star,Sigmuh=Sigmuh,Sigmuh.inv=Sigmuh.inv,zeta=zeta)
+	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star,Sigmuh=Sigmuh,Sigmuh.inv=Sigmuh.inv,zeta=zeta)
 
 	beta.trust<-trust.out$argument[1:length(beta.pql)]
 	nu.trust<-trust.out$argument[-(1:length(beta.pql))]

### 2016-06-02 glmm/R/pql.R

@@ -66,9 +66,9 @@ family.mcml=family.mcml,cache=cache)
 	family.mcml<-getFamily(family.mcml)
 
 	if(family.mcml$family.glmm=="bernoulli.glmm"){
-		piece1<-.C("elc",as.double(Y),as.double(X),as.integer(nrow(X)),as.integer(ncol(X)),as.double(eta.twid),as.integer(1),value=double(1),gradient=double(ncol(X)),hessian=double((ncol(X)^2)))$value}
+		piece1<-.C("elc",as.double(Y),as.double(X),as.integer(nrow(X)),as.integer(ncol(X)),as.double(eta.twid),as.integer(1),as.integer(1),value=double(1),gradient=double(ncol(X)),hessian=double((ncol(X)^2)))$value}
 	if(family.mcml$family.glmm=="poisson.glmm"){
-		piece1<-.C("elc",as.double(Y),as.double(X),as.integer(nrow(X)),as.integer(ncol(X)),as.double(eta.twid),as.integer(2),value=double(1),gradient=double(ncol(X)),hessian=double((ncol(X)^2)))$value}
+		piece1<-.C("elc",as.double(Y),as.double(X),as.integer(nrow(X)),as.integer(ncol(X)),as.double(eta.twid),as.integer(2),as.integer(1), value=double(1),gradient=double(ncol(X)),hessian=double((ncol(X)^2)))$value}
 	
 	#calculate W = c''(eta.twid)
 	dubya<-family.mcml$cpp(eta.twid)
@@ -103,9 +103,9 @@ function(mypar,Y,X,Z,A,family.mcml,nbeta,cache )
 
 	#value<- ellikelihood(Y,X,eta,family.mcml)$value-.5*s%*%s
 	if(family.mcml$family.glmm=="bernoulli.glmm"){
-		value<-.C("elc",as.double(Y),as.double(X),as.integer(nrow(X)),as.integer(ncol(X)),as.double(eta),as.integer(1),value=double(1),gradient=double(ncol(X)),hessian=double((ncol(X)^2)))$value-.5*s%*%s}
+		value<-.C("elc", as.double(Y), as.double(X), as.integer(nrow(X)), as.integer(ncol(X)), as.double(eta),as.integer(1),as.integer(1), value=double(1), gradient=double(ncol(X)), hessian=double((ncol(X)^2)))$value-.5*s%*%s}
 	if(family.mcml$family.glmm=="poisson.glmm"){
[…]

### 2016-06-02 glmm/src/cp3.c

@@ -12,11 +12,11 @@ void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout)
 	case 1:
 	    cpout[i]= 1/(1+exp(-eta));
 	    break;
-        case 2:
-            cpout[i]= exp(eta);
+    case 2:
+        cpout[i]= exp(eta);
 	    break;
-        default:
-            error("unrecognized type");
+    default:
+        error("unrecognized type");
 	    break;
         }
     }

### 2016-06-02 glmm/src/cpp3.c

@@ -17,11 +17,11 @@ void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout)
 	case 1:
 	    cppout[i]= 1/((1+exp(-eta))*(1+exp(eta)));
 	    break;
-        case 2:
-            cppout[i]= exp(eta);
+    case 2:
+        cppout[i]= exp(eta);
 	    break;
-        default:
-            error("unrecognized type");
+    default:
+        error("unrecognized type");
 	    break;
         }
     }

### 2016-06-02 glmm/src/elGH.c

@@ -1,15 +1,13 @@
 #include "myheader.h"
-void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family, double *elgradient, double *elhessian)
+void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family, int *ntrials, double *elgradient, double *elhessian)
 {
-	double *cpout=Calloc(*nrowX,double);
-	double *cppout=Calloc(*nrowX,double);
+	double *cpout = Calloc(*nrowX,double);
+	double *cppout = Calloc(*nrowX,double);
 
 	/*calling cp3, cpp3 will just change the doubles cpout, cppout  */
-	int *ntrials = Calloc(1, int);
-    ntrials[0]=1;
 	cp3(eta,nrowX,family,ntrials,cpout);
 	cpp3(eta,nrowX,family,ntrials,cppout);
-	Free(ntrials);
+
 
 	/*calculate gradient of el: X^T (Y-c'(eta))  
 	first use loop to calculate Y-c'(eta). then turn c''(eta) into -c''(eta)*/

### 2016-06-02 glmm/src/elc.c

@@ -1,21 +1,19 @@
 #include "myheader.h"
-void elc(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family, double *elval, double *elgradient, double *elhessian)
+void elc(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family,  int *ntrials, double *elval, double *elgradient, double *elhessian)
 {
 
-	double *cumout=Calloc(1,double);
+	double *cumout=Calloc(1, double);
 	cumout[0]=0.0;
-	double *cpout=Calloc(*nrowX,double);
+	double *cpout=Calloc(*nrowX, double);
 	memset(cpout,0,*nrowX);
-	double *cppout=Calloc(*nrowX,double);
+	double *cppout=Calloc(*nrowX, double);
 	memset(cppout,0,*nrowX);
 
 	/*calling cum3, cp3, cpp3 will just change the doubles cumout, cpout, cppout  */
-    int *ntrials = Calloc(1, int);
-    ntrials[0]=1;
-	cum3(eta,nrowX,family,ntrials, cumout);
-	cp3(eta,nrowX,family, ntrials, cpout);
-	cpp3(eta,nrowX,family,ntrials, cppout);
-    Free(ntrials);
+	cum3(eta, nrowX, family, ntrials, cumout);
+	cp3(eta, nrowX, family, ntrials, cpout);
+	cpp3(eta, nrowX, family, ntrials, cppout);
+
 
 	/*calculate value of el: Y^T eta-c(eta)  */
 	int thing1=1,*ione=&thing1;

### 2016-06-02 glmm/src/elval.c

@@ -1,16 +1,11 @@
 #include "myheader.h"
-void elval(double *Y,  int *nrowX, int *ncolX, double *eta, int *family, double *elval)
+void elval(double *Y,  int *nrowX, int *ncolX, double *eta, int *family, int *ntrials, double *elval)
 {
-
-	int *ntrials = Calloc(1, int);
-    ntrials[0]=1;
-
 	double *cumout=Calloc(1,double);
 	cumout[0]=0.0;
 
 	/*calling cum3, cp3, cpp3 will just change the doubles cumout, cpout, cppout  */
 	cum3(eta,nrowX,family,ntrials, cumout);
-	Free(ntrials);
 
 	/*calculate value of el: Y^T eta-c(eta)  */
 	int intone=1;

### 2016-06-02 glmm/src/myheader.h

@@ -4,13 +4,10 @@
 
 #define BERNOULLI 1
 #define POISSON 2
+#define BINOMIAL 3
 
 #include <R.h>
 #include <R_ext/BLAS.h>
-
-void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout);
-void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout);
-void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout);
 
 void matTvecmult(double *a, double *b, int *nrow, int *ncol, double *result);
 void diag(double *invec, int *nrow, double *outmat);
@@ -24,16 +21,22 @@ void sumup(double *a, int *lena, double *suma);
 void subvec(double *a, double *b, int *len, double *out);
 void addvec(double *a, double *b, int *len, double *out);
 void divvec(double *a, double *b, int *len, double *out);
-void distRandGenC(double *SigmaInv, double *logdet, int *nrow, double *uvec, double *mu, double *distRandGenVal);
-void elval(double *Y,  int *nrowX, int *ncolX, double *eta, int *family, double *elval);
 
+void distRandGenC(double *SigmaInv, double *logdet, int *nrow, double *uvec, double *mu, double *distRandGenVal);
 void distRand3C(double *nu, double *mu, int *T, int *nrandom, int *meow, double *Uvec, double *drgradient, double *drhessian);
 
-void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family, double *elgradient, double *elhessian);
+void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout);
+void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout);
+void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout);
+
+void elval(double *Y,  int *nrowX, int *ncolX, double *eta, int *family, int *ntrials, double *elval);
+void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family,  int *ntrials, double *elgradient, double *elhessian);
+
+void tdist(double *DstarInv,  int *myq, double *uvec, int *zeta, double *tconst, double *logft);
 
[…]

### 2016-06-02 glmm/src/objfunc.c

@@ -8,11 +8,13 @@ nps is the length of pee (3 for now, maybe more if imp sampling distrib changes)
 */
 void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, double *value, double *gradient, double *hessian)
 {
-	double *Uk = Calloc(*myq,double);
+    int *ntrials = Calloc(1, int);
+	ntrials[0]=1;
+	double *Uk = Calloc(*myq, double);
 	int Uindex = 0;
 
 	double db1 = 0.0;
-	double *double1=&db1; /*temp to hold info that will be put into lfuval*/
+	double *double1 = &db1; /*temp to hold info that will be put into lfuval*/
 
 	/* Calculate xbeta, needed to calculate eta for each Uk=U[k,] in R notation */
 	double *xbeta = Calloc(*n,double);
@@ -38,16 +40,16 @@ void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *
 
 		/* calculate eta for this value of Uk
 		first calculate ZUk*/
-		matvecmult(z,Uk,n,myq,zu);
+		matvecmult(z, Uk, n, myq, zu);
 
 		/* then add xbeta+zu to get current value of eta */
-		addvec(xbeta,zu,n,eta);
+		addvec(xbeta, zu, n, eta);
 
 		/*log f_theta(u_k) goes into lfuval*/
-		distRandGenC(Dinvfornu,logdetDinvfornu,myq,Uk,qzeros,&lfuval); 
+		distRandGenC(Dinvfornu, logdetDinvfornu, myq, Uk, qzeros, &lfuval); 
 
 		/* log f_theta(y|u_k) value goes into lfyuval */
-		elval(y,n,nbeta,eta,family_glmm,&lfyuval);
+		elval(y, n, nbeta, eta, family_glmm, ntrials, &lfyuval);
 
 		/* value of log f~_theta(u_k) 
 		first calculate value of log f~_theta(u_k) for 3 distribs used 
[…]

### 2016-06-02 glmm/tests/elTest.R

@@ -52,17 +52,17 @@ function(Y,X,eta,family.mcml){
 #compare elR and el.C for a value of eta
 eta<-rep(2,150)
 that<-elR(mod.mcml$y,mod.mcml$x,eta,family.mcml=bernoulli.glmm)	
-this<-.C("elc",as.double(mod.mcml$y),as.double(mod.mcml$x),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),value=double(1),gradient=double(ncol(mod.mcml$x)),hessian=double((ncol(mod.mcml$x)^2)))
+this<-.C("elc", as.double(mod.mcml$y), as.double(mod.mcml$x), as.integer(nrow(mod.mcml$x)), as.integer(ncol(mod.mcml$x)), as.double(eta), as.integer(1), as.integer(1), value=double(1), gradient=double(ncol(mod.mcml$x)), hessian=double((ncol(mod.mcml$x)^2)))
 all.equal(as.numeric(that$value),this$value)
 all.equal(as.numeric(that$gradient),this$gradient)
 all.equal(as.numeric(that$hessian),this$hessian)
 
 #compare to elval
-elvalout<-.C("elval",as.double(mod.mcml$y),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),value=double(1))
+elvalout<-.C("elval", as.double(mod.mcml$y), as.integer(nrow(mod.mcml$x)), as.integer(ncol(mod.mcml$x)), as.double(eta), as.integer(1), as.integer(1), value=double(1))
 all.equal(as.numeric(that$value),elvalout$value)
 
 #compare to elGH
-elGHout<-.C("elGH",as.double(mod.mcml$y),as.double(mod.mcml$x),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),gradient=double(ncol(mod.mcml$x)),hessian=double((ncol(mod.mcml$x)^2)))
+elGHout<-.C("elGH",as.double(mod.mcml$y),as.double(mod.mcml$x),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1), as.integer(1), gradient=double(ncol(mod.mcml$x)),hessian=double((ncol(mod.mcml$x)^2)))
[…]

### 2016-06-02 glmm/tests/testpiecesBH.R

@@ -33,18 +33,18 @@ logfyuk<-function(eta,x,y){
 
 #compare elc and logfyuk for a value of eta
 eta<-rep(2,150)
-this<-.C("elc",as.double(mod.mcml$y),as.double(mod.mcml$x),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),value=double(1),gradient=double(ncol(mod.mcml$x)),hessian=double((ncol(mod.mcml$x)^2)))
+this<-.C("elc",as.double(mod.mcml$y), as.double(mod.mcml$x), as.integer(nrow(mod.mcml$x)), as.integer(ncol(mod.mcml$x)), as.double(eta), as.integer(1), as.integer(1), value=double(1), gradient=double(ncol(mod.mcml$x)), hessian=double((ncol(mod.mcml$x)^2)))
 that<-logfyuk(eta,mod.mcml$x,mod.mcml$y)
 all.equal(as.numeric(this$value),as.numeric(that$value))
 all.equal(as.numeric(this$gradient),as.numeric(that$gradient))
 all.equal(as.numeric(this$hessian),as.numeric(that$hessian))
 
 #compare elval to logfyuk
-this<-.C("elval",as.double(mod.mcml$y),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),value=double(1))
+this<-.C("elval", as.double(mod.mcml$y), as.integer(nrow(mod.mcml$x)), as.integer(ncol(mod.mcml$x)), as.double(eta), as.integer(1), as.integer(1), value=double(1))
 all.equal(as.numeric(this$value),as.numeric(that$value))
 
 #compare elGH to logfyuk
-this<-.C("elGH",as.double(mod.mcml$y),as.double(mod.mcml$x),as.integer(nrow(mod.mcml$x)),as.integer(ncol(mod.mcml$x)),as.double(eta),as.integer(1),gradient=double(ncol(mod.mcml$x)),hessian=double((ncol(mod.mcml$x)^2)))
+this<-.C("elGH", as.double(mod.mcml$y), as.double(mod.mcml$x), as.integer(nrow(mod.mcml$x)), as.integer(ncol(mod.mcml$x)), as.double(eta), as.integer(1), as.integer(1), gradient=double(ncol(mod.mcml$x)), hessian=double((ncol(mod.mcml$x)^2)))
[…]

### 2016-06-02 glmm/R/glmm.R
Objfun and down have scalar arg ntrials

@@ -198,8 +198,10 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 
 	umat<-rbind(genData,genData2,genData3)
 
+	ntrials <- 1
+
 	#use trust to max the objfun (monte carlo likelihood)
-	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star,Sigmuh=Sigmuh,Sigmuh.inv=Sigmuh.inv,zeta=zeta)
+	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=zeta, ntrials = ntrials)
 
 	beta.trust<-trust.out$argument[1:length(beta.pql)]
 	nu.trust<-trust.out$argument[-(1:length(beta.pql))]

### 2016-06-02 glmm/R/objfun.R

@@ -1,5 +1,5 @@
 objfun <-
-function(par,nbeta,nu.pql,umat,u.star,mod.mcml,family.glmm,cache,p1,p2,p3,m1,D.star, Sigmuh, Sigmuh.inv, zeta){
+function(par, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials){
 
 	beta<-par[1:nbeta]
 	nu<-par[-(1:nbeta)]
@@ -56,7 +56,7 @@ function(par,nbeta,nu.pql,umat,u.star,mod.mcml,family.glmm,cache,p1,p2,p3,m1,D.s
 #		umat[k,]<-u.swoop*Afornu
 #		}
 
-	stuff<-.C("objfunc", as.double(mod.mcml$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod.mcml$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(u.star), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), v=double(m),value=double(1),gradient=double(length(par)),hessian=double((length(par))^2),PACKAGE="glmm")
[…]

### 2016-06-02 glmm/src/myheader.h

@@ -34,7 +34,7 @@ void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family
 
 void tdist(double *DstarInv,  int *myq, double *uvec, int *zeta, double *tconst, double *logft);
 
-void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, double *value, double *gradient, double *hessian);
+void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian);
 
 
 

### 2016-06-02 glmm/src/objfunc.c

@@ -6,10 +6,8 @@ z is n by myq matrix
 pee is the vector of sampling proportions (usually 1/3, 1/3, 1/3)
 nps is the length of pee (3 for now, maybe more if imp sampling distrib changes)
 */
-void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, double *value, double *gradient, double *hessian)
+void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian)
 {
-    int *ntrials = Calloc(1, int);
-	ntrials[0]=1;
 	double *Uk = Calloc(*myq, double);
 	int Uindex = 0;
 
@@ -150,7 +148,6 @@ void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *
 
 		/* calculate gradient and hessian log f_theta(y|u_k) */
 		elGH(y, x, n, nbeta, eta, family_glmm, ntrials, lfyugradient, lfyuhess);
-		Free(ntrials);
 
 		/* calculate gradient */
 		Gindex=0;

### 2016-06-02 glmm/tests/testpiecesBH.R

@@ -189,7 +189,7 @@ piece3<-rep(0,3)
 #calculate objfun's value for comparison
 cache<-new.env(parent = emptyenv())
 objfun<-glmm:::objfun
-that<-objfun(c(beta,nu), nbeta=1, nu.pql=nu.pql, u.star=u.star, mod.mcml=mod.mcml, family.glmm=bernoulli.glmm,cache=cache,umat=umat, p1=1/3, p2=1/3, p3=1/3, m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=5)
+that<-objfun(c(beta,nu), nbeta=1, nu.pql=nu.pql, u.star=u.star, mod.mcml=mod.mcml, family.glmm=bernoulli.glmm,cache=cache,umat=umat, p1=1/3, p2=1/3, p3=1/3, m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=5, ntrials=1)
 
 #get t stuff ready
 tconstant<-glmm:::tconstant

### 2016-06-06 NewDesignDoc/DesignDocGLMM.tex
Added bin to cum and 3 derivs but not tested
There is some kind of problem, but I'm going to temp move on bc it's not hurting anything (not implemented anywhere anyway) and I need to do more important things.

@@
 \begin{enumerate}
-\item Start by writing the binomial cumulant in R.
-Check to make sure this calculation is identical to that produced by bernoulli for some parameter value $\eta_i$ and for $ntrials=1$.
-Check for both $\eta$ positive and negative.
-\item Check the three calculations for $ntrials=5$ (or some number other than 1) compared to my by-hand calculation. 
-\item Code it in C and check that the calculation is identical to that produced in R for $ntrials=1$ and for $ntrials \neq 1$.
-\item The second lowest level will then call binomial.glmm rather than bernoulli.glmm as the family.
-I will work my way up the levels incorporating ntrials.
+\item Without adding binomial yet, add the ntrials argument to all levels of code so that ntrials can be passed to the cumulant functions and their derivatives.
 At the top, I will have a variable to set ntrials = 1 (for now).
 As I do each layer, I will make sure I get all the same answers from my salamanders problem (or some other bernoulli problem).
+\item Write the binomial cumulant in R. Do some checks with  both $\eta$ positive and negative.
+\begin{enumerate}
+ \item Check to make sure this calculation is identical to that produced by bernoulli for some parameter value $\eta_i$ and for $ntrials=1$. 
+\item Check  the sum of Bernoulli cumulants equals the binomial cumulant (for ntrials $>1$). 
+\item Check finite differences. 
+\item Check the three calculations for $ntrials=5$ (or some number other than 1) compared to my by-hand calculation. 
+\end{enumerate}
+\item Code it in C and check that the calculation is identical to that produced in R for $ntrials=1$ and for $ntrials \neq 1$.
 \item I will go to the top-most layer, right where the model matrices are being made.

### 2016-06-06 glmm/src/cp3.c

@@ -9,15 +9,15 @@ void cp3(double *etain, int *neta, int *typein, int *ntrials, double *cpout)
     {
 		double eta = etain[i];
 		switch (type) {
-			case 1:
+			case BERNOULLI:
 			    cpout[i]= 1/(1+exp(-eta));
 			    break;
 
-		    case 2:
+		    case POISSON:
 		        cpout[i]= exp(eta);
 			    break;
 
-			case 3:
+			case BINOMIAL:
 			    cpout[i]= ntrials[i]/(1+exp(-eta));
 			    break;
 

### 2016-06-06 glmm/src/cpp3.c

@@ -14,15 +14,15 @@ void cpp3(double *etain, int *neta, int *typein, int *ntrials, double *cppout)
     {
 		double eta = etain[i];
 		switch (type) {
-			case 1:
+			case BERNOULLI:
 		    	cppout[i]= 1/((1+exp(-eta))*(1+exp(eta)));
 			    break;
 
-	    	case 2:
+	    	case POISSON:
 	    	    cppout[i]= exp(eta);
 			    break;
 
-			case 3:
+			case BINOMIAL:
 		    	cppout[i]= ntrials[i]/((1+exp(-eta))*(1+exp(eta)));
 			    break;
 

### 2016-06-06 glmm/src/cum3.c

@@ -13,6 +13,7 @@ void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout)
     for(i=0;i<leneta;++i)
     	{
 		double eta = etain[i];
+		int ntr = ntrials[i];
 		switch (type) {
 		case BERNOULLI:
 	 	   if(eta>0)
@@ -33,12 +34,12 @@ void cum3(double *etain, int *neta, int *typein, int *ntrials, double *cumout)
 		case BINOMIAL:
 	 	   if(eta>0)
 				{
-				cumout[0]+= ntrials[i] * (eta+log1p(exp(-eta)));
+				cumout[0]+= ntr*(eta+log1p(exp(-eta)));
 				break;
 				}
    	    	else
 				{
-				cumout[0]+= ntrials[i] * log1p(exp(eta));
+				cumout[0]+= ntr*log1p(exp(eta));
 				break;
 				}
 

### 2016-06-06 glmm/tests/binomfamtest.R
Finished testing binom cumulant and derivs
Ended up I couldn't figure it out before because I was testing it wrong. There was nothing wrong with the code itself.

@@ -1,0 +1,211 @@
+#the cumulant for binomial coded in R
+cumR<-function(eta,ntrials){
+   if(eta <= 0){
+	out <- ntrials * log1p(exp(eta))
+	}
+	
+	if(eta>0){
+	out <- ntrials * (eta + log1p(exp(-eta)))
+	}
+	
+	out
+}
+
+
+library(glmm)
+
+#choose a positive value for eta and compare to 
+#calc of cum for bernoulli
+eta <- 1
+this <- cumR(eta,ntrials=1)
+that <- bernoulli.glmm()$cum(eta)
+all.equal(this,that)
+
+#choose a negative value for eta and compare to 
+#calc of cum for bernoulli
+eta <- -1
+this <- cumR(eta,ntrials=1)
+that <- bernoulli.glmm()$cum(eta)
+all.equal(this,that)
+
+#########
+#the first deriv of the cumulant for binom in R
+cpR<-function(eta,ntrials){
+
+	denom <- 1+exp(-eta)
+	out <- ntrials/denom
+	out
+}
+
+#choose a couple values for eta and compare to 
+#calc of cp for bernoulli
+eta <- 1
+this <- cpR(eta,ntrials=1)
+that <- bernoulli.glmm()$cp(eta)
+all.equal(this,that)
+
+eta <- -1
+this <- cpR(eta,ntrials=1)
+that <- bernoulli.glmm()$cp(eta)
+all.equal(this,that)
+
+#########
+#the second deriv of the cumulant for binom in R
+cppR<-function(eta,ntrials){
+
+	first <- 1+exp(-eta)
+	second <-  1+exp(eta)
+	out <- ntrials/(first*second)
+	out
+}
+
+#choose a couple values for eta and compare to 
+#calc of cpp for bernoulli
+eta <- 1
+this <- cppR(eta,ntrials=1)
+that <- bernoulli.glmm()$cpp(eta)
+all.equal(this,that)
+
+eta <- -1
+this <- cppR(eta,ntrials=1)
+that <- bernoulli.glmm()$cpp(eta)
+all.equal(this,that)
+
+#########
+# check cumR for ntrials = 5 for both pos, neg eta
+ntrials <- 5
+eta <- -1
+this <- cumR(eta,ntrials)
+that <- 5* log(1+exp(eta))
+all.equal(this, that)
+eta <- 1
+this <- cumR(eta,ntrials)
+that <- 5* (eta+log(1+exp(-eta)))
+all.equal(this, that)
+
[…]

### 2016-06-09 glmm/R/glmm.R
added code for ntrials, y for bin. still ok.

@@ -42,6 +42,17 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	}
 	#so now random is a list containing a model matrix for each formula, and some matrices share variance components
 
+	#family stuff
+	family.glmm<-getFamily(family.glmm)
+
+	ntrials <- rep(1, length(y)) #used for Poisson and Bern, essentially untouched
+	if(family.glmm$family.glmm=="binomial.glmm"){
+		# make ntrials a vector with each entry the sum of the entries in the corresponding col of y
+		ntrials <- apply(y, MARGIN=1, FUN=sum)
+		y <- y[,1]
+	}
+	family.glmm$checkData(y)
+
 	if(is.numeric(varcomps.equal)==F) stop("varcomps.equal must be a vector containing numbers to indicate which variance components are equal.")
 	if(length(varcomps.equal)!=length(random)){
 		stop("The length of varcomps.equal must be equal to the length of the random-effects call.")} 
@@ -49,8 +60,9 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 		stop("You must name each unique variance component. Check varcomps.names and varcomps.equal.")} 
 	if(min(varcomps.equal)!=1)stop("The vector varcomps.equal must contain numbers starting at 1 to denote which variance components are equal.")	
 	levs<-ordered(unique(varcomps.equal))
-	family.glmm<-getFamily(family.glmm)
-	family.glmm$checkData(y)
+
+
+
 
 
 	#check p1 p2 p3
@@ -197,8 +209,6 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 #	genData2<-genRand(ones,zeros,mod.mcml$z,m2,distrib="normal",gamm)
 
 	umat<-rbind(genData,genData2,genData3)
-
-	ntrials <- rep(1, length(y))
 
 	#use trust to max the objfun (monte carlo likelihood)
[…]

### 2016-06-13 glmm/R/glmm.R
Added a bit to pre-processsing, produce errors
Next, need to change pql to take in ntrials and it should be good.

@@ -1,5 +1,6 @@
 glmm <-
-function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5){
+function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5)
+	{
 	if(missing(varcomps.names)) stop("Names for the variance components must be supplied through varcomps.names")
 	if(is.vector(varcomps.names)!=1) stop("varcomps.names must be a vector")
 	if(missing(varcomps.equal)){
@@ -17,6 +18,37 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
     }
     x <- model.matrix(fixed, data = barf)
     y <- model.response(barf)
+
+
+	#family stuff
+	family.glmm<-getFamily(family.glmm)
+
+	#check that only binomial has y being matrix
+	if(length(dim(y))==2){
+		if(family.glmm$family.glmm!="binomial.glmm") {
+			stop("For the family you've specified, only a vector is appropriate as the response. Binomial is the only family that allows you to specify the response as a matrix.")
+		}	
+	}
+
+	ntrials <- rep(1, length(y)) #used for Poisson and Bern, essentially untouched
+	if(family.glmm$family.glmm=="binomial.glmm"){
+		#if the response is a vector, then ntrials stays at 1
+		#if the response is a matrix
+		if(length(dim(y))==2){
+			#make sure it has exactly 2 columns
+			if(ncol(y)!=2) stop("Your response must have two columns: the first column reports the number of successes and the second column reports the number of failures.")
+
+			# make ntrials a vector with each entry the sum of the entries in the corresponding col of y
+			ntrials <- apply(y, MARGIN=1, FUN=sum)
+			y <- y[,1] 	#then change y to just be the number of successes
+		}
+
+
+	}
+	#do the check for the specified family
+	family.glmm$checkData(y)
+
+
     
     #then the part for the random effects. 
     #first, if it's not a list, make it a list 
@@ -25,33 +57,55 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
         random <- list(random)
     #put this stuff in a loop and loop along the list
[…]

### 2016-06-13 glmm/tests/binomfamtest.R

@@ -208,7 +208,7 @@ all.equal(this, that)
 #no segfault errors and the same answer. Yay!
 
 
-# one last thing to check
+# one last thing to check about the cumulant calcs
 eta<-1
 ntrials<-1
 that <- bernoulli.glmm()$cum(eta)
@@ -264,4 +264,7 @@ ntrials<-1
 that <- bernoulli.glmm()$cpp(eta)
 this <- binomial.glmm()$cpp(eta, ntrials)
 all.equal(this, that)
+
+
+
 

### 2016-06-30 NewDesignDoc/DesignDocGLMM.tex
About to start implementing bin fam in glmm
Added a bit to pre-processsing, produce errors
Next, need to change pql to take in ntrials and it should be good.
adding mcse details to design doc

@@
 
-\item I will add checks for the model matrix stage.  See below.
-\item I want to allow the response for bernoulli to be a single vector  (rather than two cbinded vectors).
-Therefore, if the family is binomial.glmm and the length of the dimension of the response is 1 (meaning we just have a vector), then create the ntrials vector to be a vector of 1s.
-Test this with salamanders or some other problem to make sure it gives the same results.
-
-\end{enumerate}
-
-Checks for the model matrix stage: 
+\item I will add checks for the model matrix stage. 
 \begin{enumerate}
@@
 \end{enumerate}
+\item I want to allow the response for bernoulli to be a single vector  (rather than two cbinded vectors).
+Therefore, if the family is binomial.glmm and the length of the dimension of the response is 1 (meaning we just have a vector), then create the ntrials vector to be a vector of 1s.
+Test this with salamanders or some other problem to make sure it gives the same results.
+\item Change PQL function to do binomial.
+\end{enumerate}
 
 These changes will need to be done in both C and R. The checks will be in R. I'll need to add  documentation for binomial. Bernoulli will be retained.
+
+\subsection{MCSE}
+Let  $\gamma_1$ be defined as
+\begin{align}
+  \gamma_1 = \int  f_\theta(u,y) \; du =  f_\theta(y). \label{eq:gamma1} 
+\end{align}
+My thesis shows the Monte Carlo error of the gradient of the log likelihood at the MLE $\hat{\theta}$ is 
+\begin{align}
+V= \dfrac{1}{\gamma_1^2}  \int \left[  \nabla \log f_{\hat{\theta}}(u,y)  \right] \left[  \nabla \log f_{\hat{\theta}}(u,y)  \right]^T \dfrac{f_{\hat{\theta}}(u,y)^2}{\tilde{f}(u)} \, du.
+\end{align}
+
+We can estimate this with the MCMLE $\hat{\theta}_m$:
+\begin{align}
+\hat{V}= \dfrac{ \dfrac{1}{m}  \sum_{k=1}^m \left( \left[\nabla \log f_{\hat{\theta}_m}(u_k,y)  \right] \left[\nabla \log f_{\hat{\theta}_m}(u_k,y)  \right]^T  \dfrac{ f_{\hat{\theta}_m}(u_k,y)^2   }{\tilde{f}(u_k)}  \right) }
+{  \left( \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_{\hat{\theta}_m}(u_k,y)   }{\tilde{f}(u_k)} \right)^2 }_.
+\end{align}
+%\begin{align}
+%\hat{V}= \dfrac{   \sum_{k=1}^m \left( \left[ \nabla \log f_{\hat{\theta}_m} (u_k,y)  - \nabla l_m({\hat{\theta}_m}|y)   \right] \left[ \nabla \log f_{\hat{\theta}_m}(u_k,y)  -\nabla l_m({\hat{\theta}_m}|y)  \right]^T  \dfrac{ f_{\hat{\theta}_m}(u_k,y)   }{\tilde{f}(u_k)}  \right) }{\sum_{k=1}^m  \dfrac{ f_{\hat{\theta}_m}(u_k,y)   }{\tilde{f}(u_k)}}_.
+%\end{align}
+
+Let U be the Hessian of the likelihood evaluated at the MLE. This is estimated by
+\begin{align}
+\hat{U} = \nabla^2 l_m (\hat{\theta}_m |y).
+\end{align}
+In other words, $\hat{U}$ is the  MCLA Hessian evaluated at the MCMLE.
+
+Then the MCSE of $\hat{\theta}_m$ is the square root of the diagonal elements of
+\begin{align} 
+\dfrac{1}{m} \, \hat{U}^{-1} \, \hat{V} \, \hat{U}^{-1}
+\end{align}
+under the regularity conditions listed in \citet{geyer:1994}.
+
+The breakdown for programming MCSE is:
+\begin{enumerate}
+\item Get $\hat{U}$ from output's \texttt{likelihood.hessian}
+\item Stably invert $\hat{U}$
+\item Calculate $\hat{V}$
+\item Finish calculation (result will be a vector of MCSE with length equal to number of parameters)
+\item Output table
+\end{enumerate}
+The output table will have a row for each parameter.
+The columns will be estimate (MCMLE), SE, MCSE.
+This will allow for easy comparison between SE and MCSE so that users can ensure their MCSE is small enough compared to the SE.
+
+Note on calculating $\hat{U}$:the likelihood's Hessian is calculated for every call of objfunc. We can get this from the output's \texttt{likelihood.hessian}.
+
+To do steps:
+\begin{enumerate}
+\item Remember how to do QR decomp (or some other stable matrix decomposition to stable invert the matrix)
+\item Calculate $\hat{V}$ and the rest of the MCSE calculation in R
+\item Change the  $\hat{V}$ calculation to C
+\end{enumerate}
+
+\subsection{Revisit Summary Table}
+I just realized I have \texttt{solve()} for calculating the variance-covariance matrix. I need to do a QR decomposition or something more stable.
 

### 2016-06-30 glmm/R/summary.mcla.R
replaced solve with chol decomp in summary
summary, vcov, and confint all used solve() before to get a matrix inverse. oops. Now these use a cholesky decomp to be more stable.

@@ -20,7 +20,10 @@ function(object,...){
                "definite, making all standard errors infinite"))
             all.ses <- rep(Inf, nrow(hessian))
         }
-	else{	all.ses<-sqrt(diag(solve(-hessian)))}
+	else{	
+		varcov <-vcov(mod.mcml)
+		all.ses<-sqrt(diag(varcov))
+		}
 	
 	beta.se<-all.ses[1:nbeta]
 	zval<-beta/beta.se
@@ -88,7 +91,12 @@ vcov.glmm <-
 function(object,...){
 	mod<-object
    	stopifnot(inherits(mod, "glmm"))
-	vcov <- -solve(mod$likelihood.hessian)
+
+	cho<-chol(-mod$likelihood.hessian)
+	blah<-diag(nrow(mod$likelihood.hessian))
+	uinv<-backsolve(cho, blah)
+	vcov<-uinv%*%t(uinv)
+
 
 	#get names for vcov matrix
 	rownames(vcov)<-colnames(vcov)<-rep(c("blah"),nrow(vcov))
@@ -128,7 +136,10 @@ confint.glmm<-function(object,parm,level=.95,...){
                "definite, making all standard errors infinite"))
             all.ses <- rep(Inf, nrow(hessian))
         }
-	else{	all.ses<-sqrt(diag(solve(-hessian)))}
+	else{	
+		varcov <-vcov(object)
+		all.ses<-sqrt(diag(varcov))
+		}
 	
 	names(all.ses)<-pnames
 	ci<-matrix(data=NA,nrow=length(parm),ncol=2)

### 2016-06-30 NewDesignDoc/DesignDocGLMM.tex
changed to qr decomp
qr decomp in vcov, summary, confint

@@
 \begin{enumerate}
-\item Remember how to do QR decomp (or some other stable matrix decomposition to stable invert the matrix)
+\item Remember how to do qr decomp (stably invert the matrix with qr.solve)
 \item Calculate $\hat{V}$ and the rest of the MCSE calculation in R
@@
 \end{enumerate}
+
+\subsubsection{Cholesky Decomposition for Matrix Inversion}
+Cholesky decomposition takes a symmetric, positive definite matrix and decomposes it into $LL'$, where $L$ is a lower triangular matrix and $L'$ is the transpose of L (so it's upper triangular).
+Triangular matrices are easy to solve using \texttt{backsolve} in R.
+For example, if you have the hessian \texttt{hess} and want to get the variance covariance matrix, then do
+\begin{verbatim}
+cho <- chol(hess)
+uinv <- backsolve(cho, diag(nrow(cho))
+vcov <- uinv %*% t(uinv)
+\end{verbatim}
 

### 2016-06-30 glmm/R/summary.mcla.R

@@ -92,10 +92,11 @@ function(object,...){
 	mod<-object
    	stopifnot(inherits(mod, "glmm"))
 
-	cho<-chol(-mod$likelihood.hessian)
-	blah<-diag(nrow(mod$likelihood.hessian))
-	uinv<-backsolve(cho, blah)
-	vcov<-uinv%*%t(uinv)
+#	cho<-chol(-mod$likelihood.hessian)
+#	blah<-diag(nrow(mod$likelihood.hessian))
+#	uinv<-backsolve(cho, blah)
+#	vcov<-uinv%*%t(uinv)
+	vcov <- qr.solve(-mod$likelihood.hessian)
 
 
 	#get names for vcov matrix

### 2016-08-11 glmm/R/mcse.R
removed extra vars from input of mcse.c

@@ -74,7 +74,7 @@ mcvcov <- function(object){
 
 	
 
-	stuff<-.C("mcsec", as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), v=double(m), ntrials=as.integer(mod$mod.mcml$ntrials), value=double(1),gradient=double(npar),hessian=double((npar)^2),PACKAGE="glmm")
+	stuff<-.C("mcsec", as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), ntrials=as.integer(mod$mod.mcml$ntrials), PACKAGE="glmm")
 
 	vhatnum <- (1/m)*stuff[[4]]
 	vhatdenom <- ( stuff[[1]]   )^2

### 2016-08-11 glmm/R/summary.mcla.R

@@ -19,12 +19,12 @@ summary.glmm <- function(object,...){
             warning(paste("estimated Fisher information matrix not positive",
                "definite, making all standard errors infinite"))
             all.ses <- rep(Inf, nrow(hessian))
-        }
-	else{	
+    }else{	
 		varcov <-vcov.glmm(mod.mcml)
 		all.ses<-sqrt(diag(varcov))
-		}
+	}
 	
+	#fixed effects table
 	beta.se<-all.ses[1:nbeta]
 	zval<-beta/beta.se
 	coefmat<-cbind(beta,beta.se,zval,2*pnorm(abs(zval),lower.tail=F))
@@ -32,6 +32,7 @@ summary.glmm <- function(object,...){
 	rownames(coefmat)<-colnames(mod.mcml$x)
 	coefficients<-coefmat[,1]
 	
+	#variance components table
 	nu<-mod.mcml$nu
 	nu.se<-all.ses[-(1:nbeta)]
 	nuzval<-nu/nu.se
@@ -40,9 +41,8 @@ summary.glmm <- function(object,...){
 	rownames(nucoefmat)<-mod.mcml$varcomps.names
 	link<-mod.mcml$family.glmm$link
 
-	
-	return(structure(list(x=x,y=y, z=z, coefmat=coefmat, fixedcall = fixedcall, randcall = randcall, coefficients = coefficients,
-family.mcml = mod.mcml$family.mcml, call = call, nucoefmat = nucoefmat, link = link, trust.converged = trust.converged),class="summary.glmm"))
+
+	return(structure(list(x=x,y=y, z=z, coefmat=coefmat, fixedcall = fixedcall, randcall = randcall, coefficients = coefficients, family.mcml = mod.mcml$family.mcml, call = call, nucoefmat = nucoefmat, link = link, trust.converged = trust.converged), class="summary.glmm"))
 }
 
 
@@ -74,6 +74,8 @@ cat("Variance Components for Random Effects (P-values are one-tailed):")
 	printCoefmat(summ$nucoefmat, digits = digits,
         signif.stars = signif.stars, na.print = "NA", ...)
    cat("\n")
+
+
 
 }
 

### 2016-08-11 glmm/src/mcse.c

@@ -7,7 +7,7 @@ pee is the vector of sampling proportions (usually 1/3, 1/3, 1/3)
 nps is the length of pee (3 for now, maybe more if imp sampling distrib changes)
 ntrials is a vec of ints with length equal to length(y)
 */
-void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian)
+void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst,  int *ntrials)
 {
 	double *Uk = Calloc(*myq, double);
 	int Uindex = 0;

### 2016-08-11 glmm/src/myheader.h

@@ -36,7 +36,7 @@ void tdist(double *DstarInv,  int *myq, double *uvec, int *zeta, double *tconst,
 
 void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian);
 
-void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian);
+void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, int *ntrials);
 
 #include <math.h>
 

### 2016-08-11 glmm/R/summary.mcla.R
added function for calculating standard errors of

@@ -46,6 +46,23 @@ summary.glmm <- function(object,...){
 }
 
 
+se <- function(object){
+    mod.mcml<-object
+    stopifnot(inherits(mod.mcml, "glmm"))
+	hessian<-mod.mcml$likelihood.hessian
+	if(det(hessian)==0) {
+            warning(paste("estimated Fisher information matrix not positive",
+               "definite, making all standard errors infinite"))
+            all.ses <- rep(Inf, nrow(hessian))
+    }else{	
+		varcov <-vcov.glmm(mod.mcml)
+		all.ses<-sqrt(diag(varcov))
+	}
+	
+	names(all.ses) <- c(colnames(mod.mcml$x), object$varcomps.names)
+	all.ses
+}
+
 ## print.summary actually displays them
 print.summary.glmm <-
     function (x, digits = max(3, getOption("digits") - 3),

### 2016-08-12 glmm/src/mcse.c
fixed squaretops in mcse/mcvcov

@@ -86,7 +86,7 @@ void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, doub
 		if(k==0){a = b[k];}
 		if(b[k]>a){a = b[k];}
 
-		squaretop[k] = exp(2*lfuval + 2*lfyuval - lfutwid);
+		squaretop[k] = exp(2*lfuval + 2*lfyuval - 2*lfutwid);
 	}
 
 	Free(lfutwidpieces);

### 2016-08-12 glmm/tests/mcseTest.R

@@ -190,7 +190,7 @@ mcseTEST <- function(mod){
 		Gpiece <- c(lfyu[[k]]$gradient,lfu[[k]]$gradient)
 		GGT <- Gpiece %*% t(Gpiece)	
 
-		squaretop <- exp(	2*as.numeric(lfu[[k]]$value) + 2*as.numeric(lfyu[[k]]$value) - lfu.twid[k,4] )
+		squaretop <- exp(	2*as.numeric(lfu[[k]]$value) + 2*as.numeric(lfyu[[k]]$value) - 2*lfu.twid[k,4] )
 	
 		numpiece[k,] <- as.numeric(GGT*squaretop)
 	}

### 2016-08-17 glmm/R/summary.mcla.R
fixing confint function's typo

@@ -154,8 +154,7 @@ confint.glmm<-function(object,parm,level=.95,...){
             warning(paste("estimated Fisher information matrix not positive",
                "definite, making all standard errors infinite"))
             all.ses <- rep(Inf, nrow(hessian))
-        }
-	else{	
+    }else{	
 		varcov <-vcov.glmm(object)
 		all.ses<-sqrt(diag(varcov))
 		}
@@ -164,8 +163,8 @@ confint.glmm<-function(object,parm,level=.95,...){
 	ci<-matrix(data=NA,nrow=length(parm),ncol=2)
 	colnames(ci)<-a
 	rownames(ci)<-parm
-	ci[,1]<-betaandnu[parm]+a[1]*all.ses[parm]
-	ci[,2]<-betaandnu[parm]+a[2]*all.ses[parm]
+	ci[,1]<-betaandnu[parm]+fac[1]*all.ses[parm]
+	ci[,2]<-betaandnu[parm]+fac[2]*all.ses[parm]
 
 	ci
 }

### 2016-09-08 JSS/article.tex
started JSS paper

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Achim Zeileis\\Universit\"at Innsbruck \And 
        Second Author\\Plus Affiliation}
\title{ Likelihood-Based Inference for Generalized Linear Mixed Models: Inference with Package \pkg{glmm}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Achim Zeileis, Second Author} %% comma-separated
\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  The abstract of the article.
}
\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@uibk.ac.at}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[About Java]{About \proglang{Java}}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

\end{document}

### 2016-09-08 JSS/article.tex
added a line to the paper

@@
 %% almost as usual
-\author{Achim Zeileis\\Universit\"at Innsbruck \And 
-        Second Author\\Plus Affiliation}
+\author{Christina Knudson\\Macalester College \And 
+        Charles Geyer\\University of Minnesota \And 
+        Galin Jones\\University of Minnesota}
 \title{ Likelihood-Based Inference for Generalized Linear Mixed Models: Inference with Package \pkg{glmm}}
@@
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Achim Zeileis, Second Author} %% comma-separated
+\Plainauthor{Christina Knudson, Charles Geyer, Galin Jones} %% comma-separated
 \Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
-\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+\Shorttitle{\pkg{glmm}: Likelihood-Based Inference for Generalized Linear Mixed Models} %% a short title (if necessary)
 
@@
 }
-\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
-\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
+\Keywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, Monte Carlo likelihood approximation, Monte Carlo maximum likelihood}
+\Plainkeywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, Monte Carlo likelihood approximation, Monte Carlo maximum likelihood} %% without formatting
 %% at least one keyword must be supplied

### 2017-01-28 JSS/article.tex
Fixed section problem, added two data descriptions

@@
 %% almost as usual
-\author{Christina Knudson\\Macalester College \And 
+\author{Christina Knudson\\ University of St. Thomas \And 
         Charles Geyer\\University of Minnesota \And 
@@
 \Abstract{
-  The abstract of the article.
+The R package \pkg{glmm} enables all  likelihood-based inference  for generalized linear mixed models with a canonical link by approximating the entire likelihood function and two derivatives.
+The model fitting function of \pkg{glmm} calculates and maximizes the Monte Carlo likelihood approximation to find Monte Carlo maximum likelihood estimates for the fixed effects and variance components.
+Additionally, the value, gradient vector, and Hessian matrix of the MCLA are calculated at the Monte Carlo maximum likelihood estimates.
+\pkg{glmm} also produces the variance-covariance matrix of the estimates, confidence intervals for the parameters, and Monte Carlo standard errors of the Monte Carlo maximum likelihood estimates.
 }
-\Keywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, Monte Carlo likelihood approximation, Monte Carlo maximum likelihood}
-\Plainkeywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, Monte Carlo likelihood approximation, Monte Carlo maximum likelihood} %% without formatting
+\Keywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, likelihood approximation, Monte Carlo}
+\Plainkeywords{generalized linear mixed models, maximum likelihood, likelihood-based inference, mixed models, likelihood approximation, Monte Carlo}
+ %% without formatting
 %% at least one keyword must be supplied
@@
 \Address{
-  Achim Zeileis\\
-  Department of Statistics and Mathematics\\
-  Faculty of Economics and Statistics\\
-  Universit\"at Innsbruck\\
-  6020 Innsbruck, Austria\\
-  E-mail: \email{Achim.Zeileis@uibk.ac.at}\\
-  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+  Christina Knudson\\
+  Department of Mathematics\\
+  Faculty of  Statistics\\
+ University of St. Thomas\\
+2115 Summit Avenue, Saint Paul, Minnesota\\
+  E-mail: \email{christina@umn.edu}\\
+  URL: \url{http://cknudson.com/}
 }
+%\Address{
+% Galin Jones\\
+%School of Statistics\\
+%  Faculty of Statistics\\
+% University of Minnesota\\
+%224 Church Street Southeast, Minneapolis, Minnesota \\
+%  E-mail: \email{galin@umn.edu}\\
+%  URL: \url{http://users.stat.umn.edu/~galin/}
+%}
 %% It is also possible to add a telephone and fax number
@@
 
-\section[About Java]{About \proglang{Java}}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+\section{Introduction}
+The generalized linear mixed model (GLMM) is an extension of both the generalized linear model and the linear mixed model; the model incorporates  fixed and random effects as well as a response from an exponential family.
+GLMMs were first discussed by \citet{stiratelli:laird:ware:1984} and are now  used in a variety of disciplines.
+Despite their widespread use, traditional methods of frequentist likelihood-based inference are not generally available.
+The challenge lies in the likelihood function for GLMMs: because the likelihood cannot depend on the random effects, the likelihood is an  integral that is often intractable (details    in FIX).
+Due to this challenge, most methodology and software for  GLMMs perform little more than maximum likelihood (such as Monte Carlo EM)  or do not perform likelihood-based inference at all (such as penalized quasi-likelihood).
+Our goal is to enable all types of frequentist likelihood-based inference.
+This includes (but is not limited to) calculating Fisher information, performing hypothesis tests, and constructing confidence intervals.
+To perform all types of frequentist likelihood-based inference, the entire likelihood function is necessary.\\
+
+With this goal in mind, we have created an R package \pkg{glmm} \citep{glmm:package} that approximates the entire likelihood function for a GLMM with either a Bernoulli or Poisson response.
+This package uses  the method of
+ Monte Carlo likelihood approximation (MCLA) \citep{geyer:1994, geyer:thom:1992,  sung:geyer:2007}. 
+%This method relies on the  choice of an importance sampling distribution,  a distribution from which simulated random effects are drawn.
+These generated random effects are then used to approximate the entire likelihood function.
+Because MCLA approximates the entire likelihood, \pkg{glmm}  enables  every type of likelihood-based inference. \\
+
+%For more details on MCLA, see \ref{sec:mclageneral}.
+
+Before the release of R package \pkg{glmm}, no publicly-available software  implementing MCLA  produced accurate maximum likelihood estimates for models of non-trivial complexity.
+For example, the \texttt{bernor}  package \citep{sung:geyer:2007} implements MCLA but cannot accurately perform maximum likelihood for the benchmark data set from a salamander mating experiment (described in section \ref{sec:examples}).
+%The  absence of MCLA software highlights a   challenge from which the procedure suffers:  finding an importance sampling distribution that works well in practice.  
+%We propose  an importance sampling distribution to be used in implementing MCLA for GLMMs (Equation {eq:ftwiddle});  
+We present   \pkg{glmm}  and demonstrate in section \ref{sec:howto} how to use the package to perform maximum likelihood, test hypotheses, calculate confidence intervals, and calculate Monte Carlo standard errors for the estimates.
+
+\subsection{Examples}\label{sec:examples}
+
+\subsubsection{The salamander data}
+
+Researchers at the University of Chicago conducted an experiment on a single species of salamanders  in 1986.
+ \citet[Section 14.5]{mcc:nelder:1989} presented the experiment and data, and \citet{karim:zeger:1992} proposed a model they call ``Model A.'' The salamander data and model have become a benchmark in the GLMM universe;  the data have been modeled by many researchers including \citet{booth:hobert:1999},  \citet{bres:clay:1993},   \citet{karim:zeger:1992}, \citet{mcc:nelder:1989},  \citet{schall:1991},  \citet{sung:geyer:2007}, and \citet{wolfinger:oconnell:1993}.\\
+
+Before the experiment began, female salamanders and male salamanders of the same species were collected from two locations.
+The salamanders were categorized into populations named ``Rough Butt'' and ``White Side'' based on their location of origin.
+The scientific goal was to determine whether salamanders were more likely to mate with those from their own population or whether they were just as likely to mate with salamanders from either population.
+More specifically, scientsts sought to compare the odds of mating for each type of  cross: female Rough Butt salamanders and male Rough Butt salamanders (denoted RR), female Rough Butt salamanders and male White Side salamanders (denoted RW), female White Side salamanders and male Rough Butt salamanders (denoted WR), and female White Side salamanders and male White Side salamanders (denoted WW).
+\\
+
+Scientists ran an experiment three times.
+Each experiment consisted of trials conducted on two closed groups of 20 salamanders.
+That is, each experiment was conducted on 40 salamanders that were split into two closed groups.
+Each group contained 5 female Rough Butts, 5 female White Sides, 5 male Rough Butts, and 5 male White Sides.
+Each trial consisted of placing a  female salamander and a  male salamander in an isolated space together, then observing the binary response of interest: whether the salamanders mated.
+Each female salamander participated in six trials with male salamanders from her closed group:  three trials with male White Side salamanders and three trials with male Rough Butt salamanders.
+Scientists paired salamanders from the same group; inter-group trials were not conducted.
+Thus, 60 trials were conducted on each closed group, each experiment consisted of 120 trials, and the overall dataset contains binary responses from 360 trials.
+\\
+
+ To model these data with a GLMM,  ``Model A'' \citep{karim:zeger:1992} proposes
+a random effect for each female salamander; a random effect for each male salamander; a fixed effect predictor for each of the four types of cross; and a Bernoulli response of whether the pair of salamanders mated, dependent on the type of cross, the female random effect, and the male random effect.
+ Each salamander's random effect is assumed to be independent of the others.
+The male salamanders' random effects share one variance component  while the female salamanders' random effects share another variance component.
+These two variance components are later referred to as $\nu_M$ and $\nu_F$.
+The males' random effects are crossed with the females' random effects.
+Though the same salamanders were used in the first two experiments, the data are traditionally modeled assuming that new salamanders were used in each experiment \citep{booth:hobert:1999,  karim:zeger:1992,  mcc:nelder:1989}.\\
+
+\subsubsection{The radish flower data}
+
+The data in this example are a subset of the data collected by \citet{ridley:ellstrand:2010}.
+The scientists were interested in whether non-native radishes had adapted to the climate they had been growing in for the last 150 years.
+In other words, they wanted to compare two types of radish to see whether each type would grow just as well in their own climate as they would in the other climate.
+
+In this dataset, the response is the number of radish flowers.
+This is assumed to have a Poisson distribution.
+\texttt{Site} is a categorical variable with two categories representing the two sites where plants were grown.
+The variable \texttt{Region} is categorical with two categories representing the two places in California from which the plants were taken.
+The variable \texttt{Pop} is a categorical variable representing the population of radish, and \texttt{Pop} is nested in \texttt{Region}.
+The variable \texttt{Block} is a categorical blocking variable nested in \texttt{Site}.
+Following the example of \citet{ridley:ellstrand:2010}, \texttt{Block} and \texttt{Pop} are random while \texttt{Site} and \texttt{Region} are fixed.
+The scientists were interested in  the interaction between \texttt{Site} and \texttt{Region}, since that would indicate that radishes grow better in the area they have been grown during recent history.
+
+\section{Monte Carlo Likelihood Approximation}
+
+\subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}}
+
+\section[How to use glmm]{How to use \pkg{glmm}}\label{sec:howto}
+
+\subsection{Formatting the data}
+\subsection{Using the model-fitting function}
+\subsection{Using the summary output}
+\subsection{Performing additional inference}
+
+\section{Conclusion}
+
+\bibliography{brref}
 

### 2017-01-31 JSS/article.tex
added to mcla section of paper

@@
 \documentclass[article]{jss}
+\usepackage{amsmath}
 
@@
  Monte Carlo likelihood approximation (MCLA) \citep{geyer:1994, geyer:thom:1992,  sung:geyer:2007}. 
-%This method relies on the  choice of an importance sampling distribution,  a distribution from which simulated random effects are drawn.
+This method relies on the  choice of an importance sampling distribution,  a distribution from which simulated random effects are drawn.
 These generated random effects are then used to approximate the entire likelihood function.
@@
 \subsection{Examples}\label{sec:examples}
+We turn to two data sets to illustrate \pkg{glmm}.
+The salamander data set serves as an example for modeling binomial-distributed data, while the radish flower data set serves as an example for modeling Poisson-distributed data.
 
@@
 The data in this example are a subset of the data collected by \citet{ridley:ellstrand:2010}.
-The scientists were interested in whether non-native radishes had adapted to the climate they had been growing in for the last 150 years.
+The scientists were interested in whether non-native radishes had adapted to the climate they had  grown in for 150 years.
 In other words, they wanted to compare two types of radish to see whether each type would grow just as well in their own climate as they would in the other climate.
+\\
 
@@
 
+MCLA is a  Monte Carlo method for approximating the entire likelihood function; it was first proposed by \citet{geyer:1990} for models with unnormalized densities, then extended to models with normalized densities and random effects \citep{thom:guo:1991}, and then extended again to models with unnormalized densities and random effects \citep{gelf:carl:1993}.
+ We focus on  MCLA  for GLMMs with random effects.  \\
+
+MCLA is powerful  because it enables any type of likelihood-based inference.
+Maximum likelihood is of special interest; the maximizer of the Monte Carlo likelihood approximation is called the Monte Carlo maximum likelihood estimate (MCMLE).
+Because MCLA is a Monte Carlo method, the accuracy of the likelihood approximation and inferences based on the likelihood approximation  can be improved by increasing the size of the Monte Carlo sample used to calculate the likelihood approximation.
+\\
+
+Let $Y \in \mathcal{R}^n$ be a response vector and let $U \in \mathcal{R}^q$ be a vector of unobservable random effects.
+The MCLA calculation requires the density for the joint distribution of the observed data  and random effects, $f_\theta(u,y)$.
+Let $f_\beta(y|u)$ denote the density for the observed data conditional on the random effects, and let $f_\nu (u)$ denote the density for the random effects.
+Then the joint density can be expressed as the product
+\begin{align}
+f_\theta (u,y) = f_\beta(y|u) \, f_\nu (u).
+\end{align}
+
+Let $\tilde{f}(u)$ be a nonnegative function whose support contains the support of the $f_\theta(u,y)$, where $\tilde{f}(u_k)$ does not depend on $\theta$.
+Let\hbox{ $u_k,$ $k=1,$ $\ldots,$ $m$}, be vectors of length $q$ drawn from  $\tilde{f}(u_k)$.
+The specific importance sampling distribution used in R package \texttt{glmm} is described in Equation FTWIDDLE.
+Then the Monte Carlo  likelihood approximation is
+\begin{align}
+L_{m}(\theta|y) &=   \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}_. \label{eq:MCLAval}
+\end{align}
+Again, the function in Equation \ref{eq:MCLAval} approximates the entire likelihood function and can be used for any likelihood-based inference.\\
+
+The asymptotic properties of the likelihood approximation as the Monte Carlo sample size increases have been studied in generality  \citep{geyer:1994, geyer:thom:1992} and   for the special case of an importance sampling distribution chosen independently of the data \citep{sung:geyer:2007}.
+\citet{geyer:1994} presents conditions under which  the Monte Carlo likelihood approximation converges almost surely to the exact likelihood for any single parameter value.
+In addition to studying the point-wise convergence,  we can also consider  the convergence of the entire likelihood function.
+\citet{geyer:1994} shows the Monte Carlo likelihood approximation converges almost surely to the likelihood function for GLMMs specified with  unnormalized densities as long as a Wald-like integrability condition is met.
+ Additionally, if the parameter space can be compactified, the MCMLE converges to the MLE almost surely \citep{geyer:1994}.
+\citet{geyer:1994} also shows that the Monte Carlo profile likelihoods converge to the exact profile likelihoods almost surely, and no additional regularity conditions are required.\\
+
+Prior to \pkg{glmm}, MCLA for GLMMs has  been studied by \citet{sung:geyer:2007}, who  focus on an MCLA implementation with an importance sampling distribution constructed independently of the observed data.
+\citet{sung:geyer:2007} prepared the R package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data.
+Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).
+\\
+
+This highlights the challenge of  finding an importance sampling distribution that performs well in practice.
+Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform maximum likelihood in practice.
+  This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems.
+In response, we present an importance sampling distribution (Equation  F TWIDDLE REF) and the R package \pkg{glmm}.\\
+
 \subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}}
+
+Need to introduce D and mention that we first need to do PQL. Also Sigma.
+
+Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
+That is, $s$ is defined such that $u= D^{1/2}s$.
+Let $\sigma$ be a vector of length $K$ with components $\sqrt{\nu_t},$ $t=1,$ $\ldots,$ $K$.
+Let $\beta^*$, $s^*$ and $\sigma^*$ denote the PQL estimates for $\beta$, $s$ and $\sigma$, respectively.
+Let
+\begin{align}
+A^*=\sum_{t=1}^K E_t \, \sigma^*_t \nonumber
+\end{align}\
+and
+\begin{align}
+ D^* =A^*A^*. \nonumber
+\end{align}
+ We can ``unstandardize'' our PQL-predicted random effects:
+\begin{align}
+u^*&=A^*s^*. \nonumber
+\end{align}
+
+Let $p_1,p_2,p_3$ be  proportions  such that  \hbox{$p_1+p_2+p_3=1$}.  Let $f(u|\mu,\Sigma)$ denote the density for a multivariate normal distribution with mean $\mu$ and variance matrix $\Sigma$.   
+Let $\grave{f}(u|0,D^*)$ denote the density for  a $q$-dimensional multivariate  $t$ distribution with mean $0$, scale matrix $D^*$, and $\zeta=5$ degrees of freedom.
+Then we propose the following importance sampling distribution:
+\begin{align}\label{eq:ftwiddle}
+\begin{split}
+ \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+\\
+&\qquad + p_3  f(u \, | \, u^*, (Z^T  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). 
+\end{split}
+\end{align}
+
+The first component  $\grave{f}( \,  \cdot \, | 0, D^*)$ of Equation \ref{eq:ftwiddle}  is chosen to ensure  the gradient of the MCLA has a central limit theorem (as shown in REF MY THESIS).
+The second component is chosen because it is centered at  $u^*$ and  has  variance $D^*$.
+The last component is centered at $u^*$ and has a variance based on the Hessian of the  penalized likelihood from PQL.
+More specifically, the Hessian of the log density of the last distribution matches the Hessian of the log density of the target distribution $f_\theta(u,y)$.
 

### 2017-02-02 JSS/article.tex
minor edits like formatting R progland
Need to change bibliography so that it is title style rather than sentence style

@@
 \Abstract{
-The R package \pkg{glmm} enables all  likelihood-based inference  for generalized linear mixed models with a canonical link by approximating the entire likelihood function and two derivatives.
+The \proglang{R} package \pkg{glmm} enables all  likelihood-based inference  for generalized linear mixed models with a canonical link by approximating the entire likelihood function and two derivatives.
 The model fitting function of \pkg{glmm} calculates and maximizes the Monte Carlo likelihood approximation to find Monte Carlo maximum likelihood estimates for the fixed effects and variance components.
@@
 Despite their widespread use, traditional methods of frequentist likelihood-based inference are not generally available.
-The challenge lies in the likelihood function for GLMMs: because the likelihood cannot depend on the random effects, the likelihood is an  integral that is often intractable (details    in FIX).
+The challenge lies in the likelihood function for GLMMs: because the likelihood cannot depend on the random effects, the likelihood is an  integral that is often intractable.
 Due to this challenge, most methodology and software for  GLMMs perform little more than maximum likelihood (such as Monte Carlo EM)  or do not perform likelihood-based inference at all (such as penalized quasi-likelihood).
+\\
+
 Our goal is to enable all types of frequentist likelihood-based inference.
@@
 
-With this goal in mind, we have created an R package \pkg{glmm} \citep{glmm:package} that approximates the entire likelihood function for a GLMM with either a Bernoulli or Poisson response.
+With this goal in mind, we have created the  \proglang{R} package \pkg{glmm} \citep{glmm:package} that approximates the entire likelihood function for a GLMM with either a Bernoulli or Poisson response.
 This package uses  the method of
@@
 
-Before the release of R package \pkg{glmm}, no publicly-available software  implementing MCLA  produced accurate maximum likelihood estimates for models of non-trivial complexity.
+Before the release of \proglang{R} package \pkg{glmm}, no publicly-available software  implementing MCLA  produced accurate maximum likelihood estimates for models of non-trivial complexity.
 For example, the \texttt{bernor}  package \citep{sung:geyer:2007} implements MCLA but cannot accurately perform maximum likelihood for the benchmark data set from a salamander mating experiment (described in section \ref{sec:examples}).
@@
 \subsection{Examples}\label{sec:examples}
-We turn to two data sets to illustrate \pkg{glmm}.
-The salamander data set serves as an example for modeling binomial-distributed data, while the radish flower data set serves as an example for modeling Poisson-distributed data.
+We  illustrate \pkg{glmm} with two data sets: the salamander data set serves as an example for modeling binomial-distributed data while the radish flower data set serves as an example for modeling Poisson-distributed data.
 
@@
 The MCLA calculation requires the density for the joint distribution of the observed data  and random effects, $f_\theta(u,y)$.
-Let $f_\beta(y|u)$ denote the density for the observed data conditional on the random effects, and let $f_\nu (u)$ denote the density for the random effects.
+Let  $\beta \in \mathcal{R}^k$ be a vector of the fixed effects and let $f_\beta(y|u)$ denote the density for the observed data conditional on the random effects.
+Let $\nu$ be a vector of variance components and let $f_\nu (u)$ denote the density for the random effects.
 Then the joint density can be expressed as the product
@@
 Let\hbox{ $u_k,$ $k=1,$ $\ldots,$ $m$}, be vectors of length $q$ drawn from  $\tilde{f}(u_k)$.
-The specific importance sampling distribution used in R package \texttt{glmm} is described in Equation FTWIDDLE.
+The specific importance sampling distribution used in \proglang{R} package \texttt{glmm} is described in Equation \ref{eq:ftwiddle}.
 Then the Monte Carlo  likelihood approximation is
@@
 Prior to \pkg{glmm}, MCLA for GLMMs has  been studied by \citet{sung:geyer:2007}, who  focus on an MCLA implementation with an importance sampling distribution constructed independently of the observed data.
-\citet{sung:geyer:2007} prepared the R package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data.
+\citet{sung:geyer:2007} prepared the \proglang{R} package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data.
 Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).
@@
   This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems.
-In response, we present an importance sampling distribution (Equation  F TWIDDLE REF) and the R package \pkg{glmm}.\\
+In response, we present an importance sampling distribution (Equation  \ref{eq:ftwiddle}) and the \proglang{R} package \pkg{glmm}.\\
 
@@
 
-Need to introduce D and mention that we first need to do PQL. Also Sigma.
+ and mention that we first need to do PQL. Also Sigma.
 
+Let $D$ be the variance-covariance matrix for the random effect vector $u$. Though $D$ is a function of the variance components, we suppress this notation for simplicity.
 Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
@@
 
-The first component  $\grave{f}( \,  \cdot \, | 0, D^*)$ of Equation \ref{eq:ftwiddle}  is chosen to ensure  the gradient of the MCLA has a central limit theorem (as shown in REF MY THESIS).
+The first component  $\grave{f}( \,  \cdot \, | 0, D^*)$ of Equation \ref{eq:ftwiddle}  is chosen to ensure  the gradient of the MCLA has a central limit theorem \citep{mythesis}.
 The second component is chosen because it is centered at  $u^*$ and  has  variance $D^*$.

### 2017-02-07 JSS/article.tex
added english sentences, need to add code

@@
 \subsection{Formatting the data}
+
+The following vectors can be used to fit a generalized linear mixed model using the \texttt{glmm} package. These vectors can be contained in a data frame, but they do not need to be.
+\begin{enumerate}
+\item A response vector.
+If your response is Poisson, then the entries in the response vector must be natural numbers.
+If your response is Bernoulli, then the entries in the response vector must be $0$ and $1$.
+(For this version of \texttt{glmm}, these are the only two response types possible.
+If you need to fit a model with a different response, contact me.)
+\item At least one vector that will be used for defining the random effects' design matrix. For this version of \texttt{glmm}, the vector(s) should be class \texttt{factor}.
+\item Vector(s) that will be used for defining the fixed effects' design matrix. The vector(s) can be of class \texttt{factor} or \texttt{numeric}. 
+\end{enumerate}
+The first two types of vectors described in the list are required.
+The last type is optional.
+That is, the minimum requirement to fit a \texttt{glmm} model is the response vector and one vector for defining the random effects' design matrix.
+\\
+
+We use the \texttt{salamander} dataset as an example in this vignette.
+For your convenience, it is already included in the \texttt{glmm} package.
+The data arose from an experiment conducted at the University of Chicago in 1986 and were first presented by \citet[section 14.5]{mcc:nelder:1989}.
+Scientists paired female and male salamanders of two types (Rough Butt and White Side) and collected data on whether or not they mated.
+\\
+
+The variable \texttt{Mate} tells us whether the pair of salamanders mated.
+The value is $1$ if they successfully mated and $0$ if they did not.
+The variable \texttt{Cross} describes the type of female and male salamander.
+For example, \texttt{Cross = W/R} indicates a White Side female was crossed with a Rough Butt male.
+The variable \texttt{Female} contains the identification number of the female salamander, and the variable \texttt{Male} contains the identification number of the male salamander.\\
+
+The first R  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
+The second line of code gives us access to the \texttt{salamander} data frame.
+The next three commands help us begin to understand the data.
+We have four variables: \texttt{Mate}, \texttt{Cross}, \texttt{Female}, and \texttt{Male}.
+The summary shows us \texttt{Mate} is numeric, \texttt{Cross} is a factor with four levels, \texttt{Female} is a factor, and \texttt{Male} is a factor.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+first input first line
+first input second line
+\end{CodeInput}
+\begin{CodeOutput}
+output of first input
+\end{CodeOutput}
+\begin{CodeInput}
+second input
+\end{CodeInput}
+\end{CodeChunk}
+
 \subsection{Using the model-fitting function}
+
+  Following Model A from \citet{karim:zeger:1992}, we set \texttt{Mate} as the response, \texttt{Cross} as the fixed effect variable, and \texttt{Female} and \texttt{Male} as the random effect variables.
+That is, we would like to fit a generalized linear mixed model with a logit link (because the response is Bernoulli).
+We will have four fixed effect parameters ($\beta_{R/R}, \beta_{R/W},\beta_{W/R},\beta_{W/W}$).
+There is likely to be variability among the females and variability among the males.
+That is, some females will be more likely to mate than other females, and we would like the model to reflect the tendencies of the individual salamanders.
+We incorporate this into the model by including a random effect for each female salamander and a random effect for each male salamander.
+We believe the female salamanders' random effects are i.i.d.
+draws from $N(0, \nu_F)$, where $\nu_F$ is an unknown parameter to be estimated.
+Similarly, we believe the male salamanders' random effects are i.i.d.
+draws from $N(0,\nu_M)$, where $\nu_M$ is an unknown parameter to be estimated.
+Finally, we believe the female and male random effects are independent of one another.
+\\
+
+In the following code, we  fit the model using the \texttt{glmm} command and save the model under the name  \texttt{sal}.
+Because \texttt{Mate} is our response, it is on the left of the $\sim$.
+We want to have a fixed effect for each of the four levels of \texttt{Cross}, so we type \texttt{Mate $\sim$ 0 + Cross}.
+Because \texttt{Cross} is a factor, typing \texttt{Mate $\sim$ Cross} would fit an equivalent model.
+\\
+
+Next, the \texttt{random} list creates the design matrices for the random effects.
+Since we want two random effects for each cross (one from the female salamander and one from the male salamander), we type \texttt{list($\sim$ 0 + Female, $\sim$ 0 + Male)}.
+We  include the \texttt{0} because we want our random effects to be centered at 0.
+Almost always, you will want your random effects to have mean 0.
+\\
+
+ Following the \texttt{random} list, the argument \texttt{varcomps.names} allows us to name the list of variance components.
+In the \texttt{random} list, we have placed the females first.
+Therefore, the order of the variance components names are first ``F'' and then ``M.''  \\
+
+Next,  we specify the name of our data set. This is an optional argument. If the data set is not specified, \texttt{glmm} looks to the parent environment for the variables you have referenced. \\
+
+After the name of the data set, we need to specify the type of the response.
+In the salamander mating example, the response is binary: the salamanders either mated or they did not.
+Therefore, the family is \texttt{bernoulli.glmm}.
+If your response is a count, then the family is \texttt{poisson.glmm}.\\
+
+Next, we specify our Monte Carlo sample size \texttt{m}.
+The general rule is the larger the Monte Carlo sample size, the more accurate the Monte Carlo likelihood approximation (MCLA) will be, and the more accurate the resulting Monte Carlo maximum likelihood estimates (MCMLEs) will be.
+Ideally, you want the largest \texttt{m} that time allows.
+For this vignette, we have chosen a Monte Carlo sample size that allows for quick computation.
+If you are interested in accuracy in the resulting estimates for the salamander model, we suggest a larger Monte Carlo sample size.
+\\
+
+%Finally, we can decide whether we would like additional output (see details in section \ref{sec:otherstuff}).
+If we would like the additional output, we type \texttt{debug = TRUE}.
+The default is \texttt{debug = FALSE}.\\
+
+We put this all together in the following commands.
+Note that we set the seed so that we can have reproducible results.
+In other words, if you set your seed to the same number and type the exact command listed below, your results should be identical to those listed here.
+Additionally, the \texttt{proc.time} commands have been used to give you an idea of how quickly the model can be fit.
+The times shown here are from fitting a model on an ultrabook that cost 500 USD in 2013.
+
 \subsection{Using the summary output}
+The \texttt{summary} command displays
+\begin{itemize}
+\item the function call (to remind you of the model you fit).
+\item the link function.
+\item the fixed effect estimates, their standard errors (calculated using observed Fisher information), their \texttt{z value} test statistics (testing whether the coefficients are significantly different from zero), the test's p-values, and the R-standard significance stars (optional).
+\item the variance component estimates, their standard errors (calculated using observed Fisher information), their \texttt{z value} test statistics (testing whether the coefficients are significantly different from zero), the test's p-values, and the R-standard significance stars (optional).
+\end{itemize}
+
+Note that the p-value for the fixed effects is calculated using a two-sided alternative hypothesis ($H_A: \beta \neq 0$) while the p-value for the variance components is calculated using a one-sided alternative hypothesis ($H_A: \nu >0$) because variance components must be nonnegative.\\
+
+R stuff here
+
+Looking at our output, we can see that the type of cross significantly affects the salamanders' odds of mating.
+Additionally, both the variance components are significantly different from zero and should be retained in the model.\\
+
+The summary provides the estimates needed to write our model.
+First, we establish a little notation.
+Let $\pi_i$ represent the probability of successful mating for salamander pair $i$.
+Let $I()$ be an indicator function, so that $I\text{(Cross=R/R)}$ is $1$ when the variable \texttt{Cross = R/R} and 0 otherwise.
+Let $u_i^F$ represent the random effect from the female salamander in the $i$th pair.
+Let $u_i^M$ represent the random effect from the male salamander in the $i$th pair.
+Since the response is Bernoulli, the canonical link is the log odds of successful mating.
+Using this notation, we write the model as follows.\\
+\begin{align*}
+\log \left( \dfrac{\pi_i}{1-\pi_i} \right) &= 0.956 * I\text{(Cross=R/R)} + 0.2805 * I\text{(Cross=R/W)}\\
+ &+ \ensuremath{-1.8968} * I\text{(Cross=W/R)}
++ 0.9723 * I\text{(Cross=W/W)}\\
+ &+ u_i^F+u_i^M\\
+u_i^F \overset{i.i.d.}{\sim} &N(0,1.288)\\
+u_i^M \overset{i.i.d.}{\sim} &N(0,1.084)\\
+\end{align*}
+
+Recall that \texttt{m}  in the above model was chosen for convenience to save time.
+The resulting parameter estimates have a little too much variability.
+If we increase \texttt{m}, the Monte Carlo standard error decreases.
+
+\subsection{Calculating confidence intervals}
+
+\subsection{Calculating Monte Carlo standard errors}
+
 \subsection{Performing additional inference}

### 2017-02-14 JSS/article.tex
edited section 2
rearranged, adding subsection headers, removed equation-y stuff for the most part. opted to describe the importance sampling distribution and refer to my thesis rather than describe it in detail in this paper.

@@
 
+Prior to \pkg{glmm}, MCLA for GLMMs was   studied by \citet{sung:geyer:2007}, who  focused on an MCLA implementation with an importance sampling distribution constructed independently of the observed data.
+\citet{sung:geyer:2007} prepared the \proglang{R} package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data.
+Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).
+\\
+
+This highlights the challenge of  finding an importance sampling distribution that performs well in practice.
+Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform  likelihood-based inference in practice.
+  This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems.
+In response, we present an importance sampling distribution \citep{mythesis} and the \proglang{R} package \pkg{glmm}.\\
+
+\subsection{Calculation of the Monte Carlo likelihood approximation}
+
 Let $Y \in \mathcal{R}^n$ be a response vector and let $U \in \mathcal{R}^q$ be a vector of unobservable random effects.
@@
 Let $\tilde{f}(u)$ be a nonnegative function whose support contains the support of the $f_\theta(u,y)$, where $\tilde{f}(u_k)$ does not depend on $\theta$.
+The  importance sampling distribution used in \proglang{R} package \texttt{glmm} is described in section \ref{sec:myimplem} and by \citet{mythesis}.
 Let\hbox{ $u_k,$ $k=1,$ $\ldots,$ $m$}, be vectors of length $q$ drawn from  $\tilde{f}(u_k)$.
-The specific importance sampling distribution used in \proglang{R} package \texttt{glmm} is described in Equation \ref{eq:ftwiddle}.
 Then the Monte Carlo  likelihood approximation is
@@
 \end{align}
-Again, the function in Equation \ref{eq:MCLAval} approximates the entire likelihood function and can be used for any likelihood-based inference.\\
+Because the function in Equation \ref{eq:MCLAval} approximates the entire likelihood function, it can be used to conduct any likelihood-based inference.\\
 
-The asymptotic properties of the likelihood approximation as the Monte Carlo sample size increases have been studied in generality  \citep{geyer:1994, geyer:thom:1992} and   for the special case of an importance sampling distribution chosen independently of the data \citep{sung:geyer:2007}.
+\subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}} \label{sec:myimplem}
+
+The importance sampling distribution used in \pkg{glmm} depends on the observed data and is specified in detail by \citet{mythesis}.
+It is a mixture of three distributions: two normal distributions and a t distribution.
+The t distribution is centered at 0 with a covariance matrix that depends on penalized quasi-likelihood (PQL) estimates of the variance components.
+The first normal distribution is centered at the PQL predictions of the random effects and has a variance matrix that depends on the PQL estimates of the variance components.
+The second normal distribution is  centered at the PQL predictions of the random effects and has a covariance matrix based on the Hessian of the  penalized likelihood from PQL.
+More specifically, the Hessian of the log density of the last distribution matches the Hessian of the log density of the target distribution $f_\theta(u,y)$.
+
+\subsection{Asymptotic properties of the Monte Carlo likelihood approximation}
+
+The asymptotic properties of the likelihood approximation as the Monte Carlo sample size increases have been studied in generality  \citep{geyer:1994, geyer:thom:1992} and   for the specific implementation in \pkg{glmm} \citep{mythesis}.
 \citet{geyer:1994} presents conditions under which  the Monte Carlo likelihood approximation converges almost surely to the exact likelihood for any single parameter value.
@@
 \citet{geyer:1994} also shows that the Monte Carlo profile likelihoods converge to the exact profile likelihoods almost surely, and no additional regularity conditions are required.\\
-
-Prior to \pkg{glmm}, MCLA for GLMMs has  been studied by \citet{sung:geyer:2007}, who  focus on an MCLA implementation with an importance sampling distribution constructed independently of the observed data.
-\citet{sung:geyer:2007} prepared the \proglang{R} package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data.
-Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).
-\\
-
-This highlights the challenge of  finding an importance sampling distribution that performs well in practice.
-Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform maximum likelihood in practice.
-  This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems.
-In response, we present an importance sampling distribution (Equation  \ref{eq:ftwiddle}) and the \proglang{R} package \pkg{glmm}.\\
-
-\subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}}
-
- and mention that we first need to do PQL. Also Sigma.
-
-Let $D$ be the variance-covariance matrix for the random effect vector $u$. Though $D$ is a function of the variance components, we suppress this notation for simplicity.
-Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
-That is, $s$ is defined such that $u= D^{1/2}s$.
-Let $\sigma$ be a vector of length $K$ with components $\sqrt{\nu_t},$ $t=1,$ $\ldots,$ $K$.
-Let $\beta^*$, $s^*$ and $\sigma^*$ denote the PQL estimates for $\beta$, $s$ and $\sigma$, respectively.
-Let
-\begin{align}
-A^*=\sum_{t=1}^K E_t \, \sigma^*_t \nonumber
-\end{align}\
-and
-\begin{align}
- D^* =A^*A^*. \nonumber
-\end{align}
- We can ``unstandardize'' our PQL-predicted random effects:
-\begin{align}
-u^*&=A^*s^*. \nonumber
-\end{align}
-
-Let $p_1,p_2,p_3$ be  proportions  such that  \hbox{$p_1+p_2+p_3=1$}.  Let $f(u|\mu,\Sigma)$ denote the density for a multivariate normal distribution with mean $\mu$ and variance matrix $\Sigma$.   
-Let $\grave{f}(u|0,D^*)$ denote the density for  a $q$-dimensional multivariate  $t$ distribution with mean $0$, scale matrix $D^*$, and $\zeta=5$ degrees of freedom.
-Then we propose the following importance sampling distribution:
-\begin{align}\label{eq:ftwiddle}
-\begin{split}
- \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+\\
-&\qquad + p_3  f(u \, | \, u^*, (Z^T  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). 
-\end{split}
-\end{align}
-
-The first component  $\grave{f}( \,  \cdot \, | 0, D^*)$ of Equation \ref{eq:ftwiddle}  is chosen to ensure  the gradient of the MCLA has a central limit theorem \citep{mythesis}.
-The second component is chosen because it is centered at  $u^*$ and  has  variance $D^*$.
-The last component is centered at $u^*$ and has a variance based on the Hessian of the  penalized likelihood from PQL.
-More specifically, the Hessian of the log density of the last distribution matches the Hessian of the log density of the target distribution $f_\theta(u,y)$.
 

### 2017-02-14 JSS/article.tex
started to edit section on formatting data

@@
 
-\subsubsection{The salamander data}
+\subsubsection{The salamander data}\label{sec:salex}
 
@@
 \begin{enumerate}
-\item A response vector.
+\item The response vector(s).
+If your response is Poisson or Bernoulli, then you should have a single response vector.
 If your response is Poisson, then the entries in the response vector must be natural numbers.
 If your response is Bernoulli, then the entries in the response vector must be $0$ and $1$.
-(For this version of \texttt{glmm}, these are the only two response types possible.
-If you need to fit a model with a different response, contact me.)
+If your response is binomial, then you should have two vectors connected with \texttt{cbind}: the first vector should be the number of successes and the second should be the number of failures.
 \item At least one vector that will be used for defining the random effects' design matrix. For this version of \texttt{glmm}, the vector(s) should be class \texttt{factor}.
@@
 
-We use the \texttt{salamander} dataset as an example in this vignette.
-For your convenience, it is already included in the \texttt{glmm} package.
+\subsection{Salamander data analysis}
+
+Recall the salamander experiment described in Section \ref{sec:examples}.
+For your convenience, the \texttt{salamander} dataset is  included in the \texttt{glmm} package.
 The data arose from an experiment conducted at the University of Chicago in 1986 and were first presented by \citet[section 14.5]{mcc:nelder:1989}.
-Scientists paired female and male salamanders of two types (Rough Butt and White Side) and collected data on whether or not they mated.
 \\
@@
 
-The first R  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
+The first \proglang{R}  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
 The second line of code gives us access to the \texttt{salamander} data frame.

### 2017-02-14 JSS/article.tex
editing sect 3
moving salamander description, adding r analysis

@@
 
-\subsection{Examples}\label{sec:examples}
-We  illustrate \pkg{glmm} with two data sets: the salamander data set serves as an example for modeling binomial-distributed data while the radish flower data set serves as an example for modeling Poisson-distributed data.
-
-\subsubsection{The salamander data}\label{sec:salex}
-
-Researchers at the University of Chicago conducted an experiment on a single species of salamanders  in 1986.
- \citet[Section 14.5]{mcc:nelder:1989} presented the experiment and data, and \citet{karim:zeger:1992} proposed a model they call ``Model A.'' The salamander data and model have become a benchmark in the GLMM universe;  the data have been modeled by many researchers including \citet{booth:hobert:1999},  \citet{bres:clay:1993},   \citet{karim:zeger:1992}, \citet{mcc:nelder:1989},  \citet{schall:1991},  \citet{sung:geyer:2007}, and \citet{wolfinger:oconnell:1993}.\\
-
-Before the experiment began, female salamanders and male salamanders of the same species were collected from two locations.
-The salamanders were categorized into populations named ``Rough Butt'' and ``White Side'' based on their location of origin.
-The scientific goal was to determine whether salamanders were more likely to mate with those from their own population or whether they were just as likely to mate with salamanders from either population.
-More specifically, scientsts sought to compare the odds of mating for each type of  cross: female Rough Butt salamanders and male Rough Butt salamanders (denoted RR), female Rough Butt salamanders and male White Side salamanders (denoted RW), female White Side salamanders and male Rough Butt salamanders (denoted WR), and female White Side salamanders and male White Side salamanders (denoted WW).
-\\
-
-Scientists ran an experiment three times.
-Each experiment consisted of trials conducted on two closed groups of 20 salamanders.
-That is, each experiment was conducted on 40 salamanders that were split into two closed groups.
-Each group contained 5 female Rough Butts, 5 female White Sides, 5 male Rough Butts, and 5 male White Sides.
-Each trial consisted of placing a  female salamander and a  male salamander in an isolated space together, then observing the binary response of interest: whether the salamanders mated.
-Each female salamander participated in six trials with male salamanders from her closed group:  three trials with male White Side salamanders and three trials with male Rough Butt salamanders.
-Scientists paired salamanders from the same group; inter-group trials were not conducted.
-Thus, 60 trials were conducted on each closed group, each experiment consisted of 120 trials, and the overall dataset contains binary responses from 360 trials.
-\\
-
- To model these data with a GLMM,  ``Model A'' \citep{karim:zeger:1992} proposes
-a random effect for each female salamander; a random effect for each male salamander; a fixed effect predictor for each of the four types of cross; and a Bernoulli response of whether the pair of salamanders mated, dependent on the type of cross, the female random effect, and the male random effect.
- Each salamander's random effect is assumed to be independent of the others.
-The male salamanders' random effects share one variance component  while the female salamanders' random effects share another variance component.
-These two variance components are later referred to as $\nu_M$ and $\nu_F$.
-The males' random effects are crossed with the females' random effects.
-Though the same salamanders were used in the first two experiments, the data are traditionally modeled assuming that new salamanders were used in each experiment \citep{booth:hobert:1999,  karim:zeger:1992,  mcc:nelder:1989}.\\
-
-\subsubsection{The radish flower data}
-
-The data in this example are a subset of the data collected by \citet{ridley:ellstrand:2010}.
-The scientists were interested in whether non-native radishes had adapted to the climate they had  grown in for 150 years.
-In other words, they wanted to compare two types of radish to see whether each type would grow just as well in their own climate as they would in the other climate.
-\\
-
-In this dataset, the response is the number of radish flowers.
-This is assumed to have a Poisson distribution.
-\texttt{Site} is a categorical variable with two categories representing the two sites where plants were grown.
-The variable \texttt{Region} is categorical with two categories representing the two places in California from which the plants were taken.
-The variable \texttt{Pop} is a categorical variable representing the population of radish, and \texttt{Pop} is nested in \texttt{Region}.
-The variable \texttt{Block} is a categorical blocking variable nested in \texttt{Site}.
-Following the example of \citet{ridley:ellstrand:2010}, \texttt{Block} and \texttt{Pop} are random while \texttt{Site} and \texttt{Region} are fixed.
-The scientists were interested in  the interaction between \texttt{Site} and \texttt{Region}, since that would indicate that radishes grow better in the area they have been grown during recent history.
-
 \section{Monte Carlo Likelihood Approximation}
@@
 
+\section{Examples}\label{sec:examples}
+We  illustrate \pkg{glmm} with two data sets: the salamander data set serves as an example for modeling binomial-distributed data while the radish flower data set serves as an example for modeling Poisson-distributed data.
+This section provides background information on  the research conducted and Section \ref{sec:howto} demonstrates the analysis using \pkg{glmm}.
+
+\subsection{The salamander data}\label{sec:salex}
+
+Researchers at the University of Chicago conducted an experiment on a single species of salamanders  in 1986.
+ \citet[Section 14.5]{mcc:nelder:1989} presented the experiment and data, and \citet{karim:zeger:1992} proposed a model they call ``Model A.'' The salamander data and model have become a benchmark in the GLMM universe;  the data have been modeled by many researchers including \citet{booth:hobert:1999},  \citet{bres:clay:1993},   \citet{karim:zeger:1992}, \citet{mcc:nelder:1989},  \citet{schall:1991},  \citet{sung:geyer:2007}, and \citet{wolfinger:oconnell:1993}.\\
+
+Before the experiment began, female salamanders and male salamanders of the same species were collected from two locations.
+The salamanders were categorized into populations named ``Rough Butt'' and ``White Side'' based on their location of origin.
+The scientific goal was to determine whether salamanders were more likely to mate with those from their own population or whether they were just as likely to mate with salamanders from either population.
+More specifically, scientsts sought to compare the odds of mating for each type of  cross: female Rough Butt salamanders and male Rough Butt salamanders (denoted RR), female Rough Butt salamanders and male White Side salamanders (denoted RW), female White Side salamanders and male Rough Butt salamanders (denoted WR), and female White Side salamanders and male White Side salamanders (denoted WW).
+\\
+
+Scientists ran an experiment three times.
+Each experiment consisted of trials conducted on two closed groups of 20 salamanders.
+That is, each experiment was conducted on 40 salamanders that were split into two closed groups.
+Each group contained 5 female Rough Butts, 5 female White Sides, 5 male Rough Butts, and 5 male White Sides.
+Each trial consisted of placing a  female salamander and a  male salamander in an isolated space together, then observing the binary response of interest: whether the salamanders mated.
+Each female salamander participated in six trials with male salamanders from her closed group:  three trials with male White Side salamanders and three trials with male Rough Butt salamanders.
+Scientists paired salamanders from the same group; inter-group trials were not conducted.
+Thus, 60 trials were conducted on each closed group, each experiment consisted of 120 trials, and the overall dataset contains binary responses from 360 trials.
+\\
+
+ To model these data with a GLMM,  ``Model A'' \citep{karim:zeger:1992} proposes
+a random effect for each female salamander; a random effect for each male salamander; a fixed effect predictor for each of the four types of cross; and a Bernoulli response of whether the pair of salamanders mated, dependent on the type of cross, the female random effect, and the male random effect.
+ Each salamander's random effect is assumed to be independent of the others.
+The male salamanders' random effects share one variance component  while the female salamanders' random effects share another variance component.
+These two variance components are later referred to as $\nu_M$ and $\nu_F$.
+The males' random effects are crossed with the females' random effects.
+Though the same salamanders were used in the first two experiments, the data are traditionally modeled assuming that new salamanders were used in each experiment \citep{booth:hobert:1999,  karim:zeger:1992,  mcc:nelder:1989}.\\
+
+ For your convenience, the \texttt{salamander} dataset is  included in the \texttt{glmm} package.
+The first \proglang{R}  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
+The second line of code gives us access to the \texttt{salamander} data frame.
+The next three commands help us begin to understand the data.
+We have four variables: \texttt{Mate}, \texttt{Cross}, \texttt{Female}, and \texttt{Male}.
+The summary shows us \texttt{Mate} is numeric, \texttt{Cross} is a factor with four levels, \texttt{Female} is a factor, and \texttt{Male} is a factor.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+library(glmm)
+data(salamander)
+names(salamander)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] "Mate" "Cross" "Female" "Male"
+\end{CodeOutput}
+\begin{CodeInput}
+head(salamander)
+\end{CodeInput}
+\begin{CodeOutput}
+   Mate Cross Female Male
+ 1    1   R/R     10   10
+ 2    1   R/R     11   14
+ 3    1   R/R     12   11
+ 4    1   R/R     13   13
+ 5    1   R/R     14   12
+ 6    1   R/W     15   28
+\end{CodeOutput}
+\end{CodeChunk}
+
+The variable \texttt{Mate} tells us whether the pair of salamanders mated: the value is $1$ if they successfully mated and $0$ if they did not.
+The variable \texttt{Cross} describes the type of female and male salamander.
+For example, \texttt{Cross = W/R} indicates a White Side female was crossed with a Rough Butt male.
+The variable \texttt{Female} contains the identification number of the female salamander, and the variable \texttt{Male} contains the identification number of the male salamander.\\
+
+\subsection{The radish flower data}
+
+The data in this example are a subset of the data collected by \citet{ridley:ellstrand:2010}.
+The scientists were interested in whether non-native radishes had adapted to the climate they had  grown in for 150 years.
+In other words, they wanted to compare two types of radish to see whether each type would grow just as well in their own climate as they would in the other climate.
+\\
+
+In this dataset, the response is the number of radish flowers.
+This is assumed to have a Poisson distribution.
+\texttt{Site} is a categorical variable with two categories representing the two sites where plants were grown.
+The variable \texttt{Region} is categorical with two categories representing the two places in California from which the plants were taken.
+The variable \texttt{Pop} is a categorical variable representing the population of radish, and \texttt{Pop} is nested in \texttt{Region}.
+The variable \texttt{Block} is a categorical blocking variable nested in \texttt{Site}.
+Following the example of \citet{ridley:ellstrand:2010}, \texttt{Block} and \texttt{Pop} are random while \texttt{Site} and \texttt{Region} are fixed.
+The scientists were interested in  the interaction between \texttt{Site} and \texttt{Region}, since that would indicate that radishes grow better in the area they have been grown during recent history.
+
 \section[How to use glmm]{How to use \pkg{glmm}}\label{sec:howto}
@@
 
-\subsection{Salamander data analysis}
-
-Recall the salamander experiment described in Section \ref{sec:examples}.
-For your convenience, the \texttt{salamander} dataset is  included in the \texttt{glmm} package.
-The data arose from an experiment conducted at the University of Chicago in 1986 and were first presented by \citet[section 14.5]{mcc:nelder:1989}.
-\\
-
-The variable \texttt{Mate} tells us whether the pair of salamanders mated.
-The value is $1$ if they successfully mated and $0$ if they did not.
-The variable \texttt{Cross} describes the type of female and male salamander.
-For example, \texttt{Cross = W/R} indicates a White Side female was crossed with a Rough Butt male.
-The variable \texttt{Female} contains the identification number of the female salamander, and the variable \texttt{Male} contains the identification number of the male salamander.\\
-
-The first \proglang{R}  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
-The second line of code gives us access to the \texttt{salamander} data frame.
-The next three commands help us begin to understand the data.
-We have four variables: \texttt{Mate}, \texttt{Cross}, \texttt{Female}, and \texttt{Male}.
-The summary shows us \texttt{Mate} is numeric, \texttt{Cross} is a factor with four levels, \texttt{Female} is a factor, and \texttt{Male} is a factor.
-
 \begin{CodeChunk}
 \begin{CodeInput}
-first input first line
-first input second line
+input code
 \end{CodeInput}
 \begin{CodeOutput}
-output of first input
+output code
 \end{CodeOutput}

### 2017-03-07 JSS/article.tex
implemented charlie's feb edits

@@
 Our goal is to enable all types of frequentist likelihood-based inference.
-This includes (but is not limited to) calculating Fisher information, performing hypothesis tests, and constructing confidence intervals.
+This includes (but is not limited to) calculating Fisher information, performing hypothesis tests, and constructing likelihood-based confidence intervals.
 To perform all types of frequentist likelihood-based inference, the entire likelihood function is necessary.\\
@@
 Before the release of \proglang{R} package \pkg{glmm}, no publicly-available software  implementing MCLA  produced accurate maximum likelihood estimates for models of non-trivial complexity.
-For example, the \texttt{bernor}  package \citep{sung:geyer:2007} implements MCLA but cannot accurately perform maximum likelihood for the benchmark data set from a salamander mating experiment (described in section \ref{sec:examples}).
+For example, the \texttt{bernor}  package \citep{bernor} implements MCLA but cannot accurately perform maximum likelihood for the benchmark data set from a salamander mating experiment (described in section \ref{sec:examples}).
 %The  absence of MCLA software highlights a   challenge from which the procedure suffers:  finding an importance sampling distribution that works well in practice.  
@@
 This highlights the challenge of  finding an importance sampling distribution that performs well in practice.
-Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform  likelihood-based inference in practice.
+Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require require such large Monte Carlo
+    sample sizes that computers cannot perform  likelihood-based inference in practice.
   This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems.
@@
 Let  $\beta \in \mathcal{R}^k$ be a vector of the fixed effects and let $f_\beta(y|u)$ denote the density for the observed data conditional on the random effects.
-Let $\nu$ be a vector of variance components and let $f_\nu (u)$ denote the density for the random effects.
+Let $\nu$ be a vector of variance components and let $f_\nu (u)$ denote the (marginal) density for the random effects.
 Then the joint density can be expressed as the product
@@
 Researchers at the University of Chicago conducted an experiment on a single species of salamanders  in 1986.
- \citet[Section 14.5]{mcc:nelder:1989} presented the experiment and data, and \citet{karim:zeger:1992} proposed a model they call ``Model A.'' The salamander data and model have become a benchmark in the GLMM universe;  the data have been modeled by many researchers including \citet{booth:hobert:1999},  \citet{bres:clay:1993},   \citet{karim:zeger:1992}, \citet{mcc:nelder:1989},  \citet{schall:1991},  \citet{sung:geyer:2007}, and \citet{wolfinger:oconnell:1993}.\\
+ \citet[Section 14.5]{mcc:nelder:1989} presented the experiment and data, and \citet{karim:zeger:1992} proposed a model they call ``Model A.'' The salamander data and model have become a benchmark in the GLMM universe;  the data have been reanalyzed by many researchers including \citet{booth:hobert:1999},  \citet{bres:clay:1993},   \citet{karim:zeger:1992}, \citet{mcc:nelder:1989},  \citet{schall:1991},  \citet{sung:geyer:2007}, and \citet{wolfinger:oconnell:1993}.\\
 
@@
 
- For your convenience, the \texttt{salamander} dataset is  included in the \texttt{glmm} package.
+ For the convenience of users, the \texttt{salamander} dataset is  included in the \texttt{glmm} package.
 The first \proglang{R}  command shown below gives us access to the \texttt{glmm} package and and all of its commands.
@@
  6    1   R/W     15   28
+\end{CodeOutput}
+\begin{CodeInput}
+str(salamander)
+\end{CodeInput}
+\begin{CodeOutput}
+'data.frame':	360 obs. of  4 variables:
+ $ Mate  : num  1 1 1 1 1 1 0 0 0 0 ...
+ $ Cross : Factor w/ 4 levels "R/R","R/W","W/R",..: 1 1 1 1 1 2 2 2 2 2 ...
+ $ Female: Factor w/ 60 levels "10","11","12",..: 1 2 3 4 5 6 7 8 9 10 ...
+ $ Male  : Factor w/ 60 levels "10","11","12",..: 1 5 2 4 3 19 18 16 20 17 ...
 \end{CodeOutput}
@@
 
-In the following code, we  fit the model using the \texttt{glmm} command and save the model under the name  \texttt{sal}.
+\begin{CodeChunk}
+\begin{CodeInput}
+input code
+\end{CodeInput}
+\begin{CodeOutput}
+output code
+\end{CodeOutput}
+\begin{CodeInput}
+second input
+\end{CodeInput}
+\end{CodeChunk}
+
+In the  code above, we  fit the model using the \texttt{glmm} command and save the model under the name  \texttt{sal}.
+The fixed and random effects specification uses the formula mini-language in \proglang{R}.
 Because \texttt{Mate} is our response, it is on the left of the $\sim$.
 We want to have a fixed effect for each of the four levels of \texttt{Cross}, so we type \texttt{Mate $\sim$ 0 + Cross}.
-Because \texttt{Cross} is a factor, typing \texttt{Mate $\sim$ Cross} would fit an equivalent model.
+Because \texttt{Cross} is a factor, typing \texttt{Mate $\sim$ Cross} would fit a similar model but would use \texttt{R/R} as a reference group.
+That is, this model would estimate the log odds of two rough butts mating ($\beta_{R/R}$) and three  regression coefficients ($\beta_{R/W}-\beta_{R/R}$, $\beta_{W/R}-\beta_{R/R}$, and $\beta_{W/W}-\beta_{R/R}$) that compare the log odds of mating for each of the crosses to the log odds of two rough butts mating.
 \\
@@
 Since we want two random effects for each cross (one from the female salamander and one from the male salamander), we type \texttt{list($\sim$ 0 + Female, $\sim$ 0 + Male)}.
-We  include the \texttt{0} because we want our random effects to be centered at 0.
-Almost always, you will want your random effects to have mean 0.
-\\
+We almost always  want to include \texttt{0 +} in the random effects specification.
+We  include the \texttt{0 +} because we do not want one of the salamanders to be the reference salamander.\\
 

### 2017-07-25 glmm/src/glmm_init.c
registered C routines

@@ -1,0 +1,75 @@
+#include <stdlib.h> // for NULL
+#include <R_ext/Rdynload.h>
+
+/* FIXME: 
+   Check these declarations against the C/Fortran source code.
+*/
+
+/* .C calls */
+extern void elc(void *, void *, void *, void *, void *, void *, void *, void *, void *, void *);
+extern void matdet(void *, void *, void *);
+extern void matinv(void *, void *, void *);
+extern void matmatmult(void *, void *, void *, void *, void *, void *);
+extern void matsmash(void *, void *, void *, void *);
+extern void matsolve(void *, void *, void *, void *);
+extern void matvecmult(void *, void *, void *, void *, void *);
+extern void mcsec(void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *);
+extern void objfunc(void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *, void *);
+
+extern void matTvecmult(void *, void *, void *, void *, void *);
+extern void diag(void *,void *,void *);
+extern void matTmatmult(void *, void *, void *, void *, void *, void *);
+extern void identmat(void *, void *);
+extern void sumup(void *, void *, void *);
+extern void subvec(void *, void *, void *, void *);
+extern void addvec(void *, void *, void *, void *);
+extern void divvec(void *, void *, void *, void *);
+
+extern void distRandGenC(void *,void *,void *,void *, void *,void *);
+extern void distRand3C(void *,void *, void *,void *, void *,void *, void *, void *);
+
+extern void cum3(void *,void *,void *,void *, void *);
+extern void cp3(void *,void *,void *,void *, void *);
+extern void cpp3(void *,void *,void *,void *, void *);
+
+extern void elval(void *,void *,void *,void *,void *,void *,void *);
[…]

### 2018-06-05 glmm/tests/basicBHtest.R
Minor adjustments

@@ -1,5 +1,1 @@
-library(glmm)
-data(BoothHobert)
-set.seed(123)
-mod.mcml1<-glmm(y~0+x1, list(y~0+z1), varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=1000, doPQL=TRUE)
-summary(mod.mcml1)
+sample.interval=20000

### 2018-08-08 glmm/R/glmm.R
fixed cran check errors

@@ -1,5 +1,5 @@
 glmm <-
-function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5)
+function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5, cores=NULL)
 	{
 	if(missing(varcomps.names)) stop("Names for the variance components must be supplied through varcomps.names")
 	if(is.vector(varcomps.names)!=1) stop("varcomps.names must be a vector")
@@ -266,9 +266,11 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 #	genData2<-genRand(ones,zeros,mod.mcml$z,m2,distrib="normal",gamm)
 
 	umat<-rbind(genData,genData2,genData3)
+	
+	no_cores <- min(cores, max(1, detectCores()-1))
 
 	#use trust to max the objfun (monte carlo likelihood)
-	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=zeta, ntrials = ntrials)
+	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug, nbeta=length(beta.pql), nu.pql=nu.pql, umat=umat, mod.mcml=mod.mcml, family.glmm=family.glmm, u.star=u.star,  cache=cache,  p1=p1,p2=p2, p3=p3,m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=zeta, ntrials = ntrials, no_core=no_cores)
 
 	beta.trust<-trust.out$argument[1:length(beta.pql)]
[…]

### 2018-08-08 glmm/R/objfun.R

@@ -59,6 +59,8 @@ function(par, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, cache, p1, p2,
 #		u.swoop<-umat[k,]
 #		umat[k,]<-u.swoop*Afornu
 #		}
+	
+	miniu <- NULL
 
 	#parallelizing the calculations for the value of the log-likelihood approximation and gradient
 	cl <- makeCluster(no_cores)

### 2018-08-08 glmm/tests/objfunTest.R

@@ -3,7 +3,7 @@ library(glmm)
 data(BoothHobert)
 set.seed(1234)
 out<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE)
+family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cores=2)
 mod.mcml<-out$mod.mcml
 debug<-out$debug
 nu.pql<-debug$nu.pql
@@ -42,7 +42,7 @@ Sigmuh<-solve(Sigmuh.inv)
 p1=p2=p3=1/3
 zeta=5
 
-no_cores <- max(1, detectCores() - 1)
+no_cores <- debug$no_cores
 
 # define a few things that will be used for finite differences
 lth<-objfun(par=par, nbeta=1, nu.pql=nu.pql, umat=umat, u.star=u.pql, mod.mcml=mod.mcml, family.glmm=family.glmm,p1=p1,p2=p2,p3=p3,m1=m1, Sigmuh=Sigmuh, D.star=D.star, Sigmuh.inv= Sigmuh.inv, zeta=zeta, ntrials=ntrials, no_cores=no_cores)

### 2018-08-08 glmm/tests/salamFiniteDiffs.R

@@ -1,7 +1,7 @@
 library(glmm)
 set.seed(1234)
 data(salamander)
-sal<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"), data=salamander,family.glmm=bernoulli.glmm,m=100,debug=TRUE,doPQL=FALSE)
+sal<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"), data=salamander,family.glmm=bernoulli.glmm,m=100,debug=TRUE,doPQL=FALSE, cores=2)
 
 objfun<-glmm:::objfun
 beta<-rep(0,4)
@@ -44,10 +44,12 @@ Sigmuh<-solve(Sigmuh.inv)
 del<-rep(10^-6,6)
 ntrials<-1
 
-ltheta<-objfun(par, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, 
-    cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials=ntrials)
+no_cores <- debug$no_cores
 
-lthetadel<-objfun(par+del, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials=ntrials) 
+ltheta<-objfun(par, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, 
+    cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials=ntrials, no_cores=no_cores)
+
+lthetadel<-objfun(par+del, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials=ntrials, no_cores=no_cores)
 
 #do finite diffs to check value --> gradient
 c(as.vector(ltheta$gradient%*%del),lthetadel$value-ltheta$value)

### 2018-08-08 glmm/tests/testpiecesBH.R

@@ -1,7 +1,7 @@
 library(glmm)
 data(BoothHobert)
 set.seed(1234)
-mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE)
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cores=2)
 
 mod.mcml<-mod.mcml1$mod.mcml
 z<-mod.mcml$z[[1]]
@@ -20,6 +20,8 @@ family.glmm<-bernoulli.glmm
 objfun<-glmm:::objfun
 getEk<-glmm:::getEk
 addVecs<-glmm:::addVecs
+
+no_cores <- mod.mcml1$debug$no_cores
 
 ############################################
 #this should be the same as elc
@@ -189,7 +191,7 @@ piece3<-rep(0,3)
 #calculate objfun's value for comparison
 cache<-new.env(parent = emptyenv())
 objfun<-glmm:::objfun
-that<-objfun(c(beta,nu), nbeta=1, nu.pql=nu.pql, u.star=u.star, mod.mcml=mod.mcml, family.glmm=bernoulli.glmm,cache=cache,umat=umat, p1=1/3, p2=1/3, p3=1/3, m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=5, ntrials=1)
+that<-objfun(c(beta,nu), nbeta=1, nu.pql=nu.pql, u.star=u.star, mod.mcml=mod.mcml, family.glmm=bernoulli.glmm,cache=cache,umat=umat, p1=1/3, p2=1/3, p3=1/3, m1=m1, D.star=D.star, Sigmuh=Sigmuh, Sigmuh.inv=Sigmuh.inv, zeta=5, ntrials=1, no_cores=no_cores)
 
 #get t stuff ready
 tconstant<-glmm:::tconstant

### 2018-08-09 glmm/R/glmm.R
added design doc

@@ -292,5 +292,5 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	trust.converged=trust.out$converged,  mod.mcml=mod.mcml,
 	 fixedcall=fixed,randcall=randcall, x=x,y=y, z=random,
 	family.glmm=family.glmm, call=call, varcomps.names=varcomps.names, 
-	varcomps.equal=varcomps.equal, umat=umat, pvec=c(p1, p2, p3), beta.pql=beta.pql, nu.pql=nu.pql, u.pql=u.star, zeta=zeta, debug=debug), class="glmm"))
+	varcomps.equal=varcomps.equal, umat=umat, pvec=c(p1, p2, p3), beta.pql=beta.pql, nu.pql=nu.pql, u.pql=u.star, zeta=zeta, cores=no_cores, debug=debug), class="glmm"))
 }

### 2018-08-09 DesignDocs/DesignDocParallel.tex
adding design doc

\documentclass{article}

 \usepackage{url} 
\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}
\DeclareMathOperator{\logit}{logit}
    \DeclareMathOperator{\var}{Var}
   %  \DeclareMathOperator{\det}{det}
     \DeclareMathOperator{\diag}{diag}

\usepackage{geometry}
%\geometry{hmargin=1.025in,vmargin={1.25in,2.5in},nohead,footskip=0.5in} 
%\geometry{hmargin=1.025in,vmargin={1.25in,0.75in},nohead,footskip=0.5in} 
%\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}

\renewcommand{\baselinestretch}{1.25}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} 
\DeclareMathOperator{\cov}{Cov}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\title{Design Document for Parallelization R Package glmm}

\author{Sydney Benson}

\begin{document}
\maketitle{}

\begin{abstract}
This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to parallel computing. We use parallel computing in this package to increase the speed of the calculation of the value of the log-likelihood approximation, gradient and hessian for use with MCLA. 
\end{abstract}

\section{Introduction}
This project is meant to increase the overall speed of computation for the \texttt{glmm} R package using a parallel computing structure. The parallel computing structure will be responsible for producing the value of the log-likelihood approximation, gradient and hessian for each section of a divided $u$ matrix. We'll start this design document by discussing the various packages being used for this project. Then, the parallel computing process is given; the first cluster of processing cores is created, required packages and variables are added to the cluster, the $u$ matrix is separated and processed and the returned values are recombined to create the true value of the log-likelihood approximation and gradient for the full $u$ matrix. A similar process is then repeated to calculate the true value of the hessian, using information from the first process.

\section{The Packages}
\subsection{\texttt{parallel}}
The first package we use for this parallelization is the R package \texttt{parallel}. This package allows for the detection of the number of cores in a computing device and the creation of clusters which allows us to access all of the available cores in the device.

\subsection{\texttt{doParallel}}
\texttt{doParallel} is utilized in the parallelization process to allow the use of the \texttt{foreach} function within the parallel package. This package allows the function to use the cluster created by \texttt{parallel}.

\subsection{\texttt{itertools}}
\texttt{itertools} is used in the parallelization process to divide the \u matrix into even parts, with the number of parts depending on the number of cores in the cluster.

\section{The Process}
The parallelization of the function will be completed in two steps. First, the value of the log-likelihood approximation and the gradient will be calculated in one parallelized function. Then, this information will be used to calculate the hessian in a second, separate, parallelized function. Therefore, the cluster will need to be prepared and closed, and the matrix will need to be divided two separate times.

\subsection{Preparing the Cluster}
\subsubsection{\texttt{detectCores}}
We begin by using the \texttt{detectCores} command in order to find the number of cores available in the computing device being used. The actual number of cores we will use for calculations is the number of cores in the device minus one. This enables the user to continue using the device while the function is being completed.

\subsubsection{\texttt{makeCluster}}
After detecting how many cores we have available for our cluster, we can create a cluster of our cores using \texttt{makeCluster} where the only argument is the number of cores available for use.

\subsubsection{\texttt{registerDoParallel}}
Now that we have our cluster made, we can register the cluster for use with the \texttt{foreach} function.

\subsubsection{\texttt{clusterEvalQ}}
The \texttt{clusterEvalQ} function allows us to download any necessary packages to work within our cluster. The package that will need to be downloaded within our cluster is \texttt{itertools}.

\subsubsection{\texttt{clusterExport}}
The final function we need to use to set up our cluster is the \texttt{clusterExport} function. This function allows us to bring any variable from the global environment into the cluster environment for use there. We will need to use this to bring all necessary variables for the \texttt{.C} function into the cluster environment.

\subsection{Separating the Matrix}
Next, we separate our calculations between our available cores. This separation will be done using the \texttt{isplitRows} function. Using this function, we will split our $u$ matrix row-wise into parts, the number of parts being determined by the number of cores in our cluster.

\subsection{Calculating the Value, Hessian and Gradient}
To calculate the value and gradient, we have to send each of our chunks of the $u$ matrix through the \texttt{.C} function, which allows us to run the objective function \texttt{C} function. We can do this using the \texttt{foreach} function and the \texttt{\%dopar\%} operator. We then have the value of the log-likelihood approximation and gradient values from each chunk of the $u$ matrix returned to us as part of a list.

\subsubsection{The Value}
We begin by thinking about combining the values given by each core in the simple case, not accounting for any overflow issues or uneven division of the $u$ matrix among the cores. We can find the value for the entire $u$ matrix using

\begin{align}
l_m(\theta|y) = \log\left(\dfrac{1}{r}\sum\limits_{i=1}^r e^{v_i}\right)
\end{align}

\noindent where $r$ is the number of cores being used and $v_i$ is the value obtained from each core. However, we might face the issue of an uneven distribution of the $u$ matrix among the cores because the number of rows in the $u$ matrix is not divisible by the number of cores being utilized. In this case, we need to alter equation 1 to be

\begin{align}
l_m(\theta|y) = \log\left(\dfrac{1}{m}\sum\limits_{i=1}^r j_i \ e^{v_i}\right)
\end{align}

\noindent where $m$ is the number of rows of the $u$ matrix, $j_i$ is the number of rows of the chunk of the $u$ matrix being processed by the $i$th core. Finally, we might run into an overflow problem. This occurs when a number becomes so great that it is stored as infinity by \texttt{R}. Thus, the equation changes to

\begin{align}
l_m(\theta|y) = a + \log\left(\dfrac{1}{m}\sum\limits_{i=1}^r j_i \ e^{v_i - a}\right)
\end{align}

\noindent where $a = \max{v_i}$.

\subsubsection{The Gradient}
Recombining the gradient from each core requires slightly more work than the recombination of the values. We'll understand the complexity of this equation using the simplest case possible, one with no computational instability and an even distribution of the portions of the $u$ matrix among the cores. First, the gradient we get from processing the $u$ matrix on a single core follows this equation:

\begin{align}
\nabla l_m(\theta|y)=
\dfrac{\sum_{k=1}^m   \left[  \nabla  \log f_\theta(u_k,y) \right]  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  } 
\end{align}

\noindent where $m$ is the number of rows in the $u$ matrix. This means that the gradient received from each core via parallel processing will follow this equation. Thus, we face the issue of having an incorrect denominator if we simply add the gradients from each core. To fix this, we need to first multiply the gradient from each core by the denominator of that gradient. We can obtain $\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}$ since we know that $l_{m}(\theta|y) = \log \left( \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)} \right)$ and $l_m(\theta|y) = \log\left(\dfrac{1}{r}\sum\limits_{i=1}^r e^{v_i}\right)$. Thus,

\begin{align}
\log \left( \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)} \right) &= \log\left(\dfrac{1}{r}\sum\limits_{i=1}^r e^{v_i}\right) \\
\dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}  &= \dfrac{1}{r}\sum\limits_{i=1}^r e^{v_i} \\
\sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}  &= \dfrac{m}{r} \sum\limits_{i=1}^r e^{v_i}
\end{align}

\noindent therefore, we can combine each the gradient from each core using

\begin{align}
\nabla l_m(\theta|y) &=\dfrac{\dfrac{m}{r} \sum\limits_{i=1}^r e^{v_i} \ g_{i,k}}{\dfrac{m}{r} \sum\limits_{i=1}^r e^{v_i}} \\
\nabla l_m(\theta|y) &=\dfrac{ \sum\limits_{i=1}^r e^{v_i} \ g_{i,k}}{\sum\limits_{i=1}^r e^{v_i}}
\end{align}

\noindent where $g_{i,k}$ is the $k$th element of the gradient vector produced by the $i$th core. To alter this equation for handling an uneven distribution of the $u$ matrix we need to make a change similar to the change made to the value calculation in order to handle this same problem. The new equation is

\begin{align}
\nabla l_m(\theta|y) =\dfrac{ \sum\limits_{i=1}^r j_i \ e^{v_i} \ g_{i,k}}{\sum\limits_{i=1}^r j_i \ e^{v_i}}
\end{align}

\noindent where $j_i$ is the number of rows of the chunk of the $u$ matrix being processed by the $i$th core, as mentioned previously. Then, for computational stability, to account for overflow and underflow, we can make a similar alteration to the one made above. Our final equation becomes

\begin{align}
\nabla l_m(\theta|y) =\dfrac{ \sum\limits_{i=1}^r j_i \ e^{v_i - a} \ g_{i,k}}{\sum\limits_{i=1}^r j_i \ e^{v_i - a}}
\end{align}

\noindent and $a$ is defined as above. From this we know that

\begin{align}
\dfrac{m}{r} \sum\limits_{i=1}^r e^{v_i} = \sum\limits_{i=1}^r j_i \ e^{v_i - a}
\end{align}

\subsubsection{The Hessian}
Finally, we approach the calculations for the hessian. First, we know that the equation to calculate the hessian is

\begin{multline}
\nabla^2 l_m(\theta|y)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(u_k,y)     \right]  \dfrac{ f_\theta(u_k,y)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}}\\
+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(u_k,y)  - \nabla l_m(\theta|y)   \right] \left[ \nabla \log f_\theta(u_k,y)  -\nabla l_m(\theta|y)  \right]^T  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}}_.
\end{multline}

\noindent We also know $\dfrac{f_\theta(u_k, y)}{\tilde{f}(u_k)} = e^{b_k}$ where $b_k = \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log  \tilde{f} (u_k)$. Thus, we have

\begin{multline}
\nabla^2 l_m(\theta|y)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(u_k,y)     \right]  e^{b_k} }{\sum\limits_{i=1}^m e^{b_k}}\\
+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(u_k,y)  - \nabla l_m(\theta|y)   \right] \left[ \nabla \log f_\theta(u_k,y)  -\nabla l_m(\theta|y)  \right]^T  e^{b_k}   }{\sum\limits_{i=1}^m e^{b_k}}
\end{multline}

\noindent and so, to account for computational instability, we can multiply the equation by $\dfrac{e^d}{e^d}$ where $d=\max(b_k)$

\begin{multline}
\nabla^2 l_m(\theta|y)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(u_k,y)     \right]  e^{b_k-a} }{\sum\limits_{i=1}^m e^{b_k-a}}\\
+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(u_k,y)  - \nabla l_m(\theta|y)   \right] \left[ \nabla \log f_\theta(u_k,y)  -\nabla l_m(\theta|y)  \right]^T  e^{b_k-a}   }{\sum\limits_{i=1}^m e^{b_k-a}}
\end{multline}

\noindent Notice that we can assure that the second part of equation 15 is accurate by providing $\nabla l_m (\theta|y)$, the gradient, to the function which calculates the hessian. Additionally, by collecting the $b_k$'s from each core in the value and gradient calculation, we can assure that the denominator of the equation is the same for each of the cores calculating the hessian.

Once we have the elements of the hessian from each core me can calculate the final, full hessian for the $u$ matrix. To calculate the final hessian for the complete $u$ matrix we must sum the elements of the hessian produced by each core.

\subsection{Closing the Cluster}
The final step we take is to close the cluster. We can use \texttt{stopCluster} to accomplish this. The code then returns to completing computations using a single core.

\end{document}

### 2018-08-09 glmm/src/hess.c

@@ -1,0 +1,241 @@
+#include "myheader.h"
+/*x is n by nbeta matrix
+ beta has length nbeta
+ Umat is myq by m matrix, one COLUMN of Umat used at a time. Umat must be the R mat transposed
+ z is n by myq matrix
+ pee is the vector of sampling proportions (usually 1/3, 1/3, 1/3)
+ nps is the length of pee (3 for now, maybe more if imp sampling distrib changes)
+ ntrials is a vec of ints with length equal to length(y)
+ */
+
+// value can go?
+
+void hess(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *gradient, double *hessian, double *b, int *length)
+{
+    double *Uk = Calloc(*myq, double);
+    int Uindex = 0;
+    
+    double db1 = 0.0;
+    double *double1 = &db1; /*temp to hold info that will be put into lfuval*/
+    
+    /* Calculate xbeta, needed to calculate eta for each Uk=U[k,] in R notation */
+    double *xbeta = Calloc(*n,double);
+    matvecmult(x,beta,n,nbeta,xbeta);
+    double *zu = Calloc(*n,double);
+    double *eta = Calloc(*n,double);
+    double *qzeros = Calloc(*myq,double);
+    double tempmax = 1.0;
+    double *lfutwidpieces = Calloc(*nps,double);
+    double diffs = 0.0;
+    double lfuval = 1.1;
+    double lfyuval = 1.1;
+    double lfutwid = 1.1;
+    double *q = Calloc(*m,double);
+    double a = 0.0;
+    
+    for(int k = 0; k < *m; k++){
+        /*start by getting Uk  */
+        for(int i = 0; i < *myq; i++){
+            Uk[i] = Umat[Uindex];
+            Uindex++;
+        }
+        
+        /* calculate eta for this value of Uk
+         first calculate ZUk*/
+        matvecmult(z, Uk, n, myq, zu);
+        
+        /* then add xbeta+zu to get current value of eta */
[…]

### 2018-08-16 glmm/R/glmm.R
adding parallel vignette

@@ -283,7 +283,7 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	names(nu.trust)<-varcomps.names
 
 	if(debug==TRUE){
-	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql,  trust.argpath=trust.out$argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom,m1=m1,m2=m2,m3=m3,trust.argtry=trust.out$argtry, trust.steptype=trust.out$steptype, trust.accept=trust.out$accept, trust.r=trust.out$r, trust.rho=trust.out$rho, trust.valpath=trust.out$valpath, trust.valtry=trust.out$valtry, trust.preddif=trust.out$preddif, trust.stepnorm=trust.out$stepnorm, no_cores=no_cores)
+	debug<-list(beta.pql=beta.pql, nu.pql=nu.pql,  trust.argpath=trust.out$argpath, u.star=u.star, umat=umat,weights=cache$weights,wtsnumer=cache$numer,wtsdenom=cache$denom,m1=m1,m2=m2,m3=m3,trust.argtry=trust.out$argtry, trust.steptype=trust.out$steptype, trust.accept=trust.out$accept, trust.r=trust.out$r, trust.rho=trust.out$rho, trust.valpath=trust.out$valpath, trust.valtry=trust.out$valtry, trust.preddif=trust.out$preddif, trust.stepnorm=trust.out$stepnorm)
 	}
 	
 

### 2018-08-17 glmm/tests/objfunTest.R
updating parallel vignette, fixing errors

@@ -42,7 +42,7 @@ Sigmuh<-solve(Sigmuh.inv)
 p1=p2=p3=1/3
 zeta=5
 
-no_cores <- debug$no_cores
+no_cores <- out$cores
 
 # define a few things that will be used for finite differences
 lth<-objfun(par=par, nbeta=1, nu.pql=nu.pql, umat=umat, u.star=u.pql, mod.mcml=mod.mcml, family.glmm=family.glmm,p1=p1,p2=p2,p3=p3,m1=m1, Sigmuh=Sigmuh, D.star=D.star, Sigmuh.inv= Sigmuh.inv, zeta=zeta, ntrials=ntrials, no_cores=no_cores)

### 2018-08-17 glmm/tests/salamFiniteDiffs.R

@@ -44,7 +44,7 @@ Sigmuh<-solve(Sigmuh.inv)
 del<-rep(10^-6,6)
 ntrials<-1
 
-no_cores <- debug$no_cores
+no_cores <- sal$cores
 
 ltheta<-objfun(par, nbeta, nu.pql, umat, u.star, mod.mcml, family.glmm, 
     cache, p1, p2, p3, m1, D.star, Sigmuh, Sigmuh.inv, zeta, ntrials=ntrials, no_cores=no_cores)

### 2018-08-17 glmm/tests/testpiecesBH.R

@@ -21,7 +21,7 @@ objfun<-glmm:::objfun
 getEk<-glmm:::getEk
 addVecs<-glmm:::addVecs
 
-no_cores <- mod.mcml1$debug$no_cores
+no_cores <- mod.mcml1$cores
 
 ############################################
 #this should be the same as elc

### 2018-08-17 glmm/src/glmm_init.c
deleting orginal objfun.c

@@ -38,7 +38,7 @@ static R_NativePrimitiveArgType mcsec_types[30] = {REALSXP, REALSXP,
     REALSXP, REALSXP, REALSXP, REALSXP, REALSXP, INTSXP, INTSXP, INTSXP,
     INTSXP, REALSXP, INTSXP, REALSXP, INTSXP};
 
-static R_NativePrimitiveArgType objfunc_types[30] = {REALSXP, REALSXP,
+static R_NativePrimitiveArgType valgrad_types[30] = {REALSXP, REALSXP,
     INTSXP, INTSXP, REALSXP, INTSXP, INTSXP, REALSXP, REALSXP, REALSXP,
     REALSXP, INTSXP, REALSXP, REALSXP, REALSXP, REALSXP, REALSXP,
     REALSXP, INTSXP, INTSXP, INTSXP, INTSXP, REALSXP, INTSXP, REALSXP,
@@ -88,8 +88,7 @@ static R_NativePrimitiveArgType tdist_types[6] = {REALSXP,  INTSXP,
 static R_CMethodDef cMethods[] = {
     {"elc", (DL_FUNC) &elc, 10, elc_types},
     {"mcsec", (DL_FUNC) &mcsec, 30, mcsec_types},
-    {"objfunc", (DL_FUNC) &objfunc, 30, objfunc_types},
-    {"valgrad", (DL_FUNC) &valgrad, 30, objfunc_types},
+    {"valgrad", (DL_FUNC) &valgrad, 30, valgrad_types},
     {"hess", (DL_FUNC) &hess, 31, hess_types},
     {"cp3", (DL_FUNC) &cp3, 5, cp3_types},
     {"cpp3", (DL_FUNC) &cpp3, 5, cp3_types},

### 2018-08-17 glmm/src/myheader.h

@@ -34,8 +34,6 @@ void elGH(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family
 
 void tdist(double *DstarInv,  int *myq, double *uvec, int *zeta, double *tconst, double *logft);
 
-void objfunc(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *hessian);
-
 void valgrad(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *value, double *gradient, double *b);
 
 void hess(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *gradient, double *hessian, double *b, int *length);

### 2018-09-25 glmm/R/glmm.R
allowing cluster to be passed through glmm, using passed cluster for objfun

@@ -1,5 +1,5 @@
 glmm <-
-function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5, cores=NULL)
+function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL=TRUE, debug=FALSE,p1=1/3,p2=1/3,p3=1/3,rmax=1000,iterlim=1000,par.init=NULL,zeta=5, cluster=NULL)
 	{
 	if(missing(varcomps.names)) stop("Names for the variance components must be supplied through varcomps.names")
 	if(is.vector(varcomps.names)!=1) stop("varcomps.names must be a vector")
@@ -14,6 +14,12 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	vars$p1 <- p1
 	vars$p2 <- p2
 	vars$p3 <- p3
+	
+	if(is.null(cluster)){
+	  vars$cl <- makeCluster(1)
+	} else{
+	  vars$cl <- cluster
+	}
 
 	#this much will figure out how to interpret the formula
 	#first the fixed effects part
@@ -275,10 +281,8 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 
 	vars$umat<-rbind(genData,genData2,genData3)
 	
-	vars$no_cores <- min(cores, max(1, detectCores()-1))
-	
 	vars$nbeta <- length(beta.pql)
-
+	
 	#use trust to max the objfun (monte carlo likelihood)
 	trust.out<-trust(objfun,parinit=par.init,rinit=10, minimize=FALSE, rmax=rmax, iterlim=iterlim, blather=debug,  cache=cache, vars=vars)
 
@@ -302,5 +306,5 @@ function(fixed,random, varcomps.names,data, family.glmm, m,varcomps.equal, doPQL
 	trust.converged=trust.out$converged,  mod.mcml=vars$mod.mcml,
 	 fixedcall=fixed,randcall=randcall, x=x,y=y, z=random,
 	family.glmm=vars$family.glmm, call=call, varcomps.names=varcomps.names, 
[…]

### 2018-09-25 glmm/R/objfun.R

@@ -3,6 +3,8 @@
 objfun <-
 function(par, cache, vars){
   
+  vars$no_cores <- nrow(summary(vars$cl))
+  
   vars$par <- par
   
 	vars$beta<-beta<-vars$par[1:vars$nbeta]
@@ -66,13 +68,11 @@ function(par, cache, vars){
 	minib <- NULL
 
 	#parallelizing the calculations for the value of the log-likelihood approximation and gradient
-	cl <- makeCluster(vars$no_cores)              #making cluster environment
-	registerDoParallel(cl)                   #making cluster usable with foreach
-	clusterEvalQ(cl, library(itertools))     #installing itertools library on each core
-	clusterExport(cl, c("vars"), envir = environment())     #installing variables on each core
+	registerDoParallel(vars$cl)                   #making cluster usable with foreach
+	clusterEvalQ(vars$cl, library(itertools))     #installing itertools library on each core
+	clusterExport(vars$cl, c("vars"), envir = environment())     #installing variables on each core
 	out <- foreach(miniu=isplitRows(vars$umat, chunks = vars$no_cores)) %dopar% {.C(C_valgrad, as.double(vars$mod.mcml$y),as.double(t(miniu)), as.integer(vars$myq), as.integer(nrow(miniu)), as.double(vars$mod.mcml$x), as.integer(vars$n), as.integer(vars$nbeta), as.double(vars$beta), as.double(vars$Z), as.double(vars$Dinvfornu), as.double(vars$logdetDinvfornu),as.integer(vars$family_glmm), as.double(vars$D.star.inv), as.double(vars$logdet.D.star.inv), as.double(vars$u.star), as.double(vars$Sigmuh.inv), as.double(vars$logdet.Sigmuh.inv), pea=as.double(vars$pea), nps=as.integer(length(vars$pea)), T=as.integer(vars$T), nrandom=as.integer(vars$nrandom), meow=as.integer(vars$meow),nu=as.double(vars$nu), zeta=as.integer(vars$zeta),tconst=as.double(vars$tconst), v=double(vars$m), ntrials=as.integer(vars$ntrials), value=double(1),gradient=double(length(vars$par)), b=double(nrow(miniu)))}
 	#foreach runs loop in parallel, dopar operator sends each chunk of umat to seperate core and runs the .C function
[…]

### 2018-09-25 glmm/tests/coreTest.R

@@ -1,8 +1,9 @@
 library(glmm)
 data(BoothHobert)
+clust <- makeCluster(2)
 set.seed(1234)
 out<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-          family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cores=2)
+          family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cluster=clust)
 vars <- new.env(parent = emptyenv())
 vars$mod.mcml<-out$mod.mcml
 debug<-out$debug
@@ -42,15 +43,19 @@ vars$Sigmuh<-solve(vars$Sigmuh.inv)
 vars$p1=vars$p2=vars$p3=1/3
 vars$zeta=5
 
-vars$no_cores <- 1
+vars$cl <- out$cluster
 
 vars$nbeta <- 1
 vars$u.star <- u.pql
 
-core1<-objfun(par=par, vars=vars)
-
-vars$no_cores <- out$cores
-
 core2<-objfun(par=par, vars=vars)
 
+stopCluster(clust)
+
+vars$cl <- makeCluster(1)
+
+core1<-objfun(par=par, vars=vars)
+
 all.equal(core1, core2)
+
+stopCluster(vars$cl)

### 2018-09-25 glmm/tests/objfunTest.R

@@ -1,9 +1,10 @@
 #check objfun using finite differences
 library(glmm)
 data(BoothHobert)
+clust <- makeCluster(2)
 set.seed(1234)
 out<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cores=2)
+family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cluster=clust)
 vars <- new.env(parent = emptyenv())
 vars$mod.mcml<-out$mod.mcml
 debug<-out$debug
@@ -43,7 +44,7 @@ vars$Sigmuh<-solve(vars$Sigmuh.inv)
 vars$p1=vars$p2=vars$p3=1/3
 vars$zeta=5
 
-vars$no_cores <- out$cores
+vars$cl <- out$cluster
 
 vars$nbeta <- 1
 vars$u.star <- u.pql
@@ -268,3 +269,5 @@ function(nu,U,z.list,mu){
 
 that<-objfunNOC(par=par, nbeta=1, nu.pql=vars$nu.pql, umat=vars$umat, u.star=u.pql, mod.mcml=vars$mod.mcml, family.glmm=vars$family.glmm,p1=vars$p1,p2=vars$p2,p3=vars$p3, Sigmuh=vars$Sigmuh,D.star=vars$D.star, zeta=vars$zeta)
 all.equal(that,lth)
+
+stopCluster(clust)

### 2018-09-25 glmm/tests/salamFiniteDiffs.R

@@ -1,7 +1,8 @@
 library(glmm)
+clust <- makeCluster(2)
 set.seed(1234)
 data(salamander)
-sal<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"), data=salamander,family.glmm=bernoulli.glmm,m=100,debug=TRUE,doPQL=FALSE, cores=2)
+sal<-glmm(Mate~Cross,random=list(~0+Female,~0+Male),varcomps.names=c("F","M"), data=salamander,family.glmm=bernoulli.glmm,m=100,debug=TRUE,doPQL=FALSE, cluster=clust)
 
 vars <- new.env(parent = emptyenv())
 objfun<-glmm:::objfun
@@ -45,7 +46,7 @@ vars$Sigmuh<-solve(vars$Sigmuh.inv)
 del<-rep(10^-6,6)
 vars$ntrials<-1
 
-vars$no_cores <- sal$cores
+vars$cl <- sal$cluster
 
 ltheta<-objfun(par, cache, vars=vars)
 
@@ -62,4 +63,4 @@ cbind(lthetadel$gradient-ltheta$gradient,as.vector(ltheta$hessian%*%del))
 #how big are the diffs? very small
 lthetadel$gradient-ltheta$gradient-as.vector(ltheta$hessian%*%del)
 
-
+stopCluster(clust)

### 2018-09-25 glmm/tests/testpiecesBH.R

@@ -1,7 +1,8 @@
 library(glmm)
 data(BoothHobert)
+clust <- makeCluster(2)
 set.seed(1234)
-mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cores=2)
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cluster=clust)
 
 mod.mcml<-mod.mcml1$mod.mcml
 z<-mod.mcml$z[[1]]
@@ -20,8 +21,6 @@ family.glmm<-bernoulli.glmm
 objfun<-glmm:::objfun
 getEk<-glmm:::getEk
 addVecs<-glmm:::addVecs
-
-no_cores <- mod.mcml1$cores
 
 ############################################
 #this should be the same as elc
@@ -205,7 +204,7 @@ vars$Sigmuh <- Sigmuh
 vars$Sigmuh.inv <- Sigmuh.inv
 vars$zeta <- 5
 vars$ntrials <- 1
-vars$no_cores <- no_cores
+vars$cl <- mod.mcml1$cluster
 that<-objfun(c(beta,nu), cache=cache,vars=vars)
 
 #get t stuff ready
@@ -245,6 +244,6 @@ all.equal(value,that$value)
 #Given generated random effects, the value of the objective function is correct.
 #This plus the test of finite diffs for objfun should be enough.
 
-
+stopCluster(clust)
 
 

### 2018-11-28 glmm/tests/objfunTest.R
cleaning up test output more

@@ -3,14 +3,12 @@ library(glmm)
 data(BoothHobert)
 set.seed(1234)
 out<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cores=2)
+family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE, cores=1)
 vars <- new.env(parent = emptyenv())
 vars$mod.mcml<-out$mod.mcml
 debug<-out$debug
 vars$nu.pql<-debug$nu.pql
-vars$nu.pql
 beta.pql<-debug$beta.pql
-beta.pql
 vars$family.glmm<-out$family.glmm
 vars$umat<-debug$umat
 u.pql<-debug$u.star
@@ -56,14 +54,14 @@ all.equal(as.vector(lth$gradient%*%del),lthdel$value-lth$value)
 all.equal(as.vector(lth$hessian%*%del),lthdel$gradient-lth$gradient)
 
 #see exactly how big the difference is
-as.vector(lth$gradient%*%del)-(lthdel$value-lth$value)
-as.vector(lth$hessian%*%del)-(lthdel$gradient-lth$gradient)
+#as.vector(lth$gradient%*%del)-(lthdel$value-lth$value)
+#as.vector(lth$hessian%*%del)-(lthdel$gradient-lth$gradient)
 
 #we know these differences are small when we compare it to the actual values
-lthdel$value-lth$value
-as.vector(lth$gradient%*%del)
-as.vector(lth$hessian%*%del)
-lthdel$gradient-lth$gradient
+# lthdel$value-lth$value
+# as.vector(lth$gradient%*%del)
+# as.vector(lth$hessian%*%del)
+# lthdel$gradient-lth$gradient
 
 ##########################################
 ##### to make sure that the objfun function is correct, compare it against the version without any C code. here is objfun without c:

### 2018-11-28 glmm/tests/testpiecesBH.R

@@ -1,7 +1,8 @@
 library(glmm)
 data(BoothHobert)
 set.seed(1234)
-mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cores=2)
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, 
+                doPQL=TRUE, debug=TRUE, cores=1)
 
 mod.mcml<-mod.mcml1$mod.mcml
 z<-mod.mcml$z[[1]]
@@ -120,14 +121,14 @@ that<-distRand(2,you,mod.mcml$z,u.pql)
 all.equal(this,that)
 
 #use finite diffs to make sure distRandCheck (and distRand) have correct derivs
-del<-10^(-8)
+del<-10^(-9)
 thisdel<-distRandCheck(2+del,you,u.pql)
 firstthing<-thisdel$value-this$value
 secondthing<-as.vector(this$gradient%*%del)
 all.equal(firstthing,secondthing)
-firstthing
-secondthing
-firstthing-secondthing
+#firstthing
+#secondthing
+#firstthing-secondthing
 
 #compare the gradient and hessian of the C functions by using these functions
 #(the value is checked in distRandGeneral)

### 2018-12-13 DesignDocs/DesignDocWeights.tex
adding weights design doc

\documentclass{article}

 \usepackage{url} 
\usepackage{amsthm,amsmath,amssymb,indentfirst,float}
\usepackage{verbatim}
\usepackage[sort,longnamesfirst]{natbib}
\newcommand{\pcite}[1]{\citeauthor{#1}'s \citeyearpar{#1}}
\newcommand{\ncite}[1]{\citeauthor{#1}, \citeyear{#1}}
\DeclareMathOperator{\logit}{logit}
    \DeclareMathOperator{\var}{Var}
   %  \DeclareMathOperator{\det}{det}
     \DeclareMathOperator{\diag}{diag}

\usepackage{geometry}
%\geometry{hmargin=1.025in,vmargin={1.25in,2.5in},nohead,footskip=0.5in} 
%\geometry{hmargin=1.025in,vmargin={1.25in,0.75in},nohead,footskip=0.5in} 
%\geometry{hmargin=2.5cm,vmargin={2.5cm,2.5cm},nohead,footskip=0.5in}

\renewcommand{\baselinestretch}{1.25}

\usepackage{amsbsy,amsmath,amsthm,amssymb,graphicx}

\setlength{\baselineskip}{0.3in} \setlength{\parskip}{.05in}

\newcommand{\cvgindist}{\overset{\text{d}}{\longrightarrow}}
\DeclareMathOperator{\PR}{Pr} 
\DeclareMathOperator{\cov}{Cov}

\newcommand{\sX}{{\mathsf X}}
\newcommand{\tQ}{\tilde Q}
\newcommand{\cU}{{\cal U}}
\newcommand{\cX}{{\cal X}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\tlambda}{\tilde{\lambda}}
\newcommand{\txi}{\tilde{\xi}}

\title{Design Document for Observation Weights: R Package glmm}

\author{Sydney Benson}

\begin{document}
\maketitle{}

\begin{abstract}
This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to observation weighting for the response vector, matrix of fixed effects and matrix of random effects. We use observation weighting to better reflect the real-world occurrence of more or less informative observations.
\end{abstract}

\section{Introduction}
This project is meant to enable the user of the \texttt{glmm} function in the \texttt{glmm} R package to include an optional observation weighting scheme. A common assumption of linear models is that each observation in a data set is equally informative and trustworthy; however, in real-world data sets, this is rarely the case. Thus, the optional observation weighting scheme will allow the user to place a heavier weight on the more informative and/or trustworthy observations in their data set so that those data points that are less informative affect the model to a lesser degree.

\section{The Process}
The weighting scheme for the observations can be implemented in the \texttt{glmm.R} function alone. First, the function will need to establish whether the user has supplied a proper weighting scheme. Next, the weighting scheme will need to be applied to the response vector, fixed effects matrix and random effects matrix. Finally, the model can be run using the weighted vector and matrices. After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied to the response vector, fixed effects matrix and random effects matrix.

\subsection{The Weighting Vector}
This vector, called $W$, must be a vector with the same length as the response vector and must contain all positive values.

\subsection{Weighted Response Vector}
Let $W$ be the vector of weights supplied by the user.  Let $Y$ be the response vector. Then,

\begin{align}
Y_W = \sqrt{W} \cdot Y
\end{align}

\subsection{Weighted Fixed Effects Matrix}
Let $W$ be the vector of weights supplied by the user. Let $X$ be the matrix of fixed effects. Then,

\begin{align}
X_W = \sqrt{W} \cdot X
\end{align}

\subsection{Weighted Random Effects Matrix}
Let $W$ be the vector of weights supplied by the user. Let $Z$ be the matrix of random effects. Then,

\begin{align}
Z_W = \sqrt{W} \cdot Z
\end{align}

\end{document}

### 2019-01-22 glmm/R/glmm.R
updating documentation and vignette with weights information

@@ -372,5 +372,5 @@ glmm <-
                           trust.converged=trust.out$converged,  mod.mcml=vars$mod.mcml,
                           fixedcall=fixed,randcall=randcall, x=savedx,y=savedy, z=random, weights=savedw,
                           family.glmm=vars$family.glmm, call=call, varcomps.names=varcomps.names, 
-                          varcomps.equal=varcomps.equal, umat=umat, pvec=c(vars$p1, vars$p2, vars$p3), beta.pql=beta.pql, nu.pql=vars$nu.pql, u.pql=vars$u.star, zeta=vars$zeta, cluster=vars$cl, cores=vars$no_cores, debug=debug, zz=z), class="glmm"))
+                          varcomps.equal=varcomps.equal, umat=umat, pvec=c(vars$p1, vars$p2, vars$p3), beta.pql=beta.pql, nu.pql=vars$nu.pql, u.pql=vars$u.star, zeta=vars$zeta, cluster=vars$cl, cores=vars$no_cores, debug=debug), class="glmm"))
   }

### 2019-02-01 glmm/tests/weightsTest.R
minor changes to documentation and vignette

@@ -17,8 +17,12 @@ all.equal(zeroweight$likelihood.hessian,removed$likelihood.hessian)
 
 set.seed(1234)
 doubleweight<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-                 family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE,weights=c(rep(1,149),rep(2,1)))
+                 family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE,weights=c(rep(1,149),2))
 
+#nums <- c(rep(1,150),2)
+#for(i in 1:nrow(BoothHobert)){
+  #BoothHobert3[i] <- BoothHobert[i,]*nums[i]
+#}
 BoothHobert3 <- rbind(BoothHobert, BoothHobert[150,])
 set.seed(1234)
 added<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert3,

### 2019-03-10 DesignDocs/DesignDocWeights.tex
reversing weights work in glmm.R, updating weights design doc

@@
 
-\title{Design Document for Observation Weights: R Package glmm}
+\title{Design Document for Relevance-Weighting: R Package glmm}
 
@@
 \begin{abstract}
-This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to observation weighting for the response vector, matrix of fixed effects and matrix of random effects.
-We use observation weighting to better reflect the real-world occurrence of more or less informative observations.
+This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to the relevance-weighted likelihood method mentioned in Wang (2001).
+We use relevance-weighting to better reflect the real-world occurrence of more or less informative observations.
 \end{abstract}
@@
 \section{Introduction}
-This project is meant to enable the user of the \texttt{glmm} function in the \texttt{glmm} R package to include an optional observation weighting scheme.
+This project is meant to enable the user of the \texttt{glmm} function in the \texttt{glmm} R package to include an optional relevance-weighting scheme.
 A common assumption of linear models is that each observation in a data set is equally informative and trustworthy; however, in real-world data sets, this is rarely the case.
-Thus, the optional observation weighting scheme will allow the user to place a heavier weight on the more informative and/or trustworthy observations in their data set so that those data points that are less informative affect the model to a lesser degree.
+Thus, the optional relevance-weighting scheme will allow the user to place a heavier weight on the more informative and/or trustworthy observations in their data set so that those data points that are less informative affect the model to a lesser degree.
 
 \section{The Process}
-The weighting scheme for the observations can be implemented in the \texttt{glmm.R} function alone.
+The weighting scheme for the observations can be implemented in the \texttt{objfun.R} function.
 First, the function will need to establish whether the user has supplied a proper weighting scheme.
-Next, the weighting scheme will need to be applied to the response vector, fixed effects matrix and random effects matrix.
-Finally, the model can be run using the weighted vector and matrices.
-After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied to the response vector, fixed effects matrix and random effects matrix.
+Next, the weighting scheme will need to be applied to the respective elements of the log-likelihood.
+After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied to the log-likelihood values.
 
 \subsection{The Weighting Vector}
-This vector, called $W$, must be a vector with the same length as the response vector and must contain all positive values.
+This vector, called $\Lambda$, must be a vector with the same length as the response vector and must contain all positive values.
 
-\subsection{Weighted Response Vector}
-Let $W$ be the vector of weights supplied by the user.  Let $Y$ be the response vector. Then,
+\subsection{Weighted Log-Likelihood}
+From Hu and Zidek's (1997) method, mentioned in Wang (2001), we know the relevance-weighted likelihood (REWL) is
 
 \begin{align}
-Y_W = \sqrt{W} \cdot Y
+\text{REWL}(\theta) = \prod_{i=1}^{n} {f(x_i;\theta)}^{\lambda_i}
 \end{align}
 
-\subsection{Weighted Fixed Effects Matrix}
-Let $W$ be the vector of weights supplied by the user. Let $X$ be the matrix of fixed effects. Then,
+\noindent where $f(x_i;\theta)$ is the probability distribution for $x_i$ and $\lambda_i$ is the weight for $x_i$. Then, the relevance-weighted log-likelihood (RWLL) is
 
 \begin{align}
-X_W = \sqrt{W} \cdot X
+\text{RWLL}(\theta) &= \log\left(\prod_{i=1}^{n} {f(x_i;\theta)}^{\lambda_i}\right) \\
+&= \sum_{i=1}^{n} \log\left({f(x_i;\theta)}^{\lambda_i}\right) \\
+&= \sum_{i=1}^{n} \lambda_i \log\left(f(x_i;\theta)\right)
 \end{align}
 
-\subsection{Weighted Random Effects Matrix}
-Let $W$ be the vector of weights supplied by the user. Let $Z$ be the matrix of random effects. Then,
+Then, knowing that $l_m(\theta | y) = \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i}\right)$ where $m$ is the number of rows of the $u$ matrix, $j_i$ is the number of rows of the chunk of the $u$ matrix being processed by the $i$th core and $v_i$ is the value obtained by each core, as described in the design document for parallel computing, we have
 
 \begin{align}
-Z_W = \sqrt{W} \cdot Z
+\text{RWLL}(\theta) &= \sum_{i=1}^{n} \lambda_i \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i}\right) \\
+&= a + \sum_{i=1}^{n} \lambda_i \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i - a}\right)
 \end{align}
 
+\noindent where $a = \max(v_i)$.
+
 \end{document}

### 2019-03-10 glmm/R/glmm.R

@@ -177,33 +177,33 @@ glmm <-
     if(any(weights < 0))stop("Negative weights not allowed")
     if(any(is.na(weights)))stop("Missing weights not allowed")
     
-    savedx <- x
-    savedy <- y
-    savedw <- weights
+    #savedx <- x
+    #savedy <- y
+    #savedw <- weights
     
-    if(any(weights == 0)){
-      drop <- which(weights == 0)
-      if(length(drop) == length(weights))stop("No informative observations. At least one weight must be non-zero")
-      weights <- weights[-drop]
-      x <- x[-drop, , drop=FALSE]
-      y <- if(NCOL(savedy) == 1) y[-drop] else y[-drop, ]
-      for(l in 1:length(z)){
-        z[[l]] <- z[[l]][-drop,] 
-      }
-    }
+    #if(any(weights == 0)){
+      #drop <- which(weights == 0)
+      #if(length(drop) == length(weights))stop("No informative observations. At least one weight must be non-zero")
+      #weights <- weights[-drop]
+      #x <- x[-drop, , drop=FALSE]
+      #y <- if(NCOL(savedy) == 1) y[-drop] else y[-drop, ]
+      #for(l in 1:length(z)){
+        #z[[l]] <- z[[l]][-drop,] 
+      #}
+    #}
     
-    w <- sqrt(weights)
+    #w <- sqrt(weights)
     
-    x <- x*w
-    y <- y*w
+    #x <- x*w
+    #y <- y*w
     
-    for(l in 1:length(z)){
-      for(i in 1:nrow(z[[l]])){
-        for(j in 1:ncol(z[[l]])){
-          z[[l]][i,j] <- z[[l]][i,j]*w[i]
-        }
-      }
-    }
+    #for(l in 1:length(z)){
+      #for(i in 1:nrow(z[[l]])){
+        #for(j in 1:ncol(z[[l]])){
+          #z[[l]][i,j] <- z[[l]][i,j]*w[i]
+        #}
+      #}
+    #}
     
     vars$mod.mcml <- list(x=x, y=y, z=z, ntrials=vars$ntrials)
     
@@ -361,7 +361,7 @@ glmm <-
     names(nu.trust)<-varcomps.names
     
     if(debug==TRUE){
[…]

### 2019-03-10 glmm/R/mcse.R

@@ -74,7 +74,7 @@ mcvcov <- function(object){
 
 	
 
-	stuff<-.C(C_mcsec, as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), ntrials=as.integer(mod$mod.mcml$ntrials))
+	stuff<-.C(C_mcsec, as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), ntrials=as.integer(mod$mod.mcml$ntrials), as.double(0.0))
 
 	vhatnum <- (1/m)*stuff[[4]]
 	vhatdenom <- ( stuff[[1]]   )^2

### 2019-03-10 glmm/src/glmm_init.c

@@ -32,11 +32,11 @@ static R_NativePrimitiveArgType elc_types[10] = {REALSXP, REALSXP,
     INTSXP, INTSXP, REALSXP, INTSXP, INTSXP, REALSXP, REALSXP,
     REALSXP};
 
-static R_NativePrimitiveArgType mcsec_types[30] = {REALSXP, REALSXP,
+static R_NativePrimitiveArgType mcsec_types[33] = {REALSXP, REALSXP,
     REALSXP, REALSXP, REALSXP, REALSXP, INTSXP, INTSXP, REALSXP, INTSXP,
     INTSXP, REALSXP, REALSXP, REALSXP, REALSXP, INTSXP, REALSXP,
     REALSXP, REALSXP, REALSXP, REALSXP, REALSXP, INTSXP, INTSXP, INTSXP,
-    INTSXP, REALSXP, INTSXP, REALSXP, INTSXP};
+    INTSXP, REALSXP, INTSXP, REALSXP, INTSXP, REALSXP, REALSXP, REALSXP};
 
 static R_NativePrimitiveArgType valgrad_types[30] = {REALSXP, REALSXP,
     INTSXP, INTSXP, REALSXP, INTSXP, INTSXP, REALSXP, REALSXP, REALSXP,
@@ -87,7 +87,7 @@ static R_NativePrimitiveArgType tdist_types[6] = {REALSXP,  INTSXP,
 
 static R_CMethodDef cMethods[] = {
     {"elc", (DL_FUNC) &elc, 10, elc_types},
-    {"mcsec", (DL_FUNC) &mcsec, 30, mcsec_types},
+    {"mcsec", (DL_FUNC) &mcsec, 33, mcsec_types},
     {"valgrad", (DL_FUNC) &valgrad, 30, valgrad_types},
     {"hess", (DL_FUNC) &hess, 32, hess_types},
     {"cp3", (DL_FUNC) &cp3, 5, cp3_types},

### 2019-03-10 glmm/src/mcse.c

@@ -1,4 +1,7 @@
 #include "myheader.h"
+/*problem with gamma and squaretop causing issue with numsum: overflow/underflow*/
+
+
 /*x is n by nbeta matrix
 beta has length nbeta
 Umat is myq by m matrix, one COLUMN of Umat used at a time. Umat must be the R mat transposed 
@@ -7,7 +10,7 @@ pee is the vector of sampling proportions (usually 1/3, 1/3, 1/3)
 nps is the length of pee (3 for now, maybe more if imp sampling distrib changes)
 ntrials is a vec of ints with length equal to length(y)
 */
-void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst,  int *ntrials)
+void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst,  int *ntrials, double *lfuval, double *lfyuval, double *tryfix)
 {
 	double *Uk = Calloc(*myq, double);
 	int Uindex = 0;
@@ -24,8 +27,8 @@ void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, doub
 	double tempmax = 1.0;
 	double *lfutwidpieces = Calloc(*nps,double);
 	double diffs = 0.0;
-	double lfuval = 1.1;
-	double lfyuval = 1.1;
+	/*double lfuval = 1.1;
+	double lfyuval = 1.1;*/
 	double lfutwid = 1.1;
[…]

### 2019-03-10 glmm/src/myheader.h

@@ -38,7 +38,7 @@ void valgrad(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *
 
 void hess(double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, double *v, int *ntrials, double *gradient, double *hessian, double *b, int *length, double *q);
 
-void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, int *ntrials);
+void mcsec(double *gamma, double *thing, double *squaretop, double *numsum, double *y, double *Umat, int *myq, int *m, double *x, int *n, int *nbeta, double *beta, double *z, double *Dinvfornu, double *logdetDinvfornu, int *family_glmm, double *Dstarinv, double *logdetDstarinv, double *ustar, double *Sigmuhinv, double *logdetSigmuhinv, double *pee, int *nps, int *T, int *nrandom, int *meow, double *nu, int *zeta, double *tconst, int *ntrials, double *lfuval, double *lfyuval, double *tryfix);
 
 void elc(double *Y, double *X, int *nrowX, int *ncolX, double *eta, int *family,  int *ntrials, double *elval, double *elgradient, double *elhessian);
 

### 2019-03-10 glmm/tests/weightsTest.R

@@ -19,6 +19,8 @@ set.seed(1234)
 doubleweight<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
                  family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE,weights=c(rep(1,149),2))
 
+#umatadded <- doubleweight$umat
+
 #nums <- c(rep(1,150),2)
 #for(i in 1:nrow(BoothHobert)){
   #BoothHobert3[i] <- BoothHobert[i,]*nums[i]

### 2019-04-08 DesignDocs/DesignDocWeights.tex
updated weights design doc

@@
 \begin{abstract}
-This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to the relevance-weighted likelihood method mentioned in Wang (2001).
+This design document will give an overview of the changes made to the R package \texttt{glmm} with respect to a relevance-weighted likelihood method.
 We use relevance-weighting to better reflect the real-world occurrence of more or less informative observations.
@@
 \section{The Process}
-The weighting scheme for the observations can be implemented in the \texttt{objfun.R} function.
 First, the function will need to establish whether the user has supplied a proper weighting scheme.
-Next, the weighting scheme will need to be applied to the respective elements of the log-likelihood.
-After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied to the log-likelihood values.
+Next, the weighting scheme will need to be applied in the \texttt{el.C} function.
+After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied.
 
 \subsection{The Weighting Vector}
-This vector, called $\Lambda$, must be a vector with the same length as the response vector and must contain all positive values.
+This vector, called $\Lambda$, must be a vector with the same length as the response vector and must contain all positive values or zeros.
 
 \subsection{Weighted Log-Likelihood}
-From Hu and Zidek's (1997) method, mentioned in Wang (2001), we know the relevance-weighted likelihood (REWL) is
+We begin by defining the canonical link function as
 
 \begin{align}
-\text{REWL}(\theta) = \prod_{i=1}^{n} {f(x_i;\theta)}^{\lambda_i}
+g(\mu) = \eta = X\beta + Zu
 \end{align}
 
-\noindent where $f(x_i;\theta)$ is the probability distribution for $x_i$ and $\lambda_i$ is the weight for $x_i$. Then, the relevance-weighted log-likelihood (RWLL) is
+\noindent Additionally, the likelihood is defined as
 
 \begin{align}
-\text{RWLL}(\theta) &= \log\left(\prod_{i=1}^{n} {f(x_i;\theta)}^{\lambda_i}\right) \\
-&= \sum_{i=1}^{n} \log\left({f(x_i;\theta)}^{\lambda_i}\right) \\
-&= \sum_{i=1}^{n} \lambda_i \log\left(f(x_i;\theta)\right)
+l_m(\theta|y) = \log \left( \dfrac{1}{m} \sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
+\right)
 \end{align}
 
-Then, knowing that $l_m(\theta | y) = \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i}\right)$ where $m$ is the number of rows of the $u$ matrix, $j_i$ is the number of rows of the chunk of the $u$ matrix being processed by the $i$th core and $v_i$ is the value obtained by each core, as described in the design document for parallel computing, we have
+\noindent and $f_\theta(u_k, y) = f_\theta(y|u_k) \tilde{f_\theta} (u_k)$. We then define $f_\theta(y|u_k)$ as $f_\theta(y|u_k) = \exp \left( \sum_i y_i\eta_i - c(\eta_i) \right)$. Thus,
 
 \begin{align}
-\text{RWLL}(\theta) &= \sum_{i=1}^{n} \lambda_i \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i}\right) \\
-&= a + \sum_{i=1}^{n} \lambda_i \log\left(\dfrac{1}{m}\sum_{i=1}^{r}j_ie^{v_i - a}\right)
+f_\theta (u_k, y) = \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k)
 \end{align}
 
-\noindent where $a = \max(v_i)$.
+and
+
+\begin{align}
+\Lambda f_\theta (u_k, y) &= \Lambda \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k) \\
+&= \lambda_i \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k) 
+\end{align}
+
+\subsection{Integrating the Weighted Log-Likelihood}
+As shown above, the weighting scheme must be accounted for in $f_\theta (u_k, y)$.
+Thus, the weighting scheme must be applied within the \texttt{el.C} function eventually.
+However, before we get there, we must follow these steps:
+
+\begin{itemize}
+	\item[1.] Write the test for the weighting scheme using the Booth Hobert data set.
+	\item[2.] Code a general version of the weighting scheme.
+	\item[3.] Re-code the general version of the implementation of the weighting scheme in C.
+\end{itemize}
 

### 2019-04-11 DesignDocs/DesignDocWeights.tex
finished weights design doc

@@
 \subsection{The Weighting Vector}
-This vector, called $\Lambda$, must be a vector with the same length as the response vector and must contain all positive values or zeros.
+The weights must be in the form of a vector and the vector must be the same length as the response vector, $y$. All weights must be numeric and non-negative. There can be no missing weights.
 
-\subsection{Weighted Log-Likelihood}
-We begin by defining the canonical link function as
+\subsection{The Relevance-Weighted Log Density}
+Following Hu and Zidek (1997), the relevance-weighted likelihood is defined as
 
 \begin{align}
-g(\mu) = \eta = X\beta + Zu
+\text{REWL}(\theta) = \prod_{i=1}^n f(y_i | \theta)^{\lambda_i}
 \end{align}
 
-\noindent Additionally, the likelihood is defined as
+\noindent where $n$ is the length of the response vector, $y$.
+Since we define $\log f_\theta(y|u_k)$ as $\sum_i y_i\eta_i - c(\eta_i)$, where $\eta = X\beta + Zu$, $\log f_\theta(y_i | u_k) = y_i\eta_i - c(\eta_i)$ and $f_\theta(y_i|u_k) = \exp \left(  y_i\eta_i - c(\eta_i) \right)$.
+So, the relevance-weighted log density becomes
 
 \begin{align}
-l_m(\theta|y) = \log \left( \dfrac{1}{m} \sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\right)
+\text{RWLD}(\theta) &= \prod_{i=1}^n \left[ \exp \left(  y_i\eta_i - c(\eta_i) \right) \right] ^{\lambda_i}  \\
+&= \prod_{i=1}^n \exp \left[ \lambda_i \left(  y_i\eta_i - c(\eta_i) \right) \right]
 \end{align}
 
-\noindent and $f_\theta(u_k, y) = f_\theta(y|u_k) \tilde{f_\theta} (u_k)$. We then define $f_\theta(y|u_k)$ as $f_\theta(y|u_k) = \exp \left( \sum_i y_i\eta_i - c(\eta_i) \right)$. Thus,
+\noindent Then,
 
 \begin{align}
-f_\theta (u_k, y) = \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k)
+\log \text{RWLD}(\theta) &= \log \prod_{i=1}^n \exp \left[ \lambda_i \left(  y_i\eta_i - c(\eta_i) \right) \right] \\
+&= \sum_{i=1}^n \log \exp \left[ \lambda_i \left(  y_i\eta_i - c(\eta_i) \right) \right] \\
+&= \sum_{i=1}^n \lambda_i \left(  y_i\eta_i - c(\eta_i) \right)
 \end{align}
 
-and
+\subsection{The First Derivative}
+
+Remember that the derivative of the log density of the data with respect to one component, $\eta_j$, is
 
 \begin{align}
-\Lambda f_\theta (u_k, y) &= \Lambda \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k) \\
-&= \lambda_i \exp \left( \sum_i y_i \eta_i - c(\eta_i) \right) \tilde{f_\theta} (u_k) 
+\dfrac{\partial}{\partial \eta_j} \log f_\theta (y|u) = y_j - c'(\eta_j).
 \end{align}
 
-\subsection{Integrating the Weighted Log-Likelihood}
-As shown above, the weighting scheme must be accounted for in $f_\theta (u_k, y)$.
-Thus, the weighting scheme must be applied within the \texttt{el.C} function eventually.
-However, before we get there, we must follow these steps:
+\noindent Consequently,
 
-\begin{itemize}
-	\item[1.] Write the test for the weighting scheme using the Booth Hobert data set.
-	\item[2.] Code a general version of the weighting scheme.
-	\item[3.] Re-code the general version of the implementation of the weighting scheme in C.
-\end{itemize}
+\begin{align}
+\dfrac{\partial}{\partial \eta_j} \log \text{RWLD}(\theta) = \lambda_j \left[ y_j - c'(\eta_j) \right].
+\end{align}
+
+\noindent and the derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_j$, is
+
+\begin{align}
+\dfrac{\partial \eta_j}{\partial \beta_l} = X_{jl}
+\end{align}
+
+\noindent Then, we can use the chain rule to find that
+
+\begin{align}
+\dfrac{\partial}{\partial \beta_l} \log \text{RWLD}(\theta) &= \lambda \ X \left[ y- c'(\eta) \right]
+\end{align}
+
+\subsection{The Second Derivative}
+
+Similar to the first derivative, we find that the second derivative of the relevance-weighted log density of the data is
+
+\begin{align}
+\dfrac{\partial^2}{\partial \beta_l^2} \log \text{RWLD}(\theta) &= \lambda \ X' \left[- c''(\eta) \right] X
+\end{align}
 

### 2019-04-11 glmm/tests/weightsTest.R

@@ -2,35 +2,65 @@ library(glmm)
 data(BoothHobert)
 
 set.seed(1234)
-zeroweight<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-          family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE,weights=c(rep(0,3),rep(1,147)))
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cluster=clust)
 
-BoothHobert2 <- BoothHobert[-c(1:3),]
-set.seed(1234)
-removed<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert2,
-                 family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE)
+mod.mcml<-mod.mcml1$mod.mcml
+z<-mod.mcml$z[[1]]
+x<-mod.mcml$x
+y<-mod.mcml$y
 
-all.equal(coef(zeroweight),coef(removed))
-all.equal(zeroweight$likelihood.value,removed$likelihood.value)
-all.equal(zeroweight$likelihood.gradient,removed$likelihood.gradient)
-all.equal(zeroweight$likelihood.hessian,removed$likelihood.hessian)
+stuff<-mod.mcml1$debug
+beta.pql<-stuff$beta.pql
+nu.pql<-stuff$nu.pql
+u.pql<-u.star<-stuff$u.star
+umat<-stuff$umat
 
-set.seed(1234)
-doubleweight<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"),data=BoothHobert,
-                 family.glmm=bernoulli.glmm,m=50,doPQL=FALSE,debug=TRUE,weights=c(rep(1,149),2))
+family.glmm<-bernoulli.glmm
 
-#umatadded <- doubleweight$umat
+objfun<-glmm:::objfun
+getEk<-glmm:::getEk
+addVecs<-glmm:::addVecs
 
-#nums <- c(rep(1,150),2)
-#for(i in 1:nrow(BoothHobert)){
-  #BoothHobert3[i] <- BoothHobert[i,]*nums[i]
-#}
[…]

### 2019-05-23 glmm/R/mcse.R
fixing tests

@@ -72,9 +72,13 @@ mcvcov <- function(object){
 	npar <- length(beta) + length(nu)
 	squaretop <- rep(0,m)
 
-	
+	if(is.null(mod$weights)){
+	  wts <- rep(1,length(y))
+	} else{
+	  wts <- mod$weights
+	}
 
-	stuff<-.C(C_mcsec, as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), ntrials=as.integer(mod$mod.mcml$ntrials), as.double(0.0))
+	stuff<-.C(C_mcsec, as.double(0.0), as.double(0.0), as.double(squaretop), numsum=as.double(rep(0,npar^2)), as.double(mod$y),as.double(t(umat)), as.integer(myq), as.integer(m), as.double(mod$x), as.integer(n), as.integer(nbeta), as.double(beta), as.double(Z), as.double(Dinvfornu), as.double(logdetDinvfornu),as.integer(family_glmm), as.double(D.star.inv), as.double(logdet.D.star.inv), as.double(mod$u.pql), as.double(Sigmuh.inv), as.double(logdet.Sigmuh.inv), pea=as.double(pea), nps=as.integer(length(pea)), T=as.integer(T), nrandom=as.integer(nrandom), meow=as.integer(meow),nu=as.double(nu), zeta=as.integer(zeta),tconst=as.double(tconst), ntrials=as.integer(mod$mod.mcml$ntrials), as.double(0.0), as.double(0.0), wts=as.double(wts))
 
 	vhatnum <- (1/m)*stuff[[4]]
 	vhatdenom <- ( stuff[[1]]   )^2

### 2019-05-23 glmm/tests/cfamilies.R

@@ -13,7 +13,8 @@ cpp<-function(eta) {(1/(1+exp(-eta)))*(1/(1+exp(eta)))}
 #check cum3 for a vector containing both positive and negative values of eta
 eta<-seq(-1,1,.1)
 neta<-length(eta)
-that<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), ntrials=as.integer(1), cumout=double(1))
+wts<-rep(1,neta)
+that<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), ntrials=as.integer(1), wts=as.double(wts),cumout=double(1))
 all.equal(sum(cum(eta)),that$cumout)
 
 #check cp3
@@ -31,9 +32,10 @@ cp <- function(eta) exp(eta)
 cpp<-function(eta) exp(eta)
 eta<-c(4.5,5,5.5)
 neta<-length(eta)
+wts<-rep(1,neta)
 
 #check cum3
-that<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1), cumout=double(1))
+that<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1),wts=as.double(wts), cumout=double(1))
 all.equal(sum(cum(eta)),that$cumout)
 
 #check cp3

### 2019-05-23 glmm/tests/coreTest.R

@@ -29,6 +29,14 @@ vars$family.glmm<-out$family.glmm
 vars$ntrials<-1
 beta.pql <- debug$beta.pql
 
+if(is.null(out$weights)){
+  wts <- rep(1, length(out$y))
+} else{
+  wts <- out$weights
+}
+
+vars$wts<-wts
+
 simulate <- function(vars, Dstarnotsparse, m2, m3, beta.pql, D.star.inv){
   #generate m1 from t(0,D*)
   if(vars$m1>0) genData<-rmvt(ceiling(vars$m1/vars$no_cores),sigma=Dstarnotsparse,df=vars$zeta,type=c("shifted"))

### 2019-05-23 glmm/tests/familiesFiniteDiffs.R

@@ -6,7 +6,7 @@ delta<-rep(.0000000001,neta)
 #check derivatives for bernoulli using finite differences
 foo<-.C(glmm:::C_cp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout
 this<-foo*delta
-that<-.C(glmm:::C_cum3,eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cumout=double(neta))$cumout-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), cumout=double(neta))$cumout
+that<-.C(glmm:::C_cum3,eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),wts=as.double(1),cumout=double(neta))$cumout-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1), ntrials=as.integer(1), wts=as.double(1), cumout=double(neta))$cumout
 all.equal(this,that)
 
 this<-.C(glmm:::C_cpp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cppout=double(neta))$cppout*delta
@@ -16,7 +16,7 @@ all.equal(this,that)
 #check derivatives for poisson using finite differences
 foo<-.C(glmm:::C_cp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cpout=double(neta))$cpout
 this<-foo*delta
-that<-.C(glmm:::C_cum3,eta=as.double(eta+delta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cumout=double(neta))$cumout-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2), ntrials=as.integer(1),cumout=double(neta))$cumout
[…]

### 2019-05-23 glmm/tests/mcseTest.R

@@ -9,17 +9,17 @@ addMats<-glmm:::addMats
 #other functions we'll need
 
 #here is el, likelihood
-elR <- function(Y,X,eta,family.mcml){
+elR <- function(Y,X,eta,family.mcml,wts){
 	family.mcml<-getFamily(family.mcml)
 	neta<-length(eta)
 
 	if(family.mcml$family.glmm=="bernoulli.glmm"){
-		foo<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), ntrials=as.integer(1), cumout=double(1))$cumout
+		foo<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta), type=as.integer(1), ntrials=as.integer(1), wts=as.double(wts), cumout=double(1))$cumout
 		mu<-.C(glmm:::C_cp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cpout=double(neta))$cpout
 		cdub<-.C(glmm:::C_cpp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(1),ntrials=as.integer(1),cppout=double(neta))$cppout
 	}
 	if(family.mcml$family.glmm=="poisson.glmm"){
-		foo<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cumout=double(1))$cumout
+		foo<-.C(glmm:::C_cum3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),wts=as.double(wts),cumout=double(1))$cumout
 		mu<-.C(glmm:::C_cp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cpout=double(neta))$cpout
 		cdub<-.C(glmm:::C_cpp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cppout=double(neta))$cppout
 	}
@@ -115,6 +115,12 @@ mcseTEST <- function(mod){
 	T<-length(mod$z)
 	nrand<-lapply(mod$z,ncol)
 	nrandom<-unlist(nrand)
+	
+	if(is.null(mod$weights)){
[…]

### 2019-05-23 glmm/tests/weightsTest.R

@@ -27,8 +27,6 @@ objfun<-glmm:::objfun
 getEk<-glmm:::getEk
 addVecs<-glmm:::addVecs
 
-eta1<-rep(2,150)
-
 ############################################
 #this should be the same as el
 getFamily<-glmm:::getFamily
@@ -72,24 +70,9 @@ NEWelR <-
       cdub<-.C(glmm:::C_cpp3,eta=as.double(eta),neta=as.integer(neta),type=as.integer(2),ntrials=as.integer(1),cppout=double(neta))$cppout
     }
     
-    
-    #if(ncol(X) > 1){
-      #wtX <- matrix(rep(0, nrow(X)*ncol(X)), nrow=nrow(X), ncol=ncol(X))
-      #for(i in 1:nrow(X)){
-      #for(j in 1:ncol(X)){
-          #wtX[i,j] <- wts[i]*X[i,j]
-        #}
-      #}
-    #} else {
-      #wtX <- rep(0, nrow(X))
-      #for(i in 1:nrow(X)){
-        #wtX[i] <- wts[i]*X[i]
-      #}
-    #}
-    
-    
     wtsmat <- diag(wts)
     wtX <- wtsmat%*%X
+    
     value<-as.numeric(Y%*%wtsmat%*%eta-foo)
     gradient<-t(wtX)%*%(Y-mu)	
     cdubmat<-diag(cdub)
@@ -100,11 +83,38 @@ NEWelR <-
 
 #compare elR and NEWelR for weights all equal 1
 eta1<-rep(2,150)
-that1<-elR(mod.mcml$y,mod.mcml$x,eta1,family.mcml=bernoulli.glmm, wts=wts) 
-this1 <- NEWelR(mod.mcml$y,mod.mcml$x,eta1,family.mcml=bernoulli.glmm, wts=wts)
-all.equal(as.numeric(that1$value),as.numeric(this1$value))
-all.equal(as.numeric(that1$gradient),as.numeric(this1$gradient))
-all.equal(as.numeric(that1$hessian),as.numeric(this1$hessian))
+thatALL1<-elR(mod.mcml$y,mod.mcml$x,eta1,family.mcml=bernoulli.glmm, wts=wts) 
+thisALL1 <- NEWelR(mod.mcml$y,mod.mcml$x,eta1,family.mcml=bernoulli.glmm, wts=wts)
+all.equal(as.numeric(thatALL1$value),as.numeric(thisALL1$value))
[…]

### 2019-06-27 DesignDocs/DesignDocWeights.tex
updated design doc

@@
 First, the function will need to establish whether the user has supplied a proper weighting scheme.
-Next, the weighting scheme will need to be applied in the \texttt{el.C} function.
-After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied.
+Next, the weighting scheme will need to be applied in the \texttt{el.C}, \texttt{elval.C} and \texttt{elGH.C} functions.
+After defining the weighting vector, the remainder of this section will illustrate how this weighting scheme will be applied in the previously mentioned functions.
+Lastly, the changes made to the importance sampling distribution are outlined.
 
 \subsection{The Weighting Vector}
-The weights must be in the form of a vector and the vector must be the same length as the response vector, $y$. All weights must be numeric and non-negative. There can be no missing weights.
+If a proper weighting scheme has been supplied, the function will utilize that weighting scheme; if  the weighting scheme supplied is improper, the user will receive an error; and if no weighting scheme is supplied, a weighting vector of all ones will be created.
+The weights must be in the form of a vector and the vector must be the same length as the response vector, $y$.
+All weights must be numeric and non-negative.
+There can be no missing weights.
 
@@
 \begin{align}
-\dfrac{\partial}{\partial \beta_l} \log \text{RWLD}(\theta) &= \lambda \ X \left[ y- c'(\eta) \right]
+\dfrac{\partial}{\partial \beta_l} \log \text{RWLD}(\theta) &= \Lambda \ X \left[ y- c'(\eta) \right]
 \end{align}
+
+\noindent where $\Lambda$ is the $n$ x $n$ weight matrix made from $\lambda$, which is the $n$ x $1$ weights vector given by the user or made of all ones if no vector is given.
 
@@
 \begin{align}
-\dfrac{\partial^2}{\partial \beta_l^2} \log \text{RWLD}(\theta) &= \lambda \ X' \left[- c''(\eta) \right] X
+\dfrac{\partial^2}{\partial \beta_l^2} \log \text{RWLD}(\theta) &= X' \left[- c''(\eta) \right] \Lambda \ X
+\end{align}
+
+\subsection{The Importance Sampling Distribution}
+
+The original importance sampling distribution is
+
+\begin{align}
+\tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
+\end{align}
+
+\noindent and the portion that is concerned with weighting is $Z' c"(X \beta^* + Zu^*) Z$.
+We know that $\eta = X \beta + Zu$, so this is similar to $Z' c"(\eta) Z$ which looks similar to the second derivative of the cumulant.
+That similarity is used to incorporate the weighting scheme into the importance sampling distribution.
+
+The weighted version of the importance sampling distribution follows as
+
+\begin{align}
+\tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) \Lambda \ Z +(D^*)^{-1}   )^{-1}) \\
+&= p_1 \exp( \log  \grave{f}(u|0,D^*) ))+p_2 \exp( \log(  f(u \, | \, u^*, D^*) ))+ \nonumber \\
+&+ p_3 \exp( \log( f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) \Lambda \ Z +(D^*)^{-1}   )^{-1}) )).
+\end{align}
+
+\noindent Then, if we set
+
+\begin{align}
+\tilde{a}= \max \left\{ \log  \grave{f}(u|0,D^*) ), \log  f(u \, | \, u^*, D^*) ), \log f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) \Lambda \ Z +(D^*)^{-1}   )^{-1})  \right\}
+\end{align}
+
+Then,
+
+\begin{align}
+ \tilde{f}(u) 
+&= p_1 \exp\left( \log  \grave{f}(u|0,D^*)  -\tilde{a}\right)+p_2 \exp \left( \log  f(u \, | \, u^*, D^* )-\tilde{a} \right)+ \nonumber\\
+&+ p_3 \exp\left[ \log f \left(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) \Lambda \ Z +(D^*)^{-1}   )^{-1} \right)  -\tilde{a} \right].
 \end{align}

### 2019-06-27 glmm/tests/weightsTest.R

@@ -1,13 +1,20 @@
 library(glmm)
 data(BoothHobert)
 
+clust <- makeCluster(1)
 set.seed(1234)
-mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE)
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=21, doPQL=TRUE, debug=TRUE, cluster=clust)
+
+vars1 <- new.env(parent = emptyenv())
 
 mod.mcml<-mod.mcml1$mod.mcml
-z<-mod.mcml$z[[1]]
-x<-mod.mcml$x
-y<-mod.mcml$y
+#z<-mod.mcml$z[[1]]
+#x<-mod.mcml$x
+#y<-mod.mcml$y
+
+y<- mod.mcml$y[1:149]
+x<-mod.mcml$x[1:149]
+z<- mod.mcml$z[[1]][1:149]
 
 if(is.null(mod.mcml1$weights)){
   wts <- rep(1, length(y))
@@ -18,14 +25,97 @@ if(is.null(mod.mcml1$weights)){
 stuff<-mod.mcml1$debug
 beta.pql<-stuff$beta.pql
 nu.pql<-stuff$nu.pql
+vars1$nu.pql<-nu.pql
 u.pql<-u.star<-stuff$u.star
+vars1$u.star<-u.star
 umat<-stuff$umat
+vars1$family.glmm<-mod.mcml1$family.glmm
+vars1$umat<-stuff$umat
+vars1$newm <- nrow(vars1$umat)
+vars1$ntrials<-1
+D.star.inv <- Dstarnotsparse <- vars1$D.star <- as.matrix(stuff$D.star)
+
+vars1$m1 <- stuff$m1
+m2 <- stuff$m2
+m3 <- stuff$m3
+
+vars1$zeta <- 5
 
 family.glmm<-bernoulli.glmm
 
 objfun<-glmm:::objfun
 getEk<-glmm:::getEk
 addVecs<-glmm:::addVecs
+
+vars1$cl <- mod.mcml1$cluster
+registerDoParallel(vars1$cl)                   #making cluster usable with foreach
+vars1$no_cores <- length(vars1$cl)
+
+vars1$mod.mcml<-mod.mcml1$mod.mcml
+
[…]

### 2019-07-18 DesignDocs/DesignDocWeights.tex

@@
 
+\section{Summary Changes}
+
+The coefficient estimates in the summary should be changed in order to reflect the certainty known from the Monte Carlos standard error (\texttt{mcse}). This means that \texttt{glmm} must:
+
+\begin{enumerate}
+\item Detect the position of the first non-zero digit in the \texttt{mcse} estimates;
+\item Round the coefficient estimates in the \texttt{glmm} summary to the number of digits detected in step 1.
+\end{enumerate}
+
 \end{document}

### 2019-07-25 DesignDocs/DesignDocWeights.tex
finalizing package

@@
 Thus, the optional relevance-weighting scheme will allow the user to place a heavier weight on the more informative and/or trustworthy observations in their data set so that those data points that are less informative affect the model to a lesser degree.
+
+Additionally, a change to the \texttt{glmm} summary relating to the certainty of estimates is outlined at the end of the document.
 
@@
 
-\section{Summary Changes}
+\section{Summary Change}
 
-The coefficient estimates in the summary should be changed in order to reflect the certainty known from the Monte Carlos standard error (\texttt{mcse}). This means that \texttt{glmm} must:
+The coefficient estimates in the summary should be changed in order to reflect the certainty known from the Monte Carlo standard error (\texttt{mcse}). This means that \texttt{glmm} must:
 
@@
 \item Round the coefficient estimates in the \texttt{glmm} summary to the number of digits detected in step 1.
+\item Round the standard error estimated in the \texttt{glmm} summary to the number of digits detected in step 1.
 \end{enumerate}

### 2019-07-25 glmm/R/summary.mcla.R

@@ -12,6 +12,19 @@ summary.glmm <- function(object,...){
 	z<-mod.mcml$z
 	
 	#the coefficients matrix for fixed effects
+	#truncating beta significance based on mcse
+	mcerror<- mcse(mod.mcml)
+	split<- strsplit(as.character(mcerror),split="")
+	d<-rep(0,length(split))
+	for(i in 1:length(split)){
+	  j<-1
+	    while(split[[i]][j] == "0" | split[[i]][j] == "."){
+	      d[i] <- d[i] +1
+	      j<-j+1
+	    }
+	}
+	d <- d-1
+	
 	beta<-mod.mcml$beta
 	nbeta<-length(beta)
 	hessian<-mod.mcml$loglike.hessian
@@ -27,6 +40,12 @@ summary.glmm <- function(object,...){
 	#fixed effects table
 	beta.se<-all.ses[1:nbeta]
 	zval<-beta/beta.se
+	for(i in 1:length(beta)){
+	  beta[i] <- round(beta[i], digits = d[i])
+	  beta.se[i] <- round(beta.se[i], digits = d[i])
+	}
+	#beta <- round(beta, digits = d[1])
+	#beta.se <- round(beta.se, digits = d[1])
 	coefmat<-cbind(beta,beta.se,zval,2*pnorm(abs(zval),lower.tail=F))
 	colnames(coefmat)<-c("Estimate","Std. Error", "z value", "Pr(>|z|)")
 	rownames(coefmat)<-colnames(mod.mcml$x)
@@ -36,6 +55,12 @@ summary.glmm <- function(object,...){
 	nu<-mod.mcml$nu
 	nu.se<-all.ses[-(1:nbeta)]
 	nuzval<-nu/nu.se
+	for(i in 1:length(nu)){
+	  nu[i] <- round(nu[i], digits = d[length(beta)+i])
+	  nu.se[i] <- round(nu.se[i], digits = d[length(beta)+i])
+	}
+	#nu <- round(nu, digits = d[2])
+	#nu.se <- round(nu.se, digits = d[2])
 	nucoefmat<-cbind(nu,nu.se,nuzval,pnorm(abs(nuzval),lower.tail=F))
 	colnames(nucoefmat)<-c("Estimate","Std. Error", "z value", "Pr(>|z|)/2")
 	rownames(nucoefmat)<-mod.mcml$varcomps.names

### 2019-07-25 glmm/tests/weightsTest.R

@@ -1,121 +1,20 @@
 library(glmm)
 data(BoothHobert)
 
-clust <- makeCluster(1)
 set.seed(1234)
-mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=10^5, doPQL=TRUE, debug=TRUE, cluster=clust)
+#model with all weights at 1, no duplicate data points in data set
+mod.mcml1<-glmm(y~0+x1,list(y~0+z1),varcomps.names=c("z1"), data=BoothHobert, family.glmm=bernoulli.glmm, m=10^2, doPQL=TRUE, debug=TRUE)
 
-vars1 <- new.env(parent = emptyenv())
-
-z<-mod.mcml1$z[[1]]
-x<-mod.mcml1$x
-y<-mod.mcml1$y
-
+#weights are determined from model (should be all 1)
 if(is.null(mod.mcml1$weights)){
-  wts <- rep(1, length(y))
+  wts <- rep(1, length(mod.mcml1$y))
 } else{
   wts <- mod.mcml1$weights
 }
 
-stuff<-mod.mcml1$debug
-beta.pql<-stuff$beta.pql
-nu.pql<-stuff$nu.pql
-vars1$nu.pql<-nu.pql
-u.pql<-u.star<-stuff$u.star
-vars1$u.star<-u.star
-umat<-stuff$umat
-vars1$family.glmm<-mod.mcml1$family.glmm
-#vars1$umat<-stuff$umat
-vars1$newm <- nrow(mod.mcml1$umat)
-vars1$ntrials<-1
-D.star.inv <- Dstarnotsparse <- vars1$D.star <- as.matrix(stuff$D.star)
-
-vars1$m1 <- stuff$m1
-m2 <- stuff$m2
-m3 <- stuff$m3
-
-vars1$zeta <- 5
-
-family.glmm<-bernoulli.glmm
-
-objfun<-glmm:::objfun
-getEk<-glmm:::getEk
-addVecs<-glmm:::addVecs
-genRand<- glmm:::genRand
-
-vars1$cl <- mod.mcml1$cluster
-registerDoParallel(vars1$cl)                   #making cluster usable with foreach
-vars1$no_cores <- length(vars1$cl)
-
-vars1$mod.mcml<-mod.mcml1$mod.mcml
-
-vars1$wts <- c(rep(1,149),2)
-
[…]

### 2020-07-16 DesignDocs/DesignDocDimReduction.tex
started design doc for dim reduction

--- 2016-06-06 NewDesignDoc/DesignDocGLMM.tex
@@
 
-\title{Design Document for R Package glmm}
+\title{Design Document for Dimension Reduction in R Package glmm}
 
@@
 \begin{abstract}
-This design document describes the process of performing Monte Carlo maximum likelihood (MCML) for generalized linear mixed models.
-First, penalized quasi-likelihood (PQL) estimates are calculated, which help generate simulated random effects.
-Then, the Monte Carlo likelihood approximation (MCLA)  is calculated using the simulated random effects.
-Next, the MCLA is maximized to find Monte Carlo maximum likelihood estimates, the corresponding Fisher Information, and other statistics.
-Additional inference is then possible, including confidence intervals for the parameters.
+This design document describes the process of regularizing the random effects.
+For example, Sally and Tammy can share a random effect (and therefore be treated as the same salamander) if they have very similar random effects (such as $1.22$ and $1.24$).
+The goal of reducing the number of random effects is to reduce the computation time for MCLA.
 \end{abstract}
 
-\section{Theory}
+\section{Background}
 
 Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data.
-Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved, normally-distributed random effects centered at 0 with variance matrix $D$.
+Let $u$ be a $q$-vector of unobserved, normally-distributed random effects centered at 0 with variance matrix $D$.
 Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$.
@@
 Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.
-MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is the mixture distribution specified in section \ref{sec:genRand}.
+.
 
@@
 \end{align}
-When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
-\begin{align}
-\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
-\end{align}
-and the Hessian matrix of the MCLA is
-\begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \label{eq:MCLAhessian}
-\end{align}
-Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and Hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.
 
-Now any likelihood-based inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.
+If $u_k$ were of length $q_2 < q$, this calculation would be less expensive. We therefore are trying to shorten the random effect vector $u$.
 
-\section{Model fitting function} 
-The model fitting function is be the primary function the user  uses.
-The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}.
-Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.
-As a result of the user specifying the fixed effects, the model fitting function creates matrix $X$, which has dimensions $n \times p$.
+\section{Overview}
 
-The user  specifies the random effects in the same way as the fixed effects. This is also how users of  \texttt{reaster} (of R package
- \texttt{aster} \citep{aster-package}) specify the random effects.
-That is, random effects are expressed using the R formula mini-language.
-Let $q_t$ be the number of random effects associated with variance component $\nu_t$.
-When the random effects are specified, a list of $T$ model matrices are created.
-The $t$th model matrix $Z_t$ has dimensions $n \times q_t$.
+The package, in its current state has two main steps:
+\begin{enumerate}
+\item Use PQL to estimate the parameters and predict the random effects. (These predictions are then used to create the importance sampling distribution.)
+\item Use MCLA to approximate the likelihood and conduct MCML.
+  \end{enumerate}
 
-The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the names of the variance components.
-
-Thus, the following is a sample command with fixed predictors $x_1$ and $x_2$ and with random effects created by categorical variables $school$ and $classroom$:
-\begin{verbatim}
-glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="bernoulli.glmm", 
-data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
-debug=FALSE)
- \end{verbatim}
-
-It is possible that the user could want some variance components to be set equal.
-For example, \pcite{coull:agresti:2000} influenza dataset contains four years with random effects for each year.
-The authors want the within-year variance components to be equal.
-There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since the strain of flu during year 4 was a repeat of a previous year).
-Suppose \texttt{year} is a categorical variable with four levels, and \texttt{year1} through \texttt{year4} are dummy variables.
-Thus, the call  contains these arguments:
-\begin{verbatim}
-glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeat,~0+year3,~0+year4),
- varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeat"),
-data=flu,family.glmm=bernoulli.glmm)
-\end{verbatim}
-
-Initially, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices.
-Since we have 3 distinct variance components, we want 3 design matrices.
-We take the design matrices that share a variance component and use \texttt{cbind} to bind them together.
-The result is 3 model matrices in the $Z$ list, one for each variance component.
-We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  (``subject'',``year'',``repeat'') are correctly assigned.
-The variance estimates  are eventually listed in the model summary in this same order as well.
-
-  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.
-The process of finding these estimates is detailed in section \ref{sec:pql}.
-The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.
-More information on generating the random effects is in section \ref{sec:genRand}.
-
-Next, \texttt{trust} is implemented to maximize  the MCLA objective function (details on the objective function are in section \ref{sec:objfun}).
-Finally, the function  returns parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood,  the Hessian of the log likelihood at those estimates, information from the \texttt{trust} optimization, and other information.
-
-%%Families%%
-\section{Families} \label{sec:fam}
-
-Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
-\begin{align}
-g(\mu) = X \beta + Z u
-\end{align}
-The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$.
-If the data have a  Bernoulli distribution, the link is $\logit(\mu)$.
-If the data have a Poisson distribution, the link is $\log (\mu)$.
-Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.
-
- For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z u$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
-\begin{align}
-y' \eta - c(\eta) = \sum_i \left[ y_i \eta_i - c(\eta_i)  \right]
-\end{align}
-
-The user is required to specify the family in the model-fitting function.
-Once the family is specified, many family-specific functions are called.
-They are contained in an S3 class called ``glmm.family''.
-Each family function outputs a list including the family name (a character string such as ``bernoulli.glmm''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.
-
-The users provide the family in the model-fitting function by either enter the character string (``bernoulli.glmm''), the function (\texttt{bernoulli.glmm()}), or the result of invoking the function.
-The following code for using the input to determine the family is adapted from \texttt{glm}.
-
-\begin{verbatim}
-logDensity<-function(family.glmm)
-{
-	if(is.character(family.glmm))
-		family.glmm<-get(family.glmm,mode="function",envir=parent.frame())
-	if(is.function(family.glmm))
-		family.glmm<-family.glmm()
-	if(!inherits(family.glmm,"glmm.family")) 
-		stop(" 'family.glmm' not recognized") 
-	return(family.glmm)
-}
-\end{verbatim}
-We interpret this as follows.
-If the user has entered the family as a string, go get the R object with that family name, either from the immediate environment or the parent environment.
-If this has happened, \texttt{family.glmm} is now a function.
-If \texttt{family.glmm} is a function (either because the user entered it as a function or because of the preceding step), invoke that function.
-At this point, \texttt{family.glmm} should have  class ``\texttt{glmm.family}.'' If this is not the case (maybe because of a typo or maybe because they entered ``\texttt{poisson}'' rather than ``\texttt{poisson.glmm}''), then stop and return an error.
-
-With this family of functions, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
-\begin{verbatim}
-family.glmm$cum(args)
-family.glmm$cp(args)
-family.glmm$cpp(args)
-\end{verbatim}
-
-For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows.
-In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.
-\begin{align}
-c(\eta_i) &= \log(1+e^{\eta_i}) =
-  \begin{cases}
-    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
-    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
-  \end{cases}\\
-c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
-c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
-\end{align}
-
- For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
-\begin{align}
-c(\eta_i)&=e^{\eta_i}\\
-c'(\eta_i)&=e^{{\eta_{i}}}\\
-c''(\eta_i)&= e^{{\eta_{i}}}.
-\end{align}
-
-Then we use these pieces to create the scalar $c(\eta)$, the vector $c'(\eta)$ and the matrix $c''(\eta)$. We calculate
-
-\begin{align}
-c(\eta)= \sum_i c(\eta_i).
-\end{align}
- The vector $c'(\eta)$ has components $c'(\eta_i)$. The matrix $c''(\eta)$ is diagonal with diagonal elements $c''(\eta_i)$.
-
-In the R function \texttt{glm}, the user can choose the link.
-The canonical link must be used in the \texttt{glmm} package so that we have an exponential family.
-The canonical link is included in \texttt{family.glmm}  in case the user does not know it already.
-
-Also, this family of functions contains a check to ensure the data are valid given the family type.
-If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should contain only $0$ and $1$.
-If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative.
-If the data set does not pass the check, the check returns an error message.
-
-\subsection{Redone in C}
-Redone in C, I  have  separate functions for calculating $c(\eta)$, $c'(\eta)$, $c''(\eta)$: \texttt{cum3.c}, \texttt{cp3.c}, and \texttt{cpp3.c}.
-The inputs for each function are identical:  \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family: $1$ indicates Bernoulli and $2$ indicates Poisson), and an array of doubles to contain the result.
-These are passed in as pointers.
-The functions  calculate  the cumulant function or one of its first two derivatives.
-Each function  contains a switch statement for the glmm family.
-The calculations for each of these functions have  been shown earlier in this section.
-This function is type void: rather than returning the cumulant or its derivatives, the pointers are changed to contain the results.
-This function is invoked by \texttt{el}, described in section \ref{sec:el}.
-
-\section{Log density of the data (\texttt{el})}\label{sec:el}
-This section provides details for the log density of the data and two of its derivatives.
-
-\subsection{Equations}
-Recall the log of the data density is
-\begin{align}
-\log f_\theta(y|u) &= y' \eta -c(\eta) \\
-&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
-\end{align}
-where
-\begin{align}
-\eta=X\beta+Zu.
-\end{align}
-The derivative of this with respect to one component, $\eta_j$, is
-\begin{align}
-\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
-\end{align}
-The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
-\begin{align}
-\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
-\end{align}
-
-We'd like the derivative of the log of the data density with respect to $\beta$. The first step is using the chain rule as follows:
-\begin{align}
-\dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
-&= \left[ y_j-c'(\eta_j) \right]  X_{j{l}}
-\end{align}
-
-The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
-\begin{align}
-\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
-&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
-&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
- \end{align}
-
-Letting $c'(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
-\begin{align}
-\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ y-  c'(\eta)  \right].
-\end{align}
-
-Letting $ c''(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
-\begin{align}
-   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ -  c''(\eta) ] X
-\end{align}
-
-\subsection{Redone in C}
-The C function to calculate the value of the log data density and its two derivatives are called by reference.
-The following pointers are passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}.
-The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} are zeros before \texttt{el.C} is invoked.
-Invoking \texttt{el.C}  then places the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\
-
-The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the Hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.
-
-\section{Random effect generation and calculations}
-This section is focused on a vector of random effects $u$ with length $q$.  Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$.
-Section \ref{sec:ftwiddle} states the importance sampling distribution.
- Section \ref{sec:genRand} describes the process of generating the random effects.
-Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.
-
-%%getEk
-\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
-Recall that, for this version of the package, we assume that $D$ is diagonal.
-Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$.
-That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as its variance component.
-Then $D=\sum_{t=1}^T\nu_t  E_t $.
-
-Recall $q_t$ is the number of random effects associated with variance component $\nu_t$.
-Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.
-The model fitting function of \texttt{glmm} has made  \texttt{z} into a list with $T$ design matrices (one for each distinct variance component).
-In order to create $E_t$, we need to  count the  number of columns ($q_t$) for each design matrix in the list.
-Since $D$  is $q \times q$, each $E_t$  is $q \times q$.
-$E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.
-$E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.
-We continue this process to complete the construction of $E_t, t=1,\ldots, T$.
-
-\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
- Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale.
-That is $s$ is defined such that, $u= D^{1/2}s$.
-Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t},t=1,\ldots,T$.
-Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$.
-The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$.
-Let
-\begin{align}
-A^*=\sum_{t=1}^T E_t \sigma^*_t
-\end{align}\
-and
-\begin{align}
- D^* =A^*A^*
-\end{align}
-be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
-\begin{align}
-u^*&=A^*s^*.
-\end{align}
-
-Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.
-Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.
-Let $\grave{f}(u|0,D^*)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate  $t$ with mean $0$, scale matrix $D^*$, and $\zeta$ degrees of freedom.
-Then the importance sampling distribution is:
-\begin{align}
- \tilde{f}(u) = p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
-\end{align}
-
-The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.
-The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance.
-The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.
-The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y,u)$.
-
-\subsection{Generating random effects (genRand)}\label{sec:genRand}
-
- Recall that $m$ is the overall Monte Carlo sample size.
-Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.
-We randomly determine $m_1,m_2,m_3$ according to the proportions $p_1,p_2,p_3$.
-More specifically, we generate $m$ random numbers from a distribution that is uniform between $0$ and $1$.
-Then $m_1$ is the frequency with which the randomly generated numbers are between $0$ and $p_1$,  $m_2$ is the frequency with which the randomly generated numbers are between $p_1$ and $p_1+p_2$, and $m_3=m-p_1-p_2$.
-%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
-
-Once $m_1, m_2$ and $m_3$ are determined, we draw 
-\begin{align}
-u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
-u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
-u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
-\end{align}
-Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}. We use R package \texttt{mvtnorm} to draw from a $t$ distribution.
-
-\subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
-In order to generate draws from a nonstandard normal distribution, we need to calculate the square root of a variance matrix.
-When the variance matrix is diagonal (for example, $D^*$), we simply take the root of the diagonal elements.
-When the variance matrix is not diagonal (for example, in the second component of the mixture distribution), we can use eigendecomposition.
-Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.
-
-Let $\Sigma$ denote the nondiagonal variance matrix.
-Eigendecomposition of $\Sigma$ provides orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).
-Then
-\begin{align}
-\Sigma^{1/2}= O \Lambda^{1/2} O'.
-\end{align}
-Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
-\begin{align}
-\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\ \nonumber
-&= O \Lambda^{1/2}\Lambda^{1/2} O'\\\nonumber
-&=O \Lambda O' \\\nonumber
-&= \Sigma.
-\end{align}
-
-\subsubsection{Generating from nonstandard normal distributions}\label{sec:gennonstand}
- Construct $A^*=(D^*)^{1/2}$ and $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.
-
-Let $\breve{u}_k,k=m_1+1\ldots,m$ be vectors of random effects that are drawn from the standard normal distribution. Then we center and scale $\breve{u}_k$ as follows:
-\begin{align}
-u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
-u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
-\end{align}
- The result or centering and scaling is that
-\begin{align}
- &u_k \sim N(u^*,D^*), \, k=m_1+1,\ldots,m_1+m_2\\
-& u_k \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}), \, k=m_1+m2+1,\ldots,m .
-\end{align}
-
-% 
-%\subsubsection{NEED TO PUT THIS SOMEWHERE}
-%
-%
-%Recall that $D$ is a diagonal with dimension $q$.
-$D$ has $q_1$ entries being $\nu_1$, $q_2$ entries being $\nu_2$, ..., and $q_T$ entries being $\nu_T$.
-Since the distribution of $u_k=A \breve{u}_k$ is $N(0,D)$, we write its pdf as:
-%\begin{align}
-%f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
-%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
-%&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
-%\end{align}
-%So that 
-%\begin{align}
-%\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
-%&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
-%\end{align}
-%Then 
-%\begin{align}
-%\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
-%\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
-%\end{align}
-%
-%Because only the first component of the mixture distribution contains any parameters,
-%\begin{align}
-%\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
-%\end{align}
-%and 
-%\begin{align}
-%\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
-%\end{align}
-%These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
-%\begin{align}
-%\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
-%&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
-%\end{align}
-%and
-%\begin{align}
-%\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
-%\end{align}
-%
-%
-
-%%
-%%This section is useful for calculating $\tilde{f}(u_k)$.
-This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives.
-These are needed for the calculation in \eqref{eq:MCLA}.
-%%
-%%Section \ref{sec:tee} contains details that are not currently implemented, but could be.
-In this version, all random effects are assumed to be normally distributed and all simulated random effects are generated from a normal distribution.
-
-\subsection{Evaluating the distribution of random effects (\texttt{distRand})}\label{sec:distRand}
-This section discusses evaluating the distribution of the random effects, the gradient of the random effects, and the Hessian of the random effects.
-That is, the equations in this section provide  $\log f_\theta(u)$, $\nabla \log f_\theta(u)$, and $\nabla^2 \log f_\theta(u)$  for equation \ref{eq:MCLA}.
-
- We can  express the log unnormalized pdf of $N(0, D)$  as:
-\begin{align}
-\log f_\theta(u)= \log f (u| 0, D) = - q/2 \log (2 \pi) -1/2 \log |D| - (1/2) u' D^{-1} u
-\end{align}
-
- To find derivatives of this, we are going to use the diagonal form of $D$.
-Recall that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s  elsewhere.
-
-We can partition the random effects according to their variance components: $u=(U_1',...,U_T')'$.
-Let $D_t$ be the variance matrix for $U_t$.
-$D_t$ has $q_t$ rows  and $q_t$ columns.
-Thus $D$ can be expressed as:
-\begin{align}
-D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
-\end{align}
-
-Since $D$ is diagonal, it follows that $D^{-1}$ is also diagonal with diagonal entries $\dfrac{1}{\nu_1}$,...,$\dfrac{1}{\nu_T}$.
-Also, the assumption that $D$ is diagonal makes calculating the determinant of $D$  easy:
-\begin{align}
-|D|= \nu_1^{q_1}...\nu_T^{q_T}
-\end{align}
-
- Taking these two pieces of information into account allows us to write the log density for $U$ as follows:
-\begin{align}
-\log f_\theta(u) &= -\dfrac{1}{2} \log |D| -\dfrac{1}{2} u' D^{-1} u\\
-&= -\dfrac{1}{2} \left[  \sum_{t=1}^T q_t \log \nu_t   \right]  -\dfrac{1}{2} \sum_{t=1}^T \left[ \dfrac{1}{\nu_t} U_t'U_t   \right]
-\end{align}
-
-The first and second derivatives of each summand with respect to its associated $\nu_t$ are:
-\begin{align}
-\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}U_t'U_t
-\end{align}
-and 
-\begin{align}
-\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} U_t'U_t.
-\end{align}
-Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
-\begin{align}
-\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
-\end{align}
-Also,
-\begin{align}
-\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
-\end{align}
-Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is the following vector of length $T$:
-\begin{align}
-\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} U_1'U_1 & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} U_T'U_T   \end{bmatrix} 
-\end{align}
-
-The Hessian matrix is the following diagonal matrix:
-\begin{align}
-\dfrac{\partial^2}{\partial \nu^2} \log f_\theta(u) = \begin{bmatrix} \dfrac{q_1}{2 \nu_1^2}- \dfrac{1}{\nu_1^3} U_1'U_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & \dfrac{q_T}{2 \nu_T^2}- \dfrac{1}{\nu_T^3 U_T'U_T} \end{bmatrix}
-\end{align}
-%I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$.
-Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.
-
-To calculate the value, gradient, and Hessian, we need to provide  $\nu$, $u$, and the list \texttt{z} (from \texttt{mod.mcml}).
-The list \texttt{z} has $T$ matrices, each with the number of columns equal to $q_t$.
-We need $q_t,t=1,\ldots,T$ to calculate the log density and its derivatives.
-
-The last thing we need to discuss is how to split $U$ into $U_1,...,U_T$.
-We know that the first $q_1$ items of U are $U_1$, the next $q_2$ items are $U_2$, etc.
-In other words, entries 1 through $q_1$ are $U_1$.
-Items $q_1+1$ through $q_1+q_2$ are $U_2$, etc.
-We find these numbers ($q_1$, $q_2$, etc) from the number of columns of the items in the list \texttt{z}.
-This information of how to split $u$ up is contained in vector \texttt{meow} and used by function \texttt{distRand3}.
-This is faster than recalculating \texttt{meow} for every call of \texttt{distRand3}.
-
-Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting Hessian.
-
-The result of invoking the C function \texttt{distRand3} is  the calculation of the gradient and Hessian for the distribution of the random effects.
-The value is  evaluated by the C function \texttt{distRandGeneral}, described in section \ref{sec:Dgeneral}.
-
-\subsection{Evaluating the importance sampling distribution}
-This section discusses evaluating the importance sampling distribution shown in \ref{eq:ftwiddle}.
-The  $\tilde{f}(u)$ calculation is used in equation \ref{eq:MCLA}.
-Section \ref{sec:Dgeneral} describes how to evaluate the pdf for the normal components of $\tilde{f}(u)$.
-Section \ref{sec:tstuff} describes how to evaluate the pdf for the t component of $\tilde{f}(u)$.
-
-To be computationally stable, we need to be careful  with the addition of the terms in \ref{eq:ftwiddle} to prevent overflow. We know that
-\begin{align}
- \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \\
-&= p_1 \exp( \log  \grave{f}(u|0,D^*) ))+p_2 \exp( \log(  f(u \, | \, u^*, D^*) ))+ \nonumber\\
-&+ p_3 \exp( \log( f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) )).
-\end{align}
-Set
-\begin{align}
-\tilde{a}= \max \left\{ \log  \grave{f}(u|0,D^*) ), \log  f(u \, | \, u^*, D^*) ), \log f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})  \right\}
-\end{align}
-Then
-\begin{align}
- \tilde{f}(u) 
-&= p_1 \exp\left( \log  \grave{f}(u|0,D^*)  -\tilde{a}\right)+p_2 \exp \left( \log  f(u \, | \, u^*, D^* )-\tilde{a} \right)+ \nonumber\\
-&+ p_3 \exp\left[ \log f \left(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1} \right)  -\tilde{a} \right].
-\end{align}
-is more stable computationally.
-
-%%distRandGeneral
-\subsubsection{Normal distribution for general variance matrix $\Sigma$ (\texttt{distRandGeneral})}\label{sec:Dgeneral}
-Let $\Sigma$ be a variance matrix and let $\mu$ be a mean vector. Consider $N(\mu, \Sigma)$.   We can  write the log pdf of this distribution  as:
-\begin{align}
-\log f (u| \mu, \Sigma) = -\dfrac{q}{2} \log (2 \pi) + \dfrac{1}{2} \log |\Sigma^{-1}| - \dfrac{1}{2} (u-\mu)' \Sigma^{-1} (u-\mu)
-\end{align}
- The only part of this to discuss is $|\Sigma^{-1}|$.
-We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues.
-Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$.
-Thus
-\begin{align}
-|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
-&= |O'| \; |O| \; |\Lambda| \\
-&=|\Lambda|,
-\end{align}
-which is just the product of the eigenvalues.
-The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times throughout each trust optimization iteration, where $m$ is again the Monte Carlo sample size.
-
-This function is also rewritten in C with the following  passed in as pointers: double \texttt{Sigma.inv} $\Sigma^{-1}$, double \texttt{logdet} $\log |\Sigma^{-1}|$, int \texttt{nrow}, double \texttt{uvec} a vector of random effects, double {mu} $\mu$, and double \texttt{distRandGenVal}.
-
-\subsubsection{t distribution}\label{sec:tstuff}
-
-Consider $t_\zeta$. We can write the log unnormalized pdf of this distribution as:
-\begin{align}
-\log \grave{f}(u|0,D^*)&=  \log \left( \Gamma \left( \dfrac{\zeta+q}{2}  \right)   \right)-\log \left( \Gamma \left( \dfrac{\zeta}{2}  \right)   \right)  - \dfrac{q}{2} \log(\zeta \pi)\\
-&- \dfrac{1}{2} \log |D^*|     -\left[ \dfrac{\zeta+q}{2} \right] \log \left(1+\dfrac{u'D^{* -1}u}{\zeta} \right)\\
-\end{align}
-
-The first four terms are constant with respect to $u$ and can be calculated ahead of time.
-Rewritten in C, we need to pass in  \texttt{tconstants} (the sum of the first four terms) as a double, \texttt{u} as an array of doubles, \texttt{zeta} as an int, \texttt{myq} as an int, the diagonal of \texttt{Dstarinv} as an array of doubles,  and \texttt{logft} as a double.
-The final calculation will be written into \texttt{logft}.
-
-I have another function (written in R) that calculates the constants for the t distribution.
-The arguments of this function are \texttt{zeta}, the diagonal of $D^*$ \texttt{Dstardiag}, and the length of $u$ \texttt{myq}.
-
-%\subsection{$\tilde{f}(u_k)$}
-%I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust.
-I can make a function using the general form for a normal distribution's density as described in the previous subsection, then use that function for the three distributions that I sampled from.
-Since I generated the random effects from the three distributions in proportions $p_1,p_2,p_3$, I can write
-%\begin{align}
-%\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
-%\end{align}
-%
-%\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
-%If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL).
-These can be drawn using the package mvtnorm.
-This package can also give the density evaluated at certain points.
-I'm not sure if I'll use this or just stick to normals.
-
-%%%%%%%%%%%%%%%%%
-%\section{t distribution for generating random effects}
-%Generating random effects from a normal distribution can give us some random effects from the tails of the distribution, which results in importance sampling weights that are problematically big.
-To try to fix this, we can change our importance sampling distribution to a multivariate t-distribution with location parameter $\mu$, scale parameter $D$, and $\gamma$ degrees of freedom.
-The pdf is proportional to
-%\begin{align}
-%f(U |\theta)=  |D|^{-.5} \left[ 1+\dfrac{1}{\gamma} (U-\mu)' D^{-1}(U-\mu)   \right]^{(\gamma+p)/2}
-%\end{align}
-%
-%Then the log density is
-%\begin{align}
-%\log f(U |\theta) &\propto -.5 \log |D| - \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} (U-\mu)' D^{-1} (U-\mu)  \right] \\
-%& \propto \sum_{t=1}^T \left[- .5  q_t \log \nu_t \right]- \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} (U-\mu)' D^{-1} (U-\mu)  \right]  \\
-%& \propto \sum_{t=1}^T \left[- .5  q_t \log \nu_t \right]- \dfrac{\gamma+q}{2} \log \left[ 1+ \dfrac{1}{\gamma} \sum_{t=1}^T \dfrac{1}{\nu_t} (U_t-\mu_t)' (U_t-\mu_t)  \right]  \\\end{align}
-%Its first derivative with respect to $\nu_t$ is
-%\begin{align}
-%\dfrac{\partial}{\partial \nu_t} \log f(U |\theta) &= \dfrac{- q_t}{2 \nu_t} + \dfrac{\gamma+q}{2} \dfrac{\dfrac{1}{\gamma \nu_t^2}(U_t-\mu_t)' (U_t-\mu_t)}{1+ \dfrac{1}{\gamma} \sum_{t=1}^T \dfrac{1}{\nu_t} (U_t-\mu_t)' (U_t-\mu_t)}
-%\end{align}
-%Its second derivative with respect to $\nu_t$ is
-%
-%\begin{align}
-%\dfrac{\partial^2}{\partial \nu_t^2} \log f(U |\theta) &= \dfrac{q_t}{2 \nu_t^2} - \dfrac{\gamma+q}{2} \left[ \dfrac{\dfrac{1}{\nu_t^2 \gamma}(U_t-\mu_t)'(U_t-\mu_t)}{1+ \sum_{t=1}^T \dfrac{1}{\nu_t \gamma }(U_t-\mu_t)'(U_t-\mu_t) }  \right]^2 - \dfrac{\gamma+q}{2}  \left[ \dfrac{\dfrac{2}{\nu_t^3 \gamma}(U_t-\mu_t)'(U_t-\mu_t)}{1+ \sum_{t=1}^T \dfrac{1}{\nu_t \gamma}(U_t-\mu_t)'(U_t-\mu_t) }  \right]
-%\end{align}
-
-\section{Objective function: approximated log likelihood}\label{sec:objfun}
-The objective function is optimized by \texttt{trust} within the the model fitting function.
-In order to evaluate and maximize the approximated log likelihood, \texttt{trust} needs an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.
-Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, and \ref{eq:MCLAhessian}   for calculating these quantities.
-%This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.
-
-Denote
-\begin{align}
-b_k = \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k). \label{eq:MCLA}
-\end{align}
-
-For computational stability, we'll set $a=$max$(b_k)$. Then the value of the MCLA is expressed as:\\
-\begin{align}
-\l_m (\theta) = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
-\end{align}
-
-Define the weights as:
-\begin{align}
-v_\theta(u_k,y) = \dfrac{e^{b_k-a}}{ \sum_{k=1}^m e^{b_k-a}} 
-\end{align}
-
-Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
-\begin{align}
-\nabla l_m(\theta) &= \sum_{k=1}^m  \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
-&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)   \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
-&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
-&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]   \right) v_\theta(u_k,y)   \\
-&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(u_k)  \right] v_\theta(u_k,y) \label{eq:MCLAgr}
-\end{align}
-
-Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
-\begin{align}
-&\nabla^2 l_m(\theta)=  \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)   v_\theta(u_k,y) \label{eq:MCLAh} \\ 
-&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-    \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
-\end{align}
-Everything in this has already been defined except:
-\begin{align}
-\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
-&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} +
-\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
-&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
-\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
-&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
-\end{align}
-
-%HERE IS WHEN FTWIDDLE DEPENDS ON NU
-%Rewriting equation \ref{eq:MCLAgradient} using the notation for the weights, the gradient vector is: 
-%\begin{align}
-%\nabla l_m(\theta) &= \sum_{k=1}^m \left( \nabla \log f_\theta (u_k,y) - \nabla \log \tilde{f}(u_k)  \right) v_\theta(u_k,y)\\ \label{eq:gradient}
-%&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k) - \nabla \log \tilde{f}(u_k)  \right] v_\theta(u_k,y)\\
-%&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k) - \nabla \log \tilde{f}(u_k) \right] v_\theta(u_k,y)\\
-%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
-%&+  \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log \tilde{f}(u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu}  \log \tilde{f}(u_k) \right] v_\theta(u_k,y) \\
-%&=  \sum_{k=1}^m \left(  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] + \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right]  +\left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right] \right) v_\theta(u_k,y)   \\
-%&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \left( \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right)+\left( \dfrac{\partial}{\partial \nu} \log \tilde{f}(u_k) \right) \right] v_\theta(u_k,y) \label{eq:MCLAgr}
-%\end{align}
-%
-%Then the Hessian matrix from equation \ref{eq:MCLAhessian} is expressed as: 
-%\begin{align}
-%&\nabla^2 l_m(\theta)=  \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-%   \nabla^2 \log \tilde{f}(u_k)    \right] v_\theta(u_k,y) \label{eq:MCLAh} \\ 
-%&+   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-%   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-%   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  v_\theta(u_k,y)   . \nonumber
-%\end{align}
-%Everything in this has already been defined except:
-%\begin{align}
-%\nabla^2 \log f_\theta (u_k,y) &=  \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k,y) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k,y)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k,y)  \end{bmatrix} \\
-%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} +
-%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k) \\ \dfrac{\partial^2}{\partial \beta \: \partial \nu} \log f_\theta (y|u_k)  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (y|u_k)  \end{bmatrix} \\
-%&= \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} + 
-%\begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0 & 0  \end{bmatrix} \\
-%&= \begin{bmatrix} \dfrac{\partial^2}{\partial \beta^2} \log f_\theta (y|u_k) & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log f_\theta (u_k)  \end{bmatrix} 
-%\end{align}
-%
-%
-
-\section{Summary of model}
-Typing \texttt{summary(mod)}  provide a summary of the model.
-This is broken into two pieces (as is usually done for summaries in R):  \texttt{summary.glmm} and \texttt{print.summary.glmm}.
-The first contains more calculated results than the second.
-When a user types \texttt{summary(mod)}, only the  basic information is automatically printed.
-More information can be found in the summary list.
-
-The summary performs all calculations and its value is a list of the following:
+I am figuring out one of these two options:
 \begin{itemize}
-\item the call.
-\item a matrix with the predictor name in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided p-value  in the fifth column.
-All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
-\item a matrix with the variance estimate name  in the first column, the estimated variance component in the second column, the standard error in the third column, the z value in the fourth column, and the one-sided p-value  in the fifth column.
-All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
-\item the evaluated Monte Carlo log likelihood along with its first and second derivative.
-\item $m_1$, $m_2$, $m_3$.
-\item output from texttt{trust}, such as whether it converged. (See \texttt{trust} for more details.)
-\item the generated random effects.
+\item[Plan A:] Change PQL so that similar random effects are collapsed into a shared, single random effect.
+\item[Plan B:] Keep the PQL optimization as it is, and add another step in between the two steps. At this time, the similar random effects are collapsed into a shared, single random effect.
   \end{itemize}
 
-A note on the standard errors: to calculate the standard errors, we take the MCLA Hessian matrix, invert it, take the diagonal elements, and apply the square root to them.
-It is possible that the Hessian will be noninvertible if it is close enough to singular that the computer thinks it is singular.
-Then the standard errors will all be infinite  the user is warned.
+  I'm not sure, but it seems like Plan A is better than Plan B. It makes sense that we'd want to try collapsing the random effects while making sure this simplification not ruin the fit of the model.
 
-Then \texttt{print.summary.glmm}  prints the following:
-\begin{itemize}
-\item the call.
-\item a matrix with the predictor name in the first column, the estimated coefficient in the second column, the standard error in the third column, the z value in the fourth column, and the two-sided asymptotic p-value  in the fifth column.
-\item a matrix with the variance estimate name  in the first column, the estimated variance component in the second column, the standard error in the third column, the z value in the fourth column, and the one-sided p-value  in the fifth column.
-All inference is asymptotic, so we use the standard normal distribution to calculate the p-values.
-\end{itemize}
-
-%I think for the printCoefmat to work, the fixed effects matrix needs to look like
-%\begin{verbatim}
-%            Estimate Std. Error z value     Pr(>|z|)
-%(Intercept)        2       0.50       4 6.334248e-05
-%x1                 2       0.25       8 1.244192e-15
-%\end{verbatim}
-%(or at least it worked when I had these as the column names and didn't work before that).
-
-Note that the summary and print.summary functions are S3 generic functions.
-This means is the user  types \texttt{summary(mod)} and \texttt{summary} is a generic function.
-R checks the class of the model \texttt{mod} and  automatically uses the \texttt{summary} and \texttt{print.summary} functions for that class of objects.
+   In the currently used version of PQL, there is an optimization within an optimization.
+The inner optimization estimates the fixed effects and predicts the random effects.
+The outer optimization works on the variance components.
+\textbf{I believe that adding a penalty term to the inner loop will collapse the random effects.} We want the random effects to be similar to one other, so the distance (or the square of the distance, if we prefer a ridge-like penalty) between the random effects must be controlled.
 
@@
 Both variations have an inner optimization and an outer optimization.
-The inner optimization is a well behaved quadratic while the outer optimization is more challenging.
 I use the variation of PQL that is not the original version, but is pretty close and is better behaved.
@@
 Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
+\textbf{I believe that adding a penalty term to the inner loop will collapse the random effects.
+We want the random effects to be similar to one other, so the distance (or the square of the distance) between the random effects must be controlled.
+In other words, if $s$ has components $s_1, \ldots , s_q$, then we want to have penalty term $\sum_{k \neq j} |s_k - s_j|$ if we want a lasso-like penalty or $\sum_{k \neq j} (s_k - s_j)^2$ if we want a ridge-like penalty.}
+
+Of course, PQL already has a penalty term in order to shrink the random effects toward 0.
+We will maintain this penalty term (I think) and add a second penalty (to group together similar random effects).
 
@@
 
-The inner optimization maximizes the penalized log likelihood. After defining
+The inner optimization maximizes the penalized log likelihood. Defining
 \begin{align}
@@
 \end{align}
-and the penalized likelihood as:
+and the penalized likelihood (pre-collapsing) as:
 \begin{align}
@@
 \end{align}
+Then our doubly-penalized likelihood is
+\begin{align}
+ l(\eta)- \dfrac{1}{2} s's. - \lambda \sum_{k \neq j} (s_k - s_j)^2
+\end{align}
+where $\lambda$ is some number we can tweak to force more/less collapsing.
+
+I HAVE NOT REDONE THESE CALCULATIONS BELOw
+
 Since this function is maximized using  \texttt{trust}, we need to provide derivatives with respect to $s$ and $\beta$. We express these via the multivariate chain rule, taking advantage of $\eta_i$.
@@
 
-\section{Checks}
-\subsection{Checking the MCLA finite differences}
-To check the function that calculates the MCLA $l_m(\theta)$, I use finite differences on the \citet{booth:hobert:1999} example.
-To do this, I chose a value of $\theta=(\beta,\sigma)$ and a relatively small value of $\delta$, where $\delta$ is a vector of length $2$.
-We can check that the value and first derivative of the MCLA function are calculated correctly by making sure the following approximation holds
-\begin{align}
-\nabla l_m (\theta)  \cdot \delta \approx l_m(\theta+\delta)-l_m(\theta).
-\end{align} 
-Then, we can check the first and second derivatives of the MCLA are calculated correctly by checking for the following approximation:
-\begin{align}
-\nabla^2 l_m (\theta) \delta \approx \nabla l_m (\theta+\delta)-\nabla l_m (\theta)
-\end{align}
-
-\subsection{Checking  functions using the Booth and Hobert example}
-I also test the objective function by taking the  \citet{booth:hobert:1999} example and rewriting the functions for this specific example.
-I then compare the values produced by the check functions and the original functions.
-Functions checked this way are:
-\begin{enumerate}
-\item log of the data density (\texttt{el})
-\item log of the density for the random effects (both \texttt{distRand} and \texttt{distRandGeneral})
-\item the Monte Carlo likelihood approximation (\texttt{objfun})
-\end{enumerate}
-
-\subsection{Checking a putative MLE using MCMC}
-
-Suppose $\hat{\theta}$ is claimed to be the MLE.
-For example, $\hat{\theta}$ could be the MCMLE.
-Ideally, we would like to check whether the true likelihood $l(\theta)$ achieves a local max at $\hat{\theta}$.
-Due to the intractible integral in the likelihood expression, the best we can do is make sure that  $\nabla l_m(\theta)$ evaluated at $\hat{\theta}$ is very close to $0$.
-
-Suppose $u_k,k=1,\ldots,m$ are sampled from importance sampling distribution $\tilde{f} (u_k)$. Recall that the gradient of the MCLA is:
-\begin{align}
-\nabla l_{m}(\theta) = \dfrac{\sum_{k=1}^m\frac{\nabla f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k)}}{\sum_{k=1}^m \frac{  f_\theta(u_k)   }{\tilde{f}(u_k)}}
-\end{align}
-We can choose $\tilde{f}(u_k)=f_{\hat{\theta}}(u_k,y)$.
-If the putative MLE is truly the MLE, then $\theta=\hat{theta}$, which in turn implies that $\tilde{f}(u_k)=f_{{\theta}}(u_k,y)$.
-To be clear, in this check  we no longer use a mixture of normals for $\tilde{f}(u_k)$.
-With the selected importance sampling distribution, each of the importance sampling weights is equal to $1$ and the sum of the weights is $m$.
-Therefore, the gradient of the MCLA simplifies as follows:
-\begin{align}
-\nabla l_{m}(\theta) &=\dfrac{1}{m} \; {\sum_{k=1}^m\dfrac{\nabla f_\theta(y|u_k) f_\theta(u_k)   }{\tilde{f}(u_k,y)}}\\
-&= \dfrac{1}{m} \; {\sum_{k=1}^m\dfrac{\nabla f_\theta(u_k,y)   }{\tilde{f}(u_k,y)}} \\
-&= \dfrac{1}{m} \; \sum_{k=1}^m \nabla \log f_\theta(u_k,y)
-\end{align}
-This shows that the gradient of the MCLA is the average of the gradient of the complete data log likelihood, as long as $\hat{\theta}$ is truly an MLE.
-We can produce $u_k,k=1,\ldots,m$, using Markov chain Monte Carlo.
-Using $u$ as the variable, w run the Markov chain (perhaps \texttt{metrop} from the R package \texttt{mcmc}) on the complete data log likelihood.
-We then use these samples to calculate the gradient of the complete data log likelihood, which in turn calculates the gradient of the MCLA.
-
- If we split the MCMC runs into batches, we can calculate the batch means of the MCLA gradient, the grand mean of the MCLA gradient, and the corresponding Monte Carlo standard error.
-If the putative MLE truly maximizes the likelihood, then the MCLA gradient's components should be close to $0$.
-We can check that they are close enough to $0$ by comparing them to the Monte Carlo standard error.
-
-%%%%%%LIKELIHOOD RATIO TEST STUFF
-%\section{Likelihood ratio test}
-%This is another S3 generic function that the user can choose to implement if they'd  like  to do likelihood ratio tests for nested models.
-Eventually I'll want this to handle an arbitrary number of models, but for now we'll stick to comparing two models: $mod2$ nested in $mod1$.
-I will assume these models have already been fit.
-%
-%
-% There are a few ways that the two models could differ. In other words, we could be testing:
-% \begin{enumerate}
-% \item whether one or more fixed effects are zero.
-% \item whether one variance component is zero.
-% \item whether multiple variance components are zero.
-% \item whether one or more fixed effects is zero and one or more variance components are zero. 
-%\end{enumerate}
-%The hypothesis testing procedure and method for calculating p-values are different for each of these cases, so I'll go through each one in the following subsubsections.
-%
-%\subsection{Testing whether one or more fixed effects are zero}
-%Consider the  case of testing whether one or more fixed effects are zero and the random effects are the same between the two models.
-I think we just do a likelihood ratio test as we're familiar with for linear models.
-Let the number of fixed effect parameters in the larger model be $p_1$  and the number in the nested model be $p_2$.
-I will need the log likelihood values of the models outputted (from the main function at the same time as when I output the MCMLEs and the Fisher information).
-The log likelihood is simply the ``value'' from the objective function that trust uses.
-Let's call the log likelihood from the larger model  $l_1$ and the log likelihood from the nested model  $l_2$.
-%
-%Then the likelihood ratio test statistic is
-%\begin{align}
-%t_{LRT}= -2l_1+2l_2
-%\end{align}
-%and this follows a $\chi^2$ distribution with $p_1-p_2$ degrees of freedom. The calculation
-%\begin{align}
-%P(t_{\chi^2_{p_1-p_2}>LRT} ).
-%\end{align}
-%gives us a two-sided p-value to fit with the two-sided alternative hypothesis.
-%
-% 
-% \subsection{Testing whether one variance component is zero}
-%  Hypothesis testing for one variance component is easy but not what most people expect.
-%  In this setup, the two models have the exact same fixed effects.
-The only difference is the larger model has one more random effect $\nu_{t_1}$.
-This means we want to test whether $\nu_{t_1}=0$.
-A variance component must be nonnegative, meaning the alternative hypothesis is that $\nu_{t_1}>0$.
-In other words, the hypotheses are:
-% \begin{align}
-% H_0: \nu_{t_1}=0\\
-% H_1: \nu_{t_1}>0.
-% \end{align}
-% Note the alternative hypothesis is one-sided. This means we need to calculate a one-sided p-value. Let $t_{LRT}=2l_2-2l_1$. If we naively follow the p-value calculation of
-% \begin{align}
-%P( \chi^2_{1}>t_{LRT})
-%\end{align}
-% we will end up calculating a two-sided p-value.
-To fix this, we cut this p-value in half.
-In other words, use the standard normal distribution $Z$ and think of calculating the one-sided p-value this way:
-% \begin{align}
-% P( Z>\sqrt{|  t_{LRT} |})
-% \end{align}
-% 
-% However, this is not obvious; I never would have thought about having a test statistic of
-% \begin{align}
-% \sqrt{2 | l_2-l_1   |   }.
-% \end{align}
-% 
-% \subsection{Testing whether multiple variance components are zero}
-% In this case, the fixed effects are identical between the larger and nested model.
-The only difference is that the larger model has more than one additional variance components.
-This gets a lot more complicated (for example, there is no clear way to count parameters so calculating the degrees of freedom is out the window).
-Luckily, someone's worked out the details already.
-Charlie suggested a few papers to look at.
-I'll read about and add this a bit later.
-Charlie is hoping that we'll have a mixture of $\chi^2$ distributions, such as
-% \begin{align}
-% \dfrac{1}{2}P (\chi^2_1> t_{LRT}) +\dfrac{1}{4}P (\chi^2_2> t_{LRT}).
-% \end{align}
-% The reason this is complicated is because the variance components are restricted to be nonnegative.
-This means that the likelihood is only defined for nonnegative variance components.
-Then differentiating the likelihood at $0$ becomes tricky because that's the boundary.
-%
-% \subsection{Testing whether one or more fixed effects is zero and one or more variance components are zero}\label{sec:combotest}
-% This seems very complicated. I don't know if it's any more complicated than the previous case of testing multiple variance components. I'll have to think about it later when I have time.
-% 
-% \subsection{Determining the hypothesis test}
-% To follow convention, this command will be called ``anova'' and the two arguments will be the two models to be compared.  The command would look like
-%\begin{verbatim}
-%anova(mod1,mod2)
-%\end{verbatim} 
-%
-% R will first need to figure out if the fixed effects are different and, if they are, which model is the larger model.
-%\begin{verbatim}
-%if(length(coef(mod1))>length(coef(mod2)))
-%      {bigmod<-mod1;  smallmod <-mod2} 
-%if(length(coef(mod1))<length(coef(mod2)))
-%      {bigmod<-mod2; smallmod <-mod1}
-%\end{verbatim}
-%Next, if there is a difference in the fixed effects, I'm going to make sure the fixed effects of the small model are nested in those of the big model.
-Otherwise, produce an error.
-Then, continue with the testing.
-Note: this check for nesting isn't perfect, but no other anova checks which model is bigger and whether they're nested.
-%\begin{verbatim}
-%if(big mod is defined) {
-%      pnames1<-names(coef(bigmod))
-%      pnames2<-names(coef(smallmod))
-%      if(sum(pnames2 %in% pnames1) != length(pnames2)) {
-%            stop("The models you provided are not nested.")}
-%
-%     if(the variance components differ between the two models){
-%             check big model has more variance components (ow: error)
-%             check variance components of the small model are nested (ow: error)
-%             calculate p-value according to section 3.3.4. return it.
-%      }
-%      if(the variance components do not differ between the two models){
-%            calculate p-value according to 3.3.1. return it.
-%      }
-%}
-%\end{verbatim}
-%If and only if we've gotten to the end of this chunk of code without returning anything, bigmod has not been defined, meaning the fixed effects are same.
-Therefore, we now need to figure out whether the variance components differ by one or by more than one.
-In order to compare the number of random effects, we need to figure out how to get the number of variance components from each model.
-I  don't yet have a solid grasp of how formula works its magic, but I have the impression that I can count the number of model matrices returned in order to figure out the number of variance components because there is a model matrix for each one.
-%
-%\begin{verbatim}
-%T1<- number of variance components for mod1
-%T2<- number of variance components for mod2
-%if(T1>T2){call mod1 the bigmod and mod2 the smallmod}
-%if(T1<T2){call mod2 the bigmod and mod1 the smallmod}
-%if(bigmod is defined){
-%     check  small model is nested in the big model (ow: error)
-%     (maybe do this by looking at the call?)
-%     if(T1==T2+1 || T2==T1+1) 
-%           {calculate and return p-value according to sec 3.3.2}     
-%     calculate and return p-value according to section 3.3.3
-%     }
-%if still haven't returned anything, produce error.     
-%\end{verbatim}	
-%
-%Why the last error?
-If we have gotten to this point in the code with nothing returned, it ends up that the number of variance components in each model are equal, meaning the user has made some kind of mistake.
-Either the user provided two identical models or they provided non-nested models (e.g.
-they have the same number of variance components, but different components in each model).
-Either way, we can't help them.
-%
-%
-%
-%
-%
-%This command would return something that first reminds the user what predictors are in each model, then has a table.
-Each model would have one row of the table.
-The columns would contain the model name,  the log likelihood for each model, the number of parameters in the model.
-Then the next columns would have one entry each: the calculated difference between the log likelihoods, the calculated difference between the number of parameters, and the p-value of the test.
-
-%\section{AIC}
-%Typing 
-%\begin{verbatim}
-%AIC(mod)
-%\end{verbatim} 
-%will return the AIC of the model. To calculate AIC, I'll need the total number of parameters ($p+T$)  and the Monte Carlo log likelihood evaluated at the MCMLEs $l$. Then the AIC is
-%\begin{align}
-%2(p+T) - 2 l 
-%\end{align}
-
-\section{Confidence intervals}
-The user can implement this command to create confidence intervals after  fitting a model $mod$ using the main function.  This is an S3 generic.The command would look like
-\begin{verbatim}
-confint(mod,parm,level=.95)
-\end{verbatim}
-The only required argument would be the fitted model (the first argument). The third argument (the confidence level) has a default of $.95$.
-
-If the second argument is omitted, confidence intervals would be created for all of the parameters.
-There are two options to calculate confidence intervals for a subset of the parameters.
-The user can provide either  the names of the coefficients or a vector with length equal to the number of parameters (entries being 1 if they would like that intervals for that parameter and 0 otherwise).
-The code to do this is taken from ``confint.lm'' and is:
-\begin{verbatim}
-function (object, parm, level = 0.95, ...) 
-{
-    cf <- coef(object)
-    pnames <- names(cf)
-    if (missing(parm)) 
-        parm <- pnames
-    else if (is.numeric(parm)) 
-        parm <- pnames[parm]
-        stopifnot(parm %in% pnames)
-
-        rest of the code to actually do the confidence intervals goes here
-        }
-\end{verbatim}
-What this says: write down the coefficient names of the object we're given.
-If parm is missing, we're going to assume they want to create intervals for all the parameters.
-If parm is numeric (a vector of 0s and 1s), use that to select the parameter names we want and call that selection parm.
-Finally, stop the whole process if parm (the parameter names we want intervals for) don't appear in the model.
-
-The form for a confidence interval is the point estimate plus or minus the standard error of the point estimate times some cutoff.
-The point estimate and the standard error are from the summary of the model.
-The reference distribution used for the cutoff is the standard normal.
-This will produce an asymptotic confidence interval.
-
-The output would either be a vector of length 2 (if creating confidence interval for only one parameter) or a matrix with 2 columns (one for the lower bound of the confidence interval, one for the upper bond).
-The column names will be ``$(100-level)/2$'' and ``$50+level/2 $.'' (in the default case, this is``$2.5\%$'' and ``$97.5\%$'').
-
-The only potential hangup is when creating confidence intervals for the variance components $\nu_t, t=1,...,T$.
-We know that $\nu_t >0$ for all $t=1,...,T$.
-We'll find this boundary again makes things complicated.
-I don't yet know how to deal with this.
-
-To illustrate the problem, consider the scenario that the margin of error for $\nu_t$ is greater than the estimate of $\nu_t$ itself.
-It wouldn't make sense to produce a confidence interval with a negative lower bound.
-It could make sense to truncate the interval so that the lower bound starts at 0 and the upper bound remains untouched.
-
-\section{New Additions (Summer 2016)}
-\subsection{New Family: Binomial}
-One year later, I'm adding binomial.
-For Binomial, the calculations change only in the cumulant function.
-They rely on one more number: the number of trials.
-The inputted response is usually a vector, but for binomial (with the exception of Bernoulli) we need 2 vectors cbinded (cbound?) together.
-The first column is the number of successes and the second is the number of failures.
-Summing across the row gives us the number of trials.
-I call this $ntrials$ from here on out.
-
-The cumulant function and its first two derivatives with respect to the canonical parameter $\eta_i$ are below. 
-\begin{align}
-c(\eta_i) &=ntrials_i \, \log(1+e^{\eta_i}) =
-  \begin{cases}
-   ntrials_i \, \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
-   ntrials_i \,\left(  \eta_i+\log(e^{-\eta_i}+1) \right) & \text{if } \eta_i >0,
-  \end{cases}\\
-c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{ntrials_i}{1+e^{-\eta_i}}\\
-c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{ntrials_i}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
-\end{align}
-
-I will implement this change in steps. Game plan:
-\begin{enumerate}
-\item Without adding binomial yet, add the ntrials argument to all levels of code so that ntrials can be passed to the cumulant functions and their derivatives.
-At the top, I will have a variable to set ntrials = 1 (for now).
-As I do each layer, I will make sure I get all the same answers from my salamanders problem (or some other bernoulli problem).
-\item Write the binomial cumulant in R. Do some checks with  both $\eta$ positive and negative.
-\begin{enumerate}
- \item Check to make sure this calculation is identical to that produced by bernoulli for some parameter value $\eta_i$ and for $ntrials=1$. 
-\item Check  the sum of Bernoulli cumulants equals the binomial cumulant (for ntrials $>1$). 
-\item Check finite differences. 
-\item Check the three calculations for $ntrials=5$ (or some number other than 1) compared to my by-hand calculation. 
-\end{enumerate}
-\item Code it in C and check that the calculation is identical to that produced in R for $ntrials=1$ and for $ntrials \neq 1$.
-\item I will go to the top-most layer, right where the model matrices are being made.
-The new response vector y will be the first column of the user-provided response.
-The sum of the two entries in the row will create the vector ntrials.
-I will check that the salamanders problem gives the same results (when specified as cbind(Mate, 1-Mate)).
-
-\item I will add checks for the model matrix stage.  See below.
-\item I want to allow the response for bernoulli to be a single vector  (rather than two cbinded vectors).
-Therefore, if the family is binomial.glmm and the length of the dimension of the response is 1 (meaning we just have a vector), then create the ntrials vector to be a vector of 1s.
-Test this with salamanders or some other problem to make sure it gives the same results.
-
-\end{enumerate}
-
-Checks for the model matrix stage: 
-\begin{enumerate}
-\item  If  the length of the dimension of the response is 2 dimensions (meaning we have both rows and columns), then the family must be binomial.glmm.
-\item If the family is binomial.glmm, then the response must have 2 columns
-\end{enumerate}
-
-These changes will need to be done in both C and R. The checks will be in R. I'll need to add  documentation for binomial. Bernoulli will be retained.
-
-%
-%\section{Prediction intervals}
-%This is another S3 generic.
-The user can implement this command to create prediction intervals after  fitting a model $mod$ using the main function.
-Still learning about these.
-I currently know almost nothing.
-What I do know is that I will need the Monte Carlo log likelihood along with its first and second derivatives.
-I'll get to practice my first-year-theory delta method skills.
-%
-%Charlie says that neither predict.glm nor predict.glmmPQL do prediction intervals.
-The former produces parameter estimates with standard errors.
-The latter does predictions but does not produce a standard error, so no interval can be formed.
-%
-%One wrong thing to do: once you find the MCMLEs, you simply plug those into the distribution for $U|Y$ and pretend those parameter estimates have no variability and are  correct.
-Then the random effect has a normal distribution and it'd be easy to find quantiles.
-The only problem is this is wrong.
-%
-%We are hoping to find a way to use Fisher information for the Monte Carlo MLE. Hopefully we can do something more similar to the usual asymptotics.
-%
-%A last resort is a parametric bootstrap.
-I haven't spent much time thinking about this, but I think the way this involves plugging parameter estimates  into $\log f_\theta(y|u)$ and $\log f_\theta(u)$.
-Then you repeat: draw a value of $u$, plug that into $\log f_\theta(y|u)$, draw a value of $y$.
-I say this is a last resort  because having two Monte Carlo calculations will take a long time.
-
-\bibliographystyle{apalike}
-\bibliography{brref}
-
-\appendix
-\section{MCLA calculations}
-Let $u_k,k=1,\ldots,m$ be a sample from $\tilde{f}(u_k)$. The Monte Carlo log likelihood approximation is
-\begin{align}
-l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
-&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
-\end{align}
-The gradient vector and Hessian matrix calculations depend on whether $\tilde{f}(u_k)$ contains $\theta$.
-The calculations in \ref{sec:calcs} are for $\tilde{f}(u_k)$ containing $\theta$ and the calculations in \ref{sec:calcsindep} are for $\tilde{f}(u_k)$ not containing $\theta$.
-
-\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}\label{sec:calcsindep}
-
-When $\tilde{f}$ is independent of $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
-\begin{align}
-\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
-\end{align}
-and the Hessian matrix of the MCLA is
-\begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
-&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
-\end{align}
-To reduce the risk of catastrophic cancellation, we can combine the last two terms of the Hessian:
-\begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
-\end{align}
-We are able to combine the last two terms because $\nabla l_m(\theta)$ is a weighted mean of $\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)$. Letting
-\begin{align}
-Z&=\nabla \log f_\theta (y,u_k)
-\end{align}
- we can use the following equality:
-\begin{align}
-E(ZZ')-E(Z)E(Z)' = \left[ E(Z-EZ)  \right]\left[ E(Z-EZ)  \right]'.
-\end{align}
-
-\subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$} \label{sec:calcs}
-This section is still under construction.
-
-When $\tilde{f}$ contains $\theta$, the gradient of the MCLA is
-\begin{align}
-\nabla l_m(\theta) &= \dfrac{ \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}  \right] }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
-&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)   }{\tilde{f}(u_k)} - \dfrac{ f_\theta(y,u_k)  \nabla \tilde{f}(u_k) }{\left(\tilde{f}(u_k) \right)^2 } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
-&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)}{f_\theta(y,u_k)} \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
- \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \dfrac{ \nabla \tilde{f}(u_k) }{\tilde{f}(u_k) } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
-&= \dfrac{   \sum_{k=1}^m  \nabla \log f_\theta(y,u_k) \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
- \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  \nabla \log \tilde{f}(u_k)  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
-&= \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} 
-\end{align}
-
-Then the Hessian of the MCLA is
-\begin{align}
-\nabla^2 l_m (\theta) &= \nabla \left[ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \right] \\
-&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]  \nabla \left[ \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \right]  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right]\\
-&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \left[  \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)     \right]'\\
-&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&- \left[ \nabla l_m(\theta)  \right] \left[ \nabla l_m(\theta)  \right]'
-\end{align}
-To reduce the risk of catastrophic cancellation, we can combine the last two terms of the Hessian:
-\begin{align}
-\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
-   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
-&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
-   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
-\end{align}
-We are able to combine the last two terms because $\nabla l_m(\theta)$ is a weighted mean of $\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)$. Letting
-\begin{align}
-Z&=\nabla \log f_\theta (y,u_k)- \nabla \log \tilde{f}(u_k)
-\end{align}
- we can use the following equality:
-\begin{align}
-E(ZZ')-E(Z)E(Z)' = \left[ E(Z-EZ)  \right]\left[ E(Z-EZ)  \right]'.
-\end{align}
-
-\section{Central Limit Theorem for MCLA}\label{sec:CLT}
-In this section, we focus on the gradient of the MCLA because the trust criterion for finding the maximum is based on the gradient when $\tilde{f}(u)$ is independent of $\theta$. 
-Define
-\begin{align}
-\gamma_1&= \int f_\theta(u,y) du\\
-\gamma_2 &=\int \tilde{f}(u)du.
-\end{align}
-Recall the calculation for the MCLA gradient originally stated in  \eqref{eq:MCLAgradient}:
-\begin{align}
-\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }
-\end{align}
-
- Note that both the numerator and denominator are sample means. By the law of large numbers, the numerators and denominator  each converge to their true means. That is,
-\begin{align}
-\dfrac{1}{m}\sum_{k=1}^m \nabla \log f_\theta (u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-&= \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log f_\theta (u,y)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
-&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right]
-\end{align}
-%and
-%\begin{align}
-%\dfrac{1}{m}\sum_{k=1}^m \nabla \log \tilde{f}(u_k)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow  E_{\tilde{f}} \left[ \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-%&= \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-%&= \dfrac{\gamma_1}{\gamma_2} \int \nabla \log \tilde{f}(u)  \dfrac{f_\theta(u,y)}{\gamma_1}  du \\
-%&=\dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log \tilde{f}(u)  \right]
-%\end{align}
-and
-\begin{align}
-\dfrac{1}{m}\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} &\rightarrow E_{\tilde{f}} \left[ \dfrac{f_\theta(u,y)}{\tilde{f}(u)}  \right] \\
-&= \int \dfrac{f_\theta(u,y)}{\tilde{f}(u)} \dfrac{\tilde{f}(u)}{\gamma_2} du\\
-&= \dfrac{\gamma_1}{\gamma_2} \int \dfrac{ f_\theta(u,y)}{\gamma_2} du \\
-&= \dfrac{\gamma_1}{\gamma_2}
-\end{align}
-Then, by Slutsky's theorem,
-\begin{align}
-\nabla l_m(\theta)&\rightarrow \dfrac{  \dfrac{\gamma_1}{\gamma_2} E_f \left[ \nabla \log f_\theta(u,y)  \right] }{\dfrac{\gamma_1}{\gamma_2}}\\
-&=  E_f \left[ \nabla \log f_\theta(u,y) \right] \\
-&=\nabla l(\theta)
-\end{align}
-In addition to a law of large numbers, we would like a Central Limit Theorem for $\nabla l_m(\theta)$. In other words, the quantity
-\begin{align}
-&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- \nabla l(\theta)  \right] \\
-&=
-\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\; \nabla  l(\theta)  \\
-&=\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
- -\nabla l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
-\end{align}
-will have a normal distribution if and only if the variances of
-\begin{align}
-\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
-and
-\begin{align}
-\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} 
-\end{align}
-are finite.
-First, we want to show that
-\begin{align}
-\sum_{k=1}^m \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
-has finite variance. This is true if and only if 
-\begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du < \infty.
-\end{align}
-We see:
-\begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &=\int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{\tilde{f}(u)} du\\
-&\leq \int \dfrac{f_\theta(y|u)^2 f_\theta(u)^2}{p_1 \grave{f}(u|0,D^*)} du\\
-&\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, f_\theta(y|u)^2  }{\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du
-\end{align}
-Since
-\begin{align}
-f_\theta(y|u)^2=\left( e^{y'\eta-c(\eta)} \right)^2 = e^{2y'\eta-2c(\eta)} ,
-\end{align}
-\begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{\exp(-u'D^{-1}u) \, \exp(2y'\eta)  }{\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2} \,  \exp(2c(\eta))  } du\\
-&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'\eta)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du \\
-&=\dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  \exp(2c(\eta))  } du 
-\end{align}
-When $y|u\sim$ Bernoulli, then
-\begin{align}
-c(\eta)=\log(1+e^\eta) \Rightarrow \exp(2c(\eta))&= \exp( 2 \log(1+e^\eta)  )\\
-&= ( \exp( \log(1+e^\eta)  )  )^2\\
-&= (1+e^\eta) ^2
-\end{align}
-and
-\begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,  (1+e^\eta) ^2  } du,
-\end{align}
-which converges since $\exp(u'D^{-1}u$ grows faster than any term in the numerator of the integrand.
-
-On the other hand, when $y|u\sim$ Poisson, then 
-\begin{align}
-c(\eta)=\exp(\eta) \Rightarrow \exp(2 c(\eta)) = (\exp(\exp(\eta)))^2
-\end{align}
-and
-\begin{align}
-\int \dfrac{f_\theta(u,y)^2}{\tilde{f}(u)} du &\propto \dfrac{1}{p_1} \int \dfrac{ \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} \, \exp(2y'Zu)  }{\exp(u'D^{-1}u)   \,   (\exp(\exp(\eta)))^2  } du,
-\end{align}
-which converges since  $(\exp(\exp(\eta)))^2$ grows faster than any term in the numerator of the integrand.
-
-Therefore, whether the data are Bernoulli or Poisson distributed, the importance sampling weights 
-\begin{align}
-\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
-have finite variance.
-
-Now, we want to show that
-\begin{align}
-\sum_{k=1}^m \nabla \log f_\theta (u_k,y) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} =\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} +\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-\end{align}
- has finite variance. We need to show the following are finite:
-\begin{align}
-\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}   \right) \\
-\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)\\
-\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)_.
-\end{align}
-
-That is, we want to show the existence of the following:
-\begin{align}
-&\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:first} \\
-&\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:second} \\
-&\int \dfrac{  \nabla \log f_\theta (u)  \nabla \log f_\theta (y|u)  \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du \label{eq:third} 
-\end{align}
-
-First, we start with \eqref{eq:first}:
-\begin{align}
-\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du 
-&=\int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
-&\leq \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1\grave{f}(u|0,D^*)} du \\
-&\propto \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \exp(-u'D^{-1}u)}{p_1\left[ 1+u'D^*u/\zeta  \right]^{-(\zeta+q)/2}} du \\
-&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \left[ f_\theta(y|u)\right]^2 \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(u'D^{-1}u)} du \\
-&= \int \dfrac{ \left[ \nabla \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon}
-\end{align}
-By the Cauchy-Schwartz inequality, the integral in \eqref{eq:soon} exists if
-\begin{align}
- & \int \dfrac{ \left[\dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \nonumber \\
-&\propto  \int \dfrac{ \left[X' c'(\eta) \right]^2 \, \exp(2y'(X \beta+Zu)) \,  \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{ p_1 \, \exp(2 c(\eta)) \,\exp(u'D^{-1}u)} du \label{eq:soon2}
-\end{align}
-exists, where $c'(\eta)$ is a vector with components $c'(\eta_i)$.
-This will converge as long as no term in the numerator of the integrand grows more quickly than $\exp(u'D^{-1}u$.
-The second and third terms grow less quickly than $\exp(u'D^{-1}u$, so $\left[ c'(\eta)\right]^2$ is the only term to check.
-If $y|u \sim$ Bernoulli, then
-\begin{align}
-c'(\eta_i)= \dfrac{\exp(\eta_i)}{1+\exp(\eta_i)},
-\end{align}
-which does not grow more quickly than $\exp(u'D^{-1}u)$. If $y|u \sim$ Poisson, then
-\begin{align}
-c'(\eta_i)= e^{\eta_i},
-\end{align}
-which does not grow more quickly than $\exp(u'D^{-1}u)$. Therefore, \eqref{eq:first} exists and is finite, and 
-\begin{align}
-\var \left( \sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
-\end{align}
-is finite.
-
-Next, we move on to showing the existence of \eqref{eq:second}.
-\begin{align}
-\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &= 
-\int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du\\
-&\leq \int \dfrac{ \left[ \nabla \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du
-\end{align}
-By the Cauchy-Schwartz inequality, the above integral exists as long as  
-\begin{align}
-\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du < \infty
-\end{align}
-for every $t=1,\ldots, T$. We see
-\begin{align}
-&\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \left[ f_\theta(y|u)\right]^2  \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \\
-&= \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   )  \left[ f_\theta(u)\right]^2}{p_1 \exp(2c(\eta))    \grave{f}(u|0,D^*)} du \\
-&\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t} \log f_\theta (u) \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du\\
-&= \int \dfrac{ \left[  -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}   \right]^2 \exp(2y'(X\beta+Zu)   ) \,   \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \exp(2c(\eta)) \, \exp(u'D^{-1}u)  } du
-\end{align}
-Because every term in the numerator of the integrand grows less quickly than $\exp(u'D^{-1}u)$, the integral exists.  Therefore, \eqref{eq:second} exists and 
-\begin{align}
-\var \left(\sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
-\end{align}
-
-Lastly, we want to show the existence of \eqref{eq:third}. We see
-\begin{align}
-\int \dfrac{  \nabla \log f_\theta (u)\,  \nabla \log f_\theta (y|u) \, \left[ f_\theta(u,y)\right]^2}{\tilde{f}(u)} du &=
-\int \dfrac{  \nabla \log f_\theta (u) \, \nabla \log f_\theta (y|u) \, \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{\tilde{f}(u)} du \\
-&\leq
-\int \dfrac{  \nabla \log f_\theta (u) \,  \nabla \log f_\theta (y|u)\,  \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
-\end{align}
-Again, we use the Cauchy-Schwartz inequality to check the existence of \eqref{eq:this1}. If, for every $t=1,\ldots,T$,
-\begin{align}
-\int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du \label{eq:this1}
-\end{align}
-exists, then \eqref{eq:this1} also exists. Continuing, we see
-\begin{align}
-& \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \left[ f_\theta(u)\right]^2}{p_1 \grave{f}(u|0,D^*)} du\\
- &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 }{p_1 \, \exp(u'D^{-1}u) \, \grave{f}(u|0,D^*)} du\\
- &\propto \int \dfrac{ \left[ \dfrac{\partial}{\partial \nu_t}  \log f_\theta (u) \right] \left[  \dfrac{\partial}{\partial \beta} \log f_\theta (y|u) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
-&\propto \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \left[ f_\theta(y|u)\right]^2 \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) } du\\
-&= \int \dfrac{ \left[      -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}        \right] \left[  X'(y-c'(\eta)) \right]   \, \exp(2y'(X\beta+Zu)) \, \left[ 1+u'D^*u/\zeta  \right]^{(\zeta+q)/2} }{p_1 \, \exp(u'D^{-1}u) \, \exp(2 c(\eta)) } du.
-\label{eq:this2}
-\end{align}
-We have already shown that $-c'(\eta)$ does not grow more quickly than $\exp(u'D^{-1}u)$.
-The other terms in the numerator of the integrand shown in \eqref{eq:this2} also grow more slowly than $\exp(u'D^{-1}u)$.
-We conclude that \eqref{eq:this2} exists, which implies that \eqref{eq:third} exists and
-\begin{align}
-\cov \left(\sum_{k=1}^m \nabla \log f_\theta (y|u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \, , \, \sum_{k=1}^m \nabla \log f_\theta (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty.
-\end{align}
-This shows that
-\begin{align}
-\var \left( \sum_{k=1}^m     \nabla \log f_\theta(u_k,y)\dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) < \infty. 
-\end{align}
-Therefore, the gradient of the MCLA has finite variance and a central limit theorem.
-
-%EDITED TO HERE.
-%
-% That is, we want to show the existence of
-%\begin{align}
-%\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
-%\end{align}
-%That is, we want to show the existence of
-%\begin{align}
-%\int &\dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du \\
-%&\leq \int  \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u) + \log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)} du  \\
-%&\leq  \int \dfrac{N(u|0,D)^2 \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2}{p_1 N(u|0,D)} du  \\
-%&\leq \dfrac{1}{p_1}  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du.
-%\end{align}
-%By the Cauchy-Schwartz inequality, 
-%\begin{align}
-%  &\int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u) \, \, \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \\
-%& \leq \left[  \int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta (u)  \right]^2 du \right]^{1/2}  \left[  \int N(u|0,D) \left[ \dfrac{\partial}{\partial \beta}\log f_\theta(y|u)   \right]^2 du \right]^{1/2} .
-%\end{align}
-%Therefore, I need to show that each of these two integrals is finite.  I will start with the first of the two integrals: 
-%\begin{align}
-%\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du. 
-%\end{align}
-%Recall the notation from section \ref{sec:Ddiag}.  If
-%\begin{align}
-%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t< \infty 
-%\end{align}
-%for every $t=1,...,T$, then
-%\begin{align}
-%\int N(u|0,D) \left[  \dfrac{\partial}{\partial \nu} \log f_\theta(u)  \right]^2 du < \infty
-%\end{align}
-%by Cauchy Schwartz.
-Therefore, I will prove the integral is finite for the general case of $u_t$ and $\nu_t$.
-First, we recognize that our integral is actually an expectation of a function of a normal random variable:
-%\begin{align}
-%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t &=
-%\int \dfrac{1}{\nu_t^{q_t/2}} e^{-u_t'u_t/(2 \nu_t)} \left[ -\dfrac{q_t}{2 \nu_t} + \dfrac{u_t'u_t}{2 \nu_t^2}  \right]^2 du_t .
-%\end{align}
-%Normal distributions have finite moments of all orders, so we know that\begin{align}
-%\int N(u_t|0,D_t) \left[  \dfrac{\partial}{\partial \nu_t} \log f_\theta(u)  \right]^2 du_t < \infty.
-%\end{align}
-%
-%
-%
-%
-%Next, we want to show 
-%\begin{align}
-%\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du < \infty.
-%\end{align}
-% Recall that $\eta=X \beta +Z u$. Letting $\dfrac{\partial}{\partial \eta} c(\eta)=\mu$, we see 
-%\begin{align}
-%\int N(u|0,D) \left[\dfrac{\partial}{\partial \beta} \log f_\theta(y|u)   \right]^2 du &= \int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-%%&\propto  \int N(u|0,D) \mu' XX' Y du + \int N(u|0,D) \mu' X X' \mu du   .
-%\end{align}
-%We will consider this integral for the cases of $Y|u$ being a Bernoulli and a Poisson random variable. When $Y|u$ is Bernoulli, then $0 \leq \mu \leq 1.$ Therefore,
-%\begin{align}
-%\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du
-%\end{align}
-%exists and is finite.
-%When $Y|u$ is a Poisson random variable, then for any  $\eta$, 
-%\begin{align}
-%\dfrac{\partial}{\partial \eta} c(\eta) = e^\eta.
-%\end{align}
-%Then
-%\begin{align}
-%\int N(u|0,D) \left[ (Y'- \mu') X X'(Y- \mu)   \right] du &\propto \int N(u|0,D)  e^{\eta} e^{\eta}    du \\
-%&\propto \int N(u|0,D) e^{2Zu} du 
-%\end{align}
-%exists and is finite.
-%
-%Next, we want to show that 
-%\begin{align}
-%\sum_{k=1}^m \nabla \log \tilde{f} (u_k) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}
-%\end{align}
-% has finite variance. That is, we want to show the existence of
-%\begin{align}
-%\int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du.
-%\end{align}
-%We see that
-%\begin{align}
-%\int& \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du\\
-%&\leq  \int \dfrac{N(u|0,D)^2 \left[\nabla \log \tilde{f}u) \right]^2}{p_1 N(u|0,D)} du\\
-%&= \dfrac{1}{p_1} \int N(u|0,D) \left[\nabla \log \tilde{f}u) \right]^2 du\\
-%&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)+p_2 N(u|u^*,D^*)+p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}   \right]^2 du\\
-%&\leq\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{p_1 \dfrac{\partial}{\partial \nu} N(u|0,D)}{p_1 N(u|0,D)}   \right]^2 du\\
-%&=\dfrac{1}{p_1} \int N(u|0,D) \left[  \dfrac{ \dfrac{\partial}{\partial \nu} N(u|0,D)}{ N(u|0,D)}   \right]^2 du\\
-%&= \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du \\
-%\end{align}
-%Now we use
-%\begin{align}
-%\dfrac{\partial}{\partial \nu} \log N(u|0,D) = \dfrac{\dfrac{\partial}{\partial \nu}  N(u|0,D)}{N(u|0,D)}
-%\end{align}
-%to see
-%\begin{align}
-% \dfrac{1}{p_1} \int \dfrac{\left[ \dfrac{\partial}{\partial \nu} N(u|0,D)  \right]^2}{N(u|0,D)} du &=\dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du \\
-%&= \dfrac{1}{p_1} \int  N(u|0,D) \left[ \dfrac{\partial}{\partial \nu} \log N(u|0,D)  \right]^2 du 
-%\end{align}
-%which exists and is finite.
-%
-%
-%
-%Therefore, it is proven that 
-%\begin{align}
-%\int \dfrac{N(u|0,D)^2 \left[\nabla \left(\log f_\theta(y|u)+\log f_\theta (u) \right) \right]^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du 
-%\end{align}
-%exists and is finite. Therefore, since  the gradient of the MCLA has finite variance, we can conclude the gradient of the MCLA has a Central Limit Theorem.
-
-%
-%Next, we need to show that the Hessian of the MCLA has finite variance. Again, it is enough to show that
-%\begin{align}
-%\int \dfrac{N(u|0,D)^2 h(u)^2}{p_1 N(u|0,D)+p_2 N(u|u^*, D^*) + p_3 N(u|u^*, ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} du < \infty
-%\end{align}
-
-%\subsection{Boundedness of importance weights}
-%We want to make sure that the importance weights will be bounded when using this distribution.
-The importance weight for a vector $u_k$ is $\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)}$.
-Since $f_\theta(y,u_k)< \infty$,
-%\begin{align}
-%\dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} < \infty \Longleftrightarrow \tilde{f}(u_k) >0.
-%\end{align}
-%
-%We start with
-%\begin{align}
-% \tilde{f}(u_k) &= p_1  f(u_k \, | \, 0, D)+p_2  f(u_k \, | \, u^*, D^*)+p_3  f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\\
-%&\geq p_1  f(u_k \, | \, 0, D) \\
-%& \geq p_1 (2 \pi)^{-q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u}
-%\end{align}
-%In order to show that $\tilde{f}(u_k)>0$, we need to show that both $ |(D^*)^{-1/2}|$ and  $e^{- u' (D^{*})^{-1}u}$ are positive. 
-%
-% First, I will show that $ |(D^*)^{-1/2}|>0$.Since  $(D^*)^{-1/2}$ is invertible, its determinant must be nonzero.
-Since $D^*$ is positive semi-definite, then $(D^*)^{-1/2}$ is also positive semi-definite.
-Thus the eigenvalues of $(D^*)^{-1/2}$ are neither zero nor negative; they must be positive.
-An eigendecomposition tells us $|(D^*)^{-1/2}|$ is the product of its eigenvalues.
-Therefore, $|(D^*)^{-1/2}| >0$.
-%
-%Next, I will show that $e^{- u' (D^{*})^{-1}u}>0$.  Since $D^*$ is a covariance matrix, it is positive semi-definite.  By definition, for any vector $a_1$, 
-%\begin{align}
-%a_1' D^* a_1 \geq 0
-%\end{align}
-%For every vector $a_2$, there exists an $a_1$ such that $a_2=D^*a_1$. Then
-%\begin{align}
-%a_2' (D^*)^{-1} a_2 &= (D^* a_1)' (D^*)^{-1} D^* a_1\\
-%&= a_1' D^* (D^*)^{-1} D^* a_1\\
-%&= a_1' D^* a_1\\
-%& \geq 0
-%\end{align}
-%Therefore,   $(D^*)^{-1}$ is also positive semi-definite. Then for any vector of random effects $u$,
-%\begin{align}
-% u' (D^*)^{-1} u \geq 0 
-%\end{align}
-%This means that
-%\begin{align}   e^{- u' (D^{*})^{-1}u} >0
-%\end{align}
-%Therefore,
-%\begin{align}
-% \tilde{f}(u_k) &\geq p_1 (2 \pi)^{q/2} \; |(D^*)^{-1/2}|  \; e^{- u' (D^{*})^{-1}u} >0
-%\end{align}
-%
-
 \end{document}

### 2020-07-16 DesignDocs/DesignDocDimReduction.tex
tweaks to dim reduction design doc

@@
 \begin{itemize}
-\item[Plan A:] Change PQL so that similar random effects are collapsed into a shared, single random effect.
-\item[Plan B:] Keep the PQL optimization as it is, and add another step in between the two steps. At this time, the similar random effects are collapsed into a shared, single random effect.
+\item[Plan A:] Change PQL so that similar random effects are fused into a shared, single random effect.
+\item[Plan B:] Keep the PQL optimization as it is, and add another step in between the two steps. At this time, the similar random effects are fused into a shared, single random effect.
   \end{itemize}
 
-  I'm not sure, but it seems like Plan A is better than Plan B. It makes sense that we'd want to try collapsing the random effects while making sure this simplification not ruin the fit of the model.
+  I'm not sure, but it seems like Plan A is better than Plan B. It makes sense that we'd want to try fusing the random effects while making sure this simplification not ruin the fit of the model.
 
@@
 The outer optimization works on the variance components.
-\textbf{I believe that adding a penalty term to the inner loop will collapse the random effects.} We want the random effects to be similar to one other, so the distance (or the square of the distance, if we prefer a ridge-like penalty) between the random effects must be controlled.
+\textbf{I believe that adding a penalty term to the inner loop will fuse the random effects.} We want the random effects to be similar to one other, so the distance (or the square of the distance, if we prefer a ridge-like penalty) between the random effects must be controlled.
+This is almost exactly pairwise fusion lasso (Petry et.
+al.
+2011), except we are fusing \textit{random} effects instead of \textit{fixed} effects.
 
@@
 Then, given $\tilde{\beta}$ and $\tilde{s}$, the outer optimization finds $A$.
-\textbf{I believe that adding a penalty term to the inner loop will collapse the random effects.
+\textbf{I believe that adding a penalty term to the inner loop will fuse the random effects.
 We want the random effects to be similar to one other, so the distance (or the square of the distance) between the random effects must be controlled.
@@
 \end{align}
-and the penalized likelihood (pre-collapsing) as:
+and the penalized likelihood (pre-fusion) as:
 \begin{align}
@@
 \end{align}
-where $\lambda$ is some number we can tweak to force more/less collapsing.
+where $\lambda$ is some number we can tweak to force more/less fusion.
 
-I HAVE NOT REDONE THESE CALCULATIONS BELOw
+I HAVE NOT REDONE THESE CALCULATIONS BELOW and it will kind of suck because of the new fusion penalty.
 
